### 2021-01-06 code/StrategicVsTactical_PartI.Rmd
First commit - of the project to date

---
title: "Modeling memory and taxis to analyze migrations"
author: "E. Gurarie, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
theme: cerulean
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE)
```

# Memory driven foraging model

We intend to study how a blending of a *tactical* (i.e. direct response to resource availability or perception) and a *strategic* (i.e. memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior**.

Out model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.  Importantly, the environment is *approximately periodic*, i.e. one in which the distribution of the resource at a given time is approximately equal to that resource distribution at a fixed previous time period, i.e. $h(t) \approx h(t - \tau)$. This models annual variations, and the period corresponds to one year.  The *tactical* process is straightforward taxis up a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.  Thus:

$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial h(x,t)}{\partial x} + \sum_{i = 1}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$

where $u$ represents a population distributed in time and space, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta_i$ is the strength of advection up the gradient of the *population* distribution in an $i$'th earlier seasonal cycle (year), up to $n \in N$.  The form of $\beta_i$ should be simple, e.g., if the collective memory reaches only one year back then $n = 1$ and $\beta$ is a single constant. Alternatively, a form of memory might discounts older "traditions" via $\beta_i = \beta \exp(-i/\gamma)$, where $\gamma$ represents a generational scale of memory.

For reference and replicability, I include the incrementally constructed R code, which relies heavily on the `deSolve` (Soetaert et al. 2010) and `ReacTran` (Soetaert and Filip, 2020) packages, which efficiently implement transport ODE and PDE systems in R.

```{r, message=FALSE, warning=FALSE}
require(deSolve)
require(ReacTran)
require(magrittr)
source("functions.R")
```

# Static run

```{r}
ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, dx){
  tran.1D(C = pop, D = parms["epsilon"], 
                   flux.up = 0, flux.down = 0, 
                   v = parms["alpha"] * diff(resource)/dx + 
                       parms["beta"] * diff(pop_lag)/dx, 
                   dx = dx)
}
```

As a simple illustration of a model run, below the ultimate evolution of a forager that begins in the left part of space and uses a combination of the previous year's location (to the right) and the current resource distribution, and ultimately "splits up" into being attracted to the resource in the middle and its previous year's distribution, with a somewhat stronger taxis towards the resource than the memory:

```{r GradientExample, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
dx <- .01
X.max <- 100
x.edges <- seq(0,100, length=X.max/dx + 1)
x.mids <- (x.edges[-1] + x.edges[-length(x.edges)])/2

parameters <- c(epsilon = 0.1, alpha = 4, beta = 2)
pop <- dnorm(x.mids, mean=20, sd = 2) %>% {./sum(.)/dx}

pop_lag <- c(0,dnorm(x.mids, mean=80, sd = 2) %>% {./sum(.)/dx},0)
resource <- c(0,dnorm(x.mids, mean = 50, sd = 4) %>% {./sum(.)/dx},0)

pop.inf <- steady.1D(func = ForagingMemoryModel, 
                   y = pop, parms = parameters, 
                   resource = resource, 
                   pop_lag = pop_lag, 
                   nspec=1, dx = dx)$y %>% {./sum(.)/dx}

n <- length(x.mids)

cols <- c("black", "darkgrey", "green",  "red")
lwds <- c(1,3,3,1)
ltys = c(1,3,3,1)

par(bty = "l", mgp = c(1.5,.25,0), tck = 0.01, las = 1)

plot(x.mids, pop, type = "l", xlab = "space", ylab = "density", 
     col = cols[1], lwd = lwds[1], lty = ltys[1])
lines(x.mids, pop_lag[-c(1,n+1)], col = cols[2], lwd = lwds[2], lty = ltys[2])
lines(x.mids, resource[-c(1,n+1)], col = cols[3], lwd = lwds[3], lty = ltys[3])
lines(x.mids, pop.inf, col = cols[4], lwd = lwds[4], lty = ltys[4])

legend("top", legend = c("pop. initial", "pop. previous year", "resource", "pop. final"), 
       lty = ltys, lwd = lwds, col = cols, bty = "n")
#title(parse(text = printParameters(p)))

printParameters <- function(p) 
  parse(text = paste0("list(",paste(names(p), p, sep = "==", collapse = ", "),")"))

title(printParameters(parameters))

```

This example is completely static and symmetric around the two drivers.  Note that all the populations (previous, initial, final) are normalizes (integrate to 1) and that the space is defined between 0 and 100.

# Dynamic initial population

```{r}
source("functions.R")
```

In the following example, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is fixed in the middle of the space (as above).  The memory is a one-year memory.  In this context, migration is not, in fact, optimal.  But the important piece is to see how the reponse mixes attraction to the resource with attraction to the previous year's behavior.  Note that the year here (for simplicity, speed and symmetry), is 100 days long.

```{r setupEnvironment}

getSinePop <- function(Time, X, X.max, peak.max, peak.min, sd){
  pop <- outer(Time, X,
                     function(t,x) 
                       dnorm(x, mean = X.max/2 + (peak.max-peak.min) / 2 * sin( 2*(t * pi)/max(Time)),sd = sd))
  pop <- apply(pop, 1, function(x) x/sum(x)) %>% t
  pop / mean(diff(X))
}

dx <- 1
X.max <- 100
tau <- 100

X.edge <- seq(0,100, length=X.max/dx + 1)
X.mid <- seq(1,X.max,dx) - 1/2
Time <- 1:tau

pop.Year1 <- getSinePop(Time, X.mid, X.max = X.max, peak.max = 80, peak.min = 20, sd = 10)

fr1 <- dnorm(X.mid, mean = 80, sd = 10) %>% {./sum(.)/dx}
fr2 <- dnorm(X.mid, mean = 20, sd = 10) %>% {./sum(.)/dx}

fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
                         nrow = nrow(pop.Year1), 
                         ncol = ncol(pop.Year1))

```

The following examples illustrate how an initially migrating population evolves, even as it is attracted to a resource that is distributed in two different locations. In these examples, the first year (year 0) is a sinusoidal migration.  We run the simulations for 3 subsequent years.

## Only memory

```{r memoryOnly, fig.height = 5, fig.width = 10}
dx <- .5
parameters <- c(epsilon = 5, alpha = 0, beta = 5e2, lambda = 0, dx = dx)
eval <- FALSE
if(eval){
  MemSim <- runManyYears(pop.Year1, Time, Resource = fixed.Resource, Parameters = parameters, n.years = 3)
  save(MemSim, file = "../results/MemSim.rda")
} else load("../results/MemSim.rda")

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(MemSim, tau = tau)
plotYearList(MemSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

The population roughly echoes its initial track, but degrades from year to year.

## Resource only

In the following simulations, we add a resource distributed as follows in two concentrations:

```{r fixedResource, fig.height = 3, fig.width = 4}
image(Time, X.mid, fixed.Resource, main = "resource", ylab = "x")
```

```{r resourceOnly, fig.height = 5, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 5e2, beta = 0, lambda = 0, dx = dx)

eval <- FALSE
if(eval){
  ResourceSim <- runManyYears(pop.Year1, Time, Resource = fixed.Resource, Parameters = parameters, n.years = 3)
  save(ResourceSim, file = "../results/ResourceSim.rda")
} else load("../results/ResourceSim.rda")

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(ResourceSim, tau = tau)
plotYearList(ResourceSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

Here, the population disregards its memory and only pursues a tactic stategy. Within a year, the population has gathered around the resource.

## Resource and memory

Finally, we combine both resource following and memory:

```{r resourceAndMemory, fig.height = 5, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 3e2, beta = 3e2, lambda = 0, dx = dx)

eval <- FALSE
if(eval){
  MemResourceSim <- runManyYears(pop.Year1, Time, Resource = fixed.Resource, Parameters = parameters, n.years = 3)
  save(MemResourceSim, file = "../results/MemResourceSim.rda")
} else load("../results/MemResourceSim.rda")

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(MemResourceSim, tau = tau)
plotYearList(MemResourceSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

Here, the strong cultural memory is slowly blended with a strategy to simply concentrate in one of the two resource hot-spots.

# Seasonal resource

There is obviously nothing particularly adaptive about migrating when resources are fixed.  In the following scenaries, we explore a seasonal resource. We developed (with some fuss!) a seasonal resource function with the following properties:

1. The total amount of resource across space is constant throughout the year.
2. At the beginning, middle, and end of the year the resource is uniformly distributed. 
3. At some peak time $t_r < \tau/2$, the resource concentrates at a location $x_r < \chi/2$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
4. The resource peaks exactly symmetrically at time $\tau - t_r$ and location $\chi - x_r$ with the same variance $\sigma_r$.

To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.  Thus:

$$R(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$

where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:

$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$

where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.  These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.

The means and variances themselves are Gaussian pulses, with the mean peaking at $x_r$ at time $t_r$ with standard deviation $\sigma_t$ and at $\chi - x_t$ at time $\tau - t_r$ and the standard deviation pulsing from $\chi/\sqrt{12}$ (corresponding to a uniform distribution) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.

All of these details boil down to the ability to generate various seasonal scenarios.

```{r exampleResourcePlots, fig.width = 10, fig.height = 6}
source("functions.R")
Time <- 1:100; X <- 0:100

R1 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
R2 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5))
R3 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9))
R4 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5))

par(mfrow = c(2,2), mar = c(3,3,3,1), mgp = c(2,.25,0), las = 1, tck = 0.01)
image(Time,X,R1, main = "Broad, long peak"); contour(Time,X,R1, add = TRUE)
image(Time,X,R2, main = "Narrow, long peak"); contour(Time,X,R2, add = TRUE)
image(Time,X,R3, main = "Broad, short peak"); contour(Time,X,R3, add = TRUE)
image(Time,X,R4, main = "Narrow, short peak"); contour(Time,X,R4, add = TRUE)
```

# Running the model in a seasonal environment

In these pulsed resources, we might assume that a *combination* of memory and tactic responses would yield more optimal behavion, and might observe some shift in the migratory strategy.

## Memory only

```{r PulseSim1, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 0, beta = 4e2, lambda = 0, dx = dx)
Memory_PulsedResource_Sim <- runManyYears(pop.Year1, Time, 
                                  Resource = R1, Parameters = parameters, n.years = 6)
par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Memory_PulsedResource_Sim, tau = tau)
plotYearList(Memory_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Tactics only

This model does a surprisingly good job aggregating at the appropriate resource, but appears to only locally aggregate, not to migrate per se.

```{r PulseSim2, fig.width = 10}
parameters <- c(epsilon = 2, alpha = 4e2, beta = 0, lambda = 0, dx = dx)
Tactic_PulsedResource_Sim <- runManyYears(pop.Year1, Time, 
                                  Resource = R1, Parameters = parameters, n.years = 6)
par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Tactic_PulsedResource_Sim, tau = tau)
plotYearList(Tactic_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Memory and tactics

It is not clear that the hybrid model recovers a true migration, but it seems to retain somewhat more concentration along the migratory corridors.

```{r PulseSim3, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 2e2, beta = 3e2, lambda = 0, dx = dx)
Hybrid_PulsedResource_Sim <- runManyYears(pop.Year1, Time, 
                                  Resource = R1, Parameters = parameters, n.years = 6)
par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Hybrid_PulsedResource_Sim, tau = tau)
plotYearList(Hybrid_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Memory and tactics: Narrow resource

Here we compare a memory with and without memory exploiting the "narrow - long" resource combination, and the tendency to

### No memory - strong tactics

Notable concentrations at the resource peaks, but a lot of diffuse behavior in the interim seasons

```{r PulseSim4_tactic, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 2e2, beta = 0, lambda = 0, dx = dx)
Hybrid_PulsedResource_Sim2 <- runManyYears(pop.Year1, Time, 
                                  Resource = R4, Parameters = parameters, n.years = 6)
par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Hybrid_PulsedResource_Sim2, tau = tau)
plotYearList(Hybrid_PulsedResource_Sim2, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

### Hybrid model

Memory appears to reinforce the aggregations at the (narrow) resource peaks

```{r PulseSim4_memory, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 2e2, beta = 3e2, lambda = 0, dx = dx)
Hybrid_PulsedResource_Sim2 <- runManyYears(pop.Year1, Time, 
                                  Resource = R4, Parameters = parameters, n.years = 6)
par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Hybrid_PulsedResource_Sim2, tau = tau)
plotYearList(Hybrid_PulsedResource_Sim2, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

# Some summaries

1. The model seems to - approximately - work for exploring this dynamic.
2. It is - however - difficult to make it really repeat previous behavior strongly. I was hoping for a "knob" that perfectly replicates past behavior, but the diffusion-taxis formulation makes that somehow hard.  I haven't programmed longer time scaled memories (> 1 year) but that would probably be helpful / interesting. 
3. The memory bit does demand iteratively feeding a year's cycle back into it rather than having some analytical function, which slows the whole thing down a bit. That said, with deepthought a lot of runs could be fitted simultaneously - and I tend to be very picky about "slow" - it actually only takes about 6 seconds to run a 6 year hybrid model on a 100 x 100 spatio-temporal grid.  But sometimes, the results are more numerically stable on a finer grid. 
4. We need a metric of success!  And also - somehow - a metric of "migratoriness". 
5. What questions do we wnat to explore?  I think a big one is adaptability of the model under different kinds of resource perturbations, e.g. if the timing shifts, or location shifts.

# References

Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09

Soetaert, Karline and Meysman, Filip, 2012. Reactive transport in aquatic ecosystems: Rapid model   prototyping in the open source software R Environmental Modelling & Software, 32,

### 2021-01-06 code/StrategicVsTactical_PartI_v2.Rmd
First commit - of the project to date

--- 2021-01-06 code/StrategicVsTactical_PartI.Rmd
@@
 ```{r}
-source("functions.R")
+source("functions2.R")
 ```
@@
 ```{r setupEnvironment}
-
 getSinePop <- function(Time, X, X.max, peak.max, peak.min, sd){
@@
 
+```{r}
+ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
+  tran.1D(C = y, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/parms["dx"] +  
+            parms["gamma"] * diff(y)/parms["dx"] +
+            parms["beta"] * diff(pop_lag)/parms["dx"], 
+          dx = parms["dx"])
+}
+
+ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
+  tran.1D(C = y, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/parms["dx"] +  
+            parms["gamma"] * diff(y)/parms["dx"] +
+            parms["beta"] * diff(pop_lag)/parms["dx"], 
+          dx = parms["dx"])
+}
+
+```
+
 ```{r memoryOnly, fig.height = 5, fig.width = 10}
 dx <- .5
-parameters <- c(epsilon = 5, alpha = 0, beta = 5e2, lambda = 0, dx = dx)
+parameters <- c(epsilon = 5, alpha = 0, beta = 5e2, gamma = 20, lambda = 0, dx = dx)
+
 eval <- FALSE

### 2021-01-06 code/functions.R

@@ -1,0 +1,130 @@
+require(deSolve)
+require(ReacTran)
+require(magrittr)
+#' Foraging model
+
+ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
+  tran.1D(C = y, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/parms["dx"] +  
+            parms["beta"] * diff(pop_lag)/parms["dx"], 
+          dx = parms["dx"])
+}
+
+ForagingMemoryModel2 <- function(t, y, parms, resource, pop_lag){
+  tran.1D(C = y, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/parms["dx"] +  
+            parms["gamma"] * diff(y)/parms["dx"] +
+          parms["beta"] * diff(pop_lag)/parms["dx"], 
+          dx = parms["dx"])
+}
+
+#' Run One Year forward
+
+runNextYear <- function(pop.Year1, Time, Resource, Parameters){
+  
+  pop.Year2 <- pop.Year1*0
+  pop.Year2[1,] <- pop.Year1[nrow(pop.Year1),]
+  
+  t.max <- max(Time)
+  
+  fl <-  Parameters["lambda"]
+  
+  for(i in 2:t.max){
+    if(i < t.max-fl)
+      pop_lag <- c(0,pop.Year1[i+fl,],0) else
+        pop_lag <- c(0,pop.Year2[max(i - (t.max - fl)),],0)
+      
+      pop.Year2[i,] <- ode(y = pop.Year2[i-1,], 
+                           times = c(0,Time[i]-Time[i-1]), 
+                           parms = Parameters,
+                           func = ForagingMemoryModel, 
+                           resource = c(0,Resource[i,],0),
+                           pop_lag = pop_lag)[2,1:ncol(pop.Year2)+1]
+  }
+  
+  pop.Year2
+}
+
+#' Run several years forward
+
+runManyYears <- function(pop.Year1, Time, Resource, Parameters, n.years){
+  pop.list <- list(Year1 = pop.Year1)
+  
+  for(i in 1:n.years){
+    cat(paste("running year ", i, "\n"))
[…]

### 2021-01-06 code/functions2.R

@@ -1,0 +1,122 @@
+require(deSolve)
+require(ReacTran)
+require(magrittr)
+#' Foraging model
+
+ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
+  tran.1D(C = y, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/parms["dx"] +  
+            parms["gamma"] * diff(y)/parms["dx"] +
+            parms["beta"] * diff(pop_lag)/parms["dx"], 
+          dx = parms["dx"])
+}
+
+#' Run One Year forward
+
+runNextYear <- function(pop.Year1, Time, Resource, Parameters){
+  
+  pop.Year2 <- pop.Year1*0
+  pop.Year2[1,] <- pop.Year1[nrow(pop.Year1),]
+  
+  t.max <- max(Time)
+  
+  fl <-  Parameters["lambda"]
+  
+  for(i in 2:t.max){
+    if(i < t.max-fl)
+      pop_lag <- c(0,pop.Year1[i+fl,],0) else
+        pop_lag <- c(0,pop.Year2[max(i - (t.max - fl)),],0)
+      
+      pop.Year2[i,] <- ode(y = pop.Year2[i-1,], 
+                           times = c(0,Time[i]-Time[i-1]), 
+                           parms = Parameters,
+                           func = ForagingMemoryModel, 
+                           resource = c(0,Resource[i,],0),
+                           pop_lag = pop_lag)[2,1:ncol(pop.Year2)+1]
+  }
+  
+  pop.Year2
+}
+
+#' Run several years forward
+
+runManyYears <- function(pop.Year1, Time, Resource, Parameters, n.years){
+  pop.list <- list(Year1 = pop.Year1)
+  
+  for(i in 1:n.years){
+    cat(paste("running year ", i, "\n"))
+    pop.list[[i+1]] <- runNextYear(pop.list[[i]], Time, Resource = Resource, 
+                                   Parameters = parameters)
+  }
+  names(pop.list) <- paste0("Year",0:n.years)
+  pop.list <- lapply(pop.list, function(l) l[-1,])
+  pop.list
+}
+
+
+#' Plotting Functions
+
[…]

### 2021-01-06 code/StrategicVsTactical_PartI_v2.Rmd
Minor edits

@@
 editor_options:
-  chunk_output_type: inline
+  chunk_output_type: console
 ---
@@
 
-$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial h(x,t)}{\partial x} + \sum_{i = 1}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$
+$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial h(x,t)}{\partial x} + \sum_{i = 0}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$
 
 where $u$ represents a population distributed in time and space, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta_i$ is the strength of advection up the gradient of the *population* distribution in an $i$'th earlier seasonal cycle (year), up to $n \in N$.
-The form of $\beta_i$ should be simple, e.g., if the collective memory reaches only one year back then $n = 1$ and $\beta$ is a single constant.
+
+Note that $\beta_0$ is similar to an index of "cohesion", essentially encouraging the population to stay close together.
+Higher terms of $\beta_i$ should be simple, e.g., if the collective memory reaches only one year back then $n = 1$ and $\beta_i$ is a single constant.
 Alternatively, a form of memory might discounts older "traditions" via $\beta_i = \beta \exp(-i/\gamma)$, where $\gamma$ represents a generational scale of memory.
@@
           v = parms["alpha"] * diff(resource)/parms["dx"] +  
-            parms["gamma"] * diff(y)/parms["dx"] +
-            parms["beta"] * diff(pop_lag)/parms["dx"], 
+            parms["beta0"] * diff(y)/parms["dx"] +
+            parms["beta1"] * diff(pop_lag)/parms["dx"], 
           dx = parms["dx"])
 }
-
-ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
-  tran.1D(C = y, D = parms["epsilon"], 
-          flux.up = 0, flux.down = 0, 
-          v = parms["alpha"] * diff(resource)/parms["dx"] +  
-            parms["gamma"] * diff(y)/parms["dx"] +
-            parms["beta"] * diff(pop_lag)/parms["dx"], 
-          dx = parms["dx"])
-}
-
 ```
@@
 dx <- .5
-parameters <- c(epsilon = 5, alpha = 0, beta = 5e2, gamma = 20, lambda = 0, dx = dx)
+parameters <- c(epsilon = 5, alpha = 0, 
+                beta0 = 20, 
+                beta1 = 5e2, 
+                lambda = 0, dx = dx)
 

### 2021-01-06 code/functions2.R

@@ -7,8 +7,8 @@ ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
   tran.1D(C = y, D = parms["epsilon"], 
           flux.up = 0, flux.down = 0, 
           v = parms["alpha"] * diff(resource)/parms["dx"] +  
-            parms["gamma"] * diff(y)/parms["dx"] +
-            parms["beta"] * diff(pop_lag)/parms["dx"], 
+            parms["beta0"] * diff(y)/parms["dx"] +
+            parms["beta1"] * diff(pop_lag)/parms["dx"], 
           dx = parms["dx"])
 }
 
@@ -81,9 +81,9 @@ plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE,
     mypersp(time, x, allyears, ...) 
 }
 
-printParameters <- function(p) 
+printParameters <- function(p) {
   parse(text = paste0("list(",paste(names(p), p, sep = "==", collapse = ", "),")"))
-
+}
 
 #' getPulsedResource
 

### 2021-01-11 code/StrategicVsTactical_PartI_v2.Rmd
name
delete
Improved model structure.

@@
 title: "Modeling memory and taxis to analyze migrations"
-author: "E. Gurarie, B. Fagan, others"
+author: "E. Gurarie, S. Potluri, B. Fagan, others"
 date: "`r Sys.Date()`"
@@
 require(magrittr)
-source("functions.R")
 ```
@@
 ```{r}
-ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, dx){
+ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
+                                dx = attributes(pop)$dx){
   tran.1D(C = pop, D = parms["epsilon"], 
@@
 pop <- dnorm(x.mids, mean=20, sd = 2) %>% {./sum(.)/dx} 
+attributes(pop)$dx <- dx
 
@@
                    pop_lag = pop_lag, 
-                   nspec=1, dx = dx)$y %>% {./sum(.)/dx}
+                   nspec=1, 
+                   dx = attributes(pop)$dx)$y %>% {./sum(.)/attributes(pop)$dx}
 
@@
 ```{r setupEnvironment}
-getSinePop <- function(Time, X, X.max, peak.max, peak.min, sd){
-  pop <- outer(Time, X,
+getSinePop <- function(tau, X.min = 0, X.max = 100, dx, 
+                       peak.max, peak.min, sd){
+  Time <- 1:tau
+
+  X.edge <- seq(X.min, X.max, dx)
+  X.mid <- (X.edge[-1] + X.edge[-length(X.edge)])/2
+  pop <- outer(Time, X.mid,
                      function(t,x) 
                        dnorm(x, mean = X.max/2 + (peak.max-peak.min) / 2 * sin( 2*(t * pi)/max(Time)),sd = sd))
-  pop <- apply(pop, 1, function(x) x/sum(x)) %>% t
-  pop / mean(diff(X))
+  pop <- apply(pop, 1, function(x) x/sum(x)/dx) %>% t
+  list(pop = pop, X.mid = X.mid, time = Time, dx = dx, tau = tau)
 }
 
-dx <- 1
-X.max <- 100
-tau <- 100
+world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
+                    peak.max = 80, peak.min = 20, sd = 10)
 
-X.edge <- seq(0,100, length=X.max/dx + 1)
-X.mid <- seq(1,X.max,dx) - 1/2
-Time <- 1:tau
+fr1 <- dnorm(world$X.mid, mean = 80, sd = 10) %>% {./sum(.)/dx}
+fr2 <- dnorm(world$X.mid, mean = 20, sd = 10) %>% {./sum(.)/dx}
+fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
+                         nrow = nrow(world$pop), 
+                         ncol = ncol(world$pop))
 
-pop.Year1 <- getSinePop(Time, X.mid, X.max = X.max, peak.max = 80, peak.min = 20, sd = 10)
-
-fr1 <- dnorm(X.mid, mean = 80, sd = 10) %>% {./sum(.)/dx}
-fr2 <- dnorm(X.mid, mean = 20, sd = 10) %>% {./sum(.)/dx}
-
-fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
-                         nrow = nrow(pop.Year1), 
-                         ncol = ncol(pop.Year1))
-
+world$resource <- fixed.Resource
 ```
@@
 
-```{r}
-ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
-  tran.1D(C = y, D = parms["epsilon"], 
-          flux.up = 0, flux.down = 0, 
-          v = parms["alpha"] * diff(resource)/parms["dx"] +  
-            parms["beta0"] * diff(y)/parms["dx"] +
-            parms["beta1"] * diff(pop_lag)/parms["dx"], 
-          dx = parms["dx"])
-}
-```
+```{r memoryOnly, fig.height = 5, fig.width = 10}
 
-```{r memoryOnly, fig.height = 5, fig.width = 10}
-dx <- .5
-parameters <- c(epsilon = 5, alpha = 0, 
-                beta0 = 20, 
-                beta1 = 5e2, 
-                lambda = 0, dx = dx)
+parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2, lambda = 0)
 
@@
 if(eval){
-  MemSim <- runManyYears(pop.Year1, Time, Resource = fixed.Resource, Parameters = parameters, n.years = 3)
+  MemSim <- runManyYears(World, Parameters = parameters, 
+                         n.years = 3)
   save(MemSim, file = "../results/MemSim.rda")

### 2021-01-11 code/functions2.R

@@ -2,51 +2,51 @@ require(deSolve)
 require(ReacTran)
 require(magrittr)
 #' Foraging model
-
-ForagingMemoryModel <- function(t, y, parms, resource, pop_lag){
-  tran.1D(C = y, D = parms["epsilon"], 
+#' 
+ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
+                                dx = dx){
+  tran.1D(C = pop, D = parms["epsilon"], 
           flux.up = 0, flux.down = 0, 
-          v = parms["alpha"] * diff(resource)/parms["dx"] +  
-            parms["beta0"] * diff(y)/parms["dx"] +
-            parms["beta1"] * diff(pop_lag)/parms["dx"], 
-          dx = parms["dx"])
+          v = parms["alpha"] * diff(resource)/dx + 
+            parms["beta0"] * diff(pop)/dx + 
+            parms["beta1"] * diff(pop_lag)/dx, 
+          dx = dx)
 }
 
 #' Run One Year forward
 
-runNextYear <- function(pop.Year1, Time, Resource, Parameters){
-  
-  pop.Year2 <- pop.Year1*0
-  pop.Year2[1,] <- pop.Year1[nrow(pop.Year1),]
-  
-  t.max <- max(Time)
-  
-  fl <-  Parameters["lambda"]
+runNextYear <- function(World, Parameters){
+  pop1 <- World$pop
+  pop2 <- pop1*0
+  pop2[1,] <- pop1[nrow(pop1),]
+  t.max <- World$tau
+  Time <- World$time
+  Resource <- World$resource
   
   for(i in 2:t.max){
-    if(i < t.max-fl)
-      pop_lag <- c(0,pop.Year1[i+fl,],0) else
-        pop_lag <- c(0,pop.Year2[max(i - (t.max - fl)),],0)
-      
-      pop.Year2[i,] <- ode(y = pop.Year2[i-1,], 
-                           times = c(0,Time[i]-Time[i-1]), 
-                           parms = Parameters,
-                           func = ForagingMemoryModel, 
-                           resource = c(0,Resource[i,],0),
-                           pop_lag = pop_lag)[2,1:ncol(pop.Year2)+1]
+    if(i < t.max)
+      pop_lastyear <- c(0,pop1[i,],0) 
+      pop2[i,] <- ode(y = pop2[i-1,], 
+                     times = c(0,Time[i]-Time[i-1]), 
+                     parms = Parameters,
[…]

### 2021-01-13 code/StrategicVsTactical_PartI_v2.Rmd
Changes to match dx and world

@@
 For reference and replicability, I include the incrementally constructed R code, which relies heavily on the `deSolve` (Soetaert et al.
-2010) and `ReacTran` (Soetaert and Filip, 2020) packages, which efficiently implement transport ODE and PDE systems in R.
+2010) and `ReacTran` (Soetaert and Filip, 2010) packages, which efficiently implement transport ODE and PDE systems in R.
 
@@
 
-parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2, lambda = 0)
+parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
 
@@
 ```{r resourceOnly, fig.height = 5, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 5e2, beta = 0, lambda = 0, dx = dx)
+parameters <- c(epsilon = 5, alpha = 5e2, beta0 = 0, beta1 = 0)
 
@@
 ```{r resourceAndMemory, fig.height = 5, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 3e2, beta = 3e2, lambda = 0, dx = dx)
+parameters <- c(epsilon = 5, alpha = 3e2, beta0=20, beta1 = 5e2)
 
@@
 ```{r exampleResourcePlots, fig.width = 10, fig.height = 6}
-source("functions.R")
+source("functions2.R")
 Time <- 1:100; X <- 0:100
@@
 ```{r PulseSim1, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 0, beta = 4e2, lambda = 0, dx = dx)
-Memory_PulsedResource_Sim <- runManyYears(pop.Year1, Time, 
-                                  Resource = R1, Parameters = parameters, n.years = 6)
+parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
+Memory_PulsedResource_Sim <- runManyYears(pop.Year1, Time, Resource = R1, Parameters = parameters, n.years = 6)
+
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Memory_PulsedResource_Sim, tau = tau)
-plotYearList(Memory_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(MemSim, tau = tau)
+plotYearList(MemSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)

### 2021-01-13 code/functions2.R

@@ -2,7 +2,7 @@ require(deSolve)
 require(ReacTran)
 require(magrittr)
 #' Foraging model
-#' 
+
 ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
                                 dx = dx){
   tran.1D(C = pop, D = parms["epsilon"], 
@@ -65,6 +65,7 @@ plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE,
   
   n.years <- length(yearlist)
   allyears <- do.call(rbind, yearlist)
+  tau <- world$tau
   
   if(log) allyears <- log(allyears)
   

### 2021-01-13 code/StrategicVsTactical_PartI_v2.Rmd
minor changes to implement world

@@
 if(eval){
-  ResourceSim <- runManyYears(pop.Year1, Time, Resource = fixed.Resource, Parameters = parameters, n.years = 3)
+  ResourceSim <- runManyYears(World, Parameters = parameters, n.years = 3)
   save(ResourceSim, file = "../results/ResourceSim.rda")
@@
 if(eval){
-  MemResourceSim <- runManyYears(pop.Year1, Time, Resource = fixed.Resource, Parameters = parameters, n.years = 3)
+  MemResourceSim <- runManyYears(World, Parameters = parameters, n.years = 3)
   save(MemResourceSim, file = "../results/MemResourceSim.rda")

### 2021-01-14 code/StrategicVsTactical_PartI_v2.Rmd
Working on indices and streamlining

@@
 
-fr1 <- dnorm(world$X.mid, mean = 80, sd = 10) %>% {./sum(.)/dx}
-fr2 <- dnorm(world$X.mid, mean = 20, sd = 10) %>% {./sum(.)/dx}
+fr1 <- dnorm(world$X.mid, mean = 80, sd = 10) %>% {./sum(.)/world$dx}
+fr2 <- dnorm(world$X.mid, mean = 20, sd = 10) %>% {./sum(.)/world$dx}
 fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
@@
 ```{r fixedResource, fig.height = 3, fig.width = 4}
-image(Time, X.mid, fixed.Resource, main = "resource", ylab = "x")
+with(world, 
+  image(X.mid, 1:tau, resource, main = "resource", ylab = "x")
+)
 ```
@@
 if(eval){
-  ResourceSim <- runManyYears(World, Parameters = parameters, n.years = 3)
-  save(ResourceSim, file = "../results/ResourceSim.rda")
-} else load("../results/ResourceSim.rda")
+  FixedResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3)
+  save(FixedResourceSim, file = "../results/FixedResourceSim.rda")
+} else load("../results/FixedResourceSim.rda")
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(ResourceSim, tau = tau)
-plotYearList(ResourceSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(FixedResourceSim, tau = tau)
+plotYearList(FixedResourceSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 source("functions2.R")
-Time <- 1:100; X <- 0:100
 
-R1 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
+world.R1 <- world %>% 
+  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
+
 R2 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5))
@@
 par(mfrow = c(2,2), mar = c(3,3,3,1), mgp = c(2,.25,0), las = 1, tck = 0.01)
-image(Time,X,R1, main = "Broad, long peak"); contour(Time,X,R1, add = TRUE)
+with(world.R1, {     
+  image(1:tau,X.mid,resource, main = "Broad, long peak")
+  contour(1:tau,X.mid,resource, add = TRUE)
+  })
+
 image(Time,X,R2, main = "Narrow, long peak"); contour(Time,X,R2, add = TRUE)
@@
 
-In these pulsed resources, we might assume that a *combination* of memory and tactic responses would yield more optimal behavion, and might observe some shift in the migratory strategy.
+In these pulsed resources, we might assume that a *combination* of memory and tactic responses would yield more optimal behavior, and might observe some shift in the migratory strategy.
 
@@
 parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
-Memory_PulsedResource_Sim <- runManyYears(pop.Year1, Time, Resource = R1, Parameters = parameters, n.years = 6)
+Memory_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(MemSim, tau = tau)
-plotYearList(MemSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Memory_PulsedResource_Sim, tau = tau)
+plotYearList(Memory_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)

### 2021-01-14 code/MinimalWorkingExample.R

@@ -1,0 +1,64 @@
+source("functions2.R")
+
+world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2, 
+                    peak.max = 80, peak.min = 20, sd = 10)
+world$resource <- getPulsedResource(world$time, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
+
+
+parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
+M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+
+plotYearList(M, tau = tau)
+world.R1 <- world %>% list_modify(pop = M$Year6)
+
+M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+plotYearList(M2, tau = tau)
+
+sqrt(mean((M[[2]]-M[[1]])^2))
+sqrt(mean((M2[[6]]-M2[[5]])^2))
+
+
+
+# Migratoriness Index
+
+  findMinOverlap <- function(P, world = world){
+    
+    getOverlap <- function(i1, i2, P){
+      sum(sqrt(P[i1,] * P[i2,])) * world$dx
+    }
+    
+    Overlap <- matrix(NA, world$tau, world$tau)
+    for(i in 1:(world$tau-1))
+      for(j in (i+1):world$tau)
+        Overlap[i,j] <- sum(sqrt(P[i,] * P[j,])) * world$dx
+    
+    minOverlap <-  min(Overlap, na.rm = TRUE)
+    list(times = which(Overlap == minOverlap, arr.ind = TRUE),
+         overlap = minOverlap)
+  }
+  
+  sapply(M, function(m) findMinOverlap(m)$overlap)
+  sapply(M2, function(m) findMinOverlap(m)$overlap)
+
+# Social cohesion index
+  
+  getSD <- function(f, world){
+    EX <- sum( world$X.mid * f) * world$dx
+    EX2 <- sum( world$X.mid^2 * f) * world$dx
+    sd  = sqrt(EX2 - EX^2)
+  }
+
+  computeCohesiveness <- function(P){
+    SDs <- apply(P, 1, getSD, world = world)
+    SD.max <- (1/sqrt(12))*max(world$X.mid)
+    SC <- 1 - SDs/SD.max
+    c(SC.mean = mean(SC), SC.sd = sd(SC))
+  }
+
[…]

### 2021-01-14 code/functions2.R

@@ -1,6 +1,7 @@
 require(deSolve)
 require(ReacTran)
 require(magrittr)
+require(purrr)
 #' Foraging model
 
 ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
@@ -18,23 +19,22 @@ ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource,
 runNextYear <- function(World, Parameters){
   pop1 <- World$pop
   pop2 <- pop1*0
-  pop2[1,] <- pop1[nrow(pop1),]
-  t.max <- World$tau
+  #pop2[1,] <- pop1[nrow(pop1),]
+  
   Time <- World$time
   Resource <- World$resource
   
-  for(i in 2:t.max){
-    if(i < t.max)
-      pop_lastyear <- c(0,pop1[i,],0) 
-      pop2[i,] <- ode(y = pop2[i-1,], 
-                     times = c(0,Time[i]-Time[i-1]), 
+  for(t in Time){
+     pop_lastyear <- c(0,pop1[t,],0) 
+     if(t == 1) pop_now <- pop1[nrow(pop1),] else pop_now <- pop2[t-1,]
+     pop2[t,] <- ode(y = pop_now, 
+                     times = 0:1, 
                      parms = Parameters,
                      func = ForagingMemoryModel, 
-                     resource = c(0,Resource[i,],0),
+                     resource = c(0,Resource[t,],0),
                      pop_lag = pop_lastyear,
                      dx = World$dx)[2,1:ncol(pop2)+1]
   }
-  
   pop2
 }
 
@@ -50,7 +50,6 @@ runManyYears <- function(World, Parameters, n.years){
                                    Parameters = parameters)
   }
   names(pop.list) <- paste0("Year",0:n.years)
-  pop.list <- lapply(pop.list, function(l) l[-1,])
   pop.list
 }
 

### 2021-01-15 code/StrategicVsTactical_PartI_v2.Rmd
fixed code for pulsed resource

@@
 
-R2 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5))
-R3 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9))
-R4 <- getPulsedResource(Time, X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5))
+world.R2 <- world %>% 
+  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5)))
+
+world.R3 <- world %>% 
+  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9)))
+
+world.R4 <- world %>% 
+  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5)))
 
@@
   })
+with(world.R2, {     
+  image(1:tau,X.mid,resource, main = "Narrow, long peak")
+  contour(1:tau,X.mid,resource, add = TRUE)
+  })
+with(world.R3, {     
+  image(1:tau,X.mid,resource, main = "Broad, short peak")
+  contour(1:tau,X.mid,resource, add = TRUE)
+  })
+with(world.R4, {     
+  image(1:tau,X.mid,resource, main = "Narrow, short peak")
+  contour(1:tau,X.mid,resource, add = TRUE)
+  })
 
-image(Time,X,R2, main = "Narrow, long peak"); contour(Time,X,R2, add = TRUE)
-image(Time,X,R3, main = "Broad, short peak"); contour(Time,X,R3, add = TRUE)
-image(Time,X,R4, main = "Narrow, short peak"); contour(Time,X,R4, add = TRUE)
 ```
@@
 # Running the model in a seasonal environment
+# Note: to view any of the resource models, change the world
 
@@
 parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
-Memory_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+Memory_PulsedResource_Sim <- runManyYears(world.R3, Parameters = parameters, n.years = 6)
 
@@
 ```{r PulseSim2, fig.width = 10}
-parameters <- c(epsilon = 2, alpha = 4e2, beta = 0, lambda = 0, dx = dx)
-Tactic_PulsedResource_Sim <- runManyYears(pop.Year1, Time, 
-                                  Resource = R1, Parameters = parameters, n.years = 6)
+parameters <- c(epsilon = 2, alpha = 4e2, beta0 = 0, beta1 = 0)
+Tactic_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+
 par(mfrow = c(1,2), oma = c(0,0,3,0))
@@
 ```{r PulseSim3, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 2e2, beta = 3e2, lambda = 0, dx = dx)
-Hybrid_PulsedResource_Sim <- runManyYears(pop.Year1, Time, 
-                                  Resource = R1, Parameters = parameters, n.years = 6)
+parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
+Hybrid_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+
 par(mfrow = c(1,2), oma = c(0,0,3,0))
@@
 title(printParameters(parameters), outer = TRUE)
-```
 
@@
 ```{r PulseSim4_tactic, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 2e2, beta = 0, lambda = 0, dx = dx)
-Hybrid_PulsedResource_Sim2 <- runManyYears(pop.Year1, Time, 
-                                  Resource = R4, Parameters = parameters, n.years = 6)
+parameters <- c(epsilon = 2, alpha = 2e2, beta0 = 0, beta1 = 0)
+Tactic_PulsedResource_Sim2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+
 par(mfrow = c(1,2), oma = c(0,0,3,0))
@@
 ```{r PulseSim4_memory, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 2e2, beta = 3e2, lambda = 0, dx = dx)
-Hybrid_PulsedResource_Sim2 <- runManyYears(pop.Year1, Time, 
-                                  Resource = R4, Parameters = parameters, n.years = 6)
+parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
+Hybrid_PulsedResource_Sim3 <- runManyYears(world.R4, Parameters = parameters, n.years = 6)
+
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim2, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim2, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Hybrid_PulsedResource_Sim3, tau = tau)
+plotYearList(Hybrid_PulsedResource_Sim3, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)

### 2021-01-16 code/IndicesEquations.Rmd
Fixed pulsed resource code and added Bhattacharyya coefficient code

---
title: "Indices Equations"
output: html_notebook
---

Migratoriness Index
To determine the migratoriness of a population, we calculate overlap of the population from two different years. We calculate the overlap using the equation:
$$\int \! \psi_1(x) \psi_2(x) \, \mathrm{d}x$$
where $\psi_1(x)$ represents the population in one year and $\psi_2(x)$ represents the population in another year. After calcuating this overlap of the two matrices that contain the population, we want to find the index that contains the minimum overlap value which the function below calculates.

```{r}
 findMinOverlap <- function(P, world = world){

    getOverlap <- function(i1, i2, P){
      sum(sqrt(P[i1,] * P[i2,])) * world$dx
    }

    Overlap <- matrix(NA, world$tau, world$tau)
    for(i in 1:(world$tau-1))
      for(j in (i+1):world$tau)
        Overlap[i,j] <- sum(sqrt(P[i,] * P[j,])) * world$dx

    minOverlap <-  min(Overlap, na.rm = TRUE)
    list(times = which(Overlap == minOverlap, arr.ind = TRUE),
         overlap = minOverlap)
  }

```

Foraging Success Score Index - Bhattacharyya Distance
In this function, we determine the forgaing success score index by comparing the similarity of the population with the resource using the Bhattacharyya distance. First we calculate the Bhattacharyya Coefficient $BC(p,q) from a function found in the R package bio3d. This is done by taking the integral of the square root of the product of the two statistical distributions we wish to compare: 
$$BC(p,q) = \int \! \sqrt{p(x)q(x)} \, \mathrm{d}x$$
Then we take the negative natural log of the Bhattacharrya Coefficient to determine the Bhattacharyya distance: $$D_B(p,q) = -ln(BC(p,q))$$

The Bhattacharyya coefficient determines the probability of the overlap between the two distributions. Thus $0 \ge BC(p,q) \ge 1$. A larger Bhattacharyya coefficient indicates a smaller Bhattacharyya distance to show that the two distributions are similar. A larger distance indicates that the population is farther away from the resource at the given year.

Bhattacharyya, A. On a measure of divergence between two statistical populations defined by their probability distributions. Bull. Calcutta Math. Soc. 35 (1943), 99–109.
```{r}
require(bio3d)
  BhattacharryyaDistance <- function(i1, i2){
    BhattacharyyaCoefficient <- bhattacharyya(cov(i1), cov(i2), q=99.9) #do we need a q value?
    distance <- -log(BhattacharyyaCoefficient)
    list(distance)
  }

```

Social Cohesion
To measure the social cohesion of a population, we calculate the standard deviation, $\sigma$  using the equation:
$$\sigma = \sqrt{E[(X-\mu)^2]} = \sqrt{E[X^2]-(E[X])^2}$$
where $\mu=E[X]$ represents the expected value (the average) of the population. The maximum standard deviation of the model is $\frac{1}{\sqrt{12}}$ so the social cohesion is defined with the equation: $$1-\frac{\sigma}{\sigma_{max}}$$

where $\sigma_{max}=\frac{1}{\sqrt{12}}.$ The larger the value of social cohesion indicates that the migratory population remained close with one another. 
```{r}
 getSD <- function(f, world){
    EX <- sum( world$X.mid * f) * world$dx
    EX2 <- sum( world$X.mid^2 * f) * world$dx
    sd  = sqrt(EX2 - EX^2)
  }

  computeCohesiveness <- function(P){
    SDs <- apply(P, 1, getSD, world = world)
    SD.max <- (1/sqrt(12))*max(world$X.mid)
    SC <- 1 - SDs/SD.max
    c(SC.mean = mean(SC), SC.sd = sd(SC))
  }

```

### 2021-01-16 code/MinimalWorkingExample.R

@@ -1,11 +1,13 @@
 source("functions2.R")
 
-world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2, 
+world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
                     peak.max = 80, peak.min = 20, sd = 10)
 world$resource <- getPulsedResource(world$time, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
 
 
 parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
+world.R1 <- world %>% 
+  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
 M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
 
 plotYearList(M, tau = tau)
@@ -59,6 +61,12 @@ sqrt(mean((M2[[6]]-M2[[5]])^2))
   sapply(M2, computeCohesiveness)
   
   
-  
-  
+  #Foraging Success Score
+  require(bio3d)
+  BhattacharryyaDistance <- function(i1, i2){
+    BhattacharyyaCoefficient <- bhattacharyya(cov(i1), cov(i2), q=99.9) #do we need a q value?
+    distance <- -log(BhattacharyyaCoefficient)
+    list(distance)
+  }
+  BhattacharryyaDistance(M[[1]],world$resource)
   

### 2021-01-16 code/IndicesEquations.Rmd
Changed X.mid to X

@@
  getSD <- function(f, world){
-    EX <- sum( world$X.mid * f) * world$dx
-    EX2 <- sum( world$X.mid^2 * f) * world$dx
+    EX <- sum( world$X * f) * world$dx
+    EX2 <- sum( world$X^2 * f) * world$dx
     sd  = sqrt(EX2 - EX^2)
@@
     SDs <- apply(P, 1, getSD, world = world)
-    SD.max <- (1/sqrt(12))*max(world$X.mid)
+    SD.max <- (1/sqrt(12))*max(world$X)
     SC <- 1 - SDs/SD.max

### 2021-01-16 code/StrategicVsTactical_PartI_v2.Rmd
Fixed pulsed resource code and added Bhattacharyya coefficient code
Changed X.mid to X

@@
 
-Out model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
+Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
 Importantly, the environment is *approximately periodic*, i.e.
@@
   X.edge <- seq(X.min, X.max, dx)
-  X.mid <- (X.edge[-1] + X.edge[-length(X.edge)])/2
-  pop <- outer(Time, X.mid,
+  X <- (X.edge[-1] + X.edge[-length(X.edge)])/2
+  pop <- outer(Time, X,
                      function(t,x) 
@@
   pop <- apply(pop, 1, function(x) x/sum(x)/dx) %>% t
-  list(pop = pop, X.mid = X.mid, time = Time, dx = dx, tau = tau)
+  list(pop = pop, X = X, time = Time, dx = dx, tau = tau)
 }
@@
 
-fr1 <- dnorm(world$X.mid, mean = 80, sd = 10) %>% {./sum(.)/world$dx}
-fr2 <- dnorm(world$X.mid, mean = 20, sd = 10) %>% {./sum(.)/world$dx}
+fr1 <- dnorm(world$X, mean = 80, sd = 10) %>% {./sum(.)/world$dx}
+fr2 <- dnorm(world$X, mean = 20, sd = 10) %>% {./sum(.)/world$dx}
 fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
@@
 with(world, 
-  image(X.mid, 1:tau, resource, main = "resource", ylab = "x")
+  image(X, 1:tau, resource, main = "resource", ylab = "x")
 )
@@
 world.R1 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
+  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
 
 world.R2 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5)))
+  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5)))
 
 world.R3 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9)))
+  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9)))
 
 world.R4 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5)))
+  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5)))
 
@@
 with(world.R1, {     
-  image(1:tau,X.mid,resource, main = "Broad, long peak")
-  contour(1:tau,X.mid,resource, add = TRUE)
+  image(1:tau,X,resource, main = "Broad, long peak")
+  contour(1:tau,X,resource, add = TRUE)
   })
 with(world.R2, {     
-  image(1:tau,X.mid,resource, main = "Narrow, long peak")
-  contour(1:tau,X.mid,resource, add = TRUE)
+  image(1:tau,X,resource, main = "Narrow, long peak")
+  contour(1:tau,X,resource, add = TRUE)
   })
 with(world.R3, {     
-  image(1:tau,X.mid,resource, main = "Broad, short peak")
-  contour(1:tau,X.mid,resource, add = TRUE)
+  image(1:tau,X,resource, main = "Broad, short peak")
+  contour(1:tau,X,resource, add = TRUE)
   })
 with(world.R4, {     
-  image(1:tau,X.mid,resource, main = "Narrow, short peak")
-  contour(1:tau,X.mid,resource, add = TRUE)
+  image(1:tau,X,resource, main = "Narrow, short peak")
+  contour(1:tau,X,resource, add = TRUE)
   })

### 2021-01-16 code/MinimalWorkingExample.R

@@ -2,12 +2,12 @@ source("functions2.R")
 
 world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
                     peak.max = 80, peak.min = 20, sd = 10)
-world$resource <- getPulsedResource(world$time, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
+world$resource <- getPulsedResource(world$time, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
 
 
 parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
 world.R1 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X.mid, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
+  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
 M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
 
 plotYearList(M, tau = tau)
@@ -45,14 +45,14 @@ sqrt(mean((M2[[6]]-M2[[5]])^2))
 # Social cohesion index
   
   getSD <- function(f, world){
-    EX <- sum( world$X.mid * f) * world$dx
-    EX2 <- sum( world$X.mid^2 * f) * world$dx
+    EX <- sum( world$X * f) * world$dx
+    EX2 <- sum( world$X^2 * f) * world$dx
     sd  = sqrt(EX2 - EX^2)
   }
 
   computeCohesiveness <- function(P){
     SDs <- apply(P, 1, getSD, world = world)
-    SD.max <- (1/sqrt(12))*max(world$X.mid)
+    SD.max <- (1/sqrt(12))*max(world$X)
     SC <- 1 - SDs/SD.max
     c(SC.mean = mean(SC), SC.sd = sd(SC))
   }

### 2021-01-19 code/IndicesEquations.Rmd
Updated Indices

@@
 title: "Indices Equations"
-output: html_notebook
+author: "E. Gurarie, S. Potluri, B. Fagan, others"
+date: "`r Sys.Date()`"
+output:
+  html_document:
+    toc: yes
+    toc_float: yes
 ---
 
-Migratoriness Index
-To determine the migratoriness of a population, we calculate overlap of the population from two different years. We calculate the overlap using the equation:
-$$\int \! \psi_1(x) \psi_2(x) \, \mathrm{d}x$$
-where $\psi_1(x)$ represents the population in one year and $\psi_2(x)$ represents the population in another year.
-After calcuating this overlap of the two matrices that contain the population, we want to find the index that contains the minimum overlap value which the function below calculates.
+Simulations are run until a (quasi)-equilibrium state is achieved, where "state" refers to the population distribution across an entire year.
+Once that state is attained, we obtain three metrics that characterize the spatial distribution, the migratory behavior, and the foraging efficiency.
+All of these indices are constrained to be between 0 and 1.
+
+## Migratoriness
+
+To determine the migratoriness of a population, we calculate overlap of the population from two different times within the year, choosing those times to minimize the overlap, and reporting the overal.
+The overlap is defined as:
+
+$$O(t_1, t_2) = \int_D \sqrt{u(x, t_1) \, u(x, t_2)} \, dx; $$
+After calcuating this overlap of the two matrices that contain the population, we find the values of $t_1$ and $t_2$ that minimize $O$, corresponding to the periods of maximum distance.
+Thus, the migratoriness index is:
+
+$$MI = min(O(t_1, t_2))$$
 
@@
 
-Foraging Success Score Index - Bhattacharyya Distance
-In this function, we determine the forgaing success score index by comparing the similarity of the population with the resource using the Bhattacharyya distance.
-First we calculate the Bhattacharyya Coefficient $BC(p,q) from a function found in the R package bio3d.
-This is done by taking the integral of the square root of the product of the two statistical distributions we wish to compare:
-$$BC(p,q) = \int \! \sqrt{p(x)q(x)} \, \mathrm{d}x$$
-Then we take the negative natural log of the Bhattacharrya Coefficient to determine the Bhattacharyya distance: $$D_B(p,q) = -ln(BC(p,q))$$
+## Foraging Success
+
+To quantify the consumers’ ability to track the distribution of their resources over space and time, we use the continuous form of the Bhattacharya Coefficient (BC; Bhattacharyya 1943).
+The BC quantifies similarity between two distributions; we take the mean of the BC over the equilibrium year.
+Thus:
+$$BC = {1\over T} \int_T\int_D \sqrt{u(x,t) \, h(x,t)} \, dx\,dt$$
+Where the spatial integral is taken over the domain ($D$).
 
@@
 
-Bhattacharyya, A. On a measure of divergence between two statistical populations defined by their probability distributions. Bull. Calcutta Math. Soc. 35 (1943), 99–109.
 ```{r}
 require(bio3d)
-  BhattacharryyaDistance <- function(i1, i2){
+  BhattacharyyaDistance <- function(i1, i2){
     BhattacharyyaCoefficient <- bhattacharyya(cov(i1), cov(i2), q=99.9) #do we need a q value?
@@
   }
-
 ```
 
-Social Cohesion
-To measure the social cohesion of a population, we calculate the standard deviation, $\sigma$  using the equation:
-$$\sigma = \sqrt{E[(X-\mu)^2]} = \sqrt{E[X^2]-(E[X])^2}$$
-where $\mu=E[X]$ represents the expected value (the average) of the population.
-The maximum standard deviation of the model is $\frac{1}{\sqrt{12}}$ so the social cohesion is defined with the equation: $$1-\frac{\sigma}{\sigma_{max}}$$
+We used the `bhattacharyya` function in the `bio3d` package (Grant et al. 2006).
 
-where $\sigma_{max}=\frac{1}{\sqrt{12}}.$ The larger the value of social cohesion indicates that the migratory population remained close with one another. 
+## Social Cohesion
+
+To measure the social cohesion (SC) of a population, we simply calculate the mean standard deviation of the population across the equilibrium year:
+
+$$\overline{\sigma} = {1\over \tau} \int \sigma_u(t) \,dt$$
+
+where $$\sigma_u^2(t) = \int_D \left(x - \mu_u(t)\right)^2 \, u(x,t)\, dx$$
+
+and $\mu_u$ is the mean of $u$ at time $t$.
+If the population is uniformly distributed over the domain, the standard deviation of the model would be $\frac{X_{max}}{\sqrt{12}}$.
+Where therefore define the social cohesion index as: $$SC = 1-\frac{\sqrt{12} \, }{X_{max}}\overline{\sigma}$$
+
+where $\sigma_{max}=\frac{1}{\sqrt{12}}.$  In theory, SC can be lower than 1 if, for example, the population concentrates in two groups at the edges of the domain, but in practice the SC is positive.
+The maximum value of 1 would correspond to the entire population .
 ```{r}
@@
 ```
+
+## References
+
+Bhattacharyya, A. On a measure of divergence between two statistical populations defined by their probability distributions. Bull. Calcutta Math. Soc. 35 (1943), 99–109.
+
+Grant, B.J. et al. (2006) Bioinformatics 22, 2695--2696.

### 2021-01-20 code/IndicesEquations.Rmd
Edited BhattacharyyaCoefficient function

@@
 require(bio3d)
-  BhattacharyyaDistance <- function(i1, i2){
-    BhattacharyyaCoefficient <- bhattacharyya(cov(i1), cov(i2), q=99.9) #do we need a q value?
-    distance <- -log(BhattacharyyaCoefficient)
-    list(distance)
-  }
+  BhattacharyyaDCoefficient <- function(i1, i2){
+    BC <- bhattacharyya(cov(i1), cov(i2)) #do we need a q value? 
+    list(BC)
+  } #does this function take the mean of the BC over the year? 
 ```

### 2021-01-20 code/MinimalWorkingExample.R

@@ -38,7 +38,6 @@ sqrt(mean((M2[[6]]-M2[[5]])^2))
     list(times = which(Overlap == minOverlap, arr.ind = TRUE),
          overlap = minOverlap)
   }
-  
   sapply(M, function(m) findMinOverlap(m)$overlap)
   sapply(M2, function(m) findMinOverlap(m)$overlap)
 
@@ -63,10 +62,9 @@ sqrt(mean((M2[[6]]-M2[[5]])^2))
   
   #Foraging Success Score
   require(bio3d)
-  BhattacharryyaDistance <- function(i1, i2){
-    BhattacharyyaCoefficient <- bhattacharyya(cov(i1), cov(i2), q=99.9) #do we need a q value?
-    distance <- -log(BhattacharyyaCoefficient)
-    list(distance)
+  BhattacharyyaCoefficient <- function(i1, i2){
+    BC <- bhattacharyya(cov(i1), cov(i2)) #do we need a q value?
+    list(BC)
   }
-  BhattacharryyaDistance(M[[1]],world$resource)
+  BhattacharyyaCoefficient(M[[2]],world$resource)
   

### 2021-01-21 code/StrategicVsTactical_PartI_v2.Rmd
fixed indices

@@
     toc_float: yes
-  pdf_document:
-    toc: yes
-  word_document:
-    toc: yes
 theme: cerulean
@@
 plotYearList(Hybrid_PulsedResource_Sim, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Hybrid_PulsedResource_Sim, tau = tau, 
+             persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
+```
 

### 2021-01-21 code/MinimalWorkingExample.R

@@ -1,70 +1,48 @@
-source("functions2.R")
+  source("code/functions2.R")
 
-world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
-                    peak.max = 80, peak.min = 20, sd = 10)
-world$resource <- getPulsedResource(world$time, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
-
-
-parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
-world.R1 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
-M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
-
-plotYearList(M, tau = tau)
-world.R1 <- world %>% list_modify(pop = M$Year6)
-
-M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
-plotYearList(M2, tau = tau)
-
-sqrt(mean((M[[2]]-M[[1]])^2))
-sqrt(mean((M2[[6]]-M2[[5]])^2))
-
-
-
+# Initialize world
+  
+  world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
+                      peak.max = 80, peak.min = 20, sd = 10)
+  world$resource <- getPulsedResource(world$time, world$X, c(t.peak = 25, t.sd = 9, 
+                                                             x.peak = 80, x.sd = 9))
+  
+  
+  parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
+  world.R1 <- world %>% 
+    list_modify(resource = getPulsedResource(1:world$tau, world$X, 
+                                             c(t.peak = 25, t.sd = 9, 
+                                               x.peak = 80, x.sd = 9)))
+  
+# run simulation 12 years
+  
+  M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+  plotYearList(M, tau = tau)
+  
+  world.R1 <- world %>% list_modify(pop = M$Year6)
+  M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+  plotYearList(M2, tau = tau)
+  
+  source("code/functions_indices.R")
+  
+  computeCohesiveness(world$pop, world)["SC.mean"]
[…]

### 2021-01-21 code/functions2.R

@@ -120,3 +120,19 @@ getPulsedResource <- function(t, x, par){
   R <- apply(R, 1, function(x) x/sum(x)) %>% t
   return(R)
 }
+
+
+#' Initialize sinusoidal year 0
+
+getSinePop <- function(tau, X.min = 0, X.max = 100, dx, 
+                       peak.max, peak.min, sd){
+  Time <- 1:tau
+  
+  X.edge <- seq(X.min, X.max, dx)
+  X <- (X.edge[-1] + X.edge[-length(X.edge)])/2
+  pop <- outer(Time, X,
+               function(t,x) 
+                 dnorm(x, mean = X.max/2 + (peak.max-peak.min) / 2 * sin( 2*(t * pi)/max(Time)),sd = sd))
+  pop <- apply(pop, 1, function(x) x/sum(x)/dx) %>% t
+  list(pop = pop, X = X, time = Time, dx = dx, tau = tau)
+}

### 2021-01-21 code/functions_indices.R

@@ -1,0 +1,42 @@
+require(bio3d)
+
+getSD <- function(f, world){
+  EX <- sum( world$X * f) * world$dx
+  EX2 <- sum( world$X^2 * f) * world$dx
+  sd  = sqrt(EX2 - EX^2)
+}
+
+computeCohesiveness <- function(pop, world){
+  SDs <- apply(pop, 1, getSD, world = world)
+  SD.max <- (1/sqrt(12))*max(world$X)
+  SC <- 1 - SDs/SD.max
+  c(SC.mean = mean(SC), SC.sd = sd(SC))
+}
+
+
+computeEfficiency <- function(pop, resource){
+  bhattacharyya(cov(pop), cov(resource)) #do we need a q value? 
+}
+
+computeMigratoriness <- function(pop, world){
+
+  dx <- world$dx
+  tau <- world$tau
+  
+  Overlap <- matrix(NA, tau, tau)
+  for(i in 1:(tau-1))
+    for(j in (i+1):tau)
+      Overlap[i,j] <- sum(sqrt(pop[i,] * pop[j,])) * dx
+  
+  minOverlap <-  1-min(Overlap, na.rm = TRUE)
+  list(times = which(Overlap == minOverlap, arr.ind = TRUE),
+       overlap = minOverlap)
+}
+
+
+computeIndices <- function(pop, resource, world){
+  SC = computeCohesiveness(pop, world)["SC.mean"]
+  MI = computeMigratoriness(pop, world)$overlap
+  FE = computeEfficiency(pop, resource)
+  data.frame(SC, MI, FE)
+}

### 2021-01-22 code/IndicesEquations.Rmd
refinements
Changed foraging efficiency function and added example of using indices for a model

@@
 After calcuating this overlap of the two matrices that contain the population, we find the values of $t_1$ and $t_2$ that minimize $O$, corresponding to the periods of maximum distance.
-Thus, the migratoriness index is:
+We subtract this minimum value to calculate migratoriness of the population.
+Thus, the migratoriness index (MI) is:
 
-$$MI = min(O(t_1, t_2))$$
+$$MI = 1-min(O(t_1, t_2))$$
 
 ```{r}
- findMinOverlap <- function(P, world = world){
+ computeMigratoriness <- function(pop, world){
 
-    getOverlap <- function(i1, i2, P){
-      sum(sqrt(P[i1,] * P[i2,])) * world$dx
-    }
+  dx <- world$dx
+  tau <- world$tau
 
-    Overlap <- matrix(NA, world$tau, world$tau)
-    for(i in 1:(world$tau-1))
-      for(j in (i+1):world$tau)
-        Overlap[i,j] <- sum(sqrt(P[i,] * P[j,])) * world$dx
+  Overlap <- matrix(NA, tau, tau)
+  for(i in 1:(tau-1))
+    for(j in (i+1):tau)
+      Overlap[i,j] <- sum(sqrt(pop[i,] * pop[j,])) * dx
 
-    minOverlap <-  min(Overlap, na.rm = TRUE)
+  minOverlap <-  1-min(Overlap, na.rm = TRUE)
   list(times = which(Overlap == minOverlap, arr.ind = TRUE),
@@
 
-## Foraging Success
+## Foraging Efficiency
 
-To quantify the consumers’ ability to track the distribution of their resources over space and time, we use the continuous form of the Bhattacharya Coefficient (BC; Bhattacharyya 1943).
-The BC quantifies similarity between two distributions; we take the mean of the BC over the equilibrium year.
-Thus:
-$$BC = {1\over T} \int_T\int_D \sqrt{u(x,t) \, h(x,t)} \, dx\,dt$$
-Where the spatial integral is taken over the domain ($D$).
-
-The Bhattacharyya coefficient determines the probability of the overlap between the two distributions.
-Thus $0 \ge BC(p,q) \ge 1$.
-A larger Bhattacharyya coefficient indicates a smaller Bhattacharyya distance to show that the two distributions are similar.
-A larger distance indicates that the population is farther away from the resource at the given year.
+To quantify the consumers’ ability to track the distribution of their resources over space and time, we use the continuous form of the Bhattacharyya Coefficient (Bhattacharyya 1943).
+The Bhattacharyya Coefficient quantifies similarity between two distributions; we take the mean of the coefficient over the equilibrium year to determine foraging efficiency (FE).
+Thus, the foraging efficiency index is:
+$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{X_{max}} \sqrt{u(x,t) \, h(x,t)} \, dx\,dt$$
+Where the spatial integral is taken over the domain.
 
 ```{r}
-require(bio3d)
-  BhattacharyyaDCoefficient <- function(i1, i2){
-    BC <- bhattacharyya(cov(i1), cov(i2)) #do we need a q value? 
-    list(BC)
-  } #does this function take the mean of the BC over the year? 
+computeEfficiency <- function(pop, resource, world){
+
+  dx <- world$dx
+  tau <- world$tau
+
+  Similarity <- matrix(NA, tau, tau)
+  for(i in 1:(tau-1))
+    for(j in (i+1):tau)
+      Similarity[i,j] <- sum(sqrt(pop[i,] * resource[j,])) * dx
+
+  FS <- mean(Similarity, na.rm=TRUE)/tau
+}
 ```
-
-We used the `bhattacharyya` function in the `bio3d` package (Grant et al. 2006).
 
@@
 
-$$\overline{\sigma} = {1\over \tau} \int \sigma_u(t) \,dt$$
+$$\overline{\sigma} = {1\over \tau} \int \sigma_u(x,t) \,dt$$
 
@@
 If the population is uniformly distributed over the domain, the standard deviation of the model would be $\frac{X_{max}}{\sqrt{12}}$.
-Where therefore define the social cohesion index as: $$SC = 1-\frac{\sqrt{12} \, }{X_{max}}\overline{\sigma}$$
+We therefore define the social cohesion (SC) index as: $$SC = 1-\frac{\sqrt{12} \, }{X_{max}}\overline{\sigma}$$
 
@@
 ```{r}
- getSD <- function(f, world){
-    EX <- sum( world$X * f) * world$dx
-    EX2 <- sum( world$X^2 * f) * world$dx
+ getSD <- function(pop, world){
+    EX <- sum( world$X * pop) * world$dx
+    EX2 <- sum( world$X^2 * pop) * world$dx
     sd  = sqrt(EX2 - EX^2)
@@
 
-  computeCohesiveness <- function(P){
-    SDs <- apply(P, 1, getSD, world = world)
+  computeCohesiveness <- function(pop){
+    SDs <- apply(pop, 1, getSD, world = world)
     SD.max <- (1/sqrt(12))*max(world$X)
@@
 
+## Combining the Indices
+The following function calculates these three indices for a specific population and resource. 
+```{r}
+computeIndices <- function(pop, resource, world){
+  SC = computeCohesiveness(pop, world)["SC.mean"]
+  MI = computeMigratoriness(pop, world)$overlap
+  FE = computeEfficiency(pop, resource, world)
+  data.frame(SC, MI, FE)
+}
+```
+
 ## References  
+Bhattacharyya, A.
+(1943).
+"On a measure of divergence between two statistical populations defined by their probability distributions".
+Bulletin of the Calcutta Mathematical Society.
+35: 99–109.
+MR 0010358
 
-Bhattacharyya, A. On a measure of divergence between two statistical populations defined by their probability distributions. Bull. Calcutta Math. Soc. 35 (1943), 99–109.
-
-Grant, B.J. et al. (2006) Bioinformatics 22, 2695--2696.
+```

### 2021-01-22 code/StrategicVsTactical_PartI_v2.Rmd
Changed foraging efficiency function and added example of using indices for a model

@@
 ```{r, message=FALSE, warning=FALSE}
-require(deSolve)
-require(ReacTran)
-require(magrittr)
+source("functions2.R")
 ```
@@
 
-```{r}
-source("functions2.R")
-```
-
 In the following example, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is fixed in the middle of the space (as above).
@@
 In this context, migration is not, in fact, optimal.
-But the important piece is to see how the reponse mixes attraction to the resource with attraction to the previous year's behavior.
+But the important piece is to see how the response mixes attraction to the resource with attraction to the previous year's behavior.
 Note that the year here (for simplicity, speed and symmetry), is 100 days long.
@@
 
+## Example
+We apply the three indices to an example model.
+In this model, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is seasonal in a broad, long peak.
+The model uses the parameters $\epsilon = 5, \alpha = 300, \beta0=100, \beta1 = 300$ so that the migratory population combines both memory and a tactical strategy.
+We run the model for 12 years and compute the three indices.
+Then we plot the three indices to the view the trend of the migratory population over time.
+
+```{r}
+source("functions_indices.R")
+  world$resource <- getPulsedResource(world$time, world$X, c(t.peak = 25, t.sd = 9, 
+                                                             x.peak = 80, x.sd = 9))
+
+  parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
+  world.R1 <- world %>% 
+    list_modify(resource = getPulsedResource(1:world$tau, world$X, 
+                                             c(t.peak = 25, t.sd = 9, 
+                                               x.peak = 80, x.sd = 9)))
+  M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+
+  world.R1 <- world %>% list_modify(pop = M$Year6)
+  M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+
+  source("functions_indices.R")
+  indices1 <- sapply(M, computeIndices, resource = world$resource, world = world) %>% t
+  indices2 <- sapply(M2, computeIndices, resource = world$resource, world = world) %>% t
+  I <- rbind(indices1, indices2)
+
+  matplot(I, type="o", pch = 19, lty = 1) #year 6 and year 7 are the same so it is plotted twice 
+  legend("topright", pch = 19, col = 1:3, legend = colnames(I))
+
+```
+
 # Some summaries

### 2021-01-22 code/MinimalWorkingExample.R

@@ -1,4 +1,5 @@
-  source("code/functions2.R")
+  source("functions2.R")
+  source("functions_indices.R")
 
 # Initialize world
   
@@ -18,6 +19,8 @@
   
   M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
   plotYearList(M, tau = tau)
+
+  
   
   world.R1 <- world %>% list_modify(pop = M$Year6)
   M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
@@ -27,8 +30,9 @@
   
   computeCohesiveness(world$pop, world)["SC.mean"]
   computeMigratoriness(world$pop, world)$overlap
-  with(world, computeEfficiency(pop, resource))
+  computeEfficiency(world$pop, world$resource, world)
   computeIndices(world$pop, world$resource, world)
+  computeIndices(M[[2]], world$resource, world)
   
 # Migratoriness Index
 
@@ -45,4 +49,6 @@
          resource = world.R1$resource) %>% t
   matplot(I.v2, type="o", pch = 19, lty = 1)
   legend("topright", pch = 19, col = 1:3, legend = colnames(I))
+  
+
   

### 2021-01-22 code/functions_indices.R

@@ -1,21 +1,27 @@
-require(bio3d)
-
-getSD <- function(f, world){
-  EX <- sum( world$X * f) * world$dx
-  EX2 <- sum( world$X^2 * f) * world$dx
+getSD <- function(pop, world){
+  EX <- sum( world$X * pop) * world$dx
+  EX2 <- sum( world$X^2 * pop) * world$dx
   sd  = sqrt(EX2 - EX^2)
 }
 
-computeCohesiveness <- function(pop, world){
+computeCohesiveness <- function(pop){
   SDs <- apply(pop, 1, getSD, world = world)
   SD.max <- (1/sqrt(12))*max(world$X)
   SC <- 1 - SDs/SD.max
   c(SC.mean = mean(SC), SC.sd = sd(SC))
 }
 
-
-computeEfficiency <- function(pop, resource){
-  bhattacharyya(cov(pop), cov(resource)) #do we need a q value? 
+computeEfficiency <- function(pop, resource, world){
+  
+  dx <- world$dx
+  tau <- world$tau
+  
+  Similarity <- matrix(NA, tau, tau)
+  for(i in 1:(tau-1))
+    for(j in (i+1):tau)
+      Similarity[i,j] <- sum(sqrt(pop[i,] * resource[j,])) * dx
+  
+  FE <- mean(Similarity, na.rm=TRUE)/tau
 }
 
 computeMigratoriness <- function(pop, world){
@@ -33,10 +39,9 @@ computeMigratoriness <- function(pop, world){
        overlap = minOverlap)
 }
 
-
 computeIndices <- function(pop, resource, world){
   SC = computeCohesiveness(pop, world)["SC.mean"]
   MI = computeMigratoriness(pop, world)$overlap
-  FE = computeEfficiency(pop, resource)
+  FE = computeEfficiency(pop, resource, world)
   data.frame(SC, MI, FE)
 }

### 2021-02-03 code/IndicesEquations.Rmd
Wrote out functions and descriptions for R package

@@
 
-To determine the migratoriness of a population, we calculate overlap of the population from two different times within the year, choosing those times to minimize the overlap, and reporting the overal.
+To determine the migratoriness of a population, we calculate overlap of the population from two different times within the year, choosing those times to minimize the overlap, and reporting the overall overlap.
 The overlap is defined as:

### 2021-02-03 code/MinimalWorkingExample.R

@@ -1,6 +1,4 @@
-  source("functions2.R")
-  source("functions_indices.R")
-
+require("memorymigration")
 # Initialize world
   
   world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 

### 2021-02-03 memorymigration/R/ForagingMemoryModel.R

@@ -5,7 +5,8 @@
 #' @param t vector of times
 #' @param pop a T x X matrix describing the population distribution 
 #' across the time period.
-#' @param parms named vector of parameters. THese are \code{alpha} - resource 
+#' @param parms named vector of parameters. These are \code{epsilon} - diffusion coefficient; 
+#' \code{alpha} - resource 
 #' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
 #' memory coefficient
 #' @param pop_lag the T x X population distribution over time in the previous cycle (year)

### 2021-02-03 memorymigration/R/PlottingFunctions.R

@@ -1,4 +1,14 @@
-#' Plotting Functions
+#' Plot Year List
+#' 
+#' This function plots the migratoriness population distribution over a time period. 
+#' 
+#' @param yearlist a list of T x X matrices describing the population 
+#' distribution for each year after initial population; produced by the runManyYear function
+#' @param X.max maximum value of X-axis
+#' @param tau maximum value of time T
+#' @param log logical;
+#' @param persp logical;
+#' @seealso \link{runManyYears}; \link{printParameters}
 
 plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE, 
                          persp = FALSE, ...){
@@ -22,9 +32,26 @@ plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE,
     mypersp(time, x, allyears, ...) 
 }
 
+#' Print Parameters 
+#' 
+#' This function prints the parameters of the diffusion-advection equation used 
+#' on the model on the plot of the population distribution set up by the plotYearList function.
+#' 
+#'  @param p List of 5 a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Set up by the getSinePop function. 
+#'  @return displays the values of the parameters on the plot which are:
+#'   \code{epsilon} - diffusion coefficient; \code{alpha} - resource 
+#' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
+#' memory coefficient
+#'  @seealso \link{PlotYearList}; \link{getSinePop}
+#'  
 printParameters <- function(p) {
   parse(text = paste0("list(",paste(names(p), p, sep = "==", collapse = ", "),")"))
 }
 
+#' My Persp 
+#' 
+#' This function...  
 mypersp <- function(x,y,z,...)
   persp(x,y,z, theta=45, phi=45, bor=NA, shade=TRUE, ylab = "x", xlab="time", zlab="population", ...)

### 2021-02-03 memorymigration/R/RunningModel.R

@@ -1,4 +1,18 @@
-#' Run One Year forward
+#' Run Next Year
+#' 
+#' Based on a migratory population's set up (the World) and the values \code{alpha},
+#'\code{beta0} and \code{beta1}, this function determines the population 
+#'distribution for the following year. 
+#' 
+#' @param World List of 5 a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Set up by the getSinePop function. 
+#' @param Parameters named vector of parameters. \code{epsilon} - diffusion coefficient; 
+#' \code{alpha} - resource 
+#' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
+#' memory coefficient
+#' @return a T x X matrix describing the population distribution for the next year
+#' @seealso \link{getSinePop}, \link{runManyYears}
 
 runNextYear <- function(World, Parameters){
   pop1 <- World$pop
@@ -22,7 +36,23 @@ runNextYear <- function(World, Parameters){
   pop2
 }
 
-#' Run several years forward
+#' Run Many Years 
+#' 
+#' Based on a migratory population's set up (the World) and the values \code{alpha},
+#'\code{beta0} and \code{beta1}, this function determines the population 
+#'distribution after several years. 
+#'
+#' @param World List of 5 a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Set up by the getSinePop function. 
+#' @param Parameters named vector of parameters. These are \code{epsilon} - diffusion coefficient; 
+#' \code{alpha} - resource 
+#' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
+#' memory coefficient
+#' @param n.years number of years the population migrates 
+#' @return a list of n.years containing T x X matrices describing the population 
+#' distribution for each year after initial population
+#' @seealso \link{getSinePop}, \link{runManyYears}
 
 runManyYears <- function(World, Parameters, n.years){
   pop.list <- list(Year1 = World$pop)
@@ -36,3 +66,4 @@ runManyYears <- function(World, Parameters, n.years){
   names(pop.list) <- paste0("Year",0:n.years)
   pop.list
 }
+

### 2021-02-03 memorymigration/R/functions_indices.R

@@ -35,13 +35,29 @@ computeCohesiveness <- function(pop){
 #' @param resource The resource in an given cycle to compare to pop
 #' @param world world object
 #' @return a single foraging efficiecy (FE) index
-#' @example examples/indices_examples.
+#' @seealso \link{computeCohesiveness}, \link{computeMigratoriness}, \link{computeIndices}
+#' @example examples/indices_examples.R
 
 computeEfficiency <- function(pop, resource, world){
   dx <- world$dx
   tau <- world$tau
   sum(sqrt(pop*resource))*dx/tau
 }
+
+#' Compute Migratoriness
+#' 
+#' Returns an migratoriness index of a population by comparing the overlap 
+#' overlap of the population from two different times within the year, 
+#' choosing those times to minimize the overlap, and reporting the 
+#' overall overlap.
+#' 
+#' @param pop One cycle of population process
+#' @param world world object
+#' @return a list with the time of the minimal overlap and the Migratoriness 
+#' (MI) index 
+#' @seealso \link{computeEfficiency}, \link{computeCohesiveness}, \link{computeIndices}
+#' @example examples/indices_examples.R
+
 
 computeMigratoriness <- function(pop, world){
 
@@ -57,6 +73,21 @@ computeMigratoriness <- function(pop, world){
   list(times = which(Overlap == minOverlap, arr.ind = TRUE),
        overlap = minOverlap)
 }
+
+#' Compute Indicies
+#' 
+#' Returns the three migratoriness indeces calculated by the cumputeCohesiveness,
+#' commputeEfficiency, and computeMigratoriness functions for a population and 
+#' distribution and a resource distribution. 
+#' 
+#'
+#' @param pop One cycle of population process
+#' @param resource The resource in an given cycle to compare to pop
+#' @param world world object
+#' @return a data frame containing the Social Cohesiveness (SC) index,
+#' the Migratoriness (MI) index, and the Foraging Efficiency (FE) index
+#' @seealso \link{computeEfficiency}, \link{computeMigratoriness}, \link{computeCohesiveness}
+#' @example examples/indices_examples.R
 
 computeIndices <- function(pop, resource, world){
   SC = computeCohesiveness(pop, world)["SC.mean"]

### 2021-02-03 memorymigration/R/getPulsedResource.R

@@ -1,4 +1,25 @@
-#' getPulsedResource
+#' Get Pulsed Resource
+#' 
+#' This function generates a a seasonal resource function with the following properties:
+#' 1. The total amount of resource across space is constant throughout the year.
+#' 2. At the beginning, middle, and end of the year the resource is uniformly distributed. 
+#' 3. At some peak time $t_r < \tau/2$, the resource concentrates at a location $x_r < \chi/2$
+#'  with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ 
+#'  is the length of the year and $\chi$ is the extent of the spatial domain).
+#' 4. The resource peaks exactly symmetrically at time $\tau - t_r$ and location 
+#' $\chi - x_r$ with the same variance $\sigma_r$.
+#' 
+#' To generate a resource with these properties, we distributing the resource in 
+#' space as a beta distribution, where the two shape and scale parameters vary 
+#' sinusoidally in such a way as to fulfill the criteria above. 
+#' 
+#' @param t vector with time values 
+#' @param x vector with X-axis values
+#' @param par vector with four values: t.peak (which is the maximum peak of t), 
+#' t.sd (which is the standard deviation of t), x.peak (which is the maximum peak of x),
+#' and x.sd (which is the standard deviation of x)
+#' @return A X x X matrix containing values of the resource distribution
+
 
 getPulsedResource <- function(t, x, par){
   

### 2021-02-03 memorymigration/R/getSinePop.R

@@ -1,0 +1,28 @@
+#' Get Sine Pop
+#' 
+#' Initializes sinusoidal year 0 of a migratory population distribution 
+#' via a sinusoidal migration pattern
+#' 
+#' @param tau maximum time in days for migratory population in one year
+#' @param X.min minimum value of population distribution
+#' @param X.max maximum value of population distribution
+#' @param dx time step
+#' @param peak.max minimum value of sinusoidal peak
+#' @param peak.min maximum value of sinusoidal peak
+#' @param sd standard deviation of population distribution
+#' @return list of 5 which includes a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value.
+
+getSinePop <- function(tau, X.min = 0, X.max = 100, dx, 
+                       peak.max, peak.min, sd){
+  Time <- 1:tau
+  
+  X.edge <- seq(X.min, X.max, dx)
+  X <- (X.edge[-1] + X.edge[-length(X.edge)])/2
+  pop <- outer(Time, X,
+               function(t,x) 
+                 dnorm(x, mean = X.max/2 + (peak.max-peak.min) / 2 * sin( 2*(t * pi)/max(Time)),sd = sd))
+  pop <- apply(pop, 1, function(x) x/sum(x)/dx) %>% t
+  list(pop = pop, X = X, time = Time, dx = dx, tau = tau)
+}

### 2021-02-03 memorymigration/examples/indices_examples.R

@@ -14,6 +14,7 @@ world.R1 <- world %>%
 
 # run simulation 12 years
 
+runNextYear(world.R1, Parameters=parameters)
 M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
 plotYearList(M, tau = tau)
 
@@ -23,8 +24,6 @@ resource <- world$resource
 world.R1 <- world %>% list_modify(pop = M$Year6)
 M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
 plotYearList(M2, tau = tau)
-
-source("code/functions_indices.R")
 
 computeCohesiveness(world$pop, world)["SC.mean"]
 computeMigratoriness(world$pop, world)$overlap

### 2021-02-03 memorymigration/R/PlottingFunctions.R
debugging

@@ -7,8 +7,9 @@
 #' @param X.max maximum value of X-axis
 #' @param tau maximum value of time T
 #' @param log logical;
-#' @param persp logical;
+#' @param persp whether to plot an image / contour or perspective plot
 #' @seealso \link{runManyYears}; \link{printParameters}
+#' @example examples/plotting_examples.R
 
 plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE, 
                          persp = FALSE, ...){
@@ -25,10 +26,9 @@ plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE,
   if(!persp){
     image(time, x, allyears)
     contour(time, x, allyears, add = TRUE)  
-    
     mtext(side = 3, at = (1:n.years)*tau - tau/2, text = paste("Year", 1:n.years-1))
     abline(v = tau*1:(n.years-1), col = "grey") 
-  } else
+  } else 
     mypersp(time, x, allyears, ...) 
 }
 

### 2021-02-07 code/StrategicVsTactical_PartI_v2.Rmd
edited R package code and minor changes to running model

@@
 ```{r, message=FALSE, warning=FALSE}
-source("functions2.R")
+require("memorymigration")
 ```
@@
 if(eval){
-  MemSim <- runManyYears(World, Parameters = parameters, 
-                         n.years = 3)
+  MemSim <- runManyYears(world, Parameters = parameters, 
+                         n.years = 3, 0.99)
   save(MemSim, file = "../results/MemSim.rda")
@@
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(MemSim, tau = tau)
-plotYearList(MemSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(MemSim, world, tau = tau)
+plotYearList(MemSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 if(eval){
-  FixedResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3)
+  FixedResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3, 0.99)
   save(FixedResourceSim, file = "../results/FixedResourceSim.rda")
@@
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(FixedResourceSim, tau = tau)
-plotYearList(FixedResourceSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(FixedResourceSim, world, tau = tau)
+plotYearList(FixedResourceSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 if(eval){
-  MemResourceSim <- runManyYears(World, Parameters = parameters, n.years = 3)
+  MemResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3, 0.99)
   save(MemResourceSim, file = "../results/MemResourceSim.rda")
@@
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(MemResourceSim, tau = tau)
-plotYearList(MemResourceSim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(MemResourceSim, world, tau = tau)
+plotYearList(MemResourceSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
-Memory_PulsedResource_Sim <- runManyYears(world.R3, Parameters = parameters, n.years = 6)
+Memory_PulsedResource_Sim <- runManyYears(world.R3, Parameters = parameters, n.years = 6, 0.99)
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Memory_PulsedResource_Sim, tau = tau)
-plotYearList(Memory_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Memory_PulsedResource_Sim, world.R3, tau = tau)
+plotYearList(Memory_PulsedResource_Sim, world.R3, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 parameters <- c(epsilon = 2, alpha = 4e2, beta0 = 0, beta1 = 0)
-Tactic_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+Tactic_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Tactic_PulsedResource_Sim, tau = tau)
-plotYearList(Tactic_PulsedResource_Sim, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Tactic_PulsedResource_Sim, world.R1, tau = tau)
+plotYearList(Tactic_PulsedResource_Sim, world.R1, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
-Hybrid_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+Hybrid_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim, tau = tau, 
+plotYearList(Hybrid_PulsedResource_Sim, world.R1, tau = tau)
+plotYearList(Hybrid_PulsedResource_Sim, world.R1, tau = tau, 
              persp = TRUE, scale = FALSE, expand = 100)
@@
 parameters <- c(epsilon = 2, alpha = 2e2, beta0 = 0, beta1 = 0)
-Tactic_PulsedResource_Sim2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+Tactic_PulsedResource_Sim2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim2, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim2, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Hybrid_PulsedResource_Sim2, world.R1, tau = tau)
+plotYearList(Hybrid_PulsedResource_Sim2, world.R1, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
-Hybrid_PulsedResource_Sim3 <- runManyYears(world.R4, Parameters = parameters, n.years = 6)
+Hybrid_PulsedResource_Sim3 <- runManyYears(world.R4, Parameters = parameters, n.years = 6, 0.99)
 
 par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim3, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim3, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
+plotYearList(Hybrid_PulsedResource_Sim3, world.R4, tau = tau)
+plotYearList(Hybrid_PulsedResource_Sim3, world.R4, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
 title(printParameters(parameters), outer = TRUE)
@@
 ```{r}
-source("functions_indices.R")
-  world$resource <- getPulsedResource(world$time, world$X, c(t.peak = 25, t.sd = 9, 
+dx <- 1
+
+parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
+world.R1 <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2, 
+                    peak.max = 80, peak.min = 20, sd = 10)
+world.R1$resource <- getPulsedResource(world.R1$time, world.R1$X, c(t.peak = 25, t.sd = 9, 
                                                            x.peak = 80, x.sd = 9))
 
-  parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
-  world.R1 <- world %>% 
-    list_modify(resource = getPulsedResource(1:world$tau, world$X, 
-                                             c(t.peak = 25, t.sd = 9, 
-                                               x.peak = 80, x.sd = 9)))
-  M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+#run simulation for 12 years 
+M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 1)
 
-  world.R1 <- world %>% list_modify(pop = M$Year6)
-  M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)
+world.R2 <- world.R1 %>% list_modify(pop = M$Year6)
+M2 <- runManyYears(world.R2, Parameters = parameters, n.years = 6, 1)
 
-  source("functions_indices.R")
-  indices1 <- sapply(M, computeIndices, resource = world$resource, world = world) %>% t
-  indices2 <- sapply(M2, computeIndices, resource = world$resource, world = world) %>% t
+  indices1 <- sapply(M, computeIndices, resource = world.R1$resource, world = world.R1) %>% t
+  indices2 <- sapply(M2, computeIndices, resource = world.R2$resource, world = world.R2) %>% t
   I <- rbind(indices1, indices2)
@@
   legend("topright", pch = 19, col = 1:3, legend = colnames(I))
+
+```
+Then we run the model to stop running the simulation at a certain threshold when the population reaches a "dynamic equilibrium state".
+In this example, the model only runs three years before hitting the threshold.
+```{r}
+#simulation runs for 3 years since that is when threshold is hit
+
+M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
+plotYearList(M, world.R1, tau = tau)
+
+computeCohesiveness(M[[3]], world.R1)["SC.mean"]
+computeMigratoriness(M[[3]], world.R1)$overlap
+computeEfficiency(M[[3]], world.R1$resource, world.R1)
+computeIndices(M[[3]], world.R1$resource, world.R1)
 

### 2021-02-07 memorymigration/R/PlottingFunctions.R

@@ -4,6 +4,10 @@
 #' 
 #' @param yearlist a list of T x X matrices describing the population 
 #' distribution for each year after initial population; produced by the runManyYear function
+#' @param World List of 5: a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
+#'  Set up by the getSinePop function. 
 #' @param X.max maximum value of X-axis
 #' @param tau maximum value of time T
 #' @param log logical;
@@ -11,12 +15,12 @@
 #' @seealso \link{runManyYears}; \link{printParameters}
 #' @example examples/plotting_examples.R
 
-plotYearList <- function(yearlist, X.max = 100, tau = 360, log = FALSE, 
+plotYearList <- function(yearlist, World, X.max = 100, tau = 360, log = FALSE, 
                          persp = FALSE, ...){
   
   n.years <- length(yearlist)
   allyears <- do.call(rbind, yearlist)
-  tau <- world$tau
+  tau <- World$tau
   
   if(log) allyears <- log(allyears)
   

### 2021-02-07 memorymigration/R/RunningModel.R

@@ -4,9 +4,10 @@
 #'\code{beta0} and \code{beta1}, this function determines the population 
 #'distribution for the following year. 
 #' 
-#' @param World List of 5 a population distribution across the time period in a T x X matrix,
+#' @param World List of 5: a population distribution across the time period in a T x X matrix,
 #'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
-#'  the dx value and the tau value. Set up by the getSinePop function. 
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
+#'  Set up by the getSinePop function. 
 #' @param Parameters named vector of parameters. \code{epsilon} - diffusion coefficient; 
 #' \code{alpha} - resource 
 #' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
@@ -42,16 +43,17 @@ runNextYear <- function(World, Parameters){
 #'\code{beta0} and \code{beta1}, this function determines the population 
 #'distribution after several years. 
 #'
-#' @param World List of 5 a population distribution across the time period in a T x X matrix,
+#' @param World World List of 5: a population distribution across the time period in a T x X matrix,
 #'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
-#'  the dx value and the tau value. Set up by the getSinePop function. 
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
+#'  Set up by the getSinePop function.  
 #' @param Parameters named vector of parameters. These are \code{epsilon} - diffusion coefficient; 
 #' \code{alpha} - resource 
 #' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
 #' memory coefficient
 #' @param n.years number of years the population migrates 
 #' @param threshold the threshold for the social cohesion between two population 
-#' distributions from two consecutive years 
+#' distributions from two consecutive years. This is a number between 0 and 1. 
 #' @return a list of n.years containing T x X matrices describing the population 
 #' distribution for each year after initial population
 #' @seealso \link{getSinePop}, \link{runManyYears}

### 2021-02-07 memorymigration/R/functions_indices.R

@@ -7,7 +7,10 @@
 #' the mean standard deviation of the population.
 #'  
 #' @param pop One cycle of population process
-#' @param world world object
+#' @param world world object; list of 5: a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
+#'  Set up by the getSinePop function. 
 #' @return a names vector of mean and standard deviation of SC
 #' @seealso \link{computeEfficiency}, \link{computeMigratoriness}, \link{computeIndices}
 #' @example examples/indices_examples.R
@@ -34,7 +37,10 @@ computeCohesiveness <- function(pop, world){
 #' 
 #' @param pop One cycle of population process
 #' @param resource The resource in an given cycle to compare to pop
-#' @param world world object
+#' @param world world object; list of 5: a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
+#'  Set up by the getSinePop function. 
 #' @return a single foraging efficiecy (FE) index
 #' @seealso \link{computeCohesiveness}, \link{computeMigratoriness}, \link{computeIndices}
 #' @example examples/indices_examples.R
@@ -53,7 +59,10 @@ computeEfficiency <- function(pop, resource, world){
 #' overall overlap.
 #' 
 #' @param pop One cycle of population process
-#' @param world world object
+#' @param world world object; list of 5: a population distribution across the time period in a T x X matrix,
+#'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
+#'  Set up by the getSinePop function. 
 #' @return a list with the time of the minimal overlap and the Migratoriness 
 #' (MI) index 
 #' @seealso \link{computeEfficiency}, \link{computeCohesiveness}, \link{computeIndices}
@@ -84,7 +93,10 @@ computeMigratoriness <- function(pop, world){
 #'
 #' @param pop One cycle of population process
 #' @param resource The resource in an given cycle to compare to pop
-#' @param world world object
+#' @param world world object; list of 5: a population distribution across the time period in a T x X matrix,
[…]

### 2021-02-07 memorymigration/examples/indices_examples.R

@@ -8,20 +8,19 @@ world.R1 <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2,
 world.R1$resource <- getPulsedResource(world.R1$time, world.R1$X, c(t.peak = 25, t.sd = 9, 
                                                            x.peak = 80, x.sd = 9))
 
+#run simulation for 12 years 
+M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 1)
 
-# run simulation 12 years
+world.R2 <- world.R1 %>% list_modify(pop = M$Year6)
+M2 <- runManyYears(world.R2, Parameters = parameters, n.years = 6, 1)
+plotYearList(M2, world.R2, tau = tau)
+
+#simulation runs for 3 years since that is when threshold is hit
 
 M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
-plotYearList(M, tau = tau)
+plotYearList(M, world.R1, tau = tau)
 
-pop <- world.R1$pop
-resource <- world.R1$resource
-
-world.R1 <- world %>% list_modify(pop = M$Year6)
-M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.999)
-plotYearList(M2, tau = tau)
-
-computeCohesiveness(world$pop, world)["SC.mean"]
-computeMigratoriness(world$pop, world)$overlap
-computeEfficiency(world$pop, world$resource, world)
-computeIndices(world$pop, world$resource, world)
+computeCohesiveness(M[[6]], world.R1)["SC.mean"]
+computeMigratoriness(M[[6]], world.R1)$overlap
+computeEfficiency(M[[6]], world.R1$resource, world.R1)
+computeIndices(M[[6]], world.R1$resource, world.R1)

### 2021-03-15 presentations/deer_wolf_CWD.Rmd
Presentation and Cohesivenss fix

---
title: "Can Selective Predation Slow the Spread of CWD?"
subtitle: "Modeling disease, predation, dispersal, and population dynamics in Wisconsin"
author: "Elie Gurarie"
date: "March 30, 2020"
header-includes:	
   - \input{preamble.tex}
fontsize: 11pt
classoption: "aspectratio=43"
output: 
  beamer_presentation:
    theme: "Madrid"
    keep_tex: true
---

```{r setup, include=FALSE}

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, encoding = "utf-8", message = FALSE, warning = FALSE)
```
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
```

# Acknowledgement

\bcol

\col[.3]
\col[.3]
\PlotW{../images/VanDeelen}
\PlotW{../images/FWE}
\col[.3]
\ecol

\bc Prof. Tim Van Deelen \ec

# The setting: **Wisconsin**

\PlotW{../images/Wisconsin_in_United_States.png}

# The setting: **Wisconsin**

  \bc
  \PlotW[.45]{../images/WisconsinVegetation}
  \PlotW[.45]{../images/TensionZone}
  \ec

* **Southwest:** Mainly agricultural / ex-prairie and oak savanna.
* **Northeast:** Conifer-hardwood forest / bogs / also agriculture

# The setting: **Wisconsin**

  \bc  \PlotW[.7]{../images/Treaty1837} \ec
  Wildlife management and fishing rights largely retained by Ojibwe tribes in the North.

# Three characters

\Large

1. White-tailed deer
2. Chronic wasting disease
3. Wolves

# White-tailed deer in Wisconsin

\bcol

\col[.6]

\bbb{\emph{Odoiceulus virginiacus}}

\small
- (over)-abundant: 1.9-2.1 million ind.
- major ecological impacts
- $\approx$ 350,000 hunted annually (and falling)
- \$1.4 billion dollars / year to economy
- a big chunk of which funds research / mitigation of those ecological impacts

\ebb

\col[.4]
\PlotW{../images/deer}
\bc\scriptsize \bl{Todd Hubler - \emph{The Isthmus}}\ec
\ecol

# Chronic Wasting Disease

\bcol

\col[.6]

\small

\bbb{Transmissible spongiform encephalopathy}

- Turns brains into sponges
- Invariably fatal 
- Caused by **prion** 
   - misfolded protein found in nervous system
- Only affects cervids 
   - only TSE in *wildlife* 
- Major focus of concern / research among agencies

\ebb

\col[.4]

\PlotW{../images/misfold}
\PlotW{../images/bse_sponge}

\ecol

# Chronic Wasting Disease

\bcol

\col[.5]

\small

\bbb{Clinical Signs}

"Zombie Disease"

- Emaciation
- Lack of coordination
- Drooping head/ears
- Excessive drooling
- Excessive drinking
- Excessive urination

\ebb

\col[.5]

\PlotW{../images/sickdeer}
\ecol

**Incubation:** (asymptomatic) period lasts on average 18 months

**Transmission:** urine, feces, blood.  Direct contact. Long-term environmental persistence (even uptake by plants).

# Chronic Wasting Disease

\bcol

\col[.4]

\footnotesize

\bbb{Global Expansion}

- 1967 - First detected (mule deer) in research facility Colorado. 
- 1981 - First wild animal (elk) detected in Colorado
- 2002 - Found in wild WTD in Wisconsin
- 2011 - Found in wild WTD in Maryland
- 2017 - Appeared in 3 reindeer in Norway (!) - entire 2000 animal herd summarily executed

\ebb

\col[.6]

\PlotW{../images/CWD_NorthAmerica}

\ecol

# Chronic Wasting Disease: In Wisconsin

Concentrated in southern counties, up to 25\% prevalance.

\bc \PlotW[.6]{../images/CWDprevalence} \ec

***Good data:***  In affected counties, all hunted carcasses need to be tested.  In non-affected counties, a sample of carcasses is tested.

# Wolves in Wisconsin

\small
Extirpated early 1900's.  Re-colonized from Minnesota post-ESA.

\bcol

\col

\PlotW[.9]{../images/WolfMap2019}

\col

```{r WolfNums, fig.width = 4, fig.height = 4}
ww <- read.csv("../data/wolfcounts.csv")
require(ggplot2); require(ggthemes); require(gridExtra)
p1 <- ggplot(ww, aes(Year, Wisconsin)) + geom_point() + ylab("N. wolves in WI")
p2 <- ggplot(ww, aes(Year, Wisconsin)) + geom_point() + geom_smooth() + scale_y_log10()  + ylab("N. wolves in WI")
grid.arrange(p1, p2, ncol = 1)
```

\ecol

Currently, approx. 1000 ind. mainly in North.  Expansion slowing. ***Good Data.***

# The question

\small

Wolves selectively predate on old, young, weak or **infirm(?)** individuals ... though there is no direct evidence w.r.t. CWD (or - actually - other diseases).

Given that CWD is concentrated in the SW - and expanding - and wolves are concentrated in NE - and maybe still expanding? - What happens when they meet?

Specifically, how do wolf presence and selective predation influence:

- **CWD prevalence**
- **CWD spread**
- **Deer abundance**

# Approaches to look at this question

Lots of ***Mathematical Modeling!!***  Mainly, continuous-time, non-spatial SEIR-type ODE's.

\bcol
\col
\PlotW{../images/WildTitle}
\PlotW{../images/WildEquations}

Very influential, but no data (and no spatial structure)

\col
\bc
\PlotW{../images/TannerTitle}
\PlotW[0.7]{../images/TannerEquations}
\ec

Lots of compartments - and some data (but no spatial structure)

\ecol

# Modeling goals

* Capturing dynamics of:
  - **disease**, 
  - **predation**, 
  - **population** 
  - **dispersal**

* Biologically meaningful parameters
  - independently estimated / estimable?

* Provide spatially and temporally explicit predictions

* Balances realism with tractability

# Basic model structure:

## Discrete time / discrete space

- **Annual** - matches data collection and deer biology (birth / seasonal mortality / dispersal?)

- **County-level metapopulation** - matches data reporting and collection

## Two classes: Susceptible and Infected
\small
\begin{align}
S_{i,t+1} &= S_{i,t} - \rd{infected} & + \bl{recruited}& - \rd{died} + \bl{immigrated} - \rd{emigrated} \nonumber\\
I_{i, t+1} &= I_{i,t} + \bl{infected} & &  -\rd{died} + \bl{immigrated} - \rd{emigrated} \nonumber
\end{align}

# Complete model

\bc

\scriptsize
|   |  Susceptible $(S_{i,t+1})$|  Infected $(I_{i,t+1})$|  
|---|:---:|:---:|
| disease          | $- \gamma {S_{i,t} I_{i,t} \over area}$ | $\gamma{ S_{i,t} I_{i,t} \over area}$ |
| predation        | $-\left({S_{i,t} \over S_{i,t}+I_{i,t}}\right) \left({1 \over 1 + \alpha}\right) W_{max}$ |  $-\left({I_{i,t} \over S_{i,t}+I_{i,t}}\right) \left({\alpha \over 1 + \alpha}\right) W_{max}$      |
| other mortality  | $- \mu_s S_{i,t}$ |  $- \mu_I I_{i,t}$ |
| recruitment      |  $\rho S_{i,t}(1 - S_{i,t}/K_i)$  |  |
| immigration      | $\sum_{j} M_{S,ij}$  | $\sum_{j} M_{I,ij}$ |
| emigration       | $-\sum_{j} E_{s,ji}$ | $-\sum_{j} E_{i,ji}$ |

\PlotW[.7]{../images/gif/crudity_1}

\ec

# Recruitment

##
Approximate number of new (non-infected) individuals entering population annually:
\Large
$$\rho \, S_{i,t}(1 - S_{i,t}/K_i)$$
\normalsize

Assumptions:

1. Logistic, density dependent growth with carrying capacity: $K$
2. Intrinsic adult recruitment rate, $\rho$, uniform across range

# Mortality

## Non-Wolf:

Annual mortality rate: \rd{$\mu_S$} and \rd{$\mu_I$ }:

- Equal (!?)
- Constant across range (?))

# Mortality

## Wolf:

Assume: **Wolves gonna kill what wolves gonna kill** - (almost entirely deer).

- \small e.g. 10 wolves, each kills 20 deer / year = 200 deer/year/county killed

\normalsize Apportioning of S and I killed is proportional to the *availability* ($I \over S+I$) and the *preference* ($\alpha \over 1+\alpha$).

- \small $P(I \, killed) \propto P(I) \times P(killed \,|\, I) = \left({I \over S + I}\right)  \left(\alpha  \over 1 + \alpha \right) N_w K_{rate}$
- \small  $P(S \, killed) \propto P(S) \times P(killed \,|\, S) = \left({S \over S + I}\right)  \left(1  \over 1 + \alpha \right) N_w K_{rate}$

where \gr{$\alpha$} is the odds of selecting infected over non-infected deer, i.e. the \gr{Predation Selection Coefficient} (presumably $>$ 1). 
\nd

__This is the key, unobservable parameter of interest!__

# Disease dynamics

## Number of infected ...
(by county by year)
\Large
$$- \gamma {S \times I \over area}$$
\large

- $\gamma$ is infection rate per encounter  
- Total number of infections is higher if areas are smaller (densities are higher). 
- Lots of mechanism and structure swept under rug here!

# Dispersal

**Necessary for disease to spread! (or not spread)**

## Immigration
$$M_i  = {\kappa_m} \sum_{j} S_j \exp {\left(- {d_{ij} \over \lambda} \right)^\beta}$$

## Emigration
$$E_j  = {\kappa_e} S_i \sum_{j} \exp {\left(- {d_{ij} \over \lambda} \right)^\beta}$$

\footnotesize
- **Exponential** ($\beta = 1$) or **Gaussian** ($\beta = 2$) dispersal kernel with mean $\lambda$
- $\kappa$ - normalization, so that probability of emigrating from site $j$ sums to 1
- Assumes homogenous dispersal in all directions, dependent only on distance. 
- *Estimable (and refinable) from movement data!?*

# Data

## **Deer Abundance**
Wisconsin DNR winter population survey: [https://dnr.wi.gov/topic/hunt/maps.html](https://dnr.wi.gov/topic/hunt/maps.html)
\bc
\PlotW[.45]{../images/overwinter.pdf}
\PlotW[.45]{../images/winterpoppertotal.pdf}
\ec
Fall population estimates - total harvest, by county.

\nd Working assumption: Carrying Capacity $K_i = 2 N_i$.

# Data

## **CWD prevalence**
Wisconsin DNR CWD monitoring efforts (by county) 
\nd

\bc
\PlotW[.7]{../images/CWDprevalence}
\ec

\small https://dnr.wi.gov/wmcwd/Summary/YearCounty/2019

# Data

## **Wolves**

Latest estimate from DNR: 950 ind.

\bc \PlotH[.5]{../images/WolfMap2019.png} \,\,\, \PlotH[.5]{../images/WolfMapElie.png} \ec

County data not readily available ... so I allocated 1000 wolves across the counties north of this line.

# Model

## Shiny App.

A brand new skill!
Facilitates exploring parameters and visualizing results.

\bc
(enjoy demo)
\ec

# A Result:  Selective Predation Decreases CWD Prevalence!
\bc
\PlotW[.9]{../images/SimResults1.png}
\ec

\small
In ALL parameterizations, wolves depress CWD.  Note - dispersal scale (10 and 80 km) AND shape both important.

\footnotesize `rho = 0.5, gamma = .02, mu_S = 0.06, mu_I = 0.06, W_max = 60, lambda = 10 or 80`

# Next steps

\footnotesize

## Model structure
* Add **Male / Female** sex classes!
* Separate **Infected / Asymptomatic** from **Infected / Symptomatic**
* Assess assumptions:  Density dependence? | Disease transmission? 
\pause

## Data
- Obtain better **wolf distribution** and **predation** data
- Use **Harvest** for mortality!
- Use **GPS data** for dispersal portion
- Fit to historical data!?
  - Infer $\gamma$ by matching to observed CWD spread? 
\pause

## Larger strategy
- Thoroughly analyze / explore parameter space
- Find PhD student to do the work!?
- Get funding!

#

\bc
\Huge
Thanks!
\ec

### 2021-03-15 memorymigration/R/functions_indices.R

@@ -21,7 +21,7 @@ computeCohesiveness <- function(pop, world){
   getSD <- function(p, world){
     EX <- sum( world$X * p) * world$dx
     EX2 <- sum( world$X^2 * p) * world$dx
-    sd  = sqrt(EX2 - EX^2)
+    sqrt(EX2 - EX^2)
   }
   
   SDs <- apply(pop, 1, getSD, world = world)

### 2021-03-25 memorymigration/R/RunningModel.R
changed threshold to 60

@@ -112,7 +112,7 @@ runManyRuns <- function (parameters.df, world, filename = NULL, results.dir = NU
 {
   newresults <- data.frame()
   for (i in 1:nrow(parameters.df)) {
-    M <- try(runManyYears(world, parameters = parameters.df[i,], n.years = 30, threshold = 0.99))
+    M <- try(runManyYears(world, parameters = parameters.df[i,], n.years = 60, threshold = 0.995))
     if(!inherits(M, "try-error")){
       myR <- data.frame(parameters.df[i, ], computeIndices(M[[length(M)]], 
                                                            world$resource, world), n.runs = length(M) - 1)

### 2021-03-25 presentations/memorymigration_v2.Rmd

---
title: "Modeling tactical and memory-driven migration in dynamic environments"
author: "Sriya Potluri and Elie Gurarie"
date: "3/24/2021"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE)
```

# Premise
<div class='left' style='float:left;width:48%'>
To optimally forage in a dynamic environment, you need both **tactical** responses (adapting to local changes in resource gradients), and **strategic** responses (using memory to guide long-term decisions).

</div>
<div class='right' style='float:right;width:48%'>
**Tactics**

![](images/tacticsim.png){width=100%}

**Memory**

![](images/memorysim.png){width=100%}
</div>

# Model: Diffusion advection

$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial}{\partial x}\left(u \frac{\partial h(x,t)}{\partial x}\right) + \sum_{i = 0}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$

- $\epsilon$: diffusion, exploration / search
- $\alpha$: taxis, resource following
- $\beta_0$: cohesion, tendency to stick to group
- $\beta_1$: memory, tendency to do what was done previous year

Builds on series of projects exploring "cognitive" variations chemotaxis (e.g. Fagan et al. 2017, 2019).

# Seasonal resource

$$h(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$

<div class='left' style='float:left;width:48%'>
1. The total amount constant.
2. Between seasons, uniformly distributed. 
3. Resource concentrates symmetrically in time and space

Importantly: we explore a range of **spatial concentrations** and **temporal durations**.

</div>

<div class='right' style='float:right;width:48%'>
![](images/exampleResourcePlots-1.png){width=100%}
</div>

---

**Running the model**

<div class='left' style='float:left;width:48%'>
- R package `deSolve` and `ReacTran` - fluid transport pde solvers.
- Functions all in `memorymigration` GitHub project. 
- SESYNC cluster 
- server-based Rstudio
</div>

<div class='right' style='float:right;width:48%'>

**Assessing the model**

Three metrics:

1. Social cohesiveness
2. Foraging Efficiency
3. Migratoriness
</div>

# Metrics: Cohesiveness

<div class='left' style='float:left;width:50%'>

$$SC = 1-\frac{\sqrt{12} \, }{X_{max}}\overline{\sigma}$$

$$\overline{\sigma} = {1\over \tau} \int \sigma_u(x,t) \,dt$$
where $$\sigma_u^2(t) = \int_D \left(x - \mu_u(t)\right)^2 \, u(x,t)\, dx$$

and $\mu_u$ is the mean of $u$ at time $t$.

Definition: The mean standard deviation of the population across the equilibrium year

</div>

<div class='right' style='float:right;width:50%'>

```{r SC_example, fig.width = 4}
require(memorymigration)
data(resources)
par(mfrow = c(2,1), mar = c(3,3,3,1), mgp = c(2,.25,0), las = 1, tck = 0.01)

image(1:100,1:100,resource_R2, main = "SC = 0.36")
contour(1:100,1:100,resource_R2, add = TRUE)
#computeCohesiveness(resource_R2, world)

image(1:100,1:100,resource_R3, main = "SC = 0.11")
contour(1:100,1:100,resource_R3, add = TRUE)

#computeCohesiveness(resource_R3, world)

```

</div>

# Social Cohesiveness: Problem

For skewed of bimodal population distribution, standard deviation are a poor measure ...
$\epsilon = 8; \alpha = 200; \beta_0 = 500; \beta_1 = 0$

```{r badSCexample, cache = TRUE, fig.width = 10, eval = FALSE}
load("images/Sim1.rda")
png("images/BadSC.png", width = 2400, height = 1200, res = 300)
par(mfrow = c(1,2), bty  = "l", mgp = c(1.5,.5,0), tck = 0.01)
require(fields)
computeCohesiveness(sim[[length(sim)]], world)
image.plot(sim[[length(sim)]], main = "looks cohesive, but SC = 0.01")
snapshot <- sim[[length(sim)]][75,]
plot(snapshot, type = "l", main = "because bimodal (t = 75).", ylab = "")
dev.off()
```

![](images/BadSC.png){width=100%}

**Not sure how to deal with this ...**

# Metrics: Foraging Efficiency

$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{X_{max}} \sqrt{u(x,t) \, h(x,t)} \, dx\,dt$$

**Bhattacharyya Coefficient** quantifies similarity between two distributions.

```{r, eval = FALSE}
load("images/Sim1.rda")
png("images/FEexample.png", width = 2400, height = 1200, res = 300)
par(mfrow = c(1,2), bty  = "l", mgp = c(1.5,.5,0), tck = 0.01)
require(fields)
image.plot(1:100, 1:100, resource_R2, main = "Resource", xlab = "time", ylab = "space")
image.plot(1:100, 1:100, sim[[length(sim)]], main = "Population: FE = 0.82", xlab = "time")
dev.off()

computeEfficiency(resource_R2, sim[[length(sim)]], world)
```

<center>
$\epsilon = 8; \alpha = 200; \beta_0 = 500; \beta_1 = 0$

![](images/FEexample.png){width=70%}
</center>

# Metrics: Foraging Efficiency - Issues

Seems to always be very high!

Also - assumes that there is some sort of space inhibition ... the actually highest efficiency is when animals pile in at indefinitely high density at the location of highest resource availability.

<center>

![](images/FEexample.png){width=70%}

</center>

# Metrics: Migratoriness

$$MI = 1-min(O(t_1, t_2))$$
$$O(t_1, t_2) = \int_D \sqrt{u(x, t_1) \, u(x, t_2)} \, dx $$

Definition: MINIMUM OVERLAP of population at all possible pairs of times within year.

```{r, eval = FALSE}
load("images/Sim1.rda")
png("images/MIexample.png", width = 2400, height = 1200, res = 300)
par(mfrow = c(1,2), bty  = "l", mgp = c(1.5,.5,0), tck = 0.01)
require(fields)
image.plot(1:100, 1:100, resource_R2, main = "Resource", xlab = "time", ylab = "space")
image.plot(1:100, 1:100, sim[[length(sim)]], main = "Population: MI = 0.07", xlab = "time")
dev.off()

computeEfficiency(resource_R2, sim[[length(sim)]], world)
```

<center>

$\epsilon = 8; \alpha = 200; \beta_0 = 500; \beta_1 = 0$
![](images/MIexample.png){width=70%}

</center>

# Questions that this model can answer:

<div class='left' style='float:left;width:60%'>

Q1. How does **memory** and **sociality** influence **migratoriness** and **foraging efficiency** under different **resource dynamics**?

Q2. How long does it take for migration to emerge from a non-migratory initial condition?

Q3. As conditions CHANGE, how resilient is the population at various Diffusion, Advection, Cohesion, Memory values?  
- spatial drift
- temporal drift
- increased variability

</div>

<div class='right' style='float:right;width:40%'>

![](images/brainstorm.png){width=100%}

</div>

# Example simulations:

```{r, out.width = "33%", eval = FALSE}
include_graphics(c("images/Sim_withMemory.png","images/Sim_onlyMemory.png", "images/Sim_noMemory.png"))

```

1. Only Tactics
1. Memory and Sociality
2. Only Memory
3. Only Sociality

# Some results

<div class='left' style='float:left;width:55%'>

1. Migratoriness results from **concentrated** and **long** resource pulses. 
2. Migratoriness is **increased** by **memory**, but **attenuated** by **sociality**. 
3.  Foraging is easiest (obviously) for **diluted** and **short** resource pulses, as the resource is distributed mainly uniformly.

</div>

<div class='right' style='float:right;width:45%'>

![](images/ComparingResourceDynamics_v1.png){width=100%}

</div>

### 2021-03-25 memorymigration/R/RunningModel.R
added final_similarity to runManyRuns

@@ -115,7 +115,10 @@ runManyRuns <- function (parameters.df, world, filename = NULL, results.dir = NU
     M <- try(runManyYears(world, parameters = parameters.df[i,], n.years = 30, threshold = 0.99))
     if(!inherits(M, "try-error")){
       myR <- data.frame(parameters.df[i, ], computeIndices(M[[length(M)]], 
-                                                           world$resource, world), n.runs = length(M) - 1)
+                                                           world$resource, world), 
+                        n.runs = length(M) - 1,
+                        final_similarity = computeEfficiency(M[[length(M)-1]], 
+                                                             M[[length(M)]], world))
       newresults <- rbind(newresults, c(myR))
     }
     if(!is.null(results.dir) & (i %% 10 == 0 | i == max(i)))  

### 2021-03-25 presentations/memorymigration_v2.Rmd
added final_similarity to runManyRuns
Presentation more or less good to go

@@
 
-**Running the model**
+<div class='left' style='float:left;width:48%'>
+## Running the model
 
-<div class='left' style='float:left;width:48%'>
 - R package `deSolve` and `ReacTran` - fluid transport pde solvers.
@@
 <div class='right' style='float:right;width:48%'>
+## Assessing the model
 
-**Assessing the model**
-
-Three metrics:
+Four metrics:
 
@@
 3. Migratoriness
+4. Time to equilibrium
+
 </div>
@@
 
-```{r SC_example, fig.width = 4}
+```{r SC_example, fig.width = 6}
 require(memorymigration)
@@
 $\epsilon = 8; \alpha = 200; \beta_0 = 500; \beta_1 = 0$
+
 ![](images/MIexample.png){width=70%}
@@
 
-# Example simulations:
+# Runs
 
-```{r, out.width = "33%", eval = FALSE}
-include_graphics(c("images/Sim_withMemory.png","images/Sim_onlyMemory.png", "images/Sim_noMemory.png"))
+Lots of them! (8100)
 
-```
+## Model parameters
 
-1. Only Tactics
-1. Memory and Sociality
-2. Only Memory
-3. Only Sociality
+- $\alpha$ - tactic response: 0, 100, 1000
+- $\epsilon$ - diffusion: 10, 30, 100
+- $\beta_0$ - sociality: 0, 100, 200, 300, 400, 500
+- $\beta_1$ - memory: 0, 100, 200, 300, 400, 500
+
+## Resource pulse parameters
+
+- Duration (2 x sd): 6,12,18,24,30
+- Spatial extent (2 x sd): 6,12,18,24,30
+
+# Basic Metrics
+
+<div class='left' style='float:left;width:60%'>
+actic response**
+2. High overall mobility (high tactic + high diffusion) =  high foraging efficiency
+3. Slow diffusion leads to longer time to attain equilibrium
+
+We mainly focus on **low diffusion** and **high tactic**, where most of the (migratory) "action" is.
+
+</div>
+
+<div class='right' style='float:right;width:40%'>
+![](images/MetricComparisons.png){width=100%}
+</div>
+
+# Example simulations
+
+![](images/resource.png){width=20%}![](images/sim0.png){width=70%}
+
+# Example: Memory 0 / Sociality 500
+
+![](images/resource.png){width=20%}![](images/sim2.png){width=70%}
+
+# Example: Memory 500 / Sociality 0
+
+![](images/resource.png){width=20%}![](images/sim1.png){width=70%}
+
+# Example: Memory 200 / Sociality 200
+
+![](images/resource.png){width=20%}![](images/sim3.png){width=70%}
+
+# Example: Memory 500 / Sociality 200 (sometimes things fall apart)
+
+![](images/resource.png){width=20%}![](images/sim4.png){width=70%}
 
@@
 
-<div class='left' style='float:left;width:55%'>
+<div class='left' style='float:left;width:50%'>
 
 1. Migratoriness results from **concentrated** and **long** resource pulses. 
-2. Migratoriness is **increased** by **memory**, but **attenuated** by **sociality**. 
-3.  Foraging is easiest (obviously) for **diluted** and **short** resource pulses, as the resource is distributed mainly uniformly.
+2. Both migratoriness and sociality **increase migratoriness**. 
+3. Foraging efficiency is highest for **diluted** and **short** resource pulses (resource is distributed mainly uniformly).
+4. In a "stationary" (equilibrium) scenario, memory and sociality don't contribute much to foraging efficiency (**caveated!**).
 
@@
 
-<div class='right' style='float:right;width:45%'>
+<div class='right' style='float:right;width:50%'>
 
-![](images/ComparingResourceDynamics_v1.png){width=100%}
+![](images/ComparingResourceDynamics.png){width=100%}
 
 </div>
+
+---
+
+## Main take-away
+
+1. This is a convenient, flexible and fast tool for exploring a wide range of memory / tactic / sociality dynamics
+
+## Going forward
+
+"Stress" the system to assess plasticity of a population:
+
+- Make the pulses more isolated (weaken usefulness of tactical strategy).
+- Drift the **location** and **timing** of pulse
+- Make the pulse more **patchy** and **stochastic**

### 2021-03-25 presentations/plottingResults.R

@@ -1,0 +1,212 @@
+
+compareResourceTypes <- function(df, beta0.high = 200, beta1.high = 300){
+  p <- ggplot(df %>% 
+                mutate(extent = (space.sd*2) %>% factor,
+                       duration = (time.sd*2) %>% factor,
+                       model = ifelse(beta0 == 0 & beta1 == 500, "c) Memory: 500",
+                                      ifelse(beta0 == 500 & beta1 == 0, "b) Sociality: 500",
+                                             ifelse(beta0 == beta0.high & beta1 == beta1.high, 
+                                                    paste("d) Memory:", beta1.high,  "Sociality: ", beta0.high),
+                                                    ifelse(beta0 == 0 & beta1 == 0, "a) Tactics only", NA))))) %>% 
+                subset(!is.na(model)), 
+              aes(FE, MI, col = extent, pch = duration)) + 
+    geom_line(aes(FE, MI, group = duration), col = "grey", lty = 3) + 
+    geom_line(aes(FE, MI, group = extent), col = "grey", lty = 3) + 
+    geom_point(size = 3) + facet_wrap(.~model, ncol = 2) + 
+    theme_few() + ylab("Migratoriness") + xlab("Foraging efficiency") 
+  return(p)
+}
+
+
+
+compareResourceTypes_v2 <- function(df, beta0.high = 200, beta1.high = 300){
+  p <- ggplot(df %>% 
+                mutate(extent = (space.sd*2) %>% factor,
+                       duration = (time.sd*2) %>% factor,
+                       model = ifelse(beta0 == 0 & beta1 == 500, "c) Memory: 500",
+                                      ifelse(beta0 == 500 & beta1 == 0, "b) Sociality: 500",
+                                             ifelse(beta0 == beta0.high & beta1 == beta1.high, 
+                                                    paste("d) Memory:", beta1.high,  "Sociality: ", beta0.high),
+                                                    ifelse(beta0 == 0 & beta1 == 0, "a) Tactics only", NA))))) %>% 
+                subset(!is.na(model)), 
[…]

### 2021-03-25 presentations/memorymigration_v2.Rmd
updates

@@
 title: "Modeling tactical and memory-driven migration in dynamic environments"
-author: "Sriya Potluri and Elie Gurarie"
+author: "Elie Gurarie, Sriya Potluri, Bill Fagan"
 date: "3/24/2021"
@@
 <div class='left' style='float:left;width:60%'>
-actic response**
+1. Highest mirgation with **low diffusion** but **high tactic response**
 2. High overall mobility (high tactic + high diffusion) =  high foraging efficiency
@@
 1. This is a convenient, flexible and fast tool for exploring a wide range of memory / tactic / sociality dynamics
+2. Plusses / Minusses versus individual-based approach?
 

### 2021-04-07 text/ms.Rmd
Early text draft

--- 2021-02-07 code/StrategicVsTactical_PartI_v2.Rmd
@@
 output:
+  word_document:
+    toc: yes
   html_document:
@@
   chunk_output_type: console
+bibliography: ../library/memorymigration.bib
+csl: ecology.csl
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
 ```
 
-# Memory driven foraging model
+# Introduction
+
+Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;Okubo2001].
+Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled.
+While the links between diffusion models and movement data are somewhat tenuous (see @Gurarie2011a for quantitative links between obervations and diffusion parameters and @Ovaskainen2004 for applications to mark-recapture data), as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
+Even as equations become analytically intractable,
+
+Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.
+by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).
+
+But perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory.
+Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy.
+The interacting role of memory and sociality have been comparatively little studied.
+
+Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized.
+To perform a migration, a direct resource-following tactic (taxis) is likely to fail.
+
+# Model
 
@@
 memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior**.
-
 Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
@@
 
-$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial h(x,t)}{\partial x} + \sum_{i = 0}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$
-
+$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial}{\partial x}\left(u \frac{\partial h(x,t)}{\partial x}\right) + \sum_{i = 0}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$
 where $u$ represents a population distributed in time and space, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta_i$ is the strength of advection up the gradient of the *population* distribution in an $i$'th earlier seasonal cycle (year), up to $n \in N$.
@@
 
-For reference and replicability, I include the incrementally constructed R code, which relies heavily on the `deSolve` (Soetaert et al.
-2010) and `ReacTran` (Soetaert and Filip, 2010) packages, which efficiently implement transport ODE and PDE systems in R.
+For reference and replicability, I include the incrementally constructed R code, which relies heavily on the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages, which efficiently implement transport ODE and PDE systems in R.
 
@@
 
-# Static run
+# Results
 
@@
 
-# Some summaries
+# Discussion
 
@@
 # References
-
-Karline Soetaert, Thomas Petzoldt, R.
-Woodrow Setzer (2010).
-Solving Differential Equations in R: Package deSolve.
-Journal of Statistical Software, 33(9), 1--25.
-URL http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09
-
-Soetaert, Karline and Meysman, Filip, 2012. Reactive transport in aquatic ecosystems: Rapid model   prototyping in the open source software R Environmental Modelling & Software, 32,

### 2021-04-07 code/IslandResourcesExploration.Rmd

---
title: "Island Resources Exploration"
author: "E. Gurarie, S. Potluri, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE)
```

```{r, message=FALSE, warning=FALSE}
require("memorymigration")
```

## Gaussian Initial Conditions; Resource t3 x3

```{r, message=FALSE, warning=FALSE, eval=FALSE}

  data(world)
  data(resources_island)
  world_gaussian$resource <- resources_island[["R_t3_x3"]]
  image(world_gaussian$resource)

  sim0 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 0), verbose = TRUE)
  sim1 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 500), verbose = TRUE)
  sim2 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 0), verbose = TRUE)
  sim3 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 250, beta1 = 250), verbose = TRUE)
  sim4 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 200), verbose = TRUE)

  save(sim0, sim1, sim2, sim3, sim4, file = "results/islandresources/Gaussianalpha1000epsilon10_R_t3_x3.rda")
} else load("results/islandresources/Gaussianalpha1000epsilon10_R_t3_x3.rda")

png("presentations/images/gaussiantactict3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim0)) image(1:100, 1:100, sim0[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only tactic response", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussianmemoryt3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim1)) image(1:100, 1:100, sim1[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only memory", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussiansocialityt3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim2)) image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only sociality", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussianmemorysociality1t3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim3)) image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=beta0=250", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussianmemorysociality2t3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(2,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:14) 
  image(1:100, 1:100, sim4[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=200, beta0=500", outer = TRUE, cex.main = 2)
dev.off()

```

![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussiantactict3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussianmemoryt3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussiansocialityt3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussianmemorysociality1t3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussianmemorysociality2t3x3.png)

## Gaussian Initial Conditions; Resource t15 x15

```{r, message=FALSE, warning=FALSE, eval=FALSE}

  data(world)
  data(resources_island)
  world_gaussian$resource <- resources_island[["R_t15_x15"]]
  image(world_gaussian$resource)

  sim0 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 0), verbose = TRUE)
  sim1 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 500), verbose = TRUE)
  sim2 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 0), verbose = TRUE)
  sim3 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 250, beta1 = 250), verbose = TRUE)
  sim4 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 200), verbose = TRUE)

  save(sim0, sim1, sim2, sim3, sim4, file = "results/islandresources/Gaussianalpha1000epsilon10_R_t15_x15.rda")
} else load("results/islandresources/Gaussianalpha1000epsilon10_R_t15_x15.rda")

png("presentations/images/gaussiantactict15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim0)) image(1:100, 1:100, sim0[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only tactic response", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussianmemoryt15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim1)) image(1:100, 1:100, sim1[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only memory", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussiansocialityt15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim2)) image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only sociality", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussianmemorysociality1t15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim3)) image(1:100, 1:100, sim3[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=beta0=250", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/gaussianmemorysociality2t15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(4,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim4)) image(1:100, 1:100, sim4[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=200, beta0=500", outer = TRUE, cex.main = 2)
dev.off()
```

![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussiantactict15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussianmemoryt15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussiansocialityt15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussianmemorysociality1t15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/gaussianmemorysociality2t15x15.png)

## Sinusoidal Initial Conditions; Resource t3 x3

```{r, message=FALSE, warning=FALSE, eval=FALSE}

  data(world)
  data(resources_island)
  world_sinusoidal$resource <- resources_island[["R_t3_x3"]]
  image(world_sinusoidal$resource)

  sim0 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 0), verbose = TRUE)
  sim1 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 500), verbose = TRUE)
  sim2 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 0), verbose = TRUE)
  sim3 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 250, beta1 = 250), verbose = TRUE)
  sim4 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 200), verbose = TRUE)

  save(sim0, sim1, sim2, sim3, sim4, file = "results/islandresources/Sinusoidalalpha1000epsilon10_R_t3_x3.rda")
} else load("results/islandresources/Sinusoidalalpha1000epsilon10_R_t3_x3.rda")

png("presentations/images/sinusoidaltactict3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim0)) image(1:100, 1:100, sim0[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only tactic response", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalmemoryt3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim1)) image(1:100, 1:100, sim1[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only memory", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalsocialityt3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim2)) image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only sociality", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalmemorysociality1t3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim3)) image(1:100, 1:100, sim3[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=beta0=250", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalmemorysociality2t3x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(2,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim4)) image(1:100, 1:100, sim4[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=200, beta0=500", outer = TRUE, cex.main = 2)
dev.off()

```
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidaltactict3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalmemoryt3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalsocialityt3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalmemorysociality1t3x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalmemorysociality2t3x3.png)

## Sinusoidal Initial Conditions; Resource t15 x15

```{r, message=FALSE, warning=FALSE, eval=FALSE}

  data(world)
  data(resources_island)
  world_sinusoidal$resource <- resources_island[["R_t15_x15"]]
  image(world_sinusoidal$resource)

  sim0 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 0), verbose = TRUE)
  sim1 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 0, beta1 = 500), verbose = TRUE)
  sim2 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 0), verbose = TRUE)
  sim3 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 250, beta1 = 250), verbose = TRUE)
  sim4 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 500, beta1 = 200), verbose = TRUE)

  save(sim0, sim1, sim2, sim3, sim4, file = "results/islandresources/Sinusoidalalpha1000epsilon10_R_t15_x15.rda")
} else load("results/islandresources/Sinusoidalalpha1000epsilon10_R_t15_x15.rda")

png("presentations/images/sinusoidaltactict15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim0)) image(1:100, 1:100, sim0[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only tactic response", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalmemoryt15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim1)) image(1:100, 1:100, sim1[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only memory", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalsocialityt15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,4), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim2)) image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("Only sociality", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalmemorysociality1t15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(1,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim3)) image(1:100, 1:100, sim3[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=beta0=250", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/sinusoidalmemorysociality2t15x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(4,6), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:length(sim4)) image(1:100, 1:100, sim4[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("With memory and sociality beta1=200, beta0=500", outer = TRUE, cex.main = 2)
dev.off()
```
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidaltactict15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalmemoryt15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalsocialityt15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalmemorysociality1t15x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/sinusoidalmemorysociality2t15x15.png)

## 60 year runs Gaussian. 
```{r, message=FALSE, warning=FALSE, eval=FALSE}
  data(world)
  data(resources_island)

  world_gaussian$resource <- resources_island[["R_t12_x12"]]
  sim0 <- runManyYears(world_gaussian, parameters = c(epsilon = 5, alpha = 1000, beta0 = 100, beta1 = 100), verbose = TRUE)

  world_gaussian$resource <- resources_island[["R_t3_x15"]]
  sim1 <- runManyYears(world_gaussian, parameters = c(epsilon = 10, alpha = 1000, beta0 = 300, beta1 = 300), verbose = TRUE)

  world_gaussian$resource <- resources_island[["R_t15_x3"]]
  sim2 <- runManyYears(world_gaussian, parameters = c(epsilon = 5, alpha = 2000, beta0 = 300, beta1 = 200), verbose = TRUE)

  world_gaussian$resource <- resources_island[["R_t6_x3"]]
  sim3 <- runManyYears(world_gaussian, parameters = c(epsilon = 5, alpha = 500, beta0 = 200, beta1 = 200), verbose = TRUE)

  save(sim0, sim1, sim2, sim3, file = "results/islandresources/60yearruns_gaussianisland.rda")
} else load("results/islandresources/60yearruns_gaussianisland.rda")

png("presentations/images/60yearrungaussiant12x12.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim0[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T12X12, alpha=1000, epsilon=10, beta1=beta0=100", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/60yearrungaussiant3x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim1[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T3X15, alpha=1000, epsilon=10, beta1=beta0=300", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/60yearrungaussiant15x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T15X3, ealpha=2000, epsilon=10, beta0=300, beta1=200", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/60yearrungaussiant6x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim3[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T6X3, epsilon=500, alpha=5, beta1=beta0=200", outer = TRUE, cex.main = 2)
dev.off()

```
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrungaussiant12x12.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrungaussiant3x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrungaussiant15x3.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrungaussiant6x3.png)

## 60 year runs Sinusoidal. No matter the initial condition, all of the 60 years runs happen with the relatively the same parameter combinations. There are less 60 year runs with sinusoidal intial condition compared to gaussian initial condition. 
```{r, message=FALSE, warning=FALSE, eval=FALSE}

  data(world)
  data(resources_island)

  world_sinusoidal$resource <- resources_island[["R_t12_x12"]]
  sim0 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 5, alpha = 1000, beta0 = 100, beta1 = 100), verbose = TRUE)

  world_sinusoidal$resource <- resources_island[["R_t3_x15"]]
  sim1 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 10, alpha = 1000, beta0 = 300, beta1 = 300), verbose = TRUE)

  world_sinusoidal$resource <- resources_island[["R_t15_x3"]]
  sim2 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 5, alpha = 2000, beta0 = 300, beta1 = 200), verbose = TRUE)

  world_sinusoidal$resource <- resources_island[["R_t6_x3"]]
  sim3 <- runManyYears(world_sinusoidal, parameters = c(epsilon = 5, alpha = 500, beta0 = 200, beta1 = 200), verbose = TRUE)

  save(sim0, sim1, sim2, sim3, file = "results/islandresources/60yearruns_sinusoidalisland.rda")
} else load("results/islandresources/60yearruns_sinusoidalisland.rda")

png("presentations/images/60yearrunsinusoidalt12x12.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim0[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T12X12, alpha=1000, epsilon=10, beta1=beta0=100", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/60yearrunsinusoidalt3x15.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim1[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T3X15, alpha=1000, epsilon=10, beta1=beta0=300", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/60yearruntsinusoidal15x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim2[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T15X3, alpha=2000, epsilon=10, beta1=200, beta0=300", outer = TRUE, cex.main = 2)
dev.off()

png("presentations/images/60yearrunsinusoidalt6x3.png", width = 3000, height = 1500, res = 300)
par(mfrow = c(3,5), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
for(i in 1:15) 
  image(1:100, 1:100, sim3[[i]], main = paste("year", i-1), yaxt = "n", xaxt = "n")
title("T6X3, alpha=500, epsilon=5, beta1=beta0=200", outer = TRUE, cex.main = 2)
dev.off()

```
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrunsinusoidalt12x12.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrunsinusoidalt3x15.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrunsinusoidalt15x6.png)
![](/Users/sriyapotluri/Documents/Research/memorymigration/presentations/images/60yearrunsinusoidalt6x3.png)

## NA Runs There were two runs that had missing values of foraging efficiency and the final similatiry in the sinusoidal intitial condition. 
```{r, message=FALSE, warning=FALSE, results='asis'}

load("/Users/sriyapotluri/Documents/Research/memorymigration/results/Results_Island_Gaussian.rda")
r.gaussian <- results
r.gaussian[!complete.cases(r.gaussian),]

load("/Users/sriyapotluri/Documents/Research/memorymigration/results/Results_Island_Sinusoidal.rda")
r.sinusoidal <- results
naruns <- r.sinusoidal[!complete.cases(r.sinusoidal),]

library(knitr)
kable(naruns, caption="NaN runs")
```

```

### 2021-04-10 text/ms.Rmd
Changed the model Wrote some text Copied CWD shiny app

@@
 output:
-  word_document:
-    toc: yes
   html_document:
@@
 
-We intend to study how a blending of a *tactical* (i.e.
+We study how a blending of a *tactical* (i.e.
 direct response to resource availability or perception) and a *strategic* (i.e.
-memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior**.
-Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
-Importantly, the environment is *approximately periodic*, i.e.
+memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment, i.e.
 one in which the distribution of the resource at a given time is approximately equal to that resource distribution at a fixed previous time period, i.e.
 $h(t) \approx h(t - \tau)$.
-This models annual variations, and the period corresponds to one year.
+
+Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
 The *tactical* process is straightforward taxis up a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.
@@
 
-$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + \alpha \frac{\partial}{\partial x}\left(u \frac{\partial h(x,t)}{\partial x}\right) + \sum_{i = 0}^n \beta_i \frac{\partial u(x, t - i\tau)}{\partial x}$$
-where $u$ represents a population distributed in time and space, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta_i$ is the strength of advection up the gradient of the *population* distribution in an $i$'th earlier seasonal cycle (year), up to $n \in N$.
+$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + 
+\alpha \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial h(x,t)}{\partial x}\right) + 
+\beta  \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial m(x,t)}{\partial x}\right) \,\,\,\,(1)$$
 
-Note that $\beta_0$ is similar to an index of "cohesion", essentially encouraging the population to stay close together.
-Higher terms of $\beta_i$ should be simple, e.g., if the collective memory reaches only one year back then $n = 1$ and $\beta_i$ is a single constant.
-Alternatively, a form of memory might discounts older "traditions" via $\beta_i = \beta \exp(-i/\gamma)$, where $\gamma$ represents a generational scale of memory.
+where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the gradient memory.
+The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
+$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
+Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e.
+$y = \lfloor t/\tau \rfloor$.
+Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative strenfth of the long-term (reference) memory.
+When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource.
+If $\kappa = 0$, the memory is short-term, and the process responds only to the behavior from the previous year.
+With $0 < \kappa < 1$, the reference memory fades with time.
 

### 2021-04-10 code/moreModelExploration.R

@@ -5,49 +5,45 @@ data(resources_island)
 world_gaussian$resource <- resources_island[["R_t12_x3"]]
 image(world_gaussian$resource)
 
-plotSim <- function(sim, years = NULL, nrow = 1){
-  
- require(gplots)
-  
-  if(is.null(years)){
-    years <- 1:length(sim)
-    n <- length(years)
-    zmax <- max(sapply(sim, max))
-  } else{
-    zmax <- max(sapply(sim, max)[paste0("Year",c(0,years[-length(years)]))])
-    n <- length(years)
-  }
-  
-  par(mfrow = c(nrow,ceiling(n/nrow)), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
-  for(i in years) 
-    image(1:100, 1:100, sim[[i]], 
-          breaks = seq(0, zmax, length = 100), col = rich.colors(99),
-          main = paste("year", i-1), yaxt = "n", xaxt = "n")
-}
+world_sinusoidal <- getSinePop(100, peak.max = 75, peak.min = 25, sd = 2, dx = 1)
+world_sinusoidal$resource <- resources_island[["R_t12_x3"]]
+
+
+eps <- 1
+
+sim.ResourceOnly <- runManyYears(world_sinusoidal, 
+                                 threshold = 1, n.years = 5,
+                                 parameters = c(epsilon = eps, alpha = 100, 
+                                                beta = 0.01, gamma = 0), verbose = TRUE) %T>%
+plotManyRuns(nrow = 1)
 
 
 world_sinusoidal$resource <- resources_island[["R_t12_x3"]]
-sim0 <- runManyYears(world_sinusoidal, 
-                     threshold = 1, n.years = 30,
-                     parameters = c(epsilon = .3, alpha = 5, 
-                                    beta0 = 0, beta1 = 0), verbose = TRUE)
+sim.MemOnly <- runManyYears(world_sinusoidal, 
+                     threshold = 1, n.years = 5,
+                     parameters = c(epsilon = eps, alpha = 0, 
+                                    beta = 40, gamma = 1), verbose = TRUE)
+plotManyRuns(sim.MemOnly, nrow = 1)
 
-plotSim(sim0, nrow = 3)
+
+sim.Combined <- runManyYears(world_sinusoidal, 
+                                 threshold = 1, n.years = 5,
+                                 parameters = c(epsilon = eps, alpha = 100, 
[…]

### 2021-04-10 memorymigration/R/ForagingMemoryModel.R

@@ -9,18 +9,28 @@
 #' \code{alpha} - resource 
 #' following coefficient; \code{beta0} - social cohesion coefficient; \code{beta1} - 
 #' memory coefficient
-#' @param pop_lag the T x X population distribution over time in the previous cycle (year)
+#' @param memory the memory score
 #' @param resource a T x X distribution of the (potentially) dynamic resource.
 #' @param dx time step
 #' @return Model setup; output of \code{\link{tran.1D}} function. 
 #' @export
 
-ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
+ForagingMemoryModel <- function(t, pop, parms, memory, resource, 
+                                dx = dx){
+  tran.1D(C = pop, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/dx + 
+            parms["beta"] * diff(memory)/dx, 
+          dx = dx)
+}
+
+
+ForagingMemoryModel_v0 <- function(t, pop, parms, memory, resource, 
                                 dx = dx){
   tran.1D(C = pop, D = parms["epsilon"], 
           flux.up = 0, flux.down = 0, 
           v = parms["alpha"] * diff(resource)/dx + 
             parms["beta0"] * diff(c(0,pop,0))/dx + 
-            parms["beta1"] * diff(pop_lag)/dx, 
+            parms["beta1"] * diff(memory)/dx, 
           dx = dx)
 }

### 2021-04-10 memorymigration/R/PlottingFunctions.R

@@ -60,3 +60,30 @@ printParameters <- function(p) {
 #' This function...  
 mypersp <- function(x,y,z,...)
   persp(x,y,z, theta=45, phi=45, bor=NA, shade=TRUE, ylab = "x", xlab="time", zlab="population", ...)
+
+#' Plotting simulation results
+#' @export
+
+plotManyRuns <- function(sim, years = NULL, nrow = 1, outer = TRUE){
+  
+  require(gplots)
+  
+  if(is.null(years)){
+    years <- 1:length(sim)
+    n <- length(years)
+    zmax <- max(sapply(sim, max))
+  } else{
+    zmax <- max(sapply(sim, max)[paste0("Year",c(0,years[-length(years)]))])
+    n <- length(years)
+  }
+ 
+  parameters <- attributes(sim)$parameter
+   
+  par(mfrow = c(nrow,ceiling(n/nrow)), mar = c(0,0,3,0), oma = c(2,2,4,2), tck = 0.01)
+  for(i in years) 
+    image(1:100, 1:100, sim[[i]], 
+          breaks = seq(0, zmax, length = 100), col = rich.colors(99),
+          main = paste("year", i-1), yaxt = "n", xaxt = "n")
+  if(outer) 
+    title(outer = TRUE, paste(names(parameters),"=",parameters, collapse = "; "))
+}

### 2021-04-10 memorymigration/R/RunningModel.R

@@ -16,19 +16,19 @@
 #' @seealso \link{getSinePop}, \link{runManyYears}
 #' @export
 #' 
-runNextYear <- function(World, Parameters){
-  pop1 <- World$pop
-  pop2 <- pop1*0
+runNextYear <- function(World, Parameters, Pop_lastyear, Year){
+  pop0 <- World$pop
+  pop2 <- pop0*0
   
   Time <- World$time
   Resource <- World$resource
-  #p <- as.numeric(Parameters)
-  #names(p) <- names(Parameters)
+  
+  w0 <- Parameters["gamma"]^(Year-1)
+  memory <- w0 * pop0 + (1-w0) * Pop_lastyear
   
   for(t in Time){
-    pop_lastyear <- c(0,pop1[t,],0) 
     
-    if(t == 1) pop_now <- pop1[nrow(pop1),] else 
+    if(t == 1) pop_now <- Pop_lastyear[nrow(Pop_lastyear),] else 
       pop_now <- pop2[t-1,]
     
     pop2[t,] <- ode(y = pop_now, 
@@ -36,11 +36,38 @@ runNextYear <- function(World, Parameters){
                     parms = Parameters,
                     func = ForagingMemoryModel, 
                     resource = c(0,Resource[t,],0),
-                    pop_lag = pop_lastyear,
+                    memory = c(0, memory[t,], 0),
                     dx = World$dx)[2,1:ncol(pop2)+1]
   }
   pop2
 }
+
+runNextYear_v0 <- function(World, Parameters, Pop_lastyear, Year){
+  pop0 <- World$pop
+  pop2 <- pop0*0
+  
+  Time <- World$time
+  Resource <- World$resource
+  
+  w0 <- Parameters["gamma"]^(Year-1)
+  memory <- w0 * pop0 + (1-w0) * Pop_lastyear
+  
+  for(t in Time){
+    
+    if(t == 1) pop_now <- Pop_lastyear[nrow(Pop_lastyear),] else 
+      pop_now <- pop2[t-1,]
+    
+    pop2[t,] <- ode(y = pop_now, 
+                    times = 0:1, 
+                    parms = Parameters,
+                    func = ForagingMemoryModel, 
+                    resource = c(0,Resource[t,],0),
+                    memory = c(0, memory[t,], 0),
+                    dx = World$dx)[2,1:ncol(pop2)+1]
+  }
+  pop2
+}
+
 
 #' Run Many Years 
[…]

### 2021-04-10 shiny/CWD/SIP_Shiny.R

@@ -1,0 +1,99 @@
+rm(list=ls())
+try(setwd("C:/Users/Elie/box sync/UWiscLLab"))
+try(setwd("/home/elzizi/Dropbox/UWiscLLab/"))
+pcks <- c("shiny","sf","ggplot2","magrittr","plyr", "gplots")
+lapply(pcks, require, character = TRUE)
+
+source("functions.R")
+source("SIP_ShinyScripts.R")
+
+
+load("data/WI_data.rda")
+WI_data$Z <- WI_data$X.km + 1i*WI_data$Y.km
+M.distance <- with(WI_data, outer(Z,Z, function(z1,z2) Mod(z2-z1)))
+
+eval <- FALSE
+if(eval){
+  parameters <- list(rho = 0.5, gamma = .3, mu_S = 0.05, mu_I = 0.1, 
+                     alpha = 1, W_max = 40, lambda = 20, beta = 1)
+  a <- run_SIP(WI_data, parameters, Tmax = 500, M.distance) %T>% plotTimeSeries
+  x11(); mapResults(a)
+  round(a$I[,2]/max(a$I[,2])*255+1)
+}
+
+source("SIP_ShinyScripts.R"); 
+shinyApp(ui_SIP, server_SIP)
+
+
+
+
+runModel <- function(a, parameters, Tmax = 200){
+  print(a)
+  parameters$alpha <- a 
+  myfit <- run_SIP(WI_data, parameters, Tmax = Tmax, M.distance)
+  data.frame(alpha = a, S = myfit$S[,Tmax], I = myfit$I[,Tmax], county = WI_data$county)
+}
+
+summarizeSim <- function(sim){
+  merge(sim, WI_data[,c("county","Wolf_presence","area", "WTD")], keep = "all") %>% 
+  mutate(P.CWD = I/(S+I), K = 2*WTD, P.WTD = (S+I)/K) %>% 
+    ddply(c("alpha", "Wolf_presence", "M"), summarize, 
+          P.CWD.mean = mean(P.CWD), P.CWD.sd = sd(P.CWD)/sqrt(length(P.CWD)),
+          P.WTD.mean = mean(P.WTD), P.WTD.sd = sd(P.WTD)/sqrt(length(P.WTD)))
+  }
+
+
+eval <- FALSE
+if(eval){
+  alphas <- seq(1,10,1)
+  Tmax <- 200
+  
[…]

### 2021-04-10 shiny/CWD/SIP_ShinyScripts.R

@@ -1,0 +1,59 @@
+ui_SIP <- pageWithSidebar(
+  
+  headerPanel("SIP (Susceptible-Infected-Predation) model"),
+  sidebarPanel(
+    radioButtons("ImageType", "Image Type:", c("Time series" = "timeseries", "Wisconsin map" = "map")),
+    numericInput("Tmax", "Duration of simulation:", value = 100, step = 10),
+    numericInput("Tmax.map", "Map to year:", value = 100, step = 1),
+    sliderInput(inputId = "alpha",
+                label = "Selection preference",
+                min = 0, max = 10, value = 1),
+    sliderInput(inputId = "rho",
+                label = "Adult recruitment",
+                min = 0, max = 1, value = 0.4),
+    sliderInput(inputId = "W_max",
+                label = "Number of deer consumed / wolf / year",
+                min = 10, max = 100, value = 70),
+    sliderInput(inputId = "mu_S",
+                label = "Mortality rate of non-CWD deer (1/year)",
+                min = 0, max = 0.3, value = .06),
+    sliderInput(inputId = "mu_I",
+                label = "Mortality rate of CWD deer (1/year)",
+                min = 0, max = 0.3, value = .12),
+    sliderInput(inputId = "gamma",
+                label = "rate of infection",
+                min = 0, max = 0.1, value = .01),
+    sliderInput(inputId = "lambda",
+                label = "Dispersal kernel (km)",
+                min = 0, max = 200, value = 40),
+    radioButtons(inputId = "beta",
+                label = "Dispersal type:", 
+                choices = c("Gaussian" = "gaussian", "Exponential" = "exponential"), inline = TRUE)
+  ),
+  
+  mainPanel(
+    plotOutput("Image", height = "800px")
+#    plotOutput(outputId = "timeSeries", height = "800px")
+#    plotOutput(outputId = "finalWorldMap"),
+  )
+)
+
+
+server_SIP <- function(input, output) {
+    simulationRun <- reactive({
+      run_SIP(WI_data, parameters = input, Tmax = input$Tmax, M.distance)
+    })  
+    output$Image <- renderPlot({
+      simRun <- simulationRun()
+      switch(input$ImageType, 
+             timeseries = plotTimeSeries(simRun),
+             map = mapResults(simRun, year = input$Tmax.map))
+      #if(input$DrawMap) mapResults(simRun, year = input$Tmax.map)
[…]

### 2021-04-10 shiny/CWD/SIP_Shiny_MWE.R

@@ -1,0 +1,16 @@
+rm(list=ls())
+
+try(setwd("C:/Users/Elie/box sync/UWiscLLab"))
+try(setwd("/home/elzizi/Dropbox/CWD/"))
+try(setwd("/home/elie/Dropbox/CWD/"))
+
+pcks <- c("shiny","sf","ggplot2","magrittr","plyr", "gplots")
+lapply(pcks, require, character = TRUE)
+
+load("data/WI_data.rda")
+WI_data$Z <- WI_data$X.km + 1i*WI_data$Y.km
+M.distance <- with(WI_data, outer(Z,Z, function(z1,z2) Mod(z2-z1)))
+source("functions.R")
+source("SIP_ShinyScripts.R"); 
+shinyApp(ui_SIP, server_SIP)
+  

### 2021-04-14 text/ms.Rmd
Updates on Model and manuscript

@@
 
-Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;Okubo2001].
+Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].
 Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled.
-While the links between diffusion models and movement data are somewhat tenuous (see @Gurarie2011a for quantitative links between obervations and diffusion parameters and @Ovaskainen2004 for applications to mark-recapture data), as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
-Even as equations become analytically intractable,
+While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
+Even as equations analytically intractable,
 
@@
 
-But perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory.
+However, perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory.
 Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy.
@@
 
-Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized.
-To perform a migration, a direct resource-following tactic (taxis) is likely to fail.
+Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized (**refs**).
+While some migrants - both terrestrial and avian - follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015], many others "jump" the green wave, anticipating that critical resources will be available at the time of arrival (**refs**).
 
-# Model
+We anticiapted that under many conditons a blending of *tactical* (i.e.
+direct response to resource availability or perception) and *strategic* (i.e.
+memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment To explore the interaction of memory and resource following as an adaptive strategy in dynamic environments.
+Specifically, we explore a one-dimensional advection diffusion model with long-term and short-term memory, and run the model in a dynamic, seasonal environment.
 
-We study how a blending of a *tactical* (i.e.
-direct response to resource availability or perception) and a *strategic* (i.e.
-memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment, i.e.
-one in which the distribution of the resource at a given time is approximately equal to that resource distribution at a fixed previous time period, i.e.
-$h(t) \approx h(t - \tau)$.
+# Methods
+
+## Memory movement model
 
 Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
-The *tactical* process is straightforward taxis up a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.
+The *tactical* process is positive response to a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.
 Thus:
@@
 $y = \lfloor t/\tau \rfloor$.
-Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative strenfth of the long-term (reference) memory.
+Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.
 When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource.
-If $\kappa = 0$, the memory is short-term, and the process responds only to the behavior from the previous year.
+If $\kappa = 0$, the memory is entirely short-term, and the process responds only to the behavior from the previous year.
 With $0 < \kappa < 1$, the reference memory fades with time.
+The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e.
+${\frac{\partial u(0,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$,  Thus the total population is fixed at all times and found within the domain.
 
-For reference and replicability, I include the incrementally constructed R code, which relies heavily on the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages, which efficiently implement transport ODE and PDE systems in R.
+A few special cases of this model have been studied in the past.
+@Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.
+@Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process.
+Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model.
+@Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function, while @Fagan2019 added a behavioral switching behavior to the non-local information model.
+This study
 
-```{r, message=FALSE, warning=FALSE}
-require("memorymigration")
-```
+## Environment and seasonal resource
+
+We ran our numerical experiments on a spatial domain ranging from 0 to $\chi = 100$, and a period of duration $\tau = 100$ (i.e.~a 100 day "year").
+We are interested in an approximately periodic resource dynamic, i.e.
+one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process.
+to do this.
+Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.
+
+Thus, the resource model consists of two bivariate normal pulses centered at $\mu_p$ at $t_p$ and $\chi - \mu_p$ at $\tau - t_p$ (e.g.
+location 80 and 20 at times 25 and 75, respectively).
+These pulses have a shared time scale of duration $\sigma_t$ and a spatial scale of extent $\sigma_x$, the standard deviation in the time and space dimension respectively.
+This flexible formulation allows us to compare long (duration) and narrow (spatially), long and wide, short and narrow, and short and wide pulses, and to easily simulate scenarios where there is drift or randomness in the location or timing of the pulses.
+Note that in order to make the magnitude of the resource pulse comparable to the magnitude of the memory, which integrates to 1 at every time point, we normalize the resource such that its average value across the year is also 1.
+
+## Metrics
+
+The main metrics we are interested are *migratoriness* and *foraging efficiency*.
+The migratoriness is measured as the maximum distance between the median of the population distribution within a given year.
+The median is a more reliable metric than the mean, as even as distributions become highly assymetric - pulling the mean in one directon or another - the median can  remain consistently in the middle of the spatial domain indicating a net flux of animals of 0.
+This is the anticipated (and realized) result of a purely "tactical" model pursuing the periodic resource pulses.
+
+To quantify the foraging efficiency, i.e., the organisms’ ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya Coefficient [@Bhattacharyya1943] which quantifies the similarity between two distributions.
+We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE).
+Thus, the foraging efficiency index is:
+$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
+where the spatial integral is taken over the domain.  This metric is constraned to be between 0 and 1.
+
+Simulations are run until a (quasi)-equilibrium state is achieved, i.e.
+where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.
+Once that state is attained, we compute the metrics.
+We additionally collect the number of runs it
+
+## Simulation studies
+
+We numerically solved this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.
+
+We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the three main questions.
+
+Q1. What
 
@@
 
-```{r}
-ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
-                                dx = attributes(pop)$dx){
-  tran.1D(C = pop, D = parms["epsilon"], 
-                   flux.up = 0, flux.down = 0, 
-                   v = parms["alpha"] * diff(resource)/dx + 
-                       parms["beta"] * diff(pop_lag)/dx, 
-                   dx = dx)
-}
-```
+![](figures/ScanningResourceMemoryRatios_part1.png)
 
-As a simple illustration of a model run, below the ultimate evolution of a forager that begins in the left part of space and uses a combination of the previous year's location (to the right) and the current resource distribution, and ultimately "splits up" into being attracted to the resource in the middle and its previous year's distribution, with a somewhat stronger taxis towards the resource than the memory:
-
-```{r GradientExample, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
-dx <- .01
-X.max <- 100
-x.edges <- seq(0,100, length=X.max/dx + 1)
-x.mids <- (x.edges[-1] + x.edges[-length(x.edges)])/2
-
-parameters <- c(epsilon = 0.1, alpha = 4, beta = 2)
-pop <- dnorm(x.mids, mean=20, sd = 2) %>% {./sum(.)/dx} 
-attributes(pop)$dx <- dx
-
-pop_lag <- c(0,dnorm(x.mids, mean=80, sd = 2) %>% {./sum(.)/dx},0)
-resource <- c(0,dnorm(x.mids, mean = 50, sd = 4) %>% {./sum(.)/dx},0)
-
-pop.inf <- steady.1D(func = ForagingMemoryModel, 
-                   y = pop, parms = parameters, 
-                   resource = resource, 
-                   pop_lag = pop_lag, 
-                   nspec=1, 
-                   dx = attributes(pop)$dx)$y %>% {./sum(.)/attributes(pop)$dx}
-
-n <- length(x.mids)
-
-cols <- c("black", "darkgrey", "green",  "red")
-lwds <- c(1,3,3,1)
-ltys = c(1,3,3,1)
-
-par(bty = "l", mgp = c(1.5,.25,0), tck = 0.01, las = 1)
-
-plot(x.mids, pop, type = "l", xlab = "space", ylab = "density", 
-     col = cols[1], lwd = lwds[1], lty = ltys[1])
-lines(x.mids, pop_lag[-c(1,n+1)], col = cols[2], lwd = lwds[2], lty = ltys[2])
-lines(x.mids, resource[-c(1,n+1)], col = cols[3], lwd = lwds[3], lty = ltys[3])
-lines(x.mids, pop.inf, col = cols[4], lwd = lwds[4], lty = ltys[4])
-
-legend("top", legend = c("pop. initial", "pop. previous year", "resource", "pop. final"), 
-       lty = ltys, lwd = lwds, col = cols, bty = "n")
-#title(parse(text = printParameters(p)))
-
-printParameters <- function(p) 
-  parse(text = paste0("list(",paste(names(p), p, sep = "==", collapse = ", "),")"))
-
-title(printParameters(parameters))
-
-```
-
-This example is completely static and symmetric around the two drivers.
-Note that all the populations (previous, initial, final) are normalizes (integrate to 1) and that the space is defined between 0 and 100.
-
-# Dynamic initial population
-
-In the following example, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is fixed in the middle of the space (as above).
-The memory is a one-year memory.
-In this context, migration is not, in fact, optimal.
-But the important piece is to see how the response mixes attraction to the resource with attraction to the previous year's behavior.
-Note that the year here (for simplicity, speed and symmetry), is 100 days long.
-
-```{r setupEnvironment}
-getSinePop <- function(tau, X.min = 0, X.max = 100, dx, 
-                       peak.max, peak.min, sd){
-  Time <- 1:tau
-
-  X.edge <- seq(X.min, X.max, dx)
-  X <- (X.edge[-1] + X.edge[-length(X.edge)])/2
-  pop <- outer(Time, X,
-                     function(t,x) 
-                       dnorm(x, mean = X.max/2 + (peak.max-peak.min) / 2 * sin( 2*(t * pi)/max(Time)),sd = sd))
-  pop <- apply(pop, 1, function(x) x/sum(x)/dx) %>% t
-  list(pop = pop, X = X, time = Time, dx = dx, tau = tau)
-}
-
-world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
-                    peak.max = 80, peak.min = 20, sd = 10)
-
-fr1 <- dnorm(world$X, mean = 80, sd = 10) %>% {./sum(.)/world$dx}
-fr2 <- dnorm(world$X, mean = 20, sd = 10) %>% {./sum(.)/world$dx}
-fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
-                         nrow = nrow(world$pop), 
-                         ncol = ncol(world$pop))
-
-world$resource <- fixed.Resource
-```
-
-The following examples illustrate how an initially migrating population evolves, even as it is attracted to a resource that is distributed in two different locations.
-In these examples, the first year (year 0) is a sinusoidal migration.
-We run the simulations for 3 subsequent years.
-
-## Only memory
-
-```{r memoryOnly, fig.height = 5, fig.width = 10}
-
-parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
-
-eval <- FALSE
-if(eval){
-  MemSim <- runManyYears(world, Parameters = parameters, 
-                         n.years = 3, 0.99)
-  save(MemSim, file = "../results/MemSim.rda")
-} else load("../results/MemSim.rda")
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(MemSim, world, tau = tau)
-plotYearList(MemSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-The population roughly echoes its initial track, but degrades from year to year.
-
-## Resource only
-
-In the following simulations, we add a resource distributed as follows in two concentrations:
-
-```{r fixedResource, fig.height = 3, fig.width = 4}
-with(world, 
-  image(X, 1:tau, resource, main = "resource", ylab = "x")
-)
-```
-
-```{r resourceOnly, fig.height = 5, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 5e2, beta0 = 0, beta1 = 0)
-
-eval <- FALSE
-if(eval){
-  FixedResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3, 0.99)
-  save(FixedResourceSim, file = "../results/FixedResourceSim.rda")
-} else load("../results/FixedResourceSim.rda")
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(FixedResourceSim, world, tau = tau)
-plotYearList(FixedResourceSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-Here, the population disregards its memory and only pursues a tactic stategy. Within a year, the population has gathered around the resource.
-
-## Resource and memory
-
-Finally, we combine both resource following and memory:
-
-```{r resourceAndMemory, fig.height = 5, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 3e2, beta0=20, beta1 = 5e2)
-
-eval <- FALSE
-if(eval){
-  MemResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3, 0.99)
-  save(MemResourceSim, file = "../results/MemResourceSim.rda")
-} else load("../results/MemResourceSim.rda")
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(MemResourceSim, world, tau = tau)
-plotYearList(MemResourceSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-Here, the strong cultural memory is slowly blended with a strategy to simply concentrate in one of the two resource hot-spots.
-
-# Seasonal resource
-
-There is obviously nothing particularly adaptive about migrating when resources are fixed.
-In the following scenaries, we explore a seasonal resource.
-We developed (with some fuss!) a seasonal resource function with the following properties:
-
-1. The total amount of resource across space is constant throughout the year.
-2. At the beginning, middle, and end of the year the resource is uniformly distributed. 
-3.
-At some peak time $t_r < \tau/2$, the resource concentrates at a location $x_r < \chi/2$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
-4. The resource peaks exactly symmetrically at time $\tau - t_r$ and location $\chi - x_r$ with the same variance $\sigma_r$.
-
-To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.
-Thus:
-
-$$R(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$
-
-where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:
-
-$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
-$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$
-
-where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.
-These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.
-
-The means and variances themselves are Gaussian pulses, with the mean peaking at $x_r$ at time $t_r$ with standard deviation $\sigma_t$ and at $\chi - x_t$ at time $\tau - t_r$ and the standard deviation pulsing from $\chi/\sqrt{12}$ (corresponding to a uniform distribution) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.
-
-All of these details boil down to the ability to generate various seasonal scenarios.
-
-```{r exampleResourcePlots, fig.width = 10, fig.height = 6}
-source("functions2.R")
-
-world.R1 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))
-
-world.R2 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5)))
-
-world.R3 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9)))
-
-world.R4 <- world %>% 
-  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5)))
-
-par(mfrow = c(2,2), mar = c(3,3,3,1), mgp = c(2,.25,0), las = 1, tck = 0.01)
-with(world.R1, {     
-  image(1:tau,X,resource, main = "Broad, long peak")
-  contour(1:tau,X,resource, add = TRUE)
-  })
-with(world.R2, {     
-  image(1:tau,X,resource, main = "Narrow, long peak")
-  contour(1:tau,X,resource, add = TRUE)
-  })
-with(world.R3, {     
-  image(1:tau,X,resource, main = "Broad, short peak")
-  contour(1:tau,X,resource, add = TRUE)
-  })
-with(world.R4, {     
-  image(1:tau,X,resource, main = "Narrow, short peak")
-  contour(1:tau,X,resource, add = TRUE)
-  })
-
-```
-
-# Running the model in a seasonal environment
-# Note: to view any of the resource models, change the world
-
-In these pulsed resources, we might assume that a *combination* of memory and tactic responses would yield more optimal behavior, and might observe some shift in the migratory strategy.
-
-## Memory only
-
-```{r PulseSim1, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
-Memory_PulsedResource_Sim <- runManyYears(world.R3, Parameters = parameters, n.years = 6, 0.99)
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Memory_PulsedResource_Sim, world.R3, tau = tau)
-plotYearList(Memory_PulsedResource_Sim, world.R3, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-## Tactics only
-
-This model does a surprisingly good job aggregating at the appropriate resource, but appears to only locally aggregate, not to migrate per se.
-
-```{r PulseSim2, fig.width = 10}
-parameters <- c(epsilon = 2, alpha = 4e2, beta0 = 0, beta1 = 0)
-Tactic_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Tactic_PulsedResource_Sim, world.R1, tau = tau)
-plotYearList(Tactic_PulsedResource_Sim, world.R1, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-## Memory and tactics
-
-It is not clear that the hybrid model recovers a true migration, but it seems to retain somewhat more concentration along the migratory corridors.
-
-```{r PulseSim3, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
-Hybrid_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim, world.R1, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim, world.R1, tau = tau, 
-             persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-## Memory and tactics: Narrow resource
-
-Here we compare a memory with and without memory exploiting the "narrow - long" resource combination, and the tendency to
-
-### No memory - strong tactics
-
-Notable concentrations at the resource peaks, but a lot of diffuse behavior in the interim seasons
-
-```{r PulseSim4_tactic, fig.width = 10}
-parameters <- c(epsilon = 2, alpha = 2e2, beta0 = 0, beta1 = 0)
-Tactic_PulsedResource_Sim2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim2, world.R1, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim2, world.R1, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-### Hybrid model
-
-Memory appears to reinforce the aggregations at the (narrow) resource peaks
-
-```{r PulseSim4_memory, fig.width = 10}
-parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
-Hybrid_PulsedResource_Sim3 <- runManyYears(world.R4, Parameters = parameters, n.years = 6, 0.99)
-
-par(mfrow = c(1,2), oma = c(0,0,3,0))
-plotYearList(Hybrid_PulsedResource_Sim3, world.R4, tau = tau)
-plotYearList(Hybrid_PulsedResource_Sim3, world.R4, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
-title(printParameters(parameters), outer = TRUE)
-```
-
-## Example
-We apply the three indices to an example model.
-In this model, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is seasonal in a broad, long peak.
-The model uses the parameters $\epsilon = 5, \alpha = 300, \beta0=100, \beta1 = 300$ so that the migratory population combines both memory and a tactical strategy.
-We run the model for 12 years and compute the three indices.
-Then we plot the three indices to the view the trend of the migratory population over time.
-
-```{r}
-dx <- 1
-
-parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
-world.R1 <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2, 
-                    peak.max = 80, peak.min = 20, sd = 10)
-world.R1$resource <- getPulsedResource(world.R1$time, world.R1$X, c(t.peak = 25, t.sd = 9, 
-                                                           x.peak = 80, x.sd = 9))
-
-#run simulation for 12 years 
-M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 1)
-
-world.R2 <- world.R1 %>% list_modify(pop = M$Year6)
-M2 <- runManyYears(world.R2, Parameters = parameters, n.years = 6, 1)
-
-  indices1 <- sapply(M, computeIndices, resource = world.R1$resource, world = world.R1) %>% t
-  indices2 <- sapply(M2, computeIndices, resource = world.R2$resource, world = world.R2) %>% t
-  I <- rbind(indices1, indices2)
-
-  matplot(I, type="o", pch = 19, lty = 1) #year 6 and year 7 are the same so it is plotted twice 
-  legend("topright", pch = 19, col = 1:3, legend = colnames(I))
-
-```
-Then we run the model to stop running the simulation at a certain threshold when the population reaches a "dynamic equilibrium state".
-In this example, the model only runs three years before hitting the threshold.
-```{r}
-#simulation runs for 3 years since that is when threshold is hit
-
-M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
-plotYearList(M, world.R1, tau = tau)
-
-computeCohesiveness(M[[3]], world.R1)["SC.mean"]
-computeMigratoriness(M[[3]], world.R1)$overlap
-computeEfficiency(M[[3]], world.R1$resource, world.R1)
-computeIndices(M[[3]], world.R1$resource, world.R1)
-
-```
+![](figures/FinalMigratoryFollowing.png){width=50%}
 

### 2021-04-14 code/moreModelExploration.R

@@ -95,38 +95,162 @@ scanResourceMemoryRatios <- function(world, epsilon, alphabeta, threshold = 1,
 }
 
 
-ratios <- seq(0,1,.2)
-SimSet2 <- scanResourceMemoryRatios(world_sinusoidal, epsilon = 1, 
-                                    alphabeta = 200, threshold = 1, 
-                                    ratios = ratios, n.years = 6, verbose = TRUE)
-
-ns <- sapply(SimSet2, length)
 
 png("text/figures/ScanningResourceMemoryRatios_part1.png", width = 2000, height = 2000, res = 300)
 par(mfrow = c(length(ratios), max(ns)), mar = c(0,0,1,0), oma = c(4,2,4,2), tck = 0.01, xpd = NA)
-for(i in 1:length(ratios)){
+for(i in seq(11,1,-2)){
   plotManyRuns(SimSet2[[i]], par = FALSE, outer = FALSE, 
-               ylab = paste(ratios[i]*100, "%"), line = 0.5, labelyears = i == 1)
+               ylab = paste((1-ratios[i])*100, "%"), line = 0.5, labelyears = i == 6)
 }
-mtext(side = 3, "all reference memory", outer = TRUE, line = 1.5)
-mtext(side = 1, "all resource following", outer = TRUE, line = 1)
+mtext(side = 3, "all resource following", outer = TRUE, line = 1.5)
+mtext(side = 1, "all reference memory", outer = TRUE, line = 1)
 dev.off()
 
 
 
 require(gplots)
-palette(rich.colors(length(ratios)))
+
+
+getMedian.daily <- function(sim, plotme = TRUE, ...){
+  
+  get_median <- function(x){
+    f <- function(p)
+      (splinefun((1:length(x))-.5, cumsum(x))(p) - 0.5)^2
+    optimize(f, c(0, length(x)))$minimum
+  }
+  
+  median <- apply(sim, 1, get_median)
+  
+  if(plotme) plot(1:nrow(sim), median, type = "l", ..., xlab = "time", ylim = c(0,nrow(sim)))
+  data.frame(median)
+}
+
+getFE.year <- function(run, world){
+  FE <- c()
+  for(i in 2:length(run))
+    FE <- c(FE, computeEfficiency(run[[i]], world$resource, world = world))
+  return(data.frame(year = (2:length(run))-1, FE = FE))
+}
+
+summarizeSweep <- function(SimSet){
[…]

### 2021-04-14 memorymigration/R/RunningModel.R

@@ -94,8 +94,9 @@ runNextYear_v0 <- function(World, Parameters, Pop_lastyear, Year){
 runManyYears <- function(World, parameters, n.years = 60, 
                          threshold= 0.995, verbose = FALSE){
   
-  cat(paste(names(parameters), parameters, collapse = "; "))
   cat("\n")
+  cat(paste(names(parameters), parameters, collapse = "; "))
+  
   
   pop.list <- list(Year1 = World$pop)
   i <- 1
@@ -106,7 +107,7 @@ runManyYears <- function(World, parameters, n.years = 60,
   
   
   while((similarity < threshold) & (i < n.years)){
-    if(verbose){cat("\n"); cat(paste("running year ", i, "\n"))}
+    if(verbose){cat("\n"); cat(paste("running year ", i))}
     
     i <- i+1
     pop.list[[i+1]] <- runNextYear(World, 

### 2021-05-14 text/ms.Rmd
working on ms ...

@@
 
-Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized (**refs**).
-While some migrants - both terrestrial and avian - follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015], many others "jump" the green wave, anticipating that critical resources will be available at the time of arrival (**refs**).
+Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
+For many species, it is an extremely successful strategy as it allows a greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].
+The proximate and ultimate causes and mechanisms, however, vary considerably across species, and even within species [@Shaw2016].Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015] and can be considered "tactical" migrants.
+Others "jump" a green wave (or any other resource wave) anticipating that critical resources will be available at the time of arrival at the end point of migration (**refs**).
+This second behavior involves a major trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).
+In it's can be considered "strategic" in the sense that it is driven not by immediate cues but by memory-driven anticipation.
+Clearly,
 
-We anticiapted that under many conditons a blending of *tactical* (i.e.
+While migratory species are often considered to be more vulnerable to environmental change, as a disruptions in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].
+On the other hand, it has been argued that migratory species might be more resilient to dispruptions due to their wide-ranging mobility [@Robinson2009].
+
+We anticiapte that under many conditons a blending of *tactical* (i.e.
 direct response to resource availability or perception) and *strategic* (i.e.
-memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment To explore the interaction of memory and resource following as an adaptive strategy in dynamic environments.
-Specifically, we explore a one-dimensional advection diffusion model with long-term and short-term memory, and run the model in a dynamic, seasonal environment.
+memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of *migratory behavior* in a dynamic, seasonal resource environment.
+Our goal is to explore the interaction of memory and resource following as an adaptive strategy in a dynamic environments.
+To do this, we developed a one-dimensional advection diffusion model with long-term and short-term memory, and run the model in a dynamic, seasonal environment.
 

### 2021-05-14 text/ms.Rmd
working on ms

@@
 
+Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
+The ultimate evolutionary causes for migration are that the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or suitably habitat for reproduction in respective seasons [@Avgar2013].
+For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].
+
+Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Shaw2016].
+Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015] and can be considered "tactical" migrants.
+Others "jump" a green wave (or any other resource wave) anticipating that critical resources will be available at the time of arrival at the end point of migration (**refs**).
+This second behavior involves a major trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).
+This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by memory-driven anticipation [@Bracis2017].
+
+While migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].
+On the other hand, it has been argued that migratory species might be more resilient to dispruptions due to their wide-ranging mobility [@Robinson2009].
+Theoretical studies of migration indicate that some phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].
+
 Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].
 Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled.
-While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
-Even as equations analytically intractable,
-
-Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.
+While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.
 by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).
@@
 The interacting role of memory and sociality have been comparatively little studied.
-
-Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
-For many species, it is an extremely successful strategy as it allows a greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].
-The proximate and ultimate causes and mechanisms, however, vary considerably across species, and even within species [@Shaw2016].Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015] and can be considered "tactical" migrants.
-Others "jump" a green wave (or any other resource wave) anticipating that critical resources will be available at the time of arrival at the end point of migration (**refs**).
-This second behavior involves a major trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).
-In it's can be considered "strategic" in the sense that it is driven not by immediate cues but by memory-driven anticipation.
-Clearly,
-
-While migratory species are often considered to be more vulnerable to environmental change, as a disruptions in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].
-On the other hand, it has been argued that migratory species might be more resilient to dispruptions due to their wide-ranging mobility [@Robinson2009].
 

### 2021-05-19 text/ms.Rmd
still working on a new model (!?!?!)

--- 2021-04-14 text/ms.Rmd
@@
 
+Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
+For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].
+The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2013].
+
+Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Shaw2016].
+Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015].
+These migrations can be considered "tactical", as they can occur - as an extreme simplification - purely as  response to local conditions.
+Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration (**refs**).
+This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).
+This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by a remembered anticipation [@Bracis2017].
+
+Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].
+On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility [@Robinson2009].
+Clearly, the ability to perform a migration without local cues in only possible if the behavior is hard-cued or remembered.
+On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
+
+Here, we explore the relationship between memory and resource following for a migratory population under various static and dynamic seasonal resource distributions using a diffusion-advection model.
+
 Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].
-Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled.
+These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component.
 While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
-Even as equations analytically intractable,
-
 Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.
@@
 
-However, perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory.
+Perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory.
 Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy.
-The interacting role of memory and sociality have been comparatively little studied.
+The interacting role of memory and sociality, in contrast, have been comparatively little studied.
 
-Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized (**refs**).
-While some migrants - both terrestrial and avian - follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015], many others "jump" the green wave, anticipating that critical resources will be available at the time of arrival (**refs**).
-
-We anticiapted that under many conditons a blending of *tactical* (i.e.
+We anticipate that under many conditions a blending of *tactical* (i.e.
 direct response to resource availability or perception) and *strategic* (i.e.
-memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment To explore the interaction of memory and resource following as an adaptive strategy in dynamic environments.
-Specifically, we explore a one-dimensional advection diffusion model with long-term and short-term memory, and run the model in a dynamic, seasonal environment.
+memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal resource environments.
 
@@
 
-where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the gradient memory.
+where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the memory gradient.
 The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
@@
 Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model.
-@Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function, while @Fagan2019 added a behavioral switching behavior to the non-local information model.
+@Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function.
+@Fagan2019 added a behavioral switching behavior to the non-local information model.
 This study
+
+Note, this model does not contain any explicit birth or death processes, the population is fixed, and the parameters remain constant throughout time.
+Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic environment.
 
@@
 one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process.
-to do this.
 Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.
@@
 where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.
-Once that state is attained, we compute the metrics.
-We additionally collect the number of runs it
+Once that state is attained, we compute the mirgatoriness and foraging efficiency metrics, and additionally collect the number of years that were required to reach stationarity.
 
@@
 
+Our model is somewhat similar to an evolutionary individual-based model [@Anderson2013], which showed that some inherent phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].
+
 # References

### 2021-05-19 code/NewModel.R

@@ -1,0 +1,110 @@
+
+getDMem <- function(pop, world){
+  memory <- apply(t(pop)*world$X, 2, sum) * world$dx
+  dmem.raw <-  diff(c(memory[length(memory)], memory, memory[1])) 
+  (dmem.raw[-1] + dmem.raw[-length(dmem.raw)])/2
+}
+
+ForagingMemoryModel_v2 <- function(t, pop, parms, 
+                                   dmem.ref, dmem.new,
+                                   resource, 
+                                   dx = dx){
+  tran.1D(C = pop, D = parms["epsilon"], 
+          flux.up = 0, flux.down = 0, 
+          v = parms["alpha"] * diff(resource)/dx + 
+            parms["beta"] * (parms["kappa"] * dmem.ref + (1-parms["kappa"]) * dmem.new), 
+          dx = dx)
+}
+
+runNextYear <- function(World, Parameters, Pop_lastyear, Year){
+  pop0 <- Pop_lastyear
+  pop2 <- pop0*0
+  Time <- World$time
+  dmem.ref <- World$dmem.ref
+  dmem.new <- getDMem(pop0, World)
+  
+  if(length(dim(World$resource)) == 3)
+    myresource <- World$resource[Year,,] else
+      myresource <- World$resource
+  
+  for(t in Time){
+    if(t == 1) pop_now <- Pop_lastyear[nrow(Pop_lastyear),] else 
+      pop_now <- pop2[t-1,]
+    pop2[t,] <- ode(y = pop_now, 
+                    times = 0:1, 
+                    parms = Parameters,
+                    func = ForagingMemoryModel_v2, 
+                    resource = c(0,myresource[t,],0),
+                    dmem.ref = dmem.ref[t],
+                    dmem.new = dmem.new[t],
+                    dx = World$dx)[2,1:ncol(pop2)+1]
+  }
+  pop2 
+}
+
+runManyYears <- function(World, parameters, n.years = 60, 
+                         threshold= 0.995, verbose = FALSE){
+  
+  cat("\n")
+  cat(paste(names(parameters), parameters, collapse = "; "))
+  
+  pop.list <- list(Year1 = World$pop)
+  i <- 1
+  
+  pop.list[[i+1]] <- runNextYear(World, Parameters = parameters, 
+                                 Pop_lastyear = World$pop, Year = 1)
[…]

### 2021-05-19 code/NewModel_v3.R

@@ -1,0 +1,142 @@
+
+getMem <- function(pop, world){
+  apply(t(pop)*world$X, 2, sum) * world$dx
+}
+
+extend <- function(x){
+  x1 <- c(x[length(x)],x,x[1])
+  (x1[-1] + x1[-length(x1)])/2
+}
+
+ddx.edge <- function(x)
+  diff(c(x[1], x, x[length(x)])) 
+
+
+runManyYears <- function(world, parameters, n.years = 60, 
+                         threshold= 0.995, verbose = FALSE){
+  cat("\n")
+  cat(paste(names(parameters), parameters, collapse = "; "))
+  
+  pop.list <- list(Year1 = world$pop)
+  i <- 1
+  pop.list[[i+1]] <- runNextYear(world, Parameters = parameters, 
+                                 Pop_lastyear = world$pop, Year = 1)
+  similarity <- computeEfficiency(pop.list[[i]], pop.list[[i+1]], World)
+  
+  while((similarity < threshold) & (i < n.years)){
+    if(verbose){cat("\n"); cat(paste("running year ", i))}
+    
+    i <- i+1
+    pop.list[[i+1]] <- runNextYear(World = world, 
+                                   Parameters = parameters, 
+                                   Pop_lastyear = pop.list[[i]], 
+                                   Year = i)
+    similarity <- computeEfficiency(pop.list[[i]], pop.list[[i+1]], World)
+  }
+  names(pop.list) <- paste0("Year",0:(length(pop.list)-1))
+  attr(pop.list, "parameters") <- parameters
+  return(pop.list)
+}
+
+
+
+dm.adjust <- function(x, m.hat, dm, beta, lambda){
+  x0 <- ifelse(dm >= 0, 
+    m.hat + log(lambda/dm - 1)/beta,
+    m.hat - log(lambda/(-dm) - 1)/beta) 
+    
+  ifelse(dm >=0, 
+         lambda / (1 + exp(-beta*(x - x0))),
+         -lambda / (1 + exp(beta*(x - x0)))
+  )
+}
+
+lines(0:10, dm.adjust(0:10, m.hat = 5, dm = 0.2, beta = 1, lambda = 3), xlim = c(0,10))
+
+
+Parameters <- c(epsilon = 1, alpha = 100, beta=.01, kappa = 1, lambda = 10)
+
+
+par(mfrow = c(2,2))
+image.plot(X.matrix)
+image.plot(memory.matrix)
+image.plot(dmemory.matrix)
[…]

### 2021-05-19 memorymigration/R/PlottingFunctions.R

@@ -64,7 +64,7 @@ mypersp <- function(x,y,z,...)
 #' Plotting simulation results
 #' @export
 
-plotManyRuns <- function(sim, years = NULL, nrow = 1, outer = TRUE, 
+plotManyRuns <- function(sim, world, years = NULL, nrow = 1, outer = TRUE, 
                          labelyears = FALSE, 
                          par = NULL, ylab = "", ...){
   
@@ -84,7 +84,7 @@ plotManyRuns <- function(sim, years = NULL, nrow = 1, outer = TRUE,
     par(mfrow = c(nrow,ceiling(n/nrow)), mar = c(1,0,1,0), oma = c(2,2,4,2), tck = 0.01)
   
   for(i in years){
-    image(1:100, 1:100, sim[[i]], 
+    image(1:world$tau, world$X, sim[[i]], 
           breaks = seq(0, zmax, length = 100), col = rich.colors(99),
           yaxt = "n", xaxt = "n", ylab = "", xlab = "")
     if(labelyears) title(paste("year", i-1), line = 0.3)

### 2021-05-19 memorymigration/R/getPulsedResource.R

@@ -20,6 +20,7 @@
 #' @return A X x X matrix containing values of the resource distribution
 #' @example examples/WorldsandResources.R
 #' @export
+#' @aliases getPulsedResource
 
 getCCpars <- function(mu_x0, mu_t0 = 0, beta_x, beta_t, sigma_x, sigma_t, n.years){
   cbind(x.peak = mu_x0 - beta_x*1:n.years,

### 2021-05-19 memorymigration/examples/indices_examples.R

@@ -3,10 +3,10 @@
 dx <- 1
 
 parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
-world.R1 <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2, 
+world.R1 <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=.5, 
                     peak.max = 80, peak.min = 20, sd = 10)
-world.R1$resource <- getPulsedResource(world.R1$time, world.R1$X, c(t.peak = 25, t.sd = 9, 
-                                                           x.peak = 80, x.sd = 9))
+world.R1$resource <- getPulsedResource(world.R1, 
+                                       c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9))
 
 #run simulation for 12 years 
 M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 1)

### 2021-05-28 text/ms.Rmd
Ms

@@
 
-Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory.
-The *tactical* process is positive response to a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.
-Thus:
+In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or *reference*) memory of large-scale movement behavior, (4) a short-term (or *working*) memory that updates movement behavior based on recent experience, (5) some social aspect to the learned behavior.
+
+A diffusion-advection equation provided a computationally efficient and versatile framework for examining just such a system.
+We consider a population moving in one dimension in a constrained domain ${\cal D}$ and distributing itself according to the following equation:
 
 $${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + 
-\alpha \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial h(x,t)}{\partial x}\right) + 
-\beta  \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial m(x,t)}{\partial x}\right) \,\,\,\,(1)$$
+\alpha \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial h(x,t)}{\partial x}\right) + v_m(x, t,y)
+$$
 
-where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the memory gradient.
-The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
-$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
+where  $u(x,t)$ represents the population distributed in time and space, $h(x,t)$ is the distribution of the resource, which in dynamic and patchy as detailed below, $\varepsilon$ is the diffusion rate and$\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), $\beta$ is the strength of advection up the memory gradient.
+Finally, there is a  memory based speed term $v_m(x,t,y)$, described below.
+
+The population has a "social memory" of the mean location and speed at which the population moved exactly one year prior, i.e.
+$\widehat{u}(t-\tau)$, where $\tau$ is the duration of a single year and the $\hat{\,}$ operator signifies the expected value of the population distribution.
+This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:
+
+$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t - y\tau) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$
+
 Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e.
-$y = \lfloor t/\tau \rfloor$.
-Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.
-When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource.
-If $\kappa = 0$, the memory is entirely short-term, and the process responds only to the behavior from the previous year.
-With $0 < \kappa < 1$, the reference memory fades with time.
+$y = \lfloor t/\tau \rfloor$ and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.
+When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.
+Importantly, the individuals remember the movement rate of their mean location:  $d\widehat{u_m}(t)/dt$.
+
+The overarching behavior of the population is to attempt to follow the memory.
+Thus the speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e.
+for those animals that are lagging behind the optimal remembered location, the population's speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.
+
+The specific relationship is essentially a logistic curve:
+
+$$v_m(t) =  \begin{cases}
+\frac{\lambda}{1 + \exp((x - x_+)/\beta)}  &  \frac{d\widehat{u_m}}{dt} > 0\\
+\frac{\lambda}{1 - \exp(-(x - x_-)/\beta)} & \frac{d\widehat{u_m}}{dt} < 0
+\end{cases}$$
+
+where $x_+$ and $x_-$ are fixed such that the value of $v_m(x,t) = d{u_m}/dt$ at $x = \widehat{u_m}$.
+
+$$
+x_+ = \widehat{u_m} - \log\left(\lambda / \frac{d\widehat{u_m}}{dt} - 1\right) \beta \\
+x_- = \widehat{u_m} + \log\left(-\lambda / \frac{d\widehat{u_m}}{dt} - 1\right) \beta
+$$
+
 The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e.
@@
 
+Note, this model does not contain any explicit birth or death processes, the population is fixed, and the parameters remain constant throughout time.
+Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic environment.
+
+<!--
 A few special cases of this model have been studied in the past.
@@
 This study
-
-Note, this model does not contain any explicit birth or death processes, the population is fixed, and the parameters remain constant throughout time.
-Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic environment.
+-->
 

### 2021-05-28 text/appendix.Rmd
More ms

---
title: "Appendix"
author: "Elie"
date: "5/28/2021"
output: html_document
---

# 1. Drifting resource

The drifting resource function has the following properties

1. The total amount of resource across space is constant throughout the year.
2. At the beginning, middle, and end of the year the resource is uniformly distributed. 
3. At some peak time $\mu_t < \tau/2$, the resource concentrates at a location $\mu_x < \chi$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
4. The resource peaks exactly symmetrically at time $\tau - \mu_t$ and location $-\mu_x$ with the same variances.

To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.  Thus:

$$h(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$

where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:

$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$

where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.  These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.

The means and variances themselves are Gaussian pulses, with the mean peaking at $\mu_x$ at time $\mu_t$ with standard deviation $\sigma_t$ and at $-\mu_x$ at time $\tau - \mu_t$ and the standard deviation pulsing from $2 \chi/\sqrt{12}$ (corresponding to a uniform distribution over the domain $-\chi$ to $\chi$) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.

Examples of the drifting resource are below.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
require(memorymigration)
```

```{r exampleResourcePlots, fig.width = 10, fig.height = 6, echo = FALSE}
world <- getSinePop(tau = 100, peak.max = 0, peak.min = 0, sd = 10)

R1 <- getResource_drifting(world, par = c(t.peak = 25, t.sd = 12, x.peak = 80, x.sd = 9))
R2 <- getResource_drifting(world, par = c(t.peak = 25, t.sd = 12, x.peak = 80, x.sd = 5))
R3 <- getResource_drifting(world, par = c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9))
R4 <- getResource_drifting(world, par =  c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5))

par(mfrow = c(2,2), mar = c(3,3,3,1), mgp = c(2,.25,0), las = 1, tck = 0.01)
with(world,{
  image(time,X,R1, main = "Broad, long peak")
  contour(time,X,R1, add = TRUE)})
with(world,{
  image(time,X,R2, main = "Narrow, long peak")
  contour(time,X,R2, add = TRUE)})
with(world,{
  image(time,X,R3, main = "Broad, short peak")
  contour(time,X,R3, add = TRUE)})
with(world,{
  image(time,X,R4, main = "Narrow, short peak")
  contour(time,X,R4, add = TRUE)})
```

### 2021-05-28 text/ms.Rmd
More ms

@@
 
-Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
+Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
 For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].
@@
 
-where  $u(x,t)$ represents the population distributed in time and space, $h(x,t)$ is the distribution of the resource, which in dynamic and patchy as detailed below, $\varepsilon$ is the diffusion rate and$\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), $\beta$ is the strength of advection up the memory gradient.
-Finally, there is a  memory based speed term $v_m(x,t,y)$, described below.
+where  $u(x,t)$ represents the population distributed in time and space, $\varepsilon$ is the diffusion rate and $h(x,t)$ is the distribution of the resource.
+The resource is structured and seasonal with period $\tau$, such that $h(x, t) \approx h(x, t-\tau)$.
+$\alpha$ represents the strength of advection in the direction of the resource $h(x,t)$ gradient, as standard in chemotaxis equations.
+Finally, there is the social memory based speed term $v_m(x,t,y)$, which evolves with the year $y \in \{0,1,2,...\}$, i.e.
+the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.
 
-The population has a "social memory" of the mean location and speed at which the population moved exactly one year prior, i.e.
-$\widehat{u}(t-\tau)$, where $\tau$ is the duration of a single year and the $\hat{\,}$ operator signifies the expected value of the population distribution.
+The social memory term works as follows:  The population retains a memory of the mean location of the population exactly one year prior, i.e.
+$\widehat{u}(t-\tau)$, where $\hat{\,}$ operator signifies the expected value of the population distribution.
 This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:
 
-$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t - y\tau) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$
+$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$
 
-Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e.
-$y = \lfloor t/\tau \rfloor$ and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.
+The long-term, reference memory term, which does not change throughout the evolution of the process, can be thought of a strictly periodic function.
+The parameter $\kappa \in (0,1)$ is the relative contribution of that long-term memory.
 When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.
-Importantly, the individuals remember the movement rate of their mean location:  $d\widehat{u_m}(t)/dt$.
+Importantly, the individuals remember not just the location of the actionable memory term, but the movement rate of that term as well, i.e.
+$d\widehat{u_m}(t)/dt$.
 
-The overarching behavior of the population is to attempt to follow the memory.
-Thus the speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e.
-for those animals that are lagging behind the optimal remembered location, the population's speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.
+The overarching behavior of the population is to attempt to repeat that memory.
+Thus the memory speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e.
+for those animals that are lagging behind the optimal remembered location, the memory speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.
 
-The specific relationship is essentially a logistic curve:
+The specific relationship is a logistic curve:
 
@@
 
-The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e.
-${\frac{\partial u(0,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$,  Thus the total population is fixed at all times and found within the domain.
-
-Note, this model does not contain any explicit birth or death processes, the population is fixed, and the parameters remain constant throughout time.
-Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic environment.
+The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.
+${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.
+Thus the total population is fixed at all times - there are no birth or death processes - and found within the domain.
+Furthermore, the parameters remain constant throughout time.
+Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic and changing environment.
 
@@
 
-## Environment and seasonal resource
+## Seasonal resource
 
-We ran our numerical experiments on a spatial domain ranging from 0 to $\chi = 100$, and a period of duration $\tau = 100$ (i.e.~a 100 day "year").
-We are interested in an approximately periodic resource dynamic, i.e.
-one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process.
-Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.
+We ran our numerical experiments on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day years).
+We were interested in an approximately periodic resource dynamic, i.e.
+one in which $h(x,t \approx h(x, t-\tau))$.
+We generated two types of resource distributions.
+A "non-surfable" resource (*island resource*), and weakly surfable resource (*drifting resource*).
+Both are characterized by a peak in time and space centered at $M_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively).
+These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.
+The island resource is simply two uncorrelated bivariate normal distributions $$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$
+where the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e. ${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.
 
-Thus, the resource model consists of two bivariate normal pulses centered at $\mu_p$ at $t_p$ and $\chi - \mu_p$ at $\tau - t_p$ (e.g.
-location 80 and 20 at times 25 and 75, respectively).
-These pulses have a shared time scale of duration $\sigma_t$ and a spatial scale of extent $\sigma_x$, the standard deviation in the time and space dimension respectively.
-This flexible formulation allows us to compare long (duration) and narrow (spatially), long and wide, short and narrow, and short and wide pulses, and to easily simulate scenarios where there is drift or randomness in the location or timing of the pulses.
-Note that in order to make the magnitude of the resource pulse comparable to the magnitude of the memory, which integrates to 1 at every time point, we normalize the resource such that its average value across the year is also 1.
+The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1.
+This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).
+
+Within a given year, the resource is entirely symmetric: $h(x,t) = h(-x, \tau-t)$.
+However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {\cal N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {\cal N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
+Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
+These trends mirror the poleward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
+In the scenarios we explore, the spatial scales ($s_x$ and $s_t$) remain constant.
 
@@
 $$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
-where the spatial integral is taken over the domain.  This metric is constraned to be between 0 and 1.
+where the spatial integral is taken over the domain.  This metric is constrained to be between 0 and 1.
 
-Simulations are run until a (quasi)-equilibrium state is achieved, i.e.
-where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.
+For simulations with a constant resource, we run those until a quasi-equilibrium state is achieved, i.e.
+where the Bhattacharya index of the population distribution across subsequent years reaches a value of 0.9999.
 Once that state is attained, we compute the mirgatoriness and foraging efficiency metrics, and additionally collect the number of years that were required to reach stationarity.
+
+For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.
+
+We also record the actual memory.
 

### 2021-06-02 text/ms.Rmd
Working on a new model

@@
 editor_options:
-  chunk_output_type: console
+  chunk_output_type: inline
 bibliography: ../library/memorymigration.bib
@@
 Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].
-For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].
-The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2013].
+For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1988].
+The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, a suitable physical, biotic or social environment for reproduction [@Avgar2014].
 
@@
 
-Here, we explore the relationship between memory and resource following for a migratory population under various static and dynamic seasonal resource distributions using a diffusion-advection model.
-
 Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].
@@
 
+Here, we develop a diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
+In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
+This includes the ability to learn to migrate from a non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon (**refs**).
+Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing patterns in resource distribution that are the consequences of rapid climate change.
+
 We anticipate that under many conditions a blending of *tactical* (i.e.
 direct response to resource availability or perception) and *strategic* (i.e.
-memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal resource environments.
+memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal environments.
+Over-reliance on either strategy should be maladaptive.
+We further anticipate that a a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are increasingly stochastic.
 
@@
 In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or *reference*) memory of large-scale movement behavior, (4) a short-term (or *working*) memory that updates movement behavior based on recent experience, (5) some social aspect to the learned behavior.
-
 A diffusion-advection equation provided a computationally efficient and versatile framework for examining just such a system.
@@
 The social memory term works as follows:  The population retains a memory of the mean location of the population exactly one year prior, i.e.
-$\widehat{u}(t-\tau)$, where $\hat{\,}$ operator signifies the expected value of the population distribution.
+$\widehat{u}(t-\tau)$.
 This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:
 
-$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)\,\,\,\,(2)$$
+$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)$$
 
-The long-term, reference memory term, which does not change throughout the evolution of the process, can be thought of a strictly periodic function.
-The parameter $\kappa \in (0,1)$ is the relative contribution of that long-term memory.
+The long-term, reference memory term, which does not change throughout the evolution of the process is a strictly periodic function: $\widehat{u_o}(t)$ = $\widehat{u_o}(t - \tau)$.
+The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory.
 When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.
@@
 
-The overarching behavior of the population is to attempt to repeat that memory.
+The overarching behavior of the population is to attempt to echo that social memory.
 Thus the memory speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e.
@@
 
-$$v_m(t) =  \begin{cases}
-\frac{\lambda}{1 + \exp((x - x_+)/\beta)}  &  \frac{d\widehat{u_m}}{dt} > 0\\
-\frac{\lambda}{1 - \exp(-(x - x_-)/\beta)} & \frac{d\widehat{u_m}}{dt} < 0
+$$v_m(x,t) =  \begin{cases}
+\frac{\lambda}{1 + \exp((x - x_+(t))/\beta)}  &\text{if} &  \frac{d\widehat{u_m}}{dt} > 0\\
+\frac{\lambda}{1 - \exp(-(x - x_-(t))/\beta)} &\text{if} & \frac{d\widehat{u_m}}{dt} < 0
 \end{cases}$$
 
-where $x_+$ and $x_-$ are fixed such that the value of $v_m(x,t) = d{u_m}/dt$ at $x = \widehat{u_m}$.
+where $x_+(t)$ and $x_-(t)$ are fixed such that the value of $v_m(x,t) = d{u_m}/dt$ at $x = \widehat{u_m}$.
 
@@
 
+The parameter $\beta$, the reciprocal of the maximum slope of the logistic regression, can be considered a spatial scale of social cohesion.
+If it is large, the population responds more or less similarly across the domain, with the extreme case $\beta = \infty$ corresponding to a shared velocity of $\frac{d\widehat{u_m}}{dt}$.
+If it is small, then there is a very strong discrepancy in speeds at a short scale (Figure 1).
+The parameter $\lambda$ - corresponding to the maximum possible advective speed - is needed to constrain the speed at large distances from $u_m$ for numerical stability, but is not of central interest in our study.
+
 The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.
 ${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.
-Thus the total population is fixed at all times - there are no birth or death processes - and found within the domain.
-Furthermore, the parameters remain constant throughout time.
-Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' a dynamic and changing environment.
+As there are no birth or death processes, to total population remains fixed and constant, for convenience integrating to 1.
+Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process.
+Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'' an intra- and interannually dynamic, seasonal environment.
 
-<!--
-A few special cases of this model have been studied in the past.
+<!-- A few special cases of this model have been studied in the past.
 @Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.
 @Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process.
-Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model.
+Importantly, the authors prove the well-posedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model.
 @Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function.
 @Fagan2019 added a behavioral switching behavior to the non-local information model.
-This study
+The essential innovation of our study is the inclusion of an explicit, slowly-varying migratory advection term.
 -->
+
+```{r, eval = FALSE}
+require(memorymigration)
+
+x <- -100:100
+m.hat <- 40
+dm <- 5
+beta <- 20
+lambda <- 40
+
+v1 <- getMemoryVelocity(x, m.hat, dm, beta = 20, lambda = 40)
+v2 <- getMemoryVelocity(x, m.hat, dm, beta = 100, lambda = 40)
+v3 <- getMemoryVelocity(x, m.hat, dm, beta = 5, lambda = 40)
+
+png("figures/MemoryVelocity.png", width = 1600, height = 1000, res = 300)
+par(mar = c(3,3,1,1), tck = 0.01, mgp = c(1.5,.25,0), bty = "l", 
+    cex.lab = 1.25)
+plot(x, v1, typ = "l", lwd = 2, ylab= "memory velocity v(t)", 
+     col = "darkred", xlab = "spatial domain (x)", ylim = c(0,45))
+lines(x, v2, col = "darkgreen", lwd = 2)
+lines(x, v3, col = "darkblue", lwd = 2)
+segments(-100,dm,100,dm, col = "darkgrey", lty = 3, lwd = 2)
+abline(v = m.hat, col = "darkgrey", lty = 3, lwd = 2)
+points(m.hat, dm, bg = "darkgrey", pch= 21, cex = 1.5)
+legend("topright", col = c("darkblue", "darkred", "darkgreen"),
+       title = expression(beta), lwd = 2, legend = c(5,20,100))
+text(-85, dm, expression(d~widehat(u[m])/dt), pos = 3, col = grey(.3))
+text(m.hat, lambda, expression(widehat(u[m])(t)), pos = 2, col = grey(.3), xpd = NA)
+dev.off()
+```
+![](figures/MemoryVelocity.png)
+**Figure 1.** *Illustration of memory-based velocity term as a logistic function of space (x-axis), mean memory ($\widehat{u_m}(t)$ - vertical dashed line) and mean movement rate ($d\widehat{u_m}/dt$ - horizontal dashed line).
+Various values of the social spatial scale parameter $\beta$ reflects the maximum slope of the logistic curve such that at small values of the spatial scale, there is a steep difference in the speed around the mean location (blue curve), and at large values of the spatial scale, there as a weak response of the speed to the mean location.
+In all cases, at locations beyond the mean, the speed is lower than at locations lagging behind the mean.*
 
@@
 The island resource is simply two uncorrelated bivariate normal distributions $$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$
-where the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e. ${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.
+where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e.
+${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.
 
-The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1.
+The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1 ($\int_X h(x,t) dx = 1$).
 This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).
 
-Within a given year, the resource is entirely symmetric: $h(x,t) = h(-x, \tau-t)$.
+Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$.
 However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {\cal N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {\cal N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
-Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
-These trends mirror the poleward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
-In the scenarios we explore, the spatial scales ($s_x$ and $s_t$) remain constant.
+Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
+These trends mirror the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
+The spatial scales ($s_x$ and $s_t$) remain constant.
 
@@
 For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.
-
-We also record the actual memory.
+We also record the migration memory $\widehat{u_m}$ for each year.
 
@@
 
-We numerically solved this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.
+We numerically solve this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.
+The complete code is available as an R package (`memorymigration`) available on GitHub at https://github.com/EliGurarie/memorymigration and as an interactive Shiny application at https://url.to.be.determined.
 
 We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the three main questions.
-
-Q1. What
 

### 2021-06-02 code/NewModel_v4.R

@@ -1,0 +1,93 @@
+  require(memorymigration)
+  
+  buildOnRuns <- function(M, world, ...){
+    world$pop <- M[[length(M)]]
+    M2 <- runManyYears(world, threshold = 1, ...)
+    M2[[1]] <- NULL
+    M3 <- c(M, M2)
+    names(M3) <- paste0("Year",1:length(M3))
+    M3
+  }
+  
+  source("code/functions_v4.R")
+  source("memorymigration/R/RunningModel.R")
+  source("memorymigration/R/PlottingFunctions.R")
+  source("memorymigration/R/functions_indices.R")
+  
+  convolvePop <- function(X, pop, lambda){
+    K <- function(x, l) -(x / (2*l^2)) * exp(-x^2/(2*l^2)) 
+    ck <- function(x, pop, X, l) sum(K(x-X, l) * pop)
+    sapply(X, FUN = ck, pop = pop, X = X, l = lambda)
+  }
+  
+  
+  pop <- extend(world$pop[25,]) %>% {./sum(.)}
+  X.edge <- with(world, seq(X.min, X.max, dx))
+  dp1 <- convolvePop(X.edge, pop = pop, lambda = 10)
+  dp2 <- convolvePop(X.edge, pop = pop, lambda = 5)
+  
+  plot(pop, ylim = c(-.02, .02))
+  lines(dp1, col = "red")
+  lines(dp2, col = "green")
+  
+  #. Example 0: null
+  
+  world <- getSinePop(tau = 100, X.min = -50, X.max = 50, 
+                      peak.max = 20, peak.min = -20, sd = 10)
+  world$resource <- getResource_island(world, 
+                                       c(t.peak = 25, t.sd = 9, 
+                                         x.peak = 20, x.sd = 5))
+  p0 <- c(epsilon = 5, alpha = 100, beta = 10,
+          kappa = 0, lambda = 5, gamma = 0.5)
+  M0 <- runManyYears(world, parameters = p0, n.years = 10, 1, verbose = TRUE)
+  #M0 <- buildOnRuns(M0, world, parameters = p0, n.years = 10)
+  plotManyRuns(M0, world, nrow = 2, outer = FALSE, labelyears = TRUE)
+  
+  {
+    par(mfrow = c(1,2), mar = c(2,2,1,1), 
+        tck = 0.01, mgp = c(1.5,.25,0), 
+        bty = "l", cex.lab = 1.25, las = 1, xpd = NA)
[…]

### 2021-06-02 code/NewModel_v5.R

@@ -1,0 +1,50 @@
+require(memorymigration)
+source("code/functions_v5.R")
+
+Parameters <- c(epsilon = 1, alpha = 100, kappa = 0, beta = 100, lambda = 5)
+
+world <- getSinePop(tau = 100, peak.max = 50, peak.min = -50, sd = 10)
+
+
+world$resource <- getResource_island(world, 
+                                     c(t.peak = 25, t.sd = 7, 
+                                       x.peak = 30, x.sd = 5))
+
+image(world$resource)
+
+p0 <- c(epsilon = 1, alpha = 200, kappa = 0, beta = 20, lambda = 30)
+M2 <- runManyYears(world, parameters = p0, n.years = 50, 1, FUN = runNextYear_v2, verbose = TRUE)
+M2 <- buildOnRuns(M2, world, parameters = p0, n.years = 40, verbose = TRUE, FUN = runNextYear_v2)
+
+plotManyRuns(M2, world, nrow = 3)
+doublePlot(M2, world)
+
+
+
+
+
+
+#. Example 2: learning migration
+
+
+world <- getSinePop(tau = 100, peak.max = 0, 
+                    peak.min = 0, sd = 10)
+
+world$resource <- getResource_drifting(world, 
+                                       c(t.peak = 25, t.sd = 12, 
+                                         x.peak = 40, x.sd = 5))
+image(world$resource, col = rich.colors(101))
+
+p0 <- c(epsilon = 2, alpha = 300, kappa = 0, beta = 100, lambda = 30)
+
+M <- runManyYears(world, parameters = p0, FUN = runNextYear_v2, 
+                  n.years = 31, 1, verbose = TRUE)
+
+plotManyRuns(M, world, nrow = 2, outer = FALSE)
+doublePlot(M, world)
+
+
+#plotMemory(M)
+#save(M, file = "results/msexamples/learningtomigrate.rda")
+
+#plot(FE, type = "o")

### 2021-06-02 code/functions_v4.R

@@ -1,0 +1,110 @@
+
+#' Foraging with memory model
+#' 
+#' This function sets up the diffusion advection model 
+#' 
+#' @param t vector of times
+#' @param pop a T x X matrix describing the population distribution 
+#' across the time period.
+#' @param parms named vector of parameters. These are \code{epsilon} - 
+#' diffusion coefficient; \code{alpha} - resource following coefficient; 
+#' \code{beta} - spatial scale of sociality; \code{kappa} - memory 
+#' following coefficient; \code{lambda} - maximum speed
+#' @param memory the memory score
+#' @param dh a row of a resource matrix (a T x X distribution of the 
+#' (potentially) dynamic resource.)
+#' @param dx time step
+#' @return Model setup; output of \code{\link{tran.1D}} function. 
+#' @export
+
+
+runNextYear <- function(world, Parameters, Pop_lastyear, Year){
+  pop0 <- world$pop
+  pop1 <- Pop_lastyear
+  pop2 <- pop0*0
+  Time <- world$time
+  
+  memory <- Parameters["kappa"]^Year * getMem(pop0, world) + 
+    (1 - Parameters["kappa"]^Year) * getMem(pop1, world)
+  # dmemory <- diff(extend(memory))
+  dm1 <- diff(memory)
+  dmemory <- c(dm1[length(dm1)], dm1)
+  
+  if(length(dim(world$resource)) == 3)
+    resource <- world$resource[Year,,] else
+      resource <- world$resource
+  
+  #dp.matrix <- dresource * 0
+  
+  dresource <- apply(resource, 1, ddx.edge) %>% t
+  for(t in Time){
+    if(t == 1) 
+      pop_now <- Pop_lastyear[nrow(Pop_lastyear),] else 
+        pop_now <- pop2[t-1,]
+      dp <- convolvePop(seq(world$X.min, world$X.max, world$dx), 
+                        pop = extend(pop_now), 
+                        lambda = Parameters["lambda"])
+      
+      # dp.matrix[t,] <- extend(dp)
+      pop2[t,] <- ode(y = pop_now, 
+                      times = 0:1, 
+                      parms = Parameters,
+                      func = ForagingMemoryModel, 
+                      dh = dresource[t,],
[…]

### 2021-06-02 code/functions_v5.R

@@ -1,0 +1,164 @@
+runManyYears <- function(world, parameters, n.years = 20, 
+                         threshold= 0.9999, verbose = FALSE, 
+                         FUN = runNextYear){
+  cat("\n")
+  cat(paste(names(parameters), parameters, collapse = "; "))
+  
+  pop.list <- list(Year1 = world$pop)
+  i <- 1
+  pop.list[[2]] <- FUN(world, Parameters = parameters, 
+                         pop.lastyear = world$pop, Year = 1)
+  similarity <- computeEfficiency(pop.list[[i]], pop.list[[i+1]], world)
+  
+  while((similarity < threshold) & (i < n.years)){
+    if(verbose){cat("\n"); cat(paste("running year ", i))}
+    
+    i <- i+1
+    pop.list[[i+1]] <- FUN(world = world, 
+                           Parameters = parameters, 
+                           pop.lastyear = pop.list[[i]], 
+                           Year = i)
+    similarity <- computeEfficiency(pop.list[[i]], pop.list[[i+1]], world)
+  }
+  names(pop.list) <- paste0("Year",0:(length(pop.list)-1))
+  attr(pop.list, "parameters") <- parameters
+  return(pop.list)
+}
+
+
+runNextYear_v2 <- function(world, Parameters, pop.lastyear, Year){
+  pop.ref <- world$pop
+  Time <- world$time
+  
+  memory <- Parameters["kappa"]^Year * getMem(pop.ref, world) + 
+    (1 - Parameters["kappa"]^Year) * getMem(pop.lastyear, world)
+  
+  if(length(dim(world$resource)) == 3)
+    resource <- world$resource[Year,,] else
+      resource <- world$resource
+  
+  # get new X frame of reference
+  maxmem <- max(memory)
+  minmem <- min(memory)
+  X <- world$X
+  X.t <- seq(min(X) + minmem, max(X) + maxmem, world$dx)
+  
+  
+  resource.t <- transformToMemoryFrame(resource, X.t, memory, world)
+  pop.lastyear.t <- transformToMemoryFrame(pop.lastyear, X.t, memory, world)
+  pop.thisyear.t <- resource.t*0
+  dresource.t <- apply(resource.t, 1, ddx.edge) %>% t
+  
+  for(t in Time){
+    if(t == 1) 
[…]

### 2021-06-02 code/ms01_ExamplePlots.R

@@ -1,0 +1,111 @@
+require(memorymigration)
+
+#. Example 1: adaptation
+
+eval <- FALSE
+if(eval){
+  world <- getSinePop(tau = 100, peak.max = 50, peak.min = -50, sd = 10)
+  world$resource <- getResource_island(world, 
+                                       c(t.peak = 25, t.sd = 12, 
+                                         x.peak = 30, x.sd = 5))
+  parameters <- c(epsilon = 5, alpha = 100, 
+                  beta=50, kappa = 0, lambda = 20)
+  M <- runManyYears(world, parameters = parameters, n.years = 22, 1)
+  save(world, M, file = "results/msexamples/adaptation.rda")
+} else load("results/msexamples/adaptation.rda")
+
+
+
+world <- getSinePop(tau = 100, peak.max = 50, peak.min = -50, sd = 10)
+world$resource <- getResource_island(world, 
+                                     c(t.peak = 25, t.sd = 12, 
+                                       x.peak = 30, x.sd = 5))
+parameters <- c(epsilon = 5, alpha = 100, 
+                beta=50, kappa = 0, lambda = 20, gamma = 1)
+M <- runManyYears(world, parameters = parameters, n.years = 10, 1, verbose = TRUE)
+parameters["gamma"] <- 0.95
+M1 <- runManyYears(world, parameters = parameters, n.years = 10, 1, verbose = TRUE)
+
+plotManyRuns(M1, world, nrow = 3, 
+             outer = FALSE, labelyears = TRUE)
+{
+  par(mfrow = c(1,2), mar = c(2,2,1,1), 
+      tck = 0.01, mgp = c(1.5,.25,0), 
+      bty = "l", cex.lab = 1.25, las = 1, xpd = NA)
+  with(world, image(time, X, resource, col = grey.colors(100)))
+  plotMemory(M1, add = TRUE)
+  FE1 <- ldply(M, computeEfficiency, 
+               resource = world$resource, world = world,
+               .id = "year") %>% mutate(year = 1:length(year))
+  plot(FE1, type = "o")
+}
+
+#world$pop <- M[[length(M)]]
+#M2[[1]] <- NULL
+#M <- c(M, M2)
+#names(M) <- paste0("Year", 1:length(M)-1)
+#M2 <- runManyYears(world, parameters = parameters, 
+#                   n.years = 11, 1)
+
+
+
+par(mfrow = c(2,1))
+contour(world$time, -100:100, memory_velocity)
[…]

### 2021-06-02 memorymigration/R/ForagingMemoryModel.R

@@ -18,7 +18,8 @@ ForagingMemoryModel <- function(t, pop, parms,
                                 memory, dh, dx = dx){
   tran.1D(C = pop, D = parms["epsilon"], 
           flux.up = 0, flux.down = 0, 
-          v = parms["alpha"] * dh/dx + memory,
+          v = parms["alpha"] * dh/dx + 
+            parms["gamma"] * memory,
           dx = dx)
 }
 

### 2021-06-02 memorymigration/R/PlottingFunctions.R

@@ -6,8 +6,8 @@
 #' distribution for each year after initial population; produced by the runManyYear function
 #' @param World List of 5: a population distribution across the time period in a T x X matrix,
 #'  a vector with midpoint X-values, the time points for the population as integers 1:tau,
-#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 6.
-#'  Set up by the getSinePop function. 
+#'  the dx value and the tau value. Can incorporate resource attribute into the world to make 
+#'  a list of 6.  Set up by the getSinePop function. 
 #' @param X.max maximum value of X-axis
 #' @param tau maximum value of time T
 #' @param log logical;
@@ -42,11 +42,11 @@ plotYearList <- function(yearlist, world, X.max = 100, tau = 360, log = FALSE,
 #' This function prints the parameters of the diffusion-advection equation used 
 #' on the model on the plot of the population distribution set up by the plotYearList function.
 #' 
-#'  @param p world object; list of 7: a population distribution across the time period in a T x X matrix,
-#'  a vector with midpoint X-values, the time points for the population as integers 1:tau, the minimum value of population distribution
-#'  (X.min), the maximum value of population distribution (X.max),
-#'  the dx value and the tau value. Can incorporate resource attribute into the world to make a list of 8.
-#'  Set up by the getSinePop/getOptimal function. 
+#'  @param p world object; list of 7: a population distribution across the time period in a 
+#'  T x X matrix, a vector with midpoint X-values, the time points for the population as integers 
+#'  1:tau, the minimum value of population distribution (X.min), the maximum value of population 
+#'  distribution (X.max), the dx value and the tau value. Can incorporate resource attribute into 
+#'  the world to make a list of 8.  Set up by the getSinePop/getOptimal function. 
 #'  @return displays the values of the parameters on the plot which are:
 #'   \code{epsilon} - diffusion coefficient; 
 #' \code{alpha} - resource following coefficient; \code{beta} - spatial scale of sociality; \code{kappa} - 

### 2021-06-02 memorymigration/R/RunningModel.R

@@ -23,13 +23,14 @@
 #' @export
 #' 
 runManyYears <- function(world, parameters, n.years = 20, 
-                         threshold= 0.9999, verbose = FALSE){
+                         threshold= 0.9999, verbose = FALSE, 
+                         FUN = runNextYear){
   cat("\n")
   cat(paste(names(parameters), parameters, collapse = "; "))
   
   pop.list <- list(Year1 = world$pop)
   i <- 1
-  pop.list[[i+1]] <- runNextYear(world, Parameters = parameters, 
+  pop.list[[i+1]] <- FUN(world, Parameters = parameters, 
                                  Pop_lastyear = world$pop, Year = 1)
   similarity <- computeEfficiency(pop.list[[i]], pop.list[[i+1]], world)
   
@@ -37,7 +38,7 @@ runManyYears <- function(world, parameters, n.years = 20,
     if(verbose){cat("\n"); cat(paste("running year ", i))}
     
     i <- i+1
-    pop.list[[i+1]] <- runNextYear(world = world, 
+    pop.list[[i+1]] <- FUN(world = world, 
                                    Parameters = parameters, 
                                    Pop_lastyear = pop.list[[i]], 
                                    Year = i)
@@ -119,3 +120,13 @@ runManyRuns <- function (parameters.df, resource_param, world, resource,
   return(newresults) 
 }
 
+
+#' @export
+buildOnRuns <- function(M, world, ...){
+  world$pop <- M[[length(M)]]
+  M2 <- runManyYears(world, threshold = 1, ...)
+  M2[[1]] <- NULL
+  M3 <- c(M, M2)
+  names(M3) <- paste0("Year",1:length(M3))
+  M3
+}

### 2021-06-08 text/ms.Rmd
working on ms

@@
 
-$${\frac{\partial u(x,t)}{\partial t}} = -\varepsilon {\frac{\partial^2 u(x,t)}{\partial x^2}} + 
-\alpha \frac{\partial}{\partial x}\left(u(x,t) \frac{\partial h(x,t)}{\partial x}\right) + v_m(x, t,y)
+$${\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
+\alpha \frac{\partial}{\partial x}\left(u \frac{\partial h}{\partial x}\right) + 
+\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + 
+v_m(t)
 $$
 
-where  $u(x,t)$ represents the population distributed in time and space, $\varepsilon$ is the diffusion rate and $h(x,t)$ is the distribution of the resource.
-The resource is structured and seasonal with period $\tau$, such that $h(x, t) \approx h(x, t-\tau)$.
-$\alpha$ represents the strength of advection in the direction of the resource $h(x,t)$ gradient, as standard in chemotaxis equations.
-Finally, there is the social memory based speed term $v_m(x,t,y)$, which evolves with the year $y \in \{0,1,2,...\}$, i.e.
+where  $u$ represents the population distributed in time and space.
+The first term is the diffusion term, capturing the fast time-scale exploration and "random" movements of individuals, with $\varepsilon$ is the diffusion rate.
+
+The second term represents the attraction to a dynamic resource $h$, with the proportionality of the advection to the gradient of the resource given by the parameter $\alpha$ (note, the population and resource distributions are functions of both space and time $u(x,t)$ and $h(x,t)$ - we omit the dependent variables in the notation for brevity).
+This is the well-studied standard chemotaxic resource-following behavior.
+We borrow the general notation from our earlier related work [@Fagan2017;@Fagan2019].
+
+The third term captures the collective or social advection term of the population via a non-local, density dependent function $v_s(u,x)$.
+If this function takes the form of a convolution around a non-local kernel $k$, i.e.~
+$$v_s(u) = k(x) * u(x)$$
+and that kernel is odd, an attractive or "swarming" behavior can be generated [@Mogilner1999]. We use the kernel analyzed by @Mogilner1999: 
+$$k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$$ 
+The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \widehat{u}$, and in a negative direction when $x > \widehat{u}$.
+The parameter $\lambda$ is a length scale of sociality - roughly one-half the size of the "swarm", which $\beta$ is a parameter which quantifies the overall strength of sociality.
+
+Finally, the last term captures the direct advection that emerges from a memory-driven migratory behavior.
+This term evolves with a set of parameters $\theta_y$ that slowly change each year $y \in \{0,1,2,...\}$, i.e.
 the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.
 
-The social memory term works as follows:  The population retains a memory of the mean location of the population exactly one year prior, i.e.
-$\widehat{u}(t-\tau)$.
-This working memory is combined with a baseline reference memory, $\widehat{u_o}(t)$ via:
+The migration speed is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (nominally, winter and summer) $t_1$, $dt_1$, $t_2$, $dt_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$.
+Thus the remembered migratory speed term is a simple step function given by:
 
-$$\widehat{u_m}(t) = \kappa^y \, \widehat{u_o}(t) + \left(1-\kappa^y\right)\,\widehat{u}(t-\tau)$$
+$$
+v_m(t, \theta_y) = \begin{cases}
+0 &;  & t > t_1 &\text{and} & t \leq t_1 + dt_1 \\ 
+s_{12} &;  &  t > t_1 + dt_1 &\text{and} & t \leq t_2 \\ 
+0 &; & t> t_2 &\text{and} &t \leq t_2 + dt_2 \\ 
+s_{21} &; &  t > t_2+dt_2 & \text{or} & t \leq t_1\\ 
+\end{cases}
+$$
 
-The long-term, reference memory term, which does not change throughout the evolution of the process is a strictly periodic function: $\widehat{u_o}(t)$ = $\widehat{u_o}(t - \tau)$.
+where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + dt_1$, arrive at $x_2$ at $t = t_2$.
+Thus, $s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}$ and $s_{21} =  \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}$.
+We consider these six parameters to be the known or remembered determinants of the migratory behavior, with an initial set $\theta_0$ determining the reference migration behavior, which is updated each year by the experience of the population.
+To perform this updating, we estimate a new set of parameters $\widehat{\theta_y}$ after each year, and combine these new parameters with the reference parameters according to the following weighted mean:
+
+$$\theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y}$$
+where each of the six parameters is updated identically.
+The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimazation of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'$) against the spatial mean of the population process in year $y$ (i.e.
+$\widehat{u}(t) = \int_X u_y(t, x) dx$).
 The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory.
-When $\kappa = 0$, all of the actionable memory is from the preceding year, when $\kappa = 1$, the actionable memory is entirely the reference memory.
-Importantly, the individuals remember not just the location of the actionable memory term, but the movement rate of that term as well, i.e.
-$d\widehat{u_m}(t)/dt$.
+When $\kappa = 0$, all of the actionable memory is from the preceding year.
+When $\kappa = 1$, the actionable memory is entirely the reference memory.
 
@@
 
-The specific relationship is a logistic curve:
-
-$$v_m(x,t) =  \begin{cases}
-\frac{\lambda}{1 + \exp((x - x_+(t))/\beta)}  &\text{if} &  \frac{d\widehat{u_m}}{dt} > 0\\
-\frac{\lambda}{1 - \exp(-(x - x_-(t))/\beta)} &\text{if} & \frac{d\widehat{u_m}}{dt} < 0
-\end{cases}$$
-
-where $x_+(t)$ and $x_-(t)$ are fixed such that the value of $v_m(x,t) = d{u_m}/dt$ at $x = \widehat{u_m}$.
-
-$$
-x_+ = \widehat{u_m} - \log\left(\lambda / \frac{d\widehat{u_m}}{dt} - 1\right) \beta \\
-x_- = \widehat{u_m} + \log\left(-\lambda / \frac{d\widehat{u_m}}{dt} - 1\right) \beta
-$$
-
-The parameter $\beta$, the reciprocal of the maximum slope of the logistic regression, can be considered a spatial scale of social cohesion.
-If it is large, the population responds more or less similarly across the domain, with the extreme case $\beta = \infty$ corresponding to a shared velocity of $\frac{d\widehat{u_m}}{dt}$.
-If it is small, then there is a very strong discrepancy in speeds at a short scale (Figure 1).
-The parameter $\lambda$ - corresponding to the maximum possible advective speed - is needed to constrain the speed at large distances from $u_m$ for numerical stability, but is not of central interest in our study.
-
 The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.
 ${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.
-As there are no birth or death processes, to total population remains fixed and constant, for convenience integrating to 1.
+As there are no birth or death processes, the total population remains fixed and constant, for convenience integrating to 1.
 Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process.

### 2021-06-08 text/ms.Rmd
plugging away at examples

@@
 
-where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + dt_1$, arrive at $x_2$ at $t = t_2$.
-Thus, $s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}$ and $s_{21} =  \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}$.
-We consider these six parameters to be the known or remembered determinants of the migratory behavior, with an initial set $\theta_0$ determining the reference migration behavior, which is updated each year by the experience of the population.
+  where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + dt_1$, arrive at $x_2$ at $t = t_2$, i.e.
+$s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}$ and $s_{21} =  \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}$.
+This  step-like migration function is a one-dimensional version of the migration parameters estimated individually [@Gurarie2017] and for populations [@Gurarie2019] in empirical studies.
+
+We consider these six parameters to be the known or remembered determinants of the migratory behavior, with an initial set $\theta_0$ determining the reference migration behavior.
+This reference migration is updated each year by the experience of the population.
 To perform this updating, we estimate a new set of parameters $\widehat{\theta_y}$ after each year, and combine these new parameters with the reference parameters according to the following weighted mean:
@@
 where each of the six parameters is updated identically.
-The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimazation of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'$) against the spatial mean of the population process in year $y$ (i.e.
+The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'$) against the spatial mean of the population process in year $y$ (i.e.
 $\widehat{u}(t) = \int_X u_y(t, x) dx$).
@@
 When $\kappa = 1$, the actionable memory is entirely the reference memory.
-
-The overarching behavior of the population is to attempt to echo that social memory.
-Thus the memory speed of the population at $x = \widehat{u_m}(t)$ is $du_m(t)/dt$; if $du_m/dt > 0$ and $x < \widehat{u_m}(t)$, i.e.
-for those animals that are lagging behind the optimal remembered location, the memory speed is greater than $du_m(t)/dt$, and if $x > \widehat{u_m}(t)$ the speed is somewhat lower, to allow the remainder of the population to catch up.
 
@@
 The essential innovation of our study is the inclusion of an explicit, slowly-varying migratory advection term.
--->
 
-```{r, eval = FALSE}
-require(memorymigration)
-
-x <- -100:100
-m.hat <- 40
-dm <- 5
-beta <- 20
-lambda <- 40
-
-v1 <- getMemoryVelocity(x, m.hat, dm, beta = 20, lambda = 40)
-v2 <- getMemoryVelocity(x, m.hat, dm, beta = 100, lambda = 40)
-v3 <- getMemoryVelocity(x, m.hat, dm, beta = 5, lambda = 40)
-
-png("figures/MemoryVelocity.png", width = 1600, height = 1000, res = 300)
-par(mar = c(3,3,1,1), tck = 0.01, mgp = c(1.5,.25,0), bty = "l", 
-    cex.lab = 1.25)
-plot(x, v1, typ = "l", lwd = 2, ylab= "memory velocity v(t)", 
-     col = "darkred", xlab = "spatial domain (x)", ylim = c(0,45))
-lines(x, v2, col = "darkgreen", lwd = 2)
-lines(x, v3, col = "darkblue", lwd = 2)
-segments(-100,dm,100,dm, col = "darkgrey", lty = 3, lwd = 2)
-abline(v = m.hat, col = "darkgrey", lty = 3, lwd = 2)
-points(m.hat, dm, bg = "darkgrey", pch= 21, cex = 1.5)
-legend("topright", col = c("darkblue", "darkred", "darkgreen"),
-       title = expression(beta), lwd = 2, legend = c(5,20,100))
-text(-85, dm, expression(d~widehat(u[m])/dt), pos = 3, col = grey(.3))
-text(m.hat, lambda, expression(widehat(u[m])(t)), pos = 2, col = grey(.3), xpd = NA)
-dev.off()
-```
-![](figures/MemoryVelocity.png)
 **Figure 1.** *Illustration of memory-based velocity term as a logistic function of space (x-axis), mean memory ($\widehat{u_m}(t)$ - vertical dashed line) and mean movement rate ($d\widehat{u_m}/dt$ - horizontal dashed line).
@@
 In all cases, at locations beyond the mean, the speed is lower than at locations lagging behind the mean.*
+-->
 
@@
 These trends mirror the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
-The spatial scales ($s_x$ and $s_t$) remain constant.
+The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.
 
@@
 For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.
-We also record the migration memory $\widehat{u_m}$ for each year.
+We also record the migration parameters $\widehat{\theta_y}$ for each year.
 
@@
 
-We numerically solve this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.
+We explored this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.
+We additionally used the `nlsLM` function in package `minpack.LM` [@minpack.lm] for robust and fast annual estimation of the migration parameters.
 The complete code is available as an R package (`memorymigration`) available on GitHub at https://github.com/EliGurarie/memorymigration and as an interactive Shiny application at https://url.to.be.determined.
@@
 
-![](figures/ScanningResourceMemoryRatios_part1.png)
+```{r}
+require(memorymigration)
+source("../code/")
+```
+
+![](figures/example1_adaptation.png)
 

### 2021-06-08 code/NewModel_v6.R

@@ -15,19 +15,18 @@ world$resource <- getResource_island(world,
 
 image(world$resource)
 
-p0 <- c(epsilon = 1, alpha = 100, kappa = 1, beta = 100, lambda = 20)
-M0 <- runManyYears(world, parameters = p0, n.years = 10, 1, FUN = runNextYear, verbose = TRUE)
+p0 <- c(epsilon = 1, alpha = 200, kappa = 0, beta = 100, lambda = 30)
+M0 <- runManyYears(world, parameters = p0, n.years = 40, 1, FUN = runNextYear, verbose = TRUE)
 #M0 <- buildOnRuns(M0$pop, world, parameters = p0, n.years = 20, verbose = TRUE, FUN = runNextYear)
 doublePlot(M0$pop, world)
 plotManyRuns(M0$pop, world, nrow = 3)
-
-
-plotMigrationHat(M0$m.hat, 50, 25)
+plotMigrationHat(M0$m.hat, 30, 25)
 
 
 
 
-world <- getOptimalPop(tau = 100, x1 = 50, x2 = -50, t.peak = 25, t.sd = 12, x.sd = 12)
+world <- getOptimalPop(tau = 100, x1 = 50, x2 = -50, 
+                       t.peak = 25, t.sd = 12, x.sd = 12)
 world$resource <- getResource_drifting(world, 
                                      c(t.peak = 25, t.sd = 12, 
                                        x.peak = 50, x.sd = 12))

### 2021-06-08 code/functions_v6.R

@@ -25,8 +25,9 @@ buildOnRuns <- function(M, world, ...){
   M2$pop[[1]] <- NULL
   M3 <- list()
   M3$pop <- c(M$pop, M2$pop)
-  M3$m.hat <- c(M$m.hat, M2$m.hat)
-  names(M3$pop) <- paste0("Year",1:length(M3$pop))
+  M3$m.hat <- rbind(M$m.hat, M2$m.hat)
+  M3$m.hat$year = 1:nrow(M3$m.hat) - 1
+  names(M3$pop) <- paste0("Year",1:length(M3$pop)-1)
   M3
 }
 

### 2021-06-08 text/figures/code/ExampleRuns.R

@@ -1,0 +1,85 @@
+#rm(list=ls())
+require(memorymigration)
+source("code/functions_v6.R")
+source("code/functions_discretemigration.R")
+source("code/functions_plottingresults.R")
+require(minpack.lm); require(fields); require(scales)
+
+#############################
+# Example 1: Adaptation
+#############################
+
+
+eval <- FALSE
+if(eval)
+{
+  world <- getOptimalPop(tau = 100, x1 = 50, x2 = -50, 
+                         t.peak = 25, t.sd = 12, x.sd = 5)
+  world$m0 <- fitMigration(t = world$time, x = getMem(world$pop, world))
+  world$resource <- getResource_island(world, 
+                                       c(t.peak = 25, t.sd = 5, 
+                                         x.peak = 30, x.sd = 10))
+  
+  p0 <- c(epsilon = 1, alpha = 100, kappa = 0, beta = 100, lambda = 40)
+  M0 <- runManyYears(world, parameters = p0, n.years = 40, 1, 
+                     FUN = runNextYear, verbose = TRUE)
+# M0 <- M0 %>% buildOnRuns(world, parameters = p0, n.years = 20, 
+#                     FUN = runNextYear, verbose = TRUE)
+  save(world, M0, file = "results/msexamples/adaptation.rda")
+}
+
+load("results/msexamples/adaptation.rda")
+source("text/figures/code/functions_plotting.R")
+
+n <- 14
+
+png("text/figures/example1_adaptation.png", 
+    width = 3000, height = 3000, res = 300)
+
+{
+  M <- rbind(1:n, rep(n+1:2, each = n/2), rep(n+3:4, each = n/2))
+  layout(M)
+  
+  par(oma = c(2,4,2,2), mgp = c(2,.25,0), las = 1, mar = c(4,0,0,0), 
+      xpd = NA, tck = 0, bty = "l")
+  M0.example <- M0$pop[paste0("Year",1:n - 1)] %T>% 
+    plotSomeYears(world, labelyears = TRUE, par = TRUE)
+  
+  par(mar = c(3,4,2,2), xpd = 0, cex.lab = 1.75, cex.axis = 1.25)
+  doublePlot(M0$pop, world, par = TRUE)
+  plotMigrationHat(M0$m.hat, 30, 25, par = TRUE)
+}
+
+dev.off()
+
+
+#####################################
+# Example 2: Learning to migrate
+######################################
+
+
[…]

### 2021-06-08 text/figures/code/functions_plotting.R

@@ -1,0 +1,95 @@
+
+
+plotMemory <- function(M, add = FALSE){
+  require(plyr)
+  memory.df <- ldply(M, function(l)
+    data.frame(time = 1:nrow(l), 
+               memory = getMem(l, world = world)),.id = "year")
+  
+  if(!add) with(memory.df, plot(time, memory, type= "n"))
+  n.years <- length(unique(memory.df$year))
+  palette(rich.colors(n.years))
+  ddply(memory.df, "year", function(df)
+    with(df, lines(time, memory, col = as.integer(year))))
+  #legend("topright", legend = 1:n.years, col = 1:n.years, 
+  #       lty = 1, ncol = 2, title = "year", bty = "n")
+}
+
+doublePlot <- function(M, world, par = FALSE){
+  if(!par) 
+    par(mfrow = c(1,2), mar = c(2,2,1,1), 
+        tck = 0.01, mgp = c(1.5,.25,0), 
+        bty = "l", cex.lab = 1.25, las = 1, xpd = NA)
+  
+  with(world, image(time, X, resource, col = grey.colors(100),
+                    ylab = "space", xlab = "time of year"))
+  plotMemory(M, add = TRUE)
+  FE1 <- ldply(M, computeEfficiency, 
+               resource = world$resource, world = world,
+               .id = "year") %>% mutate(year = 1:length(year))
+  plot(FE1, type = "o", ylab = "foraging efficiency")
+}
+
+
+plotSomeYears <- function(sim, world, years = NULL, nrow = 1, outer = TRUE, 
+                          labelyears = FALSE, 
+                          par = FALSE, ylab = "", ...){
+  
+  require(gplots)
+  if(is.null(years)){
+    years <- 1:length(sim)
+    n <- length(years)
+    zmax <- max(sapply(sim, max))
+  } else{
+    zmax <- max(sapply(sim, max)[paste0("Year",c(0,years[-length(years)]))])
+    n <- length(years)
+  }
+  
+  parameters <- attributes(sim)$parameter
+  if(!par)
+    par(mfrow = c(nrow,ceiling(n/nrow)), mar = c(1,0,1,0), tck = 0.01)
+  
+  for(i in years){
+    image(1:world$tau, world$X, sim[[i]], 
+          breaks = seq(0, zmax, length = 100), col = rich.colors(99),
[…]

### 2021-06-14 text/ms.Rmd
...

@@
 
-We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the three main questions.
+We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the three main questions:
+
+1. Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality?
+
+2. Can this model "acquire" a migratory behavior from a non-migratory initial condition?
+
+3. Can this model adapt when the resource peak shifts in space and time?  What is the role of the reference memory.
 
@@
 
-```{r}
-require(memorymigration)
-source("../code/")
-```
+## Adaptation
 
 ![](figures/example1_adaptation.png)
+**Figure 1.
+Example of adaptation to a shift in resource peak.
+The initial (year 0) behavior migrates to locations 50 and -50 at days 15 and 60, whereas the resource peak is at 30 and -30, peaking at times 25 and 75.
+(a) shows the first 14 years of the simulation; (b) The centroid of the annual movement of the population is shown in panel b, with dark blue to red colors indicating year 0 to year 40; (c) annual foraging efficiency across years; (d) migration timing parameters for each year, with orange segments indicating arrival and departure from the summering (northern) grounds, and the blue segments indicating timing of arrival and departure at the wintering grounds; (e) migration arrival and departure location across years, with blue and orange indicating winter (southern) and summer (northern) locations.**
 
-![](figures/FinalMigratoryFollowing.png){width=50%}
+We illustrate an example of a population with a particular migration behavior confronted with a major sudden shift in the spatio-temporal distribution of the resource in Figure 1.
+In the illustrated scenario, it takes nearly 40 years to attain an equilibrium, and the eventual "optimum" is one where the centroid of the migration lines up exactly with the centroid of the resource, and the arrival timing conincides with the *peak* of resource availability.
+Notably, the path to this equilibrium is somewhat indirect, with the summer season, in particular, taking
+
+## Learning to migrate
+
+![](figures/example2_learningtomigrate.png)
+**Figure 2.
+Example of model learning to migrate.
+The resource is a "weakly drifting" resource and the initial (year 0) condition is non migratory.
+The simulation was run for 100 years, and a sampling of those years (labeled) are presented in panel a:  all years from 0 to 10, followerd by 20, 40 and 100.
+Otherwise, panels are as in Figure 1.
+Parameter values were $\epsilon  = 5$, $\alpha = 500$ and $\beta = 50$.**
+
+Figure 2 illustrates the ability of the model was able to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 50 and -50, but a uniform distribution of resource at times 0 and 50.
+In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
+The qualitative behavior of this process was to start drifting northward, while slowly developing a weak southward pulse in winter as well.
+After first locking in on the summer resource, the southward migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource could be reached consistently.
+
+## Adapting to climate change
 
@@
 
-1. The model seems to - approximately - work for exploring this dynamic.
-2.
-It is - however - difficult to make it really repeat previous behavior strongly.
-I was hoping for a "knob" that perfectly replicates past behavior, but the diffusion-taxis formulation makes that somehow hard.
-I haven't programmed longer time scaled memories (> 1 year) but that would probably be helpful / interesting.
-3.
-The memory bit does demand iteratively feeding a year's cycle back into it rather than having some analytical function, which slows the whole thing down a bit.
-That said, with deepthought a lot of runs could be fitted simultaneously - and I tend to be very picky about "slow" - it actually only takes about 6 seconds to run a 6 year hybrid model on a 100 x 100 spatio-temporal grid.
-But sometimes, the results are more numerically stable on a finer grid.
-4. We need a metric of success!  And also - somehow - a metric of "migratoriness". 
-5. What questions do we wnat to explore?  I think a big one is adaptability of the model under different kinds of resource perturbations, e.g. if the timing shifts, or location shifts.
+We found that the relatively straightforward inclusion of a migration specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
+In essense, our model allowed a population of individuals with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location.
+The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether as trends or inter-annual stochasticity.
 

### 2021-06-14 text/figures/code/ExampleRuns.R

@@ -33,7 +33,7 @@ if(eval)
 }
 
 load("results/msexamples/adaptation.rda")
-source("text/figures/code/functions_plotting.R")
+source("text/figures/code/functions_plottingForMS.R")
 
 n <- 14
 
@@ -77,15 +77,32 @@ if(eval)
   })
   p0 <- c(epsilon = 4, alpha = 500, kappa = 0, beta = 50, lambda = 40)
   M0 <- runManyYears(world, parameters = p0, n.years = 100, 1, verbose = TRUE)
-  
-  #m0 <- M0$migration.hat[nrow(M0$migration.hat),] %>% mutate(year = NULL) %>% 
-  #  as.list %>% unlist
-  #M0b <- M0 %>% buildOnRuns(world, parameters = p0, n.years = 50, verbose = TRUE, m0 = m0)
-  #M0 <- M0b
   save(world, M0, file = "results/msexamples/learningtomigrate.rda")
 }
-load("results/msexamples/learningtomigrate.rda")
 
-M0.example <- M0$pop[paste0("Year",seq(0,100,10))] %T>% plotSomeYears(world, labelyears = TRUE, nrow = 2)
-doublePlot(M0$pop, world, par = FALSE)
-plotMigrationHat(M0$memory.hat, x.peak = 50, t.peak = 25, par = FALSE, ylim1 = c(0,120))
+load("results/msexamples/learningtomigrate.rda")
+source("text/figures/code/functions_plottingForMS.R")
+require(fields)
+require(scales)
+
+require(memorymigration)
+png("text/figures/example2_learningtomigrate.png", width = 3000, height = 3000, res = 300)
+{
+  M0.example <- M0$pop[paste0("Year",c(0:10,20,40,100))] %T>% 
+    plotSomeYears(world, labelyears = TRUE, nrow = 1)
+  
+  n <- length(M0.example)
+  M <- rbind(1:n, rep(n+1:2, each = n/2), rep(n+3:4, each = n/2))
+  layout(M)
+  
+  par(oma = c(2,4,2,2), mgp = c(2,.25,0), las = 1, mar = c(4,0,0,0), 
+      xpd = NA, tck = 0, bty = "l")
+  M0.example <- M0$pop[paste0("Year",c(0:10,20,40,100))] %T>% 
+    plotSomeYears(world, labelyears = TRUE, nrow = 1, par = TRUE)
+  
[…]

### 2021-06-14 text/figures/code/functions_plottingForMS.R

@@ -55,11 +55,12 @@ plotSomeYears <- function(sim, world, years = NULL, nrow = 1, outer = TRUE,
       axis(2, at = seq(-80,80,20))
       mtext(side = 2, line = 1.5, 
             "space", las = 0, cex = 1.25)
+      title(side = 3, "year 0", line = 0.3)
     }
     if(i == round(max(years)/2))
       mtext(side = 1, line = 2, "time", cex = 1.25)
     
-    if(labelyears) title(paste("year", i-1), line = 0.3)
+    if(labelyears & i > 1) title(substring(names(sim)[i],5), line = 0.3)
     if(i == 1) mtext(side = 2, ylab, ...)
     box(col = "white", bty = "o")
   }

### 2021-06-14 text/figures/code/ExampleRuns.R
.....

@@ -101,8 +101,8 @@ png("text/figures/example2_learningtomigrate.png", width = 3000, height = 3000, 
     plotSomeYears(world, labelyears = TRUE, nrow = 1, par = TRUE)
   
   par(mar = c(3,4,2,2), xpd = 0, cex.lab = 1.75, cex.axis = 1.25)
-  doublePlot(M0$pop, world, par = TRUE, ylim1 = c(0,120))
-  plotMigrationHat(M0$migration.hat, 30, 25, par = TRUE)
+  doublePlot(M0$pop, world, par = TRUE)
+  plotMigrationHat(M0$migration.hat, 30, 25, par = TRUE, ylim1 = c(0,120))
 }
 dev.off()  
 

### 2021-06-14 text/ms.Rmd
,,,
Updates ... notably, new migration indices

@@
 
+## Migration
+
+Under what conditions is migration stable:
+
+sigma_t, sigma_x, 
+epsilon, alpha, 
+beta, lambda, 
+x_peak = 30, t_peak = 25
+
 ## Adaptation
@@
 Notably, the path to this equilibrium is somewhat indirect, with the summer season, in particular, taking
+
+*Under what conditions can we adapt.*>
 
@@
 
+*Under what conditions can be learn.*
+
 ## Adapting to climate change
@@
 
+However, the ability to migrate was highly dependent on the nature of the resource dynamics.
+In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
+**ALSO SOME STUFF WE LEARN ABOUT THE INTERACTION BETWEEN DIFFUSION, RESOURCE FOLLOWING, SOCIALITY**
+
+It is worth considering the assumptions of our model with respect to a biological population.  Notably, the fact that the memory is not easily interpreted as an
+
+**Describing conditions for adaptability.**
+
+**Describing conditions for learning migration.**
+
+**Describing conditions for climate change resilience.**
+
 Our model is somewhat similar to an evolutionary individual-based model [@Anderson2013], which showed that some inherent phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].

### 2021-06-14 code/testingResults_0614.R

@@ -1,0 +1,101 @@
+ i <- 40
+
+computeMigratoriness(M0$pop[[i]], world)
+image(M0$pop[[i]])
+
+
+MI <- sapply(M0$pop, function(p) computeMigratoriness(p, world)$overlap)
+plot(MI)
+
+
+load("results/climate_space/climate_space_betax_0.rda")
+r.betax0 <- newresults
+load("results/climate_space/climate_space_betax_1.rda")
+r.betax1 <- newresults
+load("results/climate_space/climate_space_betax_3.rda")
+r.betax3 <- newresults
+
+
+with(r.betax0, table(beta, epsilon, alpha))
+require(ggthemes)
+r.betax0 %>% 
+  ggplot(aes(alpha, beta, fill = FE)) + 
+  facet_grid(lambda~epsilon) + 
+  geom_tile() + theme_few() + ggtitle("no climate change")
+
+r.betax1 %>% 
+  ggplot(aes(alpha, beta, fill = FE)) + 
+  facet_grid(lambda~epsilon) + 
+  geom_tile() + theme_few()+ ggtitle("medium climate change")
+
+
+r.betax3 %>% 
+  ggplot(aes(alpha, beta, fill = FE)) + 
+  facet_grid(lambda~epsilon) + 
+  geom_tile() + theme_few()+ ggtitle("high climate change")
+
+
+world <- getOptimalPop(tau=100, X.min = -100, X.max = 100, dx=1, 
+                       x1 = 60, x2 = -60, t.peak=25, 
+                       x.sd=5, t.sd=12)
+world$m0 <- fitMigration(t = world$time, x = getMem(world$pop, world))
+
+
+resource_param <- expand.grid(mu_x0 = 60, mu_t0 = 25,
+                              beta_x = 0, beta_t = 0,
+                              n.years = 50, sigma_x = 10,
+                              sigma_t = 5,
+                              psi_x = 0, psi_t = 0)
+
+j <- 1
+par0 <- with(resource_param, 
+             getCCpars(mu_x0 = mu_x0[j],
+                  mu_t0 = mu_t0[j],
+                  beta_x = beta_x[j],
+                  beta_t = beta_t[j],
+                  n.years = n.years[j],
+                  sigma_x = sigma_x[j],
+                  sigma_t = sigma_t[j],
+                  psi_x = psi_x[j], 
[…]

### 2021-06-14 memorymigration/R/RunningModel.R

@@ -32,6 +32,7 @@ runManyYears <- function(world, parameters, n.years = 20,
   
   migration.list <- list(m0)
   memory.list <- list(m0)
+  
   pop.list <- list(world$pop)
   similarity <- 0
   i <- 1 
@@ -63,13 +64,18 @@ runManyYears <- function(world, parameters, n.years = 20,
 #' 
 buildOnRuns <- function(M, world, ...){
   world$pop <- M$pop[[length(M$pop)]]
-  M2 <- runManyYears(world, threshold = 1, ...)
+  
+  M2 <- runManyYears(world, threshold = 1, m0 = M0$memory.hat[1,1:6], ...)
   M2$pop[[1]] <- NULL
+  
   M3 <- list()
   M3$pop <- c(M$pop, M2$pop)
   
-  M3$migration.hat <- rbind(M$migration.hat, M2$migration.hat[-1,]) %>% mutate(year = 1:length(M3$pop)-1)
-  M3$memory.hat <- rbind(M$memory.hat, M2$memory.hat[-1,]) %>% mutate(year = 1:length(M3$pop)-1)
+  M3$migration.hat <- rbind(M$migration.hat, M2$migration.hat[-1,]) %>% 
+    mutate(year = 1:length(M3$pop)-1)
+  
+  M3$memory.hat <- rbind(M$memory.hat, M2$memory.hat[-1,]) %>% 
+    mutate(year = 1:length(M3$pop)-1)
   
   names(M3$pop) <- paste0("Year",1:length(M3$pop)-1)
   M3

### 2021-06-14 memorymigration/R/discreteMigrationFunctions.R

@@ -20,7 +20,7 @@ stepMigration <- function(t, t1, dt1, t2, dt2, x1, x2, tau = 100){
 fitMigration <- function(t, x, m.start = NULL, tau = 100){
   if(is.null(m.start)) m.start <- c(t1 = 15, dt1 = 30, t2 = 55, dt2 = 30, x1 = min(x), x2 = max(x))
   migration.fit <- nlsLM(x ~ stepMigration(t, t1, dt1, t2, dt2, x1, x2, tau = 100), start = as.list(m.start),
-                         lower = c(0,0,0,0,-100,-100), 
+                         lower = c(-100,0,0,0,-100,-100), 
                          upper = c(100,100,100,100,100,100)) 
   summary(migration.fit)$coef[,1]
 }

### 2021-06-14 memorymigration/R/functions_indices.R

@@ -147,6 +147,34 @@ computeMigratoriness <- function(pop, world){
        overlap = minOverlap)
 }
 
+
+#' @export
+computeMigrationIndices <- function(sim, world){
+  
+  dx <- world$dx
+  tau <- world$tau
+  par <- attributes(world$resource)$par
+
+  m.hat <- tail(sim$migration.hat, 1)
+  
+  if(par["t.peak"] < m.hat$t1 | par["t.peak"] > m.hat$t1 + m.hat$dt1)
+    t1.error <- min( abs(par["t.peak"] - m.hat$t1), 
+       abs(par["t.peak"] - m.hat$t1  - m.hat$dt1)) else 
+         t1.error = 0
+  
+  if(world$tau - par["t.peak"] < m.hat$t2 | 
+     world$tau - par["t.peak"] > m.hat$t2 + m.hat$dt2)
+    t2.error <- min( abs(world$tau - par["t.peak"] - m.hat$t2), 
+                     abs( world$tau - par["t.peak"] - m.hat$t2  - m.hat$dt2)) else 
+                       t2.error = 0
+  
+  x1.error <- par["x.peak"] - m.hat$x1
+  x2.error <- - par["x.peak"] - m.hat$x2
+  
+  data.frame(t1.error, t2.error, x1.error, x2.error)
+}
+
+
 #' Compute Indicies
 #' 
 #' Returns the three migratoriness indeces calculated by the cumputeCohesiveness,

### 2021-06-14 memorymigration/R/getPulsedResource.R

@@ -99,6 +99,7 @@ getResource_drifting <- function(world, par){
   colnames(R) <- x
   rownames(R) <- t
   R <- apply(R, 1, function(x) x/sum(x)) %>% t
+  attr(R, "par") <- par
   return(R)
 }
 
@@ -147,6 +148,6 @@ getResource_island <- function(world, par){
   colnames(R) <- x
   rownames(R) <- t
   R <- R/sum(R) * world$tau
-  
+  attr(R, "par") <- par
   return(R)
 }

### 2021-06-17 text/ms.Rmd
progress ....

@@
 
-Under what conditions is migration stable:
-
 sigma_t, sigma_x, 
@@
 x_peak = 30, t_peak = 25
+
+All variables contribute to the ability of the organism to find a stable, migratory state that matches the resource dynamics and maximizes foraging efficiency.
+
+- In general, greater values of $\alpha$ (resource following) lead to an improved ability to match the migration,
+
+- Resource peaks with larger spatial extent (high $\sigma_x$) are smaller duration (low $\sigma_t$) are generally better for migration matching, suggesting that broad but short pulses have the strongest impact on the migration memory.
+
+- Larger values of social attraction ($\beta$) generally lead to poorer migration matching except in those cases where the sociality scale $\lambda$ was high.
+Thus, for example, at $\lambda = 20$, no simulations at $\beta >= 200$ managed to acquire or maintain a matched migration.
+However, at $\lambda = 50$ or $100$, the migration was slightly better matched at high values of $\beta.
+(Fig.
+**X**)
+
+- Higher diffusion rate
 

### 2021-06-17 code/analysis/01_StabilityAnalysis.R

@@ -1,3 +1,5 @@
+require(memorymigration)
+
 a <- load("results/stability/stability_1.rda")
 eps1 <- newresults %>% mutate(TE = t1.error + t2.error + abs(x1.error) + abs(x2.error))
 load("results/stability/stability_4.rda")
@@ -38,3 +40,57 @@ ggResults(eps1, top = "epsilon = 1")
 ggResults(eps4, top = "epsilon = 4")
 ggResults(eps8, top = "epsilon = 8")
 dev.off()
+
+table(eps1$lambda)
+
+pdf("plots/StabilityResults_eps1_bylambda.pdf", height = 11, width = 8)
+ggResults(eps1 %>% subset(lambda == 20), top = "epsilon = 1, lambda = 20")
+ggResults(eps1 %>% subset(lambda == 50), top = "epsilon = 1, lambda = 50")
+ggResults(eps1 %>% subset(lambda == 100), top = "epsilon = 1, lambda = 100")
+dev.off()
+
+
+
+table(eps4$lambda)
+pdf("plots/StabilityResults_eps4_bylambda.pdf", height = 11, width = 8)
+ggResults(eps4 %>% subset(lambda == 20), top = "epsilon = 4, lambda = 20")
+ggResults(eps4 %>% subset(lambda == 50), top = "epsilon = 4, lambda = 50")
+#ggResults(eps1 %>% subset(lambda == 100), top = "epsilon = 1, lambda = 100")
+dev.off()
+
+
+hist(eps1$TE[eps1$TE < 7], breaks = 50)
+
+
+summary(TE.lm)
+
+require(broom)
+require("GGally")
+
+coefs.eps1 <- glm(I(TE < 2) ~ (scale(alpha) + 
+                                          scale(beta) + 
+                                          scale(sigma_t) + 
+                                          scale(sigma_x) + 
+                                          (factor(lambda) - 1))^2, 
+                           data = eps1 %>% subset(lambda < 100), family = "binomial") %>% 
+  tidy(conf.int = TRUE) 
+
+
+
+coefs.eps4 <- glm(I(TE < 2) ~ (scale(alpha) + 
+                                 scale(beta) + 
+                                 scale(sigma_t) + 
+                                 scale(sigma_x) + 
+                                 (factor(lambda) - 1))^2, 
+                  data = eps4, family = "binomial") %>% 
+  tidy(conf.int = TRUE) 
+
+
[…]

### 2021-06-17 text/figures/code/functions_plottingForMS.R
tweak

@@ -21,7 +21,7 @@ doublePlot <- function(M, world, par = FALSE){
   
   with(world, image(time, X, resource, col = grey.colors(100),
                     ylab = "space", xlab = "time of year"))
-  plotMigration(M, world, add = TRUE)
+  plotMigration(M, add = TRUE)
   FE1 <- ldply(M, computeEfficiency, 
                resource = world$resource, world = world,
                .id = "year") %>% mutate(year = 1:length(year))

### 2021-06-21 text/ms.Rmd
exploring ...

@@
 
+A random forest analysis indicated that the most important variables for attaining optimal migratory behavior were $\alpha$ and $\lambda$, and the least important, consistently, was $\sigma_t$ (table **X**).
+
 ## Adaptation
@@
 
+The model had, in general, a difficult time learning migration from a non-migratory initial condition.
+Curiously, the conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$)
+ but narrow spatial extent (low $\sigma_x$).  This was in contrast to the conditions that were most amenable for stable migratory behavior (high $\sigma_x$, intermediate $\sigma_t$).   Furthermore
+
 *Under what conditions can be learn.*
@@
 
-**Describing conditions for adaptability.**
+**Describing conditions for adaptability / stability.**
 

### 2021-06-21 code/analysis/01_StabilityAnalysis.R

@@ -1,5 +1,9 @@
 require(memorymigration)
 require(plyr)
+require(gplots)
+require(ggplot2)
+require(ggthemes)
+require(gridExtra)
 
 fixTE <- function(df){
   df %>% mutate(TE = t1.error + t2.error + abs(x1.error) + abs(x2.error), 
@@ -7,74 +11,64 @@ fixTE <- function(df){
          Migration = ifelse(is.na(Migration), "bad", Migration))
 }
 
-a <- load("results/stability/stability_1.rda")
+load("results/stability/stability_1.rda")
 eps1 <- newresults %>% fixTE
+load("results/stability/stability_1b.rda")
+eps1 <- rbind(eps1, newresults %>% fixTE)
+
 load("results/stability/stability_4.rda")
 eps4 <- newresults %>% fixTE
+load("results/stability/stability_4b.rda")
+eps4 <- rbind(eps4, newresults %>% fixTE)
+
 load("results/stability/stability_8.rda")
 eps8 <- newresults %>% fixTE
+load("results/stability/stability_8b.rda")
+eps8 <- rbind(eps8, newresults %>% fixTE)
 
 
 
-ggResults <- function(df, ...){
-  
-  require(gplots)
-  require(ggplot2)
-  require(ggthemes)
-  require(gridExtra)
-  
-  #cuts <- c(0, 2^(seq(-1,10,1)))
-  
+ggMigration <- function(df,  cols = c("darkblue", "orange", "darkgrey"), ...){
   df %>% 
     ggplot(aes(sigma_t, sigma_x, fill = factor(Migration))) + 
     facet_grid(beta~alpha) + 
-    #scale_fill_gradientn(trans = "log", breaks = cuts, colours = rich.colors(100)[100:1]) + 
-    scale_fill_manual(values = c("darkblue", "grey", "white")) + 
+    scale_fill_manual(values = cols) + 
     geom_tile() + theme_few() + 
     xlab("") + theme(legend.position = "none")
-
-  #  p2 <- df %>% subset(ss) %>% 
-  #   ggplot(aes(sigma_t, sigma_x, fill = n.runs)) + 
-  #    facet_grid(beta~alpha) + 
-  #    geom_tile() + theme_few() + ggtitle("N. runs")  + xlab("")
-    
-  #  p3 <- df %>% subset(ss) %>% 
-  #    ggplot(aes(sigma_t, sigma_x, fill = FE)) + 
[…]

### 2021-06-21 code/analysis/02_LearningMigrationAnalysis.R

@@ -3,31 +3,31 @@ require(plyr)
 
 fixTE <- function(df){
   df %>% mutate(TE = abs(t1.error) + abs(t2.error) + abs(x1.error) + abs(x2.error), 
-                Migration = cut(TE, c(0,1,5,10,1e3)))
+                Migration = cut(TE, c(1e3,32,8,2,0), label = LETTERS[1:4]) %>% 
+                  relevel("B") %>% relevel("C") %>% relevel("D"))
 }
+
+
 
 load("results/migratory/migratory_1.rda")
 eps1 <- newresults %>% fixTE
 load("results/migratory/migratory_4.rda")
 eps4 <- newresults %>% fixTE
+load("results/migratory/migratory_8.rda")
+eps8 <- newresults %>% fixTE
 
+require(gtools)
+df <- smartbind(eps1, eps4, eps8)
 
-
-ggResults <- function(df, ...){
-  
-  require(gplots)
-  require(ggplot2)
-  require(ggthemes)
-  require(gridExtra)
-
+ggMigration <- function(df,  cols = rich.colors(4), ...){
   df %>% 
     ggplot(aes(sigma_t, sigma_x, fill = factor(Migration))) + 
     facet_grid(beta~alpha) + 
-    #scale_fill_gradientn(trans = "log", breaks = cuts, colours = rich.colors(100)[100:1]) + 
-    scale_fill_manual(values = rich.colors(4)) + 
+    scale_fill_manual(values = cols) + 
     geom_tile() + theme_few() + 
-    xlab("") #+ theme(legend.position = "none")
+    xlab("") + theme(legend.position = "none")
 }
+
 
 
 ggN.runs <- function(df, ...){
@@ -41,28 +41,29 @@ ggN.runs <- function(df, ...){
     xlab("") + theme(legend.position = "none")
 }
 
-df <- smartbind(eps1, eps4)
 
-
-plist <- dlply(df, c("epsilon","lambda"), 
-               function(df) ggResults(df) + 
+migration.plist <- dlply(df, c("epsilon","lambda"), 
+               function(df) ggMigration(df) + 
                  ggtitle(paste("epsilon =", df$epsilon[1], ";", paste("lambda =", df$lambda[1]))))
 
 nruns.plist <- dlply(df, c("epsilon","lambda"), 
                      function(df) ggN.runs(df) + 
[…]

### 2021-06-23 text/figures/code/ResourceExamples.R
Resource example plot

@@ -1,0 +1,65 @@
+# Illustrating resources
+
+require(memorymigration)
+
+t.low <- 3
+x.low <- 3
+t.high <- 12
+x.high <- 15
+
+world <- getOptimalPop(tau = 100, x1 = 50, x2 = -50, 
+                       t.peak = 25, t.sd = 12, x.sd = 5)
+R1 <- getResource_island(world, 
+                   c(t.peak = 25, t.sd = t.low, 
+                     x.peak = 30, x.sd = x.high))
+R2 <- getResource_island(world, 
+                         c(t.peak = 25, t.sd = t.high, 
+                           x.peak = 30, x.sd = x.low))
+
+R3 <- getResource_drifting(world, 
+                         c(t.peak = 25, t.sd = t.low, 
+                           x.peak = 30, x.sd = x.high), x.null = 30)
+R4 <- getResource_drifting(world, 
+                         c(t.peak = 25, t.sd = t.high, 
+                           x.peak = 30, x.sd = x.low), x.null = 30)
+R.list <- list(R1, R2, R3,R4)
+R.max <-  R.list%>% sapply(max) %>% max
+
+b  <- seq(0, R.max, length = 101)
+cols <- rich.colors(100)
+  
+x <- world$X
+t <- world$time
+
+png("text/figures/ResourceExamples.png", width = 2000, height = 1200, res = 300)
+{
+  par(mfrow = c(2,2), mar = c(1,1,1,1), oma = c(2,2,1,1), 
+      tck = 0.01, mgp = c(1,.25,0), las = 1, xpd = NA)
+  
+  for(i in 1:4){ 
+    image(t,x,R.list[[i]], 
+          breaks = b, col = cols, xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+    
+    if(i == 1){
+      axis(2, at = seq(-80,80,40))
+      text(50, 80, col = "white", expression(sigma[t] == 3~";"~sigma[x]==12))
+      mtext(side = 2, line = 1.5, "space", las = 0)
+    }
+    if(i == 2){
+      text(50,80, col = "white", expression(sigma[t] == 12~";"~sigma[x]==3))
+      mtext(side = 3, at = -10, "isolated resource", font = 2, line = .25)
+    }
+    
+    if(i == 3){
+      axis(2, at = seq(-80,80,40))
+      mtext(side = 2, line = 1.5, "space", las = 0)
+    }
+    if(i %in% 3:4){
[…]

### 2021-06-23 text/figures/code/ResourceExamples.R
...

@@ -26,6 +26,7 @@ R.list <- list(R1, R2, R3,R4)
 R.max <-  R.list%>% sapply(max) %>% max
 
 b  <- seq(0, R.max, length = 101)
+require(viridis)
 cols <- rich.colors(100)
   
 x <- world$X

### 2021-06-23 memorymigration/R/RunningModel.R
updates to run on server

@@ -40,8 +40,8 @@ runManyYears <- function(world, parameters, n.years = 20,
     if(verbose){cat("\n"); cat(paste("running year ", i))}
     i <- i+1
     pop.list[[i]] <- runFWM(world = world, parameters = parameters, 
-                         pop.lastyear = pop.list[[length(pop.list)]], 
-                         Year = i, m.hat = memory.list[[i-1]])
+                            pop.lastyear = pop.list[[length(pop.list)]], 
+                            Year = i, m.hat = memory.list[[i-1]])
     
     # m.hat.lastyear  <- migration.list[[i-1]]
     if(abs(migration.list[[i-1]]["x2"] - migration.list[[i-1]]["x1"]) < 1){
@@ -51,7 +51,7 @@ runManyYears <- function(world, parameters, n.years = 20,
       lower = c(-100,0,0,0,-100,-100)
       
       m.hat <- try(fitMigration(t = world$time, x = getMem(pop.list[[i]], world), 
-                            m.start = migration.list[[i-1]]))
+                                m.start = migration.list[[i-1]]))
       if(inherits(m.hat, "try-error"))
         m.hat <- try(fitMigration(t = world$time, x = getMem(pop.list[[i]], world), 
                                   lower = c(-100,-20,0,0,-100,-100), 
@@ -61,13 +61,13 @@ runManyYears <- function(world, parameters, n.years = 20,
                                   lower = c(-100,-50,0,-20,-100,-100), 
                                   m.start = migration.list[[i-1]]))
       if(inherits(m.hat, "try-error")) stop(cat("\n We tried, and failed at year", i)) else
-          migration.list[[i]] <- m.hat
-        
+        migration.list[[i]] <- m.hat
+      
       migration.list[[i]] <- 
-      if(migration.list[[i]]["dt1"] < 0){
+        if(migration.list[[i]]["dt1"] < 0){
           migration.list[[i]]["dt1"] <- 0
           warning("\nnegative dt1 estimated!!!  coerced to 0")
-      }
+        }
       if(migration.list[[i]]["dt2"] < 0){
         migration.list[[i]]["dt2"] <- 0
         warning("\nnegative dt2 estimated!!!  coerced to 0")
@@ -277,7 +277,74 @@ runManyRuns_res <- function (world_param, parameters.df, resource_param, world, 
 #' 
 #' @export
 
-runMissedRuns <- function (world_param, parameters.df, resource_param, world, resource, 
+runMissedRuns <- function (parameters.df, resource_param, world, resource, 
[…]

### 2021-06-23 text/ms.Rmd
Resource example plot
...

@@
 This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).
+Both types of resources are illustrated in Figure 1.
 
@@
 
-sigma_t, sigma_x, 
-epsilon, alpha, 
-beta, lambda, 
-x_peak = 30, t_peak = 25
+An illustration of the ability to adjust its migration to find a stable, migratory state that matches the dynamics of the resource is illustrated in Figure X.
+Under many combinations of parameters, this kind of steady state adjustment is obtained and maintained.
 
@@
 
-- In general, greater values of $\alpha$ (resource following) lead to an improved ability to match the migration,
+In general, greater values of $\alpha$ (resource following) lead to an improved ability to match the migration,
 
-- Resource peaks with larger spatial extent (high $\sigma_x$) are smaller duration (low $\sigma_t$) are generally better for migration matching, suggesting that broad but short pulses have the strongest impact on the migration memory.
+Resource peaks with larger spatial extent (high $\sigma_x$) are smaller duration (low $\sigma_t$) are generally better for migration matching, suggesting that broad but short pulses have the strongest impact on the migration memory.
 
-- Larger values of social attraction ($\beta$) generally lead to poorer migration matching except in those cases where the sociality scale $\lambda$ was high.
+Larger values of social attraction ($\beta$) generally lead to poorer migration matching except in those cases where the sociality scale $\lambda$ was high.
 Thus, for example, at $\lambda = 20$, no simulations at $\beta >= 200$ managed to acquire or maintain a matched migration.
@@
 
-- Higher diffusion rate
+Higher diffusion rates
 

### 2021-06-23 memorymigration/R/RunningModel.R

@@ -48,22 +48,24 @@ runManyYears <- function(world, parameters, n.years = 20,
       migration.list[[i]] <- migration.list[[i-1]]
     } else{
       
-      lower = c(-100,0,0,0,-100,-100)
-      
       m.hat <- try(fitMigration(t = world$time, x = getMem(pop.list[[i]], world), 
                             m.start = migration.list[[i-1]]))
-      if(inherits(m.hat, "try-error"))
+      if(inherits(m.hat, "try-error")){
+        cat("\nRetry migration fitting 1")
         m.hat <- try(fitMigration(t = world$time, x = getMem(pop.list[[i]], world), 
                                   lower = c(-100,-20,0,0,-100,-100), 
                                   m.start = migration.list[[i-1]]))
-      if(inherits(m.hat, "try-error"))
+      }
+      if(inherits(m.hat, "try-error")){
+        cat("\nRetry migration fitting 2")
         m.hat <- try(fitMigration(t = world$time, x = getMem(pop.list[[i]], world), 
                                   lower = c(-100,-50,0,-20,-100,-100), 
                                   m.start = migration.list[[i-1]]))
+      }
       if(inherits(m.hat, "try-error")) stop(cat("\n We tried, and failed at year", i)) else
           migration.list[[i]] <- m.hat
-        
-      migration.list[[i]] <- 
+
+      
       if(migration.list[[i]]["dt1"] < 0){
           migration.list[[i]]["dt1"] <- 0
           warning("\nnegative dt1 estimated!!!  coerced to 0")

### 2021-06-23 text/ms_v2.Rmd
ms ms ms ms ms

--- 2021-06-21 text/ms.Rmd
@@
 
+![](figures/ResourceExamples.png)
+**Figure 1.
+Examples of various seasonal resource distribution functions, contrasting short duration, but wide pulses ($\sigma_t = 3, \sigma_x = 12$; left panels), long duration but spatially concentrated pulses ($\sigma_t = 12, \sigma_x = 3$; right panels), and isolated resource pulses (upper panels) from the weakly drifting resource (lower panels).
+The total amount of resource is identical across all scenarios.
+In the weakly drifting resources, the total amount is constant at all times, and uniform in the middle of the phase (time = 0, 50, 100).**
+
 ## Metrics
@@
 
-- Higher diffusion rate
+- Higher diffusion rates
 
@@
 ![](figures/example1_adaptation.png)
-**Figure 1.
+**Figure 2.
 Example of adaptation to a shift in resource peak.

### 2021-06-25 text/ms.Rmd
more progress on ms

--- 2021-06-17 text/ms.Rmd
@@
 A "non-surfable" resource (*island resource*), and weakly surfable resource (*drifting resource*).
-Both are characterized by a peak in time and space centered at $M_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively).
+Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively).
 These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.
@@
 
+![](figures/ResourceExamples.png)
+**Figure 1.
+Examples of various seasonal resource distribution functions, contrasting short duration, but wide pulses ($\sigma_t = 3, \sigma_x = 12$; left panels), long duration but spatially concentrated pulses ($\sigma_t = 12, \sigma_x = 3$; right panels), and isolated resource pulses (upper panels) from the weakly drifting resource (lower panels).
+The total amount of resource is identical across all scenarios.
+In the weakly drifting resources, the total amount is constant at all times, and uniform in the middle of the phase (time = 0, 50, 100).**
+
 ## Metrics
 
-The main metrics we are interested are *migratoriness* and *foraging efficiency*.
-The migratoriness is measured as the maximum distance between the median of the population distribution within a given year.
-The median is a more reliable metric than the mean, as even as distributions become highly assymetric - pulling the mean in one directon or another - the median can  remain consistently in the middle of the spatial domain indicating a net flux of animals of 0.
-This is the anticipated (and realized) result of a purely "tactical" model pursuing the periodic resource pulses.
+The main metrics we are interested are *migration mismatch* and *foraging efficiency*.
+
+Migration mismatch captures the similarity between the migration phenology and the resource phenology.
+To do this, we compute the total mismatch, i.e.
+is a sum of the difference between the migration coefficients and the resource peaks.
+Thus, the spatial mismatch is given by the absolute difference between the migration targets and the resource peaks, i.e.
+$MM_x = |x_1 - m_x| + |x_2 + m_x|$.
+The temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.
+$MM_t = max\{t_1 - m_1, m_1 - (t_1 + dt_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + dt_2), 0\}$.
+The total mismatch is the sum of $MM_x$ and $MM_t$.
+If the total mismatch is less than 1, we consider the migration to be perfect.
+If it is between 1 and 5, we consider it to be good.
+Otherwise, we consider the migration to be mismatched.
 
@@
 
-## Migration
+![](figures/example1_adaptation.png)
+**Figure 2.** *Example of adaptation to a shift in resource peak.
+The initial (year 0) behavior migrates to locations 50 and -50 at days 15 and 60, whereas the resource peak is at 30 and -30, peaking at times 25 and 75.
+(a) shows the first 14 years of the simulation; (b) The centroid of the annual movement of the population is shown in panel b, with dark blue to red colors indicating year 0 to year 40; (c) annual foraging efficiency across years; (d) migration timing parameters for each year, with orange segments indicating arrival and departure from the summering (northern) grounds, and the blue segments indicating timing of arrival and departure at the wintering grounds; (e) migration arrival and departure location across years, with blue and orange indicating winter (southern) and summer (northern) locations.*
 
-sigma_t, sigma_x, 
-epsilon, alpha, 
-beta, lambda, 
-x_peak = 30, t_peak = 25
+## Adaptation to resource phenology
 
-All variables contribute to the ability of the organism to find a stable, migratory state that matches the resource dynamics and maximizes foraging efficiency.
+The ability of this system to attain stable, migratory state that matches the dynamics of the resource is illustrated in Figure 2.
+In the illustrated scenario, it takes nearly 40 years to attain an equilibrium, and the eventual steady state is one where the centroid of the migration lines up exactly with the centroid of the resource, and the arrival timing coincides with the *peak* of resource availability.
+Notably, the path to this equilibrium is somewhat indirect, with the winter (later, more southerly) range taking more time to stabilize than the summer (earlier, more northern) range.
+The eventual steady state is one where the foraging efficiency is relatively high.
 
-- In general, greater values of $\alpha$ (resource following) lead to an improved ability to match the migration,
+We ran this process for 8100 parameter combinations (Figure 3).
+The fundamental dynamic of attaining and maintaining a well-matched migration phenology, can occur under many combinations of parameter values, but that all parameters play a role.
+Among the more intuitive results are that greater values of $\alpha$ (resource following) lead to an improved ability to match the migration.
+Resource peaks with larger spatial extent (higher $\sigma_x$) are generally better for migration matching.
+A larger set of parameters matched migration with low diffusion than high diffusion.
 
-- Resource peaks with larger spatial extent (high $\sigma_x$) are smaller duration (low $\sigma_t$) are generally better for migration matching, suggesting that broad but short pulses have the strongest impact on the migration memory.
+Less intuitive was the high importance of the sociality parameters, in particular the spatial scale of the swarming.
+Higher levels of social attraction ($\beta$) led to improved migration matching except in those cases where the sociality scale $\lambda$ was high.
+Thus, for example, at $\lambda = 20$, no simulations at $\beta >= 200$ managed to acquire or maintain a matched migration.
+However, at $\lambda = 50$ or $100$, the migration was slightly better matched at high values of $\beta$.
+(Fig.
+**X**)  The spatial extent of the swarm was a remarkably significant variable.
+Smaller swarms were able to match migration only at low values of social attraction ($\beta = 200$), and relatively high values of resource attraction ($\alpha \geq 600$).
 
-- Larger values of social attraction ($\beta$) generally lead to poorer migration matching except in those cases where the sociality scale $\lambda$ was high.
-Thus, for example, at $\lambda = 20$, no simulations at $\beta >= 200$ managed to acquire or maintain a matched migration.
-However, at $\lambda = 50$ or $100$, the migration was slightly better matched at high values of $\beta.
-(Fig.
-**X**)
+Random forest analyses, whether on the log of total mismatch or on the classification of a perfect match, uniformly show that the most important variables [@Breiman2001] were $\alpha$ and $\lambda$ (4.14 and 4.02 proportional increase in MSE), and the least important was $\sigma_t$, with a 0.5 proportional increase (table **X**).
 
-- Higher diffusion rate
+Overall, foraging efficiency was strongly correlated with migration matching, as expected.
+At high mismatch (> 50), foraging efficiency was low (mean 0.29, s.d.
+0.16) compared to the near-perfect matching migrations (mean 0.58, s.d.
+0.14).
+However, somewhat higher mismatch (1 to 5) showed an even higher overall foraging efficiency (mean 0.62, s.d.
+0.18 - see also Figure **X**).
 
-## Adaptation
+```{r, echo = FALSE, eval = FALSE}
+load("../results/stability/stability_compiled.rda")
+require(plyr)
+df <- stability %>% mutate(mismatch = cut(TE, c(0,1,5,50,300)))
+ddply(df, "mismatch", summarize, mean(FE), sd(FE))
+```
 
-![](figures/example1_adaptation.png)
-**Figure 1.
-Example of adaptation to a shift in resource peak.
-The initial (year 0) behavior migrates to locations 50 and -50 at days 15 and 60, whereas the resource peak is at 30 and -30, peaking at times 25 and 75.
-(a) shows the first 14 years of the simulation; (b) The centroid of the annual movement of the population is shown in panel b, with dark blue to red colors indicating year 0 to year 40; (c) annual foraging efficiency across years; (d) migration timing parameters for each year, with orange segments indicating arrival and departure from the summering (northern) grounds, and the blue segments indicating timing of arrival and departure at the wintering grounds; (e) migration arrival and departure location across years, with blue and orange indicating winter (southern) and summer (northern) locations.**
+![](figures/StabilityResults.png)
+**Figure 3.** *Migration phenology matching across six model parameters.
+Low and high diffusion ($\epsilon = 1$ and $8$ in upper and lower panel blocks), tight, medium and loose swarms ($\lambda = 20, 50, 100$) left to right panels.
+Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.
+Finally, within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right.
+Finally, the color scheme reflects the total mismatch, i.e.
+the sum of the absolute differences between the migration timing and locations from the resource peak.*
 
-We illustrate an example of a population with a particular migration behavior confronted with a major sudden shift in the spatio-temporal distribution of the resource in Figure 1.
-In the illustrated scenario, it takes nearly 40 years to attain an equilibrium, and the eventual "optimum" is one where the centroid of the migration lines up exactly with the centroid of the resource, and the arrival timing conincides with the *peak* of resource availability.
-Notably, the path to this equilibrium is somewhat indirect, with the summer season, in particular, taking
+![](figures/ForagingEfficiency.png)
 
-*Under what conditions can we adapt.*>
+**Figure 4.** *Box-plots of foraging efficiency against mismatch across several values of foraging patch duration.*
 
@@
 ![](figures/example2_learningtomigrate.png)
-**Figure 2.
-Example of model learning to migrate.
+**Figure X.** *Example of model learning to migrate.
 The resource is a "weakly drifting" resource and the initial (year 0) condition is non migratory.
@@
 Otherwise, panels are as in Figure 1.
-Parameter values were $\epsilon  = 5$, $\alpha = 500$ and $\beta = 50$.**
+Parameter values were $\epsilon  = 5$, $\alpha = 500$ and $\beta = 50$.*
 
-Figure 2 illustrates the ability of the model was able to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 50 and -50, but a uniform distribution of resource at times 0 and 50.
+Figure **X** illustrates the ability of the model was able to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 50 and -50, but a uniform distribution of resource at times 0 and 50.
 In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
@@
 
-*Under what conditions can be learn.*
+The model had, in general, a difficult time learning migration from a non-migratory initial condition.
+Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5 (Appendix Figure **Z**).
+Conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$), but *smaller* in scope (low $\sigma_x$).
+but narrow spatial extent (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact but long enough in duration to lock in to the memory.
+
+```{r, echo = FALSE, eval = FALSE}
+load("../results/migratory/learningmigration.rda")
+df <- learningmigration
+table(df$TE %>% cut(c(0,1,5,50,300)))
+nrow(df)
+```
 

### 2021-06-25 code/analysis/01_StabilityAnalysis.R

@@ -47,7 +47,6 @@ ggMigration <- function(df,  cols = c("darkblue", "orange", "darkgrey"), ...){
     xlab("") + theme(legend.position = "none")
 }
 
-
 ggFE <- function(df, ...){
   df %>% 
     ggplot(aes(sigma_t, sigma_x, fill = FE)) + 
@@ -57,153 +56,10 @@ ggFE <- function(df, ...){
     xlab("") + theme(legend.position = "none")
 }
 
-plotBlock <- function(block, cols = c("darkblue", "orange", "lightgrey"), 
-                      side1 = FALSE, side2 = FALSE){
-  require(reshape2)
-  b <- block[,c("TE", "alpha", "beta", "sigma_t", "sigma_x")]
-  ns <- apply(b[,-1], 2, function(x) length(unique(x)))
-  
-  M <-   b %>% acast(alpha + sigma_t ~ beta +  sigma_x, 
-                     value.var = "TE")
-  
-  #M[is.na(M)] <- max(M, na.rm = TRUE)
-  image(1:ncol(M)-.5, 1:nrow(M)-.5, M, breaks = c(0,1,5,1000), 
-        col = cols, xaxt ="n", xlab = "", ylab = "", yaxt = "n")
-  abline(v = 1:ncol(M), col = "white")
-  abline(h = 1:nrow(M), col = "white")
-  abline(v = (0:ns["alpha"]) * ns["sigma_t"], lwd = 2)
-  abline(h = (0:ns["beta"]) * ns["sigma_x"], lwd = 2)
-  
-  if(side1)
-    text(1:ns["alpha"] * ns["sigma_t"] - ns["sigma_t"]/2, -1, 
-        unique(b$alpha), xpd = NA)
-  
-  if(side2)
-    text(nrow(M)+ 1, 
-        1:ns["beta"] * ns["sigma_x"] - ns["sigma_x"]/2,
-        unique(b$beta), xpd = NA, srt = 270)
-}
-
-
-
-PlotMiniblock <- function(cols = c("darkblue", "orange", "lightgrey")){
-  miniM <- subset(df, 
-                  lambda == 100 & epsilon == 1 & 
-                    beta == 400 & alpha == 200)[,c("sigma_t", "sigma_x", "TE")] %>% 
-    acast(sigma_t ~ sigma_x)
-  
-  sigs <- seq(3,15,3)
[…]

### 2021-06-25 code/analysis/02_LearningMigrationAnalysis.R

@@ -1,5 +1,8 @@
 require(memorymigration)
 require(plyr)
+require(ggthemes)
+require(gridExtra)
+
 
 fixTE <- function(df){
   df %>% mutate(TE = abs(t1.error) + abs(t2.error) + abs(x1.error) + abs(x2.error), 
@@ -8,16 +11,30 @@ fixTE <- function(df){
 }
 
 
+eval <- FALSE
+if(eval){
+  require(gtools)
+  
+  load("results/migratory/migratory_1.rda")
+  eps1 <- newresults %>% fixTE
+  load("results/migratory/migratory_1b.rda")
+  eps1 <- eps1 %>% smartbind(newresults %>% fixTE)
+  
+  load("results/migratory/migratory_4.rda")
+  eps4 <- newresults %>% fixTE
+  load("results/migratory/migratory_4b.rda")
+  eps4 <- eps4 %>% smartbind(newresults %>% fixTE)
+  
+  load("results/migratory/migratory_8.rda")
+  eps8 <- newresults %>% fixTE
+  load("results/migratory/migratory_8b.rda")
+  eps8 <- eps8 %>% smartbind(newresults %>% fixTE)
 
-load("results/migratory/migratory_1.rda")
-eps1 <- newresults %>% fixTE
-load("results/migratory/migratory_4.rda")
-eps4 <- newresults %>% fixTE
-load("results/migratory/migratory_8.rda")
-eps8 <- newresults %>% fixTE
+  learningmigration <- rbind(eps1, eps4, eps8)  
+  save(learningmigration, file = "results/migratory/learningmigration.rda")
+}
 
-require(gtools)
-df <- smartbind(eps1, eps4, eps8)
+load("results/migratory/learningmigration.rda")
 
 ggMigration <- function(df,  cols = rich.colors(4), ...){
   df %>% 
@@ -42,11 +59,11 @@ ggN.runs <- function(df, ...){
 }
 
 
-migration.plist <- dlply(df, c("epsilon","lambda"), 
+migration.plist <- dlply(learningmigration, c("epsilon","lambda"), 
                function(df) ggMigration(df) + 
                  ggtitle(paste("epsilon =", df$epsilon[1], ";", paste("lambda =", df$lambda[1]))))
 
-nruns.plist <- dlply(df, c("epsilon","lambda"), 
+nruns.plist <- dlply(learningmigration, c("epsilon","lambda"), 
                      function(df) ggN.runs(df) + 
[…]

### 2021-06-25 text/figures/code/ResourceExamples.R

@@ -39,7 +39,7 @@ png("text/figures/ResourceExamples.png", width = 2000, height = 1200, res = 300)
   
   for(i in 1:4){ 
     image(t,x,R.list[[i]], 
-          breaks = b, col = cols, xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+          col = cols, xaxt = "n", yaxt = "n", xlab = "", ylab = "")
     
     if(i == 1){
       axis(2, at = seq(-80,80,40))

### 2021-06-25 text/ms.Rmd
starting in on climate change

@@
 
-Figure **X** illustrates the ability of the model was able to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 50 and -50, but a uniform distribution of resource at times 0 and 50.
+Figure **X** illustrates the ability of the model to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 50 and -50, but a uniform distribution of resource at times 0 and 50.
 In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).

### 2021-06-25 code/analysis/02_LearningMigrationAnalysis.R

@@ -126,13 +126,14 @@ varImpPlot(learning.rf)
 require(rpart)
 
 load("results/migratory/learningmigration.rda")
+require(magrittr)
+require(plyr)
+df <- learningmigration %>% mutate(mismatch = factor(TE<20))
 
+features <- c("epsilon","alpha","beta","lambda","sigma_x","sigma_t", "mismatch")
 
-
-features <- c("epsilon","alpha","beta","lambda","sigma_x","sigma_t", "Migration")
-
-learning.dt <- rpart(Migration ~ ., 
-                      data=learningmigration[,features] %>% mutate(epsilon = factor(epsilon),
+learning.dt <- rpart(mismatch ~ ., 
+                      data=df[,features] %>% mutate(epsilon = factor(epsilon),
                                                     lambda = factor(lambda)))
 plot(learning.dt)
 text(learning.dt, use.n = TRUE)

### 2021-06-25 code/analysis/03_ClimateChangeTimeAnalysis.R

@@ -1,0 +1,48 @@
+require(memorymigration)
+require(plyr)
+require(ggthemes)
+require(gridExtra)
+
+
+fixTE <- function(df){
+  df %>% mutate(TE = abs(t1.error) + abs(t2.error) + abs(x1.error) + abs(x2.error), 
+                Migration = cut(TE, c(1e3,32,8,2,0), label = LETTERS[1:4]) %>% 
+                  relevel("B") %>% relevel("C") %>% relevel("D"))
+}
+
+
+eval <- FALSE
+if(eval){
+  require(gtools)
+  
+  files <- list.files("results/climate_time/raw/")
+  
+  cctime <- data.frame()
+  for(f in files[grepl("rda", files)]){ 
+    print(f)
+   load(paste0("results/climate_time/raw/", f))
+   cctime <- cctime %>% smartbind(newresults %>% fixTE)
+  }
+  
+  save(cctime, file = "results/climate_time/cctime.rda")
+}
+
+load("results/climate_time/cctime.rda")
+
+
+with(cctime, table(epsilon, beta_t, lambda))
+head(cctime)
+
+require(ggplot2)
+cctime %>% subset(epsilon > 4 & beta_t == -1) %>% 
+  ggplot(aes(alpha, beta, fill = cut(FE, seq(0,1,.2)))) + 
+  geom_tile() + facet_grid(epsilon~lambda~beta_t) + ggtitle("Foraging efficiency")
+
+
+cctime %>% subset(epsilon > 4 & beta_t == -1) %>% 
+  ggplot(aes(alpha, beta, fill = cut(TE, seq(0,750,150)))) + 
+  geom_tile() + facet_grid(epsilon~lambda~beta_t) + ggtitle("Foraging efficiency")
+
+ 
+cctime %>% subset(epsilon > 4 & beta_t == -1) %>% 
+  ggplot(aes(FE, TE, col = alpha)) + geom_point()  + facet_grid(.~lambda)

### 2021-07-07 text/ms.Rmd
progress

@@
 output:
+  word_document:
   html_document:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, eval = TRUE)
 ```
@@
 On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility [@Robinson2009].
-Clearly, the ability to perform a migration without local cues in only possible if the behavior is hard-cued or remembered.
+Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered.
 On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
@@
 In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from a non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon (**refs**).
-Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing patterns in resource distribution that are the consequences of rapid climate change.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon (**refs**).
+Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing patterns in resource distribution, whether inter-annual variability or trends in resource distribution or phenology reflecting climate change.
 
-We anticipate that under many conditions a blending of *tactical* (i.e.
+We further anticipate that under many conditions a blending of *tactical* (i.e.
 direct response to resource availability or perception) and *strategic* (i.e.
@@
 Over-reliance on either strategy should be maladaptive.
-We further anticipate that a a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are increasingly stochastic.
+We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic.
 
@@
 A "non-surfable" resource (*island resource*), and weakly surfable resource (*drifting resource*).
-Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 40 and -40 at times 25 and 75, respectively).
+Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 30 and -30 at times 25 and 75, respectively).
 These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.
@@
 
-The main metrics we are interested are *migration mismatch* and *foraging efficiency*.
+The main metrics we are interested in are *migration mismatch*, *foraging efficiency* and *adaptation to trends*.
 
@@
 The total mismatch is the sum of $MM_x$ and $MM_t$.
-If the total mismatch is less than 1, we consider the migration to be perfect.
-If it is between 1 and 5, we consider it to be good.
-Otherwise, we consider the migration to be mismatched.
+A mismatch of less than 1 is essentially perfect, we consider a mismatch of 1-5 to be "good", and beyond 30 the system can be said to have failed to keep up with climate change.
 
@@
 Thus, the foraging efficiency index is:
-$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
+$$FE = {1\over \tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
 where the spatial integral is taken over the domain.  This metric is constrained to be between 0 and 1.
@@
 where the Bhattacharya index of the population distribution across subsequent years reaches a value of 0.9999.
-Once that state is attained, we compute the mirgatoriness and foraging efficiency metrics, and additionally collect the number of years that were required to reach stationarity.
+Once that state is attained, we compute the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.
 
-For numerical runs with climate change, we run the simulation for 30 years, since equilibrium is never achieved) and retain the foraging efficiency for the final 10 years, reflecting something like the dynamic long-term average of the system's ability to keep track of climate change.
-We also record the migration parameters $\widehat{\theta_y}$ for each year.
+For numerical runs with climate change, we first run a simulation for 20 years to attain an equilibrium, and then begin shifting the location of the resource with a trend (slow 0.25, medium 0.5, or rapid 1 units per year), or by adding stochasticity (s.d.
+3, 6, 12).
+For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this index as the *spatial adaptation* (SA) index.
 
@@
 
-We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the three main questions:
+We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions:
 
-1. Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality?
+1. Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation?
 
-2. Can this model "acquire" a migratory behavior from a non-migratory initial condition?
+2. Can this model acquire a migratory behavior from a non-migratory initial condition?
 
-3. Can this model adapt when the resource peak shifts in space and time?  What is the role of the reference memory.
+3. What is the role of a reference memory for dealing with stochastic resource dynamics?
+
+4. Can this model adapt when the resource peaks shifts in space?
 
@@
 We ran this process for 8100 parameter combinations (Figure 3).
-The fundamental dynamic of attaining and maintaining a well-matched migration phenology, can occur under many combinations of parameter values, but that all parameters play a role.
+The fundamental dynamic of attaining and maintaining a well-matched migration phenology, can occur under many combinations of parameter values, but all parameters play a role.
 Among the more intuitive results are that greater values of $\alpha$ (resource following) lead to an improved ability to match the migration.
@@
 Finally, the color scheme reflects the total mismatch, i.e.
-the sum of the absolute differences between the migration timing and locations from the resource peak.*
+the sum of the absolute differences between the migration timing and locations from the resource peak.
+White squares represent runs that numerically failed to estimate migration parameters.*
 
@@
 
-Figure **X** illustrates the ability of the model to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 50 and -50, but a uniform distribution of resource at times 0 and 50.
+Figure **X** illustrates the ability of the model to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.
 In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
@@
 Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5 (Appendix Figure **Z**).
-Conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$), but *smaller* in scope (low $\sigma_x$).
-but narrow spatial extent (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact but long enough in duration to lock in to the memory.
+Conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$), but *smaller* in scope (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact in space but long enough in duration to lock in to the memory.
 
@@
 
+![](figures/SpatialClimateChange.png)
+
+**Figure Z.** *Adaptation to a steadily drifting resource.
+In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).
+The y-axis is the index of spatial adaptiation (SA), i.e.
+the trend of the memory-driven migration divided by the resource drift trend.
+Values near 1 indicate a behavior that keeps up with climate change, values near indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend.
+We compare across varying spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels).
+The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations.
+Finally, the bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicte the highest values, orage and grey lower values.*
+
+To assess the ability of the system to adapt to a trending climate, we generated scenarios with slow (0.25 units / year), medium (0.5 units / year), and fast (1 unit / year) drift outward of the two resource pulses.
+We then assessed 40 parameter combinations for each of those scenarios, a high and low value of resource following ($\alpha = 400$ and $100$), a high and low value of sociality ($\beta = 400$ and $100$), and 10 values of the spatial scale of sociality ($\lambda = 20$ to $200$).
+The spatial and temporal scale of the resource pulses were fixed to a relatively "easy" to adapt to $\sigma_x = 12$ and $\sigma_t = 6$.
+We computed the adaptation index and foraging efficiency for each of the 120 runs (figure **Z**).
+We were interested in the dynamics against $\lambda$ due to the uniformly high importance of this parameter for determining the success of this process to match migration in steady states.
+
+As figure Z shows, higher values of resource following ($\alpha = 400$; orange circles) are nearly universally better for keeping up with climate change (SA values near 1).
+Furthermore, when combined with high sociality ($\beta = 400$; right panels), nearly all parameter combinations do a good job keeping up with climate change (SA values ranging between 0.53 and 0.85 for a swarm size greater than 50).
+However, that maximum value is still less than 1, suggesting that truly matching a steadily drifting trend is very difficult.
+Small swarms ($\lambda < 50$) have a very hard time adapting when the social attraction is high, but do fairly well when social attraction is lower.
+But larger sized swarms do progressively worse across more parameterizations, e.g.
+in the most rapid climate change scenario, the SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to 200 (essentially, the entire spatial domain).
+
+A rather more dramatic pattern is visible for the lower foraging attraction scenario ($\alpha = 100$; blue circles).
+Notably, no parameter combination at this value can remotely keep up with the rapid climate change (SA range -0.64 to 0.13).
+For slower climate change, however, there is a window of values for the swarm size - relatively small, between 40 and 80, where the SA *exceeds* 1, and then crashing quite rapidly to negative values of SA as that swarm size increases.
+These "super-adapting" processes indicate a unique sweet spot where a swarm is large enough to capture and adapt to the drifting resource, but not so large that the information gathered in a given year is too weak to adjust the migratory behavior in a following year.
+
+As anticipated, better adaptation to the drifting resource correlated strongly with higher foraging efficiency (inset boxplots).
+
+```{r, echo = FALSE, eval = FALSE}
+load("../results/climate_space/ccspace.rda")
+nrow(ccspace)
+unique(ccspace$lambda)
+(ccspace %>% subset(alpha == 400& beta == 400 & lambda > 50))$SAI_recent %>% range
+(ccspace %>% subset(alpha == 400& beta == 100 & lambda >= 40))$SAI_recent %>% range
+(ccspace %>% subset(alpha == 100 & beta_x == -1))$SAI_recent %>% range
+```
+
+## Role of reference memory in adapting to stochasticity
+
+*This section still rough*
+
+We ran simulations with various levels of stochasticity (inter-annual standard deviation in location of resource pulse 3, 6, and 12) notably against various values of reference memory $\kappa$, where $\kappa = 0$ refers to only recent memory, and $\kappa = 1$ refers to all reference memory, i.e.
+that never updates.
+We ran simulations for 20 years with constant resource to stabilize, and then for 30 more years to let the process evolve.
+In all cases, unsurprisingly, the conservative $\kappa = 1$ limit yielded the highest interannual foraging efficiency (see figure ZZ below).
+
+```{r KappaPlot1, cache  = TRUE, message = FALSE, warning = FALSE}
+load("../results/stochasticity/stochasticity.rda")
+require(magrittr)
+require(plyr)
+require(ggplot2)
+p1 <- list()
+for(a in c(100,400)) for(b in c(100,400)){
+  p1[[length(p1)+1]] <- ccsigma %>% subset(alpha == a & beta == b & lambda > 40) %>% 
+    ggplot(aes(kappa, avgFE, col = factor(lambda))) +
+    facet_grid(.~psi_x) + geom_path() + theme_few() + ggtitle(bquote(alpha == .(a) ~ beta == .(b))) + 
+    ylim(c(0.5,0.75))
+}
+print(p1[[1]])
+```
+**Figure ZZ.** (*to be refined*) *Results of foraging efficiency against kappa.
+Higher stochasticity leads to  generally lower efficiency.
+Higher use of reference memory leads to higher resource efficiency as well.*
+
+The final question (simulations currently runnning) is for a system to deal with BOTH trend AND stochasticity, does an intermediate blending of recent and reference memory improve the outcome?
+
 # Discussion
 
-We found that the relatively straightforward inclusion of a migration specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
+*Highly sketchy and incomplete.  Some odd thoughts here and there.*
+
+We found that the relatively straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
 In essense, our model allowed a population of individuals with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location.
@@
 
-However, the ability to migrate was highly dependent on the nature of the resource dynamics.
+Note, we are interested here entirely in mechanisms that do *not* correspond to the "green surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.
+
+The ability to migrate was highly dependent on the nature of the resource dynamics.
 In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
-**ALSO SOME STUFF WE LEARN ABOUT THE INTERACTION BETWEEN DIFFUSION, RESOURCE FOLLOWING, SOCIALITY**
+Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than, e.g, the beginning of the resource dynamic.
+Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is very much the case for the rapid green-up that occurs in high latitudes as the snow recedes even as day length is high, leading to a rapid and intense green-up.
 
-It is worth considering the assumptions of our model with respect to a biological population.  Notably, the fact that the memory is not easily interpreted as an
-
-**Describing conditions for adaptability / stability.**
-
-**Describing conditions for learning migration.**
-
-**Describing conditions for climate change resilience.**
+It is worth considering the assumptions of our model with respect to a biological population.
+The population is explicitly social and collective, and the memory is a memory that belongs to the entire population.
+The importance of the spatial scale of the collective was an unexpected and very strong result, suggesting that for a population to adequately learn socially, it must have a large enough exploratory radius to find new information to adapt to, while being small enough that the entire population can then learn from this.
 
@@
 
+*Some comments on systems that may or may not be more resilient to climate change based on our results*.
+
 # References

### 2021-07-07 code/analysis/04_ClimateChangeSpaceAnalysis.R

@@ -11,6 +11,8 @@ fixTE <- function(df){
 }
 
 
+#  Climate change trend --------
+
 eval <- FALSE
 if(eval){
   require(gtools)
@@ -30,24 +32,142 @@ if(eval){
 load("results/climate_space/ccspace.rda")
 
 
-with(ccspace, table(epsilon, beta_x, lambda))
+with(ccspace, table(beta_x, lambda, alpha, beta))
 head(ccspace)
 
 require(ggplot2)
-ccspace %>% subset(beta_x == -1) %>% 
-  ggplot(aes(alpha, beta, fill = cut(avgFE, seq(0,1,.2)))) + 
-  geom_tile() + facet_grid(epsilon~lambda) + ggtitle("Foraging efficiency")
+ccspace %>% 
+  ggplot(aes(lambda, SAI_recent, col = factor(alpha))) +
+             # beta, fill = cut(SAI_recent, c(-3,-1,0,.5,.7,.8,.9,1,10)))) + geom_tile() 
+  facet_grid(beta_x~beta) + geom_path() + geom_point(aes(alpha = FE), size = 3) + theme_few()
+
+ccspace %>% 
+  ggplot(aes(lambda,  FE, col = factor(alpha))) +
+  facet_grid(beta_x~beta) + geom_path() + geom_point() + theme_few()
 
 
-ccspace %>% subset(beta_x == -1) %>% 
-  ggplot(aes(alpha, beta, fill = cut(n.runs, c(0,10,20,40,80,160)))) + 
-  geom_tile() + facet_grid(epsilon~lambda) + ggtitle("N.runs")
+#  Climate change stochasticity --------
+
+eval <- FALSE
+if(eval){
+  require(gtools)
+  
+  files <- list.files("results/stochasticity/raw/")
+  
+  ccsigma <- data.frame()
+  for(f in files[grepl("rda", files)]){ 
+    print(f)
+    load(paste0("results/stochasticity/raw/", f))
+    ccsigma <- ccsigma %>% smartbind(newresults %>% fixTE)
+  }
+  
+  save(ccsigma, file = "results/stochasticity/stochasticity.rda")
+}
+
+load("results/stochasticity/stochasticity.rda")
+
+with(ccsigma, table(alpha, beta, lambda, kappa, psi_x))
+
+require(ggplot2)
+ccsigma %>% subset(alpha == 100) %>% 
[…]

### 2021-07-07 text/figures/code/ClimateChangeSpace.R

@@ -1,0 +1,104 @@
+
+plotCCspacepanel <- function(df, cols = c("darkblue", "darkorange")){
+  with(df, {
+    plot(lambda, SAI_recent, type = "n", ylab = "", xlab = "", 
+         ylim = c(-2.3,2), xlim = c(0,200), xaxt = "n", yaxt = "n")
+    abline(h = c(0,1), lty = 1:2, col = "grey", lwd = 2)
+    points(lambda[alpha == 100], SAI_recent[alpha == 100], 
+           col = cols[1], bg = alpha(cols[1], .4),
+           type = "o", pch = 21, cex = FE[alpha == 100]*4)
+    points(lambda[alpha == 400], SAI_recent[alpha == 400], 
+           col = cols[2], bg = alpha(cols[2], .4), 
+           type = "o", pch = 21, cex = FE[alpha == 400]*4)
+  })
+}
+
+load("results/climate_space/ccspace.rda")
+
+boplotFE <- function(){
+  cols <- c("grey", "orange", "purple", "darkblue")
+  par(mar = c(0,0,2,0), bty = "l", tck = 0.02, 
+      mgp = c(1,.25,0), las = 1)
+  with(ccspace %>% subset(alpha == 100), {
+    boxplot(avgFE ~ cut(SAI_recent, c(-1,0,.5,1,10)), varwidth = TRUE, 
+            border = cols, col = alpha(cols, .5), 
+            at = 1:4, xaxt = "n", ylim = c(0,0.8), xlim = c(.5,8))
+  })
+  
+  with(ccspace %>% subset(alpha == 400), {
+    boxplot(avgFE ~ cut(SAI_recent, c(-1,0,.5,1)), varwidth = TRUE, 
+            border = cols, col = alpha(cols, .5), 
+            at = 5:7 + 0.5, add = TRUE, xaxt = "n")
+    
+  })
+  
+  mtext(side = 3, at = 2.5, expression(alpha == 100), cex = 1.25)
+  mtext(side = 3, at = 6.5, expression(alpha == 400), cex = 1.25)
+  
+  axis(1, at = c(1:4, 5.5:7.5), c("< 0", "0 - 0.5", "0.5 - 1", "> 1", "< 0", "0 - 0.5", "0.5 - 1"))
+  abline(v = 5)
+  
+  mtext(side = 1, line = 1.5, "spatial adaptation index", cex = 1.25, outer = TRUE)
[…]

### 2021-07-07 memorymigration/R/RunningModel.R
fix

@@ -182,13 +182,13 @@ runManyRuns <- function (parameters.df, resource_param, world, resource,
                                                                M$pop[[length(M$pop)]], world), 
                           resource_param[j,],
                           resource = resource)
-        if(resource_param[1,]$beta_x != 0){ 
-          myR$SAI_total <- computeSpatialAdaptationIndex(M, resource_param[1,])
-          myR$SAI_recent <- computeSpatialAdaptationIndex(M, resource_param[1,], trim = 10)
+        if(resource_param[j,]$beta_x != 0){ 
+          myR$SAI_total <- computeSpatialAdaptationIndex(M, resource_param[j,])
+          myR$SAI_recent <- computeSpatialAdaptationIndex(M, resource_param[j,], trim = 10)
         }
-        if(resource_param[1,]$beta_t != 0){ 
-          myR$TAI_total <- computeTemporalAdaptationIndex(M, resource_param[1,])
-          myR$TAI_recent <- computeTemporalAdaptationIndex(M, resource_param[1,], trim = 10)
+        if(resource_param[j,]$beta_t != 0){ 
+          myR$TAI_total <- computeTemporalAdaptationIndex(M, resource_param[j,])
+          myR$TAI_recent <- computeTemporalAdaptationIndex(M, resource_param[j,], trim = 10)
         }
         
         newresults <- rbind(newresults, c(myR))

### 2021-07-07 text/ms.Rmd

@@
 On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
+Because the scenarios underlying migration are multifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.
 
@@
 Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy.
-The interacting role of memory and sociality, in contrast, have been comparatively little studied.
+The interacting role of memory and sociality, in contrast, has been comparatively little studied.
 
@@
 In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon (**refs**).
-Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing patterns in resource distribution, whether inter-annual variability or trends in resource distribution or phenology reflecting climate change.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, which is also a commonly observed phenomenon (**refs**).
+Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, whether inter-annual variability or trends in resource distribution or phenology reflecting climate change.
 
@@
 direct response to resource availability or perception) and *strategic* (i.e.
-memory-driven and forward-thinking) behavior help foragers navigate dynamic, seasonal environments.
+memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.
 Over-reliance on either strategy should be maladaptive.
-We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic.
+We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic (see also @Lin2021).
 
@@
 
-In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or *reference*) memory of large-scale movement behavior, (4) a short-term (or *working*) memory that updates movement behavior based on recent experience, (5) some social aspect to the learned behavior.
+In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or *reference*) memory of large-scale movement behavior, (4) a short-term (or *working*) memory that updates movement behavior based on recent experience, and (5) some social aspect to the learned behavior.
 A diffusion-advection equation provided a computationally efficient and versatile framework for examining just such a system.
-We consider a population moving in one dimension in a constrained domain ${\cal D}$ and distributing itself according to the following equation:
+We consider a population moving in one dimension in a constrained domain $D$ and distributing itself according to the following equation:
 
@@
 The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \widehat{u}$, and in a negative direction when $x > \widehat{u}$.
-The parameter $\lambda$ is a length scale of sociality - roughly one-half the size of the "swarm", which $\beta$ is a parameter which quantifies the overall strength of sociality.
+The parameter $\lambda$ is a length scale of sociality - roughly one-half the size of the "swarm", where $\beta$ is a parameter that quantifies the overall strength of sociality.
 
@@
 $s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}$ and $s_{21} =  \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}$.
-This step-like migration function is a one-dimensional version of the migration parameters estimated individually [@Gurarie2017] and for populations [@Gurarie2019] in empirical studies.
+This step-like migration function is a one-dimensional version of the migration parameters estimated in empirical studies for individuals [@Gurarie2017] and, more relevantly, for populations [@Gurarie2019] in empirical studies.
 
@@
 
-We ran our numerical experiments on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day years).
+We solved the model numerically on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~years of 100 days).
 We were interested in an approximately periodic resource dynamic, i.e.
@@
 where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e.
-${1 \over \tau} \int_T\int_X h(x,t) dx\,dt = 1$.
+$\frac{1}{\tau} \int_T\int_X h(x,t) dx\,dt = 1$.
 
@@
 ![](figures/ResourceExamples.png)
+
 **Figure 1.
@@
 Thus, the foraging efficiency index is:
-$$FE = {1\over \tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
+$$FE = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
 where the spatial integral is taken over the domain.  This metric is constrained to be between 0 and 1.
 
-For simulations with a constant resource, we run those until a quasi-equilibrium state is achieved, i.e.
-where the Bhattacharya index of the population distribution across subsequent years reaches a value of 0.9999.
-Once that state is attained, we compute the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.
+For simulations with a constant resource, we ran the model until a quasi-equilibrium state was achieved, i.e.
+where the Bhattacharya index of the population distribution across subsequent years reached a value of 0.9999.
+Once that state was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.
 
@@
 ![](figures/example1_adaptation.png)
+
 **Figure 2.** *Example of adaptation to a shift in resource peak.
 The initial (year 0) behavior migrates to locations 50 and -50 at days 15 and 60, whereas the resource peak is at 30 and -30, peaking at times 25 and 75.
-(a) shows the first 14 years of the simulation; (b) The centroid of the annual movement of the population is shown in panel b, with dark blue to red colors indicating year 0 to year 40; (c) annual foraging efficiency across years; (d) migration timing parameters for each year, with orange segments indicating arrival and departure from the summering (northern) grounds, and the blue segments indicating timing of arrival and departure at the wintering grounds; (e) migration arrival and departure location across years, with blue and orange indicating winter (southern) and summer (northern) locations.*
+The panels show (a) the first 14 years of the simulation; (b) the centroid of the annual movement of the population is shown in panel b, with dark blue to red colors indicating year 0 to year 40; (c) annual foraging efficiency across years; (d) migration timing parameters for each year, with orange segments indicating arrival and departure from the summering (northern) grounds, and the blue segments indicating timing of arrival and departure at the wintering grounds; (e) migration arrival and departure location across years, with blue and orange indicating winter (southern) and summer (northern) locations.*
 
@@
 In the illustrated scenario, it takes nearly 40 years to attain an equilibrium, and the eventual steady state is one where the centroid of the migration lines up exactly with the centroid of the resource, and the arrival timing coincides with the *peak* of resource availability.
-Notably, the path to this equilibrium is somewhat indirect, with the winter (later, more southerly) range taking more time to stabilize than the summer (earlier, more northern) range.
+Notably, the path to this equilibrium is somewhat indirect, with the later winter range taking more time to stabilize than the earlier summer range.
 The eventual steady state is one where the foraging efficiency is relatively high.
@@
 
-Figure **X** illustrates the ability of the model to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.
+Figure **X** illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.
 In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
-The qualitative behavior of this process was to start drifting northward, while slowly developing a weak southward pulse in winter as well.
-After first locking in on the summer resource, the southward migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource could be reached consistently.
+The qualitative behavior of this process was to start drifting towards the summer resource, while slowly developing a weak pulse towards the winter resource as well.
+After first locking in on the summer resource, the winter migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource can be consistently reached.
 
@@
 In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).
-The y-axis is the index of spatial adaptiation (SA), i.e.
+The y-axis is the index of spatial adaptation (SA), i.e.
 the trend of the memory-driven migration divided by the resource drift trend.
 Values near 1 indicate a behavior that keeps up with climate change, values near indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend.
-We compare across varying spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels).
+We compare across spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels).
 The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations.
-Finally, the bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicte the highest values, orage and grey lower values.*
+Finally, the bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.*
 
 To assess the ability of the system to adapt to a trending climate, we generated scenarios with slow (0.25 units / year), medium (0.5 units / year), and fast (1 unit / year) drift outward of the two resource pulses.
-We then assessed 40 parameter combinations for each of those scenarios, a high and low value of resource following ($\alpha = 400$ and $100$), a high and low value of sociality ($\beta = 400$ and $100$), and 10 values of the spatial scale of sociality ($\lambda = 20$ to $200$).
+We then assessed 40 parameter combinations for each of those scenarios, a high and low value of resource following ($\alpha = 400$ and $100$) crossed with a high and low value of sociality ($\beta = 400$ and $100$) crossed with 10 values of the spatial scale of sociality ($\lambda = 20$ to $200$).
 The spatial and temporal scale of the resource pulses were fixed to a relatively "easy" to adapt to $\sigma_x = 12$ and $\sigma_t = 6$.
@@
 A rather more dramatic pattern is visible for the lower foraging attraction scenario ($\alpha = 100$; blue circles).
-Notably, no parameter combination at this value can remotely keep up with the rapid climate change (SA range -0.64 to 0.13).
+Notably, no parameter combination at this value comes close to keeping up with the rapid climate change (SA range -0.64 to 0.13).
 For slower climate change, however, there is a window of values for the swarm size - relatively small, between 40 and 80, where the SA *exceeds* 1, and then crashing quite rapidly to negative values of SA as that swarm size increases.
@@
 We found that the relatively straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
-In essense, our model allowed a population of individuals with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location.
+In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location.
 The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether as trends or inter-annual stochasticity.
 
-Note, we are interested here entirely in mechanisms that do *not* correspond to the "green surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.
+Note, we are interested here in a process that do *not* correspond to the "green wave surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.
 
-The ability to migrate was highly dependent on the nature of the resource dynamics.
+The ability to migrate depended on the nature of the resource dynamics.
 In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
-Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than, e.g, the beginning of the resource dynamic.
+Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.
 Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is very much the case for the rapid green-up that occurs in high latitudes as the snow recedes even as day length is high, leading to a rapid and intense green-up.

### 2021-07-08 text/ms.Rmd
ms ms ms ms ms ms ms ms

@@
 output:
-  word_document:
   html_document:
@@
 Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$.
-However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {\cal N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {\cal N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
+However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to:  $m_x(y) \sim {N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
 Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
@@
 
-**Figure Z.** *Adaptation to a steadily drifting resource.
+**Figure 5.** *Adaptation to a steadily drifting resource.
 In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).
@@
 
-## Role of reference memory in adapting to stochasticity
+## Reference memory and stochasticity
 
-*This section still rough*
+While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, inter-annual stochasticity can
 
@@
 
-```{r KappaPlot1, cache  = TRUE, message = FALSE, warning = FALSE}
-load("../results/stochasticity/stochasticity.rda")
-require(magrittr)
-require(plyr)
-require(ggplot2)
-p1 <- list()
-for(a in c(100,400)) for(b in c(100,400)){
-  p1[[length(p1)+1]] <- ccsigma %>% subset(alpha == a & beta == b & lambda > 40) %>% 
-    ggplot(aes(kappa, avgFE, col = factor(lambda))) +
-    facet_grid(.~psi_x) + geom_path() + theme_few() + ggtitle(bquote(alpha == .(a) ~ beta == .(b))) + 
-    ylim(c(0.5,0.75))
-}
-print(p1[[1]])
+![](figures/TrendStochasticity.png)
+
+**Figure 6**.
+*Role of reference memory in adaptying to climate change for increasingly stochastic resource dynamics.
+We ran the model with slow (mean shift 0.25 / year; blue colors) and moderate outward trend (0.5 / year; red colors), at four levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels).
+For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.
+Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and social spatial scale $\lambda = 80$.*
+
+```{r KappaPlot1, cache  = TRUE, message = FALSE, warning = FALSE, eval = FALSE}
+# **Figure ZZ.** (*to be refined*) *Results of foraging 
+# efficiency against kappa.  Higher stochasticity leads to  
+# generally lower efficiency. Higher use of reference memory 
+# leads to higher resource efficiency as well.*
+
 ```
-**Figure ZZ.** (*to be refined*) *Results of foraging efficiency against kappa.
-Higher stochasticity leads to  generally lower efficiency.
-Higher use of reference memory leads to higher resource efficiency as well.*
-
-The final question (simulations currently runnning) is for a system to deal with BOTH trend AND stochasticity, does an intermediate blending of recent and reference memory improve the outcome?
 
@@
 The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether as trends or inter-annual stochasticity.
-
-Note, we are interested here in a process that do *not* correspond to the "green wave surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.
+Importantly, we were not interested here in a process that do *not* correspond to the "green wave surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.
 
@@
 Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.
-Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is very much the case for the rapid green-up that occurs in high latitudes as the snow recedes even as day length is high, leading to a rapid and intense green-up.
+Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.
 
@@
 
+Collective knowledge has been shown to be important, if not essential, to the evolution and process of migration [@Shaw2013;@Guttal2010;@Berdahl2018].
+In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior.
+Synchrony of migration timing and high site fidelity are well documented for migratory species [@Gurarie2019;@Joly2021] **more**.
+
 *Some comments on systems that may or may not be more resilient to climate change based on our results*.

### 2021-07-08 code/analysis/04_ClimateChangeSpaceAnalysis.R

@@ -68,7 +68,7 @@ load("results/stochasticity/stochasticity.rda")
 
 with(ccsigma, table(alpha, beta, lambda, kappa, psi_x))
 
-require(ggplot2)
+require(ggplot2); require(ggthemes)
 ccsigma %>% subset(alpha == 100) %>% 
   ggplot(aes(lambda, avgFE, col = factor(kappa))) +
   facet_grid(beta~psi_x) + geom_path() + theme_few() + ggtitle(expression(alpha == 100)) + 
@@ -110,44 +110,84 @@ if(eval){
   for(i in 1:length(files)){ 
     print(files[i])
     load(paste0("results/trendstochasticity/raw/", files[i]))
-    ccmusigma <- ccmusigma %>% smartbind(newresults %>% mutate(run = i))
+    ccmusigma <- ccmusigma %>% 
+      smartbind(newresults %>% mutate(run = i))
   }
+  
+  ccmusigma <- mutate(ccmusigma, 
+                      SAI_total = as.numeric(SAI_total),
+                      SAI_recent = as.numeric(SAI_recent))
   
   save(ccmusigma, file = "results/trendstochasticity/trendstochasticity.rda")
 }
 
 load("results/trendstochasticity/trendstochasticity.rda")
 
-with(ccmusigma %>% subset(alpha == 100 & beta == 400 & lambda > 40), 
-     table(lambda, kappa, psi_x, beta_x))
+with(ccmusigma %>% subset(alpha == 100 & beta == 400 & lambda == 80), 
+     table(kappa, psi_x, beta_x))
 
 require(ggplot2); require(ggthemes)
-ccmusigma %>% subset(beta_x < 0 & lambda == 80) %>% 
-  ggplot(aes(factor(kappa), SAI_total)) +
+
+ccmusigma %>% subset(beta_x < 0) %>% 
+  subset(alpha == 100 & beta == 400) %>% 
+  ggplot(aes(factor(kappa), SAI_total, col = factor(lambda))) +
   facet_grid(beta_x~psi_x) + geom_boxplot(alpha = 0.5) + theme_few() + 
   geom_jitter(alpha = 0.5) + ylim(c(-1,1.5))
 
+ccmusigma %>% subset(alpha == 100 & 
+                       beta == 400 & lambda == 80) %>% 
[…]

### 2021-07-08 results/simulationexamples.R

@@ -18,24 +18,41 @@ world$resource <- aaply(par0, 1, function(p) getResource_island(world, p))
 
 parameters <- c(epsilon = 4, alpha = 100, beta=400, kappa = 1, lambda = 80)
 
-sim <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
-sim1 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
-sim2 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
-sim3 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
+eval <- FALSE
+if(eval){  
+  sim <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
+  sim1 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
+  sim2 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
+  sim3 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
+}
 
-plotManyRuns(sim$pop, world)
+load("results/kappa06.rda")  # sim2
+load("results/kappa1.rda")  # sim3
+
+
+
 plotManyRuns(sim1$pop, world, nrow = 3)
-plotManyRuns(sim2$pop, world, nrow = 3)
 plotManyRuns(sim3$pop, world, nrow = 3)
 
 FE <- data.frame()
-for(i in 2:length(sim$pop)){
+for(i in 2:length(sim2$pop)){
 compFE <- computeEfficiency(sim2$pop[[i]], world$resource[i-1,,],world)
 FE <- rbind(FE, compFE)
 }
 
 FE2 <- data.frame()
-for(i in 2:length(sim$pop)){
+for(i in 2:length(sim2$pop)){
   compFE <- computeEfficiency(sim3$pop[[i]], world$resource[i-1,,],world)
   FE2 <- rbind(FE2, compFE)
 }
+
+plot(FE[,1], type = "o")
+lines(FE2[,1], type = "o", col = 2)
+
+
+image(sim3$pop[[50]])
+contour(world$resource[47,,], add = TRUE)
+
+
+image(sim2$pop[[50]])
+contour(world$resource[50,,], add = TRUE)

### 2021-07-08 text/figures/code/Stochasticity.R

@@ -1,0 +1,100 @@
+require(magrittr)
+require(plyr)
+require(ggplot2)
+require(ggthemes)
+
+rm(list=ls())
+load("results/trendstochasticity/trendstochasticity.rda")
+
+cc.summary <- ccmusigma %>% subset(alpha == 100 & beta == 400 & beta_x == 0) %>% 
+
+p1 <- list()
+  df %>% 
+    ggplot(aes(kappa, avgFE, col = factor(lambda))) +
+    facet_grid(.~psi_x) + geom_path() + theme_few() + 
+    ylim(c(0.5,0.75))
+print(p1[[2]])
+
+
+
+
+rm(list=ls())
+
+load("results/trendstochasticity/trendstochasticity.rda")
+
+
+cc.summary <- ccmusigma %>% subset(alpha == 100 & beta == 400 & lambda == 80) %>% 
+  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
+        summarize, 
+        n = sum(!is.na(SAI_total)), 
+        ST_mean = mean(SAI_total, na.rm = TRUE), 
+        ST_sd = sd(SAI_total, na.rm = TRUE),
+        SR_mean = mean(SAI_recent, na.rm = TRUE), 
+        SR_sd = sd(SAI_recent, na.rm = TRUE)) %>% 
+  mutate(ST_se = ST_sd/sqrt(n),
+         ST_high = ST_mean + 2*ST_se, 
+         ST_low = ST_mean - 2*ST_se,
+         SR_se = SR_sd/sqrt(n),
+         SR_high = SR_mean + 2*SR_se,
+         SR_low = SR_mean - 2*SR_se)
+
+head(cc.summary)
+
+ylim <- range(cc.summary[,c("SR_high", "SR_low", "ST_high", "ST_low")], na.rm = TRUE)
+
+cc.summary %>% subset(beta_x < 0) %>% 
+  ggplot(aes(kappa, SR_mean, 
+             ymin = SR_low, 
+             ymax = SR_high,
+             col = factor(beta_x))) +
+  facet_grid(.~psi_x) + theme_few() + 
+  geom_point() + geom_errorbar() + geom_hline(yintercept = c(0,1)) + 
+  ggtitle("Adaptation - recent years") +
+  geom_path()
+
+
+
+cc.summary %>% subset(psi_x == 0)
+
+df <- subset(cc.summary, psi_x == 3)
+cols <-  c("red", "blue")
[…]

### 2021-07-08 text/figures/code/TrendStochasticity.R

@@ -1,0 +1,80 @@
+rm(list=ls())
+
+load("results/trendstochasticity/trendstochasticity.rda")
+
+
+cc.summary <- ccmusigma %>% subset(alpha == 100 & beta == 400 & lambda == 80) %>% 
+  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
+        summarize, 
+        n = sum(!is.na(SAI_total)), 
+        ST_mean = mean(SAI_total, na.rm = TRUE), 
+        ST_sd = sd(SAI_total, na.rm = TRUE),
+        SR_mean = mean(SAI_recent, na.rm = TRUE), 
+        SR_sd = sd(SAI_recent, na.rm = TRUE)) %>% 
+  mutate(ST_se = ST_sd/sqrt(n),
+         ST_high = ST_mean + 2*ST_se, 
+         ST_low = ST_mean - 2*ST_se,
+         SR_se = SR_sd/sqrt(n),
+         SR_high = SR_mean + 2*SR_se,
+         SR_low = SR_mean - 2*SR_se)
+
+head(cc.summary)
+
+ylim <- range(cc.summary[,c("SR_high", "SR_low", "ST_high", "ST_low")], na.rm = TRUE)
+
+cc.summary %>% subset(beta_x < 0) %>% 
+  ggplot(aes(kappa, SR_mean, 
+             ymin = SR_low, 
+             ymax = SR_high,
+             col = factor(beta_x))) +
+  facet_grid(.~psi_x) + theme_few() + 
+  geom_point() + geom_errorbar() + geom_hline(yintercept = c(0,1)) + 
+  ggtitle("Adaptation - recent years") +
+  geom_path()
+
+
+
+cc.summary %>% subset(psi_x == 0)
+
+df <- subset(cc.summary, psi_x == 3)
+cols <-  c("red", "blue")
+
+
+plotSR <- function(df, ...){
+  plot(c(0,1), ylim, type = "n", xlab = "", ylab = "")
+  abline(h = c(0,1), lty = c(1,3), lwd = 3, col = "grey", xpd = FALSE)
+  with(df, {
+    points(kappa, SR_mean, col = cols[factor(beta_x)], pch = 19, ...)
+    segments(kappa, SR_low, kappa, SR_high, col = cols[factor(beta_x)])
+    lines(kappa[beta_x == -.25], SR_mean[beta_x == -.25], col = cols[2])
[…]

### 2021-07-08 text/figures/code/Stochasticity.R
stochasticity plotting

@@ -7,14 +7,20 @@ rm(list=ls())
 load("results/trendstochasticity/trendstochasticity.rda")
 
 cc.summary <- ccmusigma %>% subset(alpha == 100 & beta == 400 & beta_x == 0) %>% 
+  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
+        summarize, 
+        n = sum(!is.na(avgTE)), 
+        FE_mean = mean(avgTE, na.rm = TRUE), 
+        FE_sd = sd(avgTE, na.rm = TRUE)) %>% 
+  mutate(FE_se = FE_sd/sqrt(n),
+         FE_high = FE_mean + 2*FE_se,
+         FE_low = FE_mean - 2*FE_se)
 
-p1 <- list()
-  df %>% 
-    ggplot(aes(kappa, avgFE, col = factor(lambda))) +
-    facet_grid(.~psi_x) + geom_path() + theme_few() + 
-    ylim(c(0.5,0.75))
-print(p1[[2]])
 
+cc.summary %>% subset(lambda == 80) %>% 
+    ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
+    facet_grid(.~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
+    geom_path()
 
 
 

### 2021-07-09 text/ms.Rmd
...

@@
 
-While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, inter-annual stochasticity can
+While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, we hypothesized that more conservative approach, relying on a "reference memory" may be more beneficial when conditions change stochastically.
+We tested this hypothesis by solving a set of models across a range of $\kappa$ values fom 0 (all recent memory) to 1 (all reference memory).
+In these scenarios, we ran the system for as many years as needed with no stochasticity to acquire a stationary state (i.e.
+similarity index greater than 0.9999), used the stationary state to define the reference migration memory, and then ran the process for an additional 50 years with a stochasticity (i.e.
+standard deviation in peak location of the resource) ranging from 0 to 12 units.
 
-We ran simulations with various levels of stochasticity (inter-annual standard deviation in location of resource pulse 3, 6, and 12) notably against various values of reference memory $\kappa$, where $\kappa = 0$ refers to only recent memory, and $\kappa = 1$ refers to all reference memory, i.e.
-that never updates.
-We ran simulations for 20 years with constant resource to stabilize, and then for 30 more years to let the process evolve.
-In all cases, unsurprisingly, the conservative $\kappa = 1$ limit yielded the highest interannual foraging efficiency (see figure ZZ below).
+Figure 6 illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency; however this is highly sensitive to the spatial scale of sociality.
+When that scale of sociality is high enough (e..g $\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.
+However, if that scale is too small, then - counterintuitively - the process is overly fixed on repeatability and misses the opportunity to increase its foraging by "randomizing" its migration movements via memory of improved foraging in a previous year.
+Overall, as anticipated, the greater the stochasticity, the lower the foraging efficiency; and the greater the stochasticity, the greater the benefit of a conservative reference memory driven strategy.
 
-![](figures/TrendStochasticity.png)
+```{r StochasticityOnly, cache = TRUE}
+require(magrittr); require(plyr); require(ggplot2); require(ggthemes); require(gtools)
+a <- load("../results/stochasticity/stochasticity.rda")
+b <- load("../results/trendstochasticity/trendstochasticity.rda")
 
-**Figure 6**.
+cc <- smartbind(ccsigma, 
+                subset(ccmusigma, psi_x == 0)) %>% 
+  subset(alpha == 100 & beta == 400 & beta_x == 0)
+
+cc.summary <- cc %>% 
+  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
+        summarize, 
+        n = sum(!is.na(avgFE)), 
+        FE_mean = mean(avgFE, na.rm = TRUE), 
+        FE_sd = sd(avgFE, na.rm = TRUE)) %>% 
+  mutate(FE_se = FE_sd/sqrt(n),
+         FE_high = FE_mean + 2*FE_se,
+         FE_low = FE_mean - 2*FE_se)
+
+cc.summary %>% subset(lambda > 40) %>% 
+    ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
+    facet_grid(.~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
+    geom_path()
+
+```
+
+**Figure 6** *Mean and standard error of foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$.
+For the stochastic processes ($\psi_x > 0$), the process was run 30 times, and the mean and standard error of the foraging efficiency over a 50 year post-stabilization period is presented (points and bars).
+In these scenarios, the resource following parameter $\alpha = 100$ and  social attraction $\beta = 400$.*
+
+## Balancing stochasticity with trends
+
+We added trends to the stochastic process described above, adding 30 years post-stabilization with slower (0.25 / year) and moderate (0.5 / year) shifts in the peak of resource, with the goal of assessing the ability of the process to keep up with climate change as a function of the reference memory parameter (Figure 7).
+Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change.
+However, over-reliance on recent memory is also sub-optimal, and - in fact - a balancing of recent and reference memory (in our relatively narrow set of similutions with a $\kappa$ value between 0.8 and 0.9) helps the system keep up with climate change better than chasing each recent shift.
+Note also, that at the highest levels of stochasticity ($\psi_x = 12$), the system fails completely to keep up with the climate change.
+The negative spatial adaptation index indicated that the migration behavior, in fact, can completely fall apart.
+
+![](figures/TrendStochasticity_v2.png)
+
+**Figure 7**.
 *Role of reference memory in adaptying to climate change for increasingly stochastic resource dynamics.
@@
 Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and social spatial scale $\lambda = 80$.*
-
-```{r KappaPlot1, cache  = TRUE, message = FALSE, warning = FALSE, eval = FALSE}
-# **Figure ZZ.** (*to be refined*) *Results of foraging 
-# efficiency against kappa.  Higher stochasticity leads to  
-# generally lower efficiency. Higher use of reference memory 
-# leads to higher resource efficiency as well.*
-
-```
 

### 2021-07-09 code/analysis/04_ClimateChangeSpaceAnalysis.R

@@ -60,7 +60,8 @@ if(eval){
     load(paste0("results/stochasticity/raw/", f))
     ccsigma <- ccsigma %>% smartbind(newresults)
   }
-  
+  ccsigma <- mutate(ccsigma, 
+         annualFE = as.numeric(annualFE))
   save(ccsigma, file = "results/stochasticity/stochasticity.rda")
 }
 
@@ -116,7 +117,8 @@ if(eval){
   
   ccmusigma <- mutate(ccmusigma, 
                       SAI_total = as.numeric(SAI_total),
-                      SAI_recent = as.numeric(SAI_recent))
+                      SAI_recent = as.numeric(SAI_recent),
+                      annualFE = as.numeric(annualFE))
   
   save(ccmusigma, file = "results/trendstochasticity/trendstochasticity.rda")
 }

### 2021-07-09 text/figures/code/Stochasticity.R

@@ -11,7 +11,6 @@ cc <- smartbind(ccsigma,
                 subset(ccmusigma, psi_x == 0)) %>% 
   subset(alpha == 100 & beta == 400 & beta_x == 0)
 
-
 cc.summary <- cc %>% 
   ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
         summarize, 
@@ -23,7 +22,7 @@ cc.summary <- cc %>%
          FE_low = FE_mean - 2*FE_se)
 
 
-cc.summary %>% subset(lambda %in% c(80,120)) %>% 
+cc.summary %>% subset(lambda > 40) %>% 
     ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
     facet_grid(.~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
     geom_path()
@@ -42,13 +41,18 @@ cc.summary <- ccmusigma %>% subset(alpha == 100 & beta == 400 & lambda == 80) %>
         ST_mean = mean(SAI_total, na.rm = TRUE), 
         ST_sd = sd(SAI_total, na.rm = TRUE),
         SR_mean = mean(SAI_recent, na.rm = TRUE), 
-        SR_sd = sd(SAI_recent, na.rm = TRUE)) %>% 
+        SR_sd = sd(SAI_recent, na.rm = TRUE),
+        FE_mean = mean(avgFE, na.rm = TRUE),
+        FE_sd = sd(avgFE, na.rm = TRUE)) %>% 
   mutate(ST_se = ST_sd/sqrt(n),
          ST_high = ST_mean + 2*ST_se, 
          ST_low = ST_mean - 2*ST_se,
          SR_se = SR_sd/sqrt(n),
          SR_high = SR_mean + 2*SR_se,
-         SR_low = SR_mean - 2*SR_se)
+         SR_low = SR_mean - 2*SR_se,
+         FE_se = FE_sd/sqrt(n),
+         FE_high = FE_mean + 2*FE_se,
+         FE_low = FE_mean - 2*FE_se)
 
 head(cc.summary)
 
@@ -66,13 +70,7 @@ cc.summary %>% subset(beta_x < 0) %>%
 
 
 
-cc.summary %>% subset(psi_x == 0)
-
-df <- subset(cc.summary, psi_x == 3)
-cols <-  c("red", "blue")
-
-
-plotSR <- function(df, ...){
+plotST <- function(df, ...){
[…]

### 2021-07-09 text/ms.Rmd
...

@@
 
-cc.summary %>% subset(lambda > 40) %>% 
+cc.summary %>% subset(lambda > 40 & psi_x > 0) %>% 
     ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +

### 2021-07-09 code/analysis/04_ClimateChangeSpaceAnalysis.R

@@ -47,7 +47,7 @@ ccspace %>%
 
 
 #  Climate change stochasticity --------
-
+require(memorymigration)
 eval <- FALSE
 if(eval){
   require(gtools)
@@ -61,7 +61,9 @@ if(eval){
     ccsigma <- ccsigma %>% smartbind(newresults)
   }
   ccsigma <- mutate(ccsigma, 
-         annualFE = as.numeric(annualFE))
+         annualFE = as.numeric(annualFE), 
+         SAI_total = NULL, SAI_recent = NULL, 
+         TAI_total = NULL, TAI_recent = NULL)
   save(ccsigma, file = "results/stochasticity/stochasticity.rda")
 }
 

### 2021-07-09 memorymigration/R/RunningModel.R

@@ -14,7 +14,7 @@ runManyYearsWithStabilization <- function(world, parameters,
   parameters.null <- parameters
   parameters.null['kappa'] <- 0
   sim.null <- runManyYears(world.null, parameters = parameters.null, 
-                           n.years = n.years.null, threshold = 0.9999, 
+                           n.years = n.years.null, threshold = threshold, 
                            verbose = verbose, m0 = world$m0)
   
   # extract pieces to run for the remainder
@@ -25,7 +25,7 @@ runManyYearsWithStabilization <- function(world, parameters,
                   as.matrix)[1,]
   world$pop <- sim.null$pop[[length(sim.null$pop)]]
   sim <- runManyYears(world, parameters = parameters, 
-                      n.years = n.years, threshold = threshold, 
+                      n.years = n.years, threshold = 1, 
                       verbose = verbose, m0 = world$m0)
   return(sim)
 }

### 2021-07-09 memorymigration/R/datadocumentation.R

@@ -1,3 +1,19 @@
+#' Climate change worlds
+#'
+#' memorymigration set up for running climate change models.  Contains two worlds: 
+#' world_cclambda80 and world_cclambda120.
+#' 
+#' These worlds are "stabilized" at precision 1e-6.  
+#' See generating code for other parameter values. 
+#'
+#' @usage
+#' data(world_cc)
+#'
+#' @example examples/creatingClimateChangeWorlds.R
+#' @keywords data
+#' @export
+NULL
+
 #' Basic World object
 #'
 #' memorymigration set up for running models

### 2021-07-09 memorymigration/examples/creatingClimateChangeWorlds.R

@@ -1,0 +1,55 @@
+world <- getOptimalPop(tau=100, X.min = -100, X.max = 100, dx=1, 
+                       x1 = 30, x2 = -30, t.peak=25, 
+                       x.sd=12, t.sd=6)
+world$m0 <- fitMigration(t = world$time, x = getMem(world$pop, world))
+resource.null <- getResource_island(world, c(x.peak = 30, t.peak = 25, x.sd = 12, t.sd = 6))
+world$resource <- resource.null
+
+# run baseline no climate change for 100 years
+sim3 <- runManyYearsWithStabilization(world, parameters = model.pars[[3]], 
+                                        n.years = 30,  
+                                        n.years.null = 100, 
+                                        resource.null = resource.null, 
+                                        threshold = 0.999999, 
+                                        verbose=TRUE)
+plotMigrationHat(sim3$migration.hat)
+
+
+# create 2 worlds for climate change runs
+
+world_cc <- getOptimalPop(tau=100, X.min = -100, X.max = 100, dx=1, 
+                          x1 = 30, x2 = -30, t.peak=25, 
+                          x.sd=12, t.sd=6)
+
+makeMhatvector <- function(df){
+  ((df %>% tail(1))[1:6] %>% as.matrix)[1,]
+}
+
+world_cc$m0 <- makeMhatvector(sim3$migration.hat)
+world_cc$resource <- getResource_island(world_cc, c(x.peak = 30, t.peak = 25, x.sd = 12, t.sd = 6))
+
+params_cclambda80 <- c(epsilon = 4, alpha = 100, beta=400, kappa = 0, lambda = 80)
+params_cclambda120 <- c(epsilon = 4, alpha = 100, beta=400, kappa = 0, lambda = 120)
+
+world_cc$pop <- sim3$pop[[length(sim3$pop)]]
+sim0_cclambda80 <- runManyYears(world_cc, 
+                                parameters = params_cclambda80, 
+                                n.years = 100, 
+                                threshold= 1-1e-6, 
+                                verbose = TRUE)
+
+sim0_cclambda120 <- runManyYears(world_cc, 
+                                 parameters = params_cclambda120, 
+                                 n.years = 100, 
+                                 threshold= 1-1e-6, 
+                                 verbose = TRUE)
+
+world_cclambda120 <- world_cclambda80 <- world_cc
+
[…]

### 2021-07-09 results/simulationexamples.R

@@ -5,30 +5,96 @@ world$m0 <- fitMigration(t = world$time, x = getMem(world$pop, world))
 
 par0 <- getCCpars(mu_x0 = 30, 
                   mu_t0 = 25,
-                  beta_x = -0.5,
+                  beta_x = 0,
                   beta_t = 0,
                   n.years = 30,
                   sigma_x = 12,
                   sigma_t = 6,
-                  psi_x = 3, 
+                  psi_x = 0, 
                   psi_t = 0,
-                  n.years.null = 20) 
-world$resource <- aaply(par0, 1, function(p) getResource_island(world, p))
+                  n.years.null = 0) 
+
+resource <- aaply(par0, 1, function(p) getResource_island(world, p))
 
 
-parameters <- c(epsilon = 4, alpha = 100, beta=400, kappa = 1, lambda = 80)
+rm(list=ls())
 
-eval <- FALSE
-if(eval){  
-  sim <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
-  sim1 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
-  sim2 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
-  sim3 <- runManyYears(world, parameters = parameters, n.years = 50, threshold = 1, verbose=TRUE)
+model.pars80 <- list(
+  c(epsilon = 4, alpha = 100, beta=400, kappa = 0, lambda = 80),
+  c(epsilon = 4, alpha = 100, beta=400, kappa = .8, lambda = 80),
+  c(epsilon = 4, alpha = 100, beta=400, kappa = 1, lambda = 80))
+  
+  
+model.pars120 <- list(
+  c(epsilon = 4, alpha = 100, beta=400, kappa = 0, lambda = 120),
+  c(epsilon = 4, alpha = 100, beta=400, kappa = .8, lambda = 120),
+  c(epsilon = 4, alpha = 100, beta=400, kappa = 1, lambda = 120))
+
+
+data(world_cc)
+
+sims.80 <- lapply(model.pars80, function(p)
+    runManyYears(world_cclambda80, parameters = p, n.years = 10,  
+               threshold = 1, verbose=TRUE)
+  )
+
+sims.120 <- lapply(model.pars120, function(p)
+  runManyYears(world_cclambda120, parameters = p, n.years = 10,  
[…]

### 2021-07-09 text/table.Rmd
table

---
title: "Tables"
author: "Elie"
date: "7/9/2021"
output: html_document
---

**Memory migration model**

| Parameter   | | Description  
| ----------- | ----------- | 
| $\alpha$  | Forage following | 
| $\beta$  | Strength of sociality |
| $\lambda$  | Spatial scale of sociality |

**Resource**

**Metrics**

### 2021-07-10 text/table.Rmd
updated shiny + table for MS

@@
 
-| Parameter   | | Description  
+| Parameter   | | Description |
 | ----------- | ----------- |
+| $\epsilon$  | Diffusion | 
 | $\alpha$  | Forage following | 
@@
 | $\lambda$  | Spatial scale of sociality |
+| $\kappa$  | Proportion of reference versus working memory |
 
 **Resource**
+| Parameter   | | Description |
+| ----------- | ----------- |
+| $h$  | Dynamic resource | 
+| $\tau$  | Periodicity | 
+| $M_x$  | Spatial coordinate of resource peak for the summer season | 
+| $-m_x$  | Spatial coordinate of resource peak for the winter season |
+| $m_t$  | Timing of resource peak for the summer season | 
+| $\tau - m_t$ | Timing of resource peak for the summer season | 
+| $s_t$  | Time scale of duration of resource pulse | 
+| $s_x$  | Spatial scale of extent of resource pulse |
+
+**Population**
+| Parameter   | | Description |
+| ----------- | ----------- |
+| $t_1$  | Timing of the summer season | 
+| $dt_1$  | Duration of the summer season | 
+| $t_2$  | Timing of the winter season | 
+| $dt_2$  | Duration of the winter season | 
+| $x_1$  |  Spatial coordinate of the population centroid for the summer season | 
+| $x_2$  |Spatial coordinate of the population centroid for the winter season |
 
 **Metrics**
+| Parameter   | | Description |
+| ----------- | ----------- |
+| $MM_x$  | Spatial migration mismatch |
+| $MM_t$  | Temporal migration mismatch |
+| $FE$  | Foraging efficiency |
+| $SA$  | Spatial adaptation |

### 2021-07-10 memorymigration/R/functions_indices.R

@@ -245,27 +245,52 @@ computeAvgTotalError <- function(sim, world, par0){
   tau <- world$tau
   average <- data.frame()
   
-  for(i in 30:length(sim$pop)){
-    par <- par0[i,]
-    m.hat <- sim$migration.hat[i,]
-    if(par["t.peak"] < m.hat$t1 | par["t.peak"] > m.hat$t1 + m.hat$dt1)
-      t1.error <- min( abs(par["t.peak"] - m.hat$t1), 
-                       abs(par["t.peak"] - m.hat$t1  - m.hat$dt1)) else 
-                         t1.error = 0
-    
-    if(world$tau - par["t.peak"] < m.hat$t2 | 
-       world$tau - par["t.peak"] > m.hat$t2 + m.hat$dt2)
-      t2.error <- min( abs(world$tau - par["t.peak"] - m.hat$t2), 
-                       abs( world$tau - par["t.peak"] - m.hat$t2  - m.hat$dt2)) else 
-                         t2.error = 0
-    
-    x1.error <- par["x.peak"] - m.hat$x1
-    x2.error <- - par["x.peak"] - m.hat$x2
-    
-   error <- abs(t1.error) + abs(t2.error) + abs(x1.error) + abs(x1.error)
-    
-   average <- rbind(average, error)
-  }
+  if (length(sim) <= 20){
+    for(i in 2:length(sim)){
+      par <- par0[i,]
+      m.hat <- sim$migration.hat[i,]
+      if(par["t.peak"] < m.hat$t1 | par["t.peak"] > m.hat$t1 + m.hat$dt1)
+        t1.error <- min( abs(par["t.peak"] - m.hat$t1), 
+                         abs(par["t.peak"] - m.hat$t1  - m.hat$dt1)) else 
+                           t1.error = 0
+      
+      if(world$tau - par["t.peak"] < m.hat$t2 | 
+         world$tau - par["t.peak"] > m.hat$t2 + m.hat$dt2)
+        t2.error <- min( abs(world$tau - par["t.peak"] - m.hat$t2), 
+                         abs( world$tau - par["t.peak"] - m.hat$t2  - m.hat$dt2)) else 
+                           t2.error = 0
+      
+      x1.error <- par["x.peak"] - m.hat$x1
+      x2.error <- - par["x.peak"] - m.hat$x2
+      
+      error <- abs(t1.error) + abs(t2.error) + abs(x1.error) + abs(x1.error)
+      
[…]

### 2021-07-10 shiny/memorymigration/ShinyScripts.R

@@ -20,16 +20,16 @@ ui <- fluidPage(
                         value = 100, min = 0, step = 1),
            numericInput(inputId = "beta",
                         label = "Strength of Sociality - Beta",
-                        value = 100, min = 0, step = 1),
+                        value = 400, min = 0, step = 50),
            sliderInput(inputId = "kappa",
                        label = "Proportion reference vs. working memory - Kappa",
                        value = 0, min = 0, max = 1),
            numericInput(inputId = "lambda",
                         label = "Spatial Scale of Sociality - Lambda",
-                        value = 20, min = 0, step = 1),
+                        value = 80, min = 0, step = 1),
            numericInput(inputId = "epsilon",
                         label = "Diffusion Parameter - Epsilon",
-                        value = 1, min = 0, step = 1)
+                        value = 4, min = 0, step = 1)
            ),
     column(6,
            h3("Migratory Population"),
@@ -47,17 +47,18 @@ ui <- fluidPage(
                         label = "View Resource"),
            sliderInput(inputId = "x.sd",
                        label = "Resource Space Distribution",
-                       value = 5, min = 0, max = 15),
+                       value = 12, min = 0, max = 15),
            sliderInput(inputId = "t.sd",
                        label = "Resource Time Distribution",
-                       value = 12, min = 0, max = 15),
-           numericInput(inputId = "mu.x0", label = "Initial Resource Position in Space", value = 40, step = 1),
+                       value = 6, min = 0, max = 15),
+           numericInput(inputId = "mu.x0", label = "Initial Resource Position in Space", value = 30, step = 1),
            numericInput(inputId = "mu.t0", label = "Initial Resource Position in Time", value = 25, step = 1),
            numericInput(inputId = "beta.x", label = "Resource Change in Space", value = 0, step = 1),
            numericInput(inputId = "beta.t", label = "Resource Change in Time ", value = 0, step = 1),
            numericInput(inputId = "psi_x", label = "Stochasticity in Space", value = 0, step = 1),
            numericInput(inputId = "psi_t", label = "Stochasticity in Time ", value = 0, step = 1),
            numericInput(inputId = "x_null", label = "Constraint in Space ", value = 50, step = 1),
+           numericInput(inputId = "n_years_null", label = "Years of Stable Resource ", value = 0, step = 1),
[…]

### 2021-07-10 text/ms.Rmd

@@
 
-*Highly sketchy and incomplete.  Some odd thoughts here and there.*
+Animals navigate complex, dynamic, patchy environments.
+When there is a strongly localized and seasonal component to the resource dynamics, typical strategies of spatial distribution by straightforward resource-following taxis necessarily fails to maximally exploit the dynamics.
+It is in these cases - quite common in the natural world - that seasonal migration becomes a viable, even necessary, strategy.
+However, when those resources start shifting in space and time - as is occurring at an accelerated pace with recent global climate change - the migration phenology itself must exhibit some plasticity.
+It is our conjecture that this plasticity is greatly facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years.
+To our knowledge, this study is the first attempt to focus in and propose a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics.
 
-We found that the relatively straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
+We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
 In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location.
-The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether as trends or inter-annual stochasticity.
-Importantly, we were not interested here in a process that do *not* correspond to the "green wave surfing" mechanism, as there is very little trend to surf, but consider only a necessarily strategic, long-scaled displacements and the way those displacements are modified by experience.
+The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity.
+We note that our model did not include any selection, inheritance or birth or death processes, i.e.
+there was evolutionary optimization of parameters to maximize a fitness, as in several somewhat similar models that explicitly explore the evolution of migration [@Guttal2010;@Shaw2013;@Anderson2013].
+Rather, we postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior.
+The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.
 
-The ability to migrate depended on the nature of the resource dynamics.
+It is worth considering the bioloogical interpretations of the various parameters.
+The diffusion parameter ($\epsilon$) captures short time-scaled randomness and exploration of a population and the foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.
+These are the most common ingredients in diffusion-advection models of animal movement and direct parallels can be drawn between these and, respectively, individual random movement models [@Gurarie2011] and resource selection models, which are the fundamental ingredients of empirical analysis of animal movements.
+The sociality parameters, which turn out to be exceedingly important for the resiliency of this process, capture two very important but lesser-studied properties of animal movements.
+The strength of sociality ($\beta$) essentially quantified the importance of the group.
+Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion; and it appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology.
+The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, and was, somewhat unexpectedly, among the most important parameters.
+In particular, a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
+This result points to an important fundamental dynamic: that in order to adapt to any shift in resources, some individuals must experience that resource, and the two sociality-related mechanisms to have that scope is either a relatively large spatial scale or a relatively weak social cohesion.
+
+The ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics.
 In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
@@
 Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.
-
-It is worth considering the assumptions of our model with respect to a biological population.
-The population is explicitly social and collective, and the memory is a memory that belongs to the entire population.
-The importance of the spatial scale of the collective was an unexpected and very strong result, suggesting that for a population to adequately learn socially, it must have a large enough exploratory radius to find new information to adapt to, while being small enough that the entire population can then learn from this.
 

### 2021-07-13 text/ms.Rmd
98% there

@@
 ---
-title: "Modeling memory and taxis to analyze migrations"
-author: "E. Gurarie, S. Potluri, B. Fagan, others"
+title: 'Memories of migrations past: Sociality and cognition in dynamic, seasonal
+  environments'
+author: "E. Gurarie, S. Potluri, C. Cosner, R. Cantrell, W.F. Fagan"
 date: "`r Sys.Date()`"
 output:
+  pdf_document:
+    keep_tex: true
+    citation_package: natbib
   html_document:
@@
 
+# Abstract
+
+*to do.*
+
 # Introduction
@@
 
-Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Shaw2016].
+Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Berthold1999;@Shaw2016].
 Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015].
 These migrations can be considered "tactical", as they can occur - as an extreme simplification - purely as response to local conditions.
-Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration (**refs**).
-This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).
+Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration [@Abrahms2019].
+This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized.
 This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by a remembered anticipation [@Bracis2017].
@@
 These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component.
-While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
+While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
 Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.
-by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).
+by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics.
 
@@
 In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, which is also a commonly observed phenomenon (**refs**).
-Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, whether inter-annual variability or trends in resource distribution or phenology reflecting climate change.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon [@Wilcove2008].
+Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, including both variability or trends in resource distribution or phenology reflecting climate change.
 
-We further anticipate that under many conditions a blending of *tactical* (i.e.
+We anticipate that under many conditions a blending of *tactical* (i.e.
 direct response to resource availability or perception) and *strategic* (i.e.
@@
 v_m(t, \theta_y) = \begin{cases}
-0 &;  & t > t_1 &\text{and} & t \leq t_1 + dt_1 \\ 
-s_{12} &;  &  t > t_1 + dt_1 &\text{and} & t \leq t_2 \\ 
-0 &; & t> t_2 &\text{and} &t \leq t_2 + dt_2 \\ 
-s_{21} &; &  t > t_2+dt_2 & \text{or} & t \leq t_1\\ 
+0; & t > t_1  \,\, \text{and} \,\, t \leq t_1 + dt_1 \\ 
+s_{12}; & t > t_1 + dt_1 \,\, \text{and} \,\,  t \leq t_2 \\ 
+0; & t> t_2 \,\, \text{and} \,\, t \leq t_2 + dt_2 \\ 
+s_{21}; &  t > t_2+dt_2 \,\, \text{or} \,\, t \leq t_1\\ 
 \end{cases}
@@
 3, 6, 12).
-For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this index as the *spatial adaptation* (SA) index.
+For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this as the *spatial adaptation* (SA) index.
 
@@
 
-We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions:
+We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions: 1.
+Can this model adapt to a discrete shift in peak resource location and timing?
+What is the relative role of memory and sociality for adaptation?
+2.
+Can this model acquire a migratory behavior from a non-migratory initial condition?
+3.
+What is the role of a reference memory for dealing with stochastic resource dynamics?
+4.
+Can this model adapt when the resource peaks shifts in space?
 
-1. Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation?
+**Table 1** *Parameters, variables and metrics.*
 
-2. Can this model acquire a migratory behavior from a non-migratory initial condition?
-
-3. What is the role of a reference memory for dealing with stochastic resource dynamics?
-
-4. Can this model adapt when the resource peaks shifts in space?
+|| Parameter   | Description |
+|-----| ----------- | ----------- |
+| *Memory migration model* 
+|| $\epsilon$  | Diffusion | 
+|| $\alpha$  | Forage following | 
+|| $\beta$  | Strength of sociality |
+|| $\lambda$  | Spatial scale of sociality |
+|| $\kappa$  | Proportion of reference versus working memory |
+|| $x_1$, $x_2$  | location of population centroids in summer and winter| 
+|| $t_1$, $dt_1$  | start and duration of summer season |
+|| $t_2$, $dt_2$  | start and duration of winter season | 
+|| $\kappa$ | proportion memory allocated to reference (long-term) memory vs. short-term memory|
+| *Resource dynamics* 
+|| $\tau$  | duration of period (year) | 
+|| $m_x$, $-m_x$  | spatial coordinate of resource peak for summer and winter | 
+|| $m_t$, $\tau - m_t$  | timing of resource peak for the summer and winter | 
+|| $\sigma_x$, $\sigma_t$   | time duration and and spatial scale of resource pulse | 
+|| $dx$, $dt$  | rate of change of peak locaton and timing of resource | 
+|| $\psi_x$, $\psi_t$  | standard deviation of peak location and timing | 
+| *Metrics* 
+|| $MM_x$  | spatial migration mismatch |
+|| $MM_t$  | temporal migration mismatch |
+|| $TE$  | total mismatch |
+|| $FE$  | foraging efficiency |
+|| $SA$  | spatial adaptation index |
 
@@
 Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.
-Finally, within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right.
-Finally, the color scheme reflects the total mismatch, i.e.
+Within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right.
+The color scheme reflects the total mismatch, i.e.
 the sum of the absolute differences between the migration timing and locations from the resource peak.
@@
 ![](figures/example2_learningtomigrate.png)
-**Figure X.** *Example of model learning to migrate.
+**Figure 5.** *Example of model learning to migrate.
 The resource is a "weakly drifting" resource and the initial (year 0) condition is non migratory.
@@
 
-Figure **X** illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.
+Figure **5** illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.
 In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
@@
 The model had, in general, a difficult time learning migration from a non-migratory initial condition.
-Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5 (Appendix Figure **Z**).
+Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5.
 Conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$), but *smaller* in scope (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact in space but long enough in duration to lock in to the memory.
@@
 
-**Figure 5.** *Adaptation to a steadily drifting resource.
+**Figure 6.** *Adaptation to a steadily drifting resource.
 In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).
@@
 The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations.
-Finally, the bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.*
+The bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.*
 
@@
 
-As figure Z shows, higher values of resource following ($\alpha = 400$; orange circles) are nearly universally better for keeping up with climate change (SA values near 1).
+As figure 6 shows, higher values of resource following ($\alpha = 400$; orange circles) are nearly universally better for keeping up with climate change (SA values near 1).
 Furthermore, when combined with high sociality ($\beta = 400$; right panels), nearly all parameter combinations do a good job keeping up with climate change (SA values ranging between 0.53 and 0.85 for a swarm size greater than 50).
@@
 In these scenarios, we ran the system for as many years as needed with no stochasticity to acquire a stationary state (i.e.
-similarity index greater than 0.9999), used the stationary state to define the reference migration memory, and then ran the process for an additional 50 years with a stochasticity (i.e.
-standard deviation in peak location of the resource) ranging from 0 to 12 units.
+similarity index greater than 1-1e-6).
+We then used the stationary state as the reference memory, and then ran the process for an additional 50 years with a stochasticity (i.e.
+standard deviation in peak location of the resource) ranging from 0 to 12.
 
-Figure 6 illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency; however this is highly sensitive to the spatial scale of sociality.
+Figure 7 illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency; however this is highly sensitive to the spatial scale of sociality.
 When that scale of sociality is high enough (e..g $\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.
-However, if that scale is too small, then - counterintuitively - the process is overly fixed on repeatability and misses the opportunity to increase its foraging by "randomizing" its migration movements via memory of improved foraging in a previous year.
 Overall, as anticipated, the greater the stochasticity, the lower the foraging efficiency; and the greater the stochasticity, the greater the benefit of a conservative reference memory driven strategy.
 
-```{r StochasticityOnly, cache = TRUE}
-require(magrittr); require(plyr); require(ggplot2); require(ggthemes); require(gtools)
-a <- load("../results/stochasticity/stochasticity.rda")
-b <- load("../results/trendstochasticity/trendstochasticity.rda")
+![](figures/stochasticity.png)
 
-cc <- smartbind(ccsigma, 
-                subset(ccmusigma, psi_x == 0)) %>% 
-  subset(alpha == 100 & beta == 400 & beta_x == 0)
-
-cc.summary <- cc %>% 
-  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
-        summarize, 
-        n = sum(!is.na(avgFE)), 
-        FE_mean = mean(avgFE, na.rm = TRUE), 
-        FE_sd = sd(avgFE, na.rm = TRUE)) %>% 
-  mutate(FE_se = FE_sd/sqrt(n),
-         FE_high = FE_mean + 2*FE_se,
-         FE_low = FE_mean - 2*FE_se)
-
-cc.summary %>% subset(lambda > 40 & psi_x > 0) %>% 
-    ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
-    facet_grid(.~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
-    geom_path()
-
-```
-
-**Figure 6** *Mean and standard error of foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$.
+**Figure 7** *Foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$.
 For the stochastic processes ($\psi_x > 0$), the process was run 30 times, and the mean and standard error of the foraging efficiency over a 50 year post-stabilization period is presented (points and bars).
-In these scenarios, the resource following parameter $\alpha = 100$ and  social attraction $\beta = 400$.*
+In these scenarios, the resource following parameter $\alpha = 100$, the  social attraction $\beta = 400$ and diffusion $\epsilon = 4$.* (NB: we are running more replicated to get stronger results / narrower error bars here)
 
@@
 
-We added trends to the stochastic process described above, adding 30 years post-stabilization with slower (0.25 / year) and moderate (0.5 / year) shifts in the peak of resource, with the goal of assessing the ability of the process to keep up with climate change as a function of the reference memory parameter (Figure 7).
-Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change.
-However, over-reliance on recent memory is also sub-optimal, and - in fact - a balancing of recent and reference memory (in our relatively narrow set of similutions with a $\kappa$ value between 0.8 and 0.9) helps the system keep up with climate change better than chasing each recent shift.
-Note also, that at the highest levels of stochasticity ($\psi_x = 12$), the system fails completely to keep up with the climate change.
-The negative spatial adaptation index indicated that the migration behavior, in fact, can completely fall apart.
+We added trends to the stochastic process described above, adding 30 years of spatial outward trend in the resource.
+Our goal was to assess the ability of the process to keep up with climate change as a function of the reference memory parameter (Figure 8).
+Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change, leading to an adaptation index of 0.
+However, in many cases a balancing of recent and reference memory ($\kappa$ value between 0.6 and 0.8) in many cases was slightly but significantly better than relying entirely on recent memory.
+The smaller spatial scale (in the selected parameter space) does a generally better job than the larger spatial scale at lower stochasticity.
+At higher level of stochasticity, however, the larger spatial scale outperforms the smaller spatial scale, which completely loses track of the climate change.
 
-![](figures/TrendStochasticity_v2.png)
+![](figures/TrendStochasticity.png)
 
-**Figure 7**.
-*Role of reference memory in adaptying to climate change for increasingly stochastic resource dynamics.
-We ran the model with slow (mean shift 0.25 / year; blue colors) and moderate outward trend (0.5 / year; red colors), at four levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels).
+**Figure 8**.
+*Role of reference memory in adapting to climate change for increasingly stochastic resource dynamics.
+We ran the model with a moderiate rate of climate change (mean shift: 0.5 / year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels).
 For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.
@@
 The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity.
-We note that our model did not include any selection, inheritance or birth or death processes, i.e.
-there was evolutionary optimization of parameters to maximize a fitness, as in several somewhat similar models that explicitly explore the evolution of migration [@Guttal2010;@Shaw2013;@Anderson2013].
-Rather, we postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior.
+We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior.
 The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.
+A few rather general conclusions can be made from our model.
+One is that migration in general can be acquired, as has been demonstrated for translocated ungulates [@Jesmer2018], even in environments where resources are difficult to "surf", but that this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years.
+On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.
+increasing stochasticity, rapidly shifting resources, or a shift in some of the system parameters, migration can collapse down into a localized non-migratory, residential behavior, at times with only a marginal cost to the foraging efficiency index.
+This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable [@Berthold1999;@Chapman2011].
 
-It is worth considering the bioloogical interpretations of the various parameters.
-The diffusion parameter ($\epsilon$) captures short time-scaled randomness and exploration of a population and the foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.
-These are the most common ingredients in diffusion-advection models of animal movement and direct parallels can be drawn between these and, respectively, individual random movement models [@Gurarie2011] and resource selection models, which are the fundamental ingredients of empirical analysis of animal movements.
-The sociality parameters, which turn out to be exceedingly important for the resiliency of this process, capture two very important but lesser-studied properties of animal movements.
-The strength of sociality ($\beta$) essentially quantified the importance of the group.
-Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion; and it appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology.
-The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, and was, somewhat unexpectedly, among the most important parameters.
-In particular, a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
-This result points to an important fundamental dynamic: that in order to adapt to any shift in resources, some individuals must experience that resource, and the two sociality-related mechanisms to have that scope is either a relatively large spatial scale or a relatively weak social cohesion.
+We also found that the ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics.
+In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
+Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.
+This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.
 
-The ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics.
-In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
-Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.
-Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.
+The sociality parameters - in particular, the spatial scale $\lambda$ - were somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior.
+In particular, a very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
+In order to adapt to any shift in resources, *some* individuals must experience that resource, and the two sociality-related mechanisms to have that scope is either a relatively large spatial scale or a relatively weak social cohesion.
+On the other hand, overly large spatial scales compromized the ability of the process to track climate change, due to a diluting of the population's ability to concentrate over available resource patches when made available.
 
-Our model is somewhat similar to an evolutionary individual-based model [@Anderson2013], which showed that some inherent phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].
+All of the parameters in our model have well-defined biological interpretations.
+The diffusion ($\epsilon$) captures short time-scaled randomness of movement, capturing the exploratory and short-term dispersive behavior.
+The foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.
+These two parameters, the basic ingredients in diffusion-advection models of animal movement, have direct parallels to empirically estimated properties of animal behavior: diffusion is closely related to families of random walk models [@Gurarie2011] while the advective taxis is related to the step- and resource selection functions that are routinely estimated from movement datas [@Potts2020].
+The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, i.e.
+a population-level range.
+The sociality parameter ($\beta$) quantifies the strength an individual's desire to approach the center of the social group.
+This is not a parameter that is typically measured, though it would - in principle - be possible to estimate in a manner analogous to step-selection function using only other conspecifics as opposed to environmental variables as a covariate.
+Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion.
+It does appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology.
+Finally, the migration timing and location parameter can be similarly straightforwardly estimated from movement data [@Cagnacci2015;@Gurarie2019].
+Thus, for example, @Gurarie2019 explicitly esimated the ranging area, timing, and seasonal range locations for migratory caribou (*Rangifer tarandus*), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration.
+Finally, we note that our  underlying assumption of the migration "urge" is consistent with the strong endogenous programs to migrate taht is well-documented, in particular, in birds that exhibit a seasonal restlessness known as *Zugunruhe* [@Berthold1999;@Helm2006].
 
@@
 In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior.
-Synchrony of migration timing and high site fidelity are well documented for migratory species [@Gurarie2019;@Joly2021] **more**.
+Synchrony of migration timing and high site fidelity are well documented for migratory species [@Gurarie2019;@Joly2021].
+We note, however, that diffusion-advection models can also be interpreted as a probabilistic description of a single individual's movement.
+In this case, $\lambda$ would correspond to an individual home-range and $\beta$ would be an individual's tendency to be drawn to the center of that home range, akin to an individual migratory Ornstein-Uhlenbeck process [@Gurarie2017].
 
-*Some comments on systems that may or may not be more resilient to climate change based on our results*.
+The reference memory parameter $\kappa$ is, of course, impossible to observe directly.
+Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity.
+In general, our model allows us to consider broadly the risks.
+
+Importantly, our model did not include any selection, inheritance or birth or death processes, i.e.
+there was no evolutionary maximization of a fitness, as in many models that explicitly explore the evolution of migration [e.g.
+@Guttal2010;@Shaw2013;@Anderson2013].
+Nonetheless, it is instructive to compare the fundamental assumptions behind such evolutionary individual-based models and a social memory-based model like ours.
+For example, @Anderson2013 similarly explored the resilience of a population under selective pressure under persistent trends and increased stochasticity of a drifting optimal resource window, showing that a certain amount of heritable phenotypic plasticity is necessary to adapt successfully to climate change even at the cost of efficiency.
+Our model underscores the fact that some level of resilience and adapatability can be attained with a purely cognitive process that balances sociality, long, and short term collective memory.
+Importantly, this knowledge must be transmitted through social and cultural, rather than genetic, pathways.
+The high level of sociality among migratory animals, as well as multi-annual parent offspring bonds, are an evident pathway for that kind of transmission.
+As with those evolutionary models, however, it is clear that when changes are too rapid, no amount of cognition can help entirely mitigate against adverse outcomes.
+Furthermore, if behaviors are not sufficiently plastic (i.e.
+if $\kappa$ is too close to 1), then adaptation is that much more difficult.
+
+Rapid environmental change, both global warming and increased anthropogenic development, is causing severe and dramatic impacts to the widespread and generally successful strategy of seasonal migration for many taxa, and the fate of many animal migrations is a topic of increasing concern [@Kauffman2021;@Wilcove2008].
+The ability of animals to respond to these changes is deeply dependent on their behavioral plasticity and cognitive abilities.
+The importance of those abilities are in direct proportion to the difficulty in studying them directly.
+By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adapatability and resilience.
+
+# Acknowledgments
+
+NSF grant.  Quentin at SESYNC.
 

### 2021-07-13 text/ms.tex
98% there

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[12pt]{article}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[left=1in,top=1.25in,right=1in,bottom=1.25in]{geometry}
\usepackage[running]{lineno}

\usepackage{authblk}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{graphicx}
\usepackage{color}				
\usepackage{multirow}	
\usepackage{setspace}
\usepackage[colorlinks]{hyperref}
\hypersetup{citecolor={blue}}
\hypersetup{urlcolor={blue}}
\hypersetup{linkcolor={blue}}

\usepackage{lineno}
\usepackage{graphicx}
\makeatletter

\usepackage[]{natbib}
\bibliographystyle{ecology}

\title{Memories of migrations past: Sociality and cognition in dynamic,
seasonal environments}

\author[a,b]{Eliezer Gurarie\footnote{Corresponding author: \texttt{egurarie@umd.edu}}}
\author[a]{Sriya Potluri}
\author[c]{G.~Christopher Cosner}
\author[c]{R.~Stephen Cantrell}
\author[a,b]{William F.~Fagan}
\renewcommand\Authands{ and }

\affil[a]{\,\, Department of Biology, University of Maryland, College Park, MD 20742, USA.}
\affil[b]{\,\,Socio-Economic Synthesis Center, SESYNC, Annapolis, MD, USA.}
\affil[c]{\,\,Department of Mathematics, The University of Miami, Coral Gables, FL 33146}
\date{\today}

\begin{document}
\onehalfspacing

\maketitle

\linenumbers

\abstract

\emph{to do.}

\section{Introduction}

Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}. For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction \citep{Avgar2014}.

Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by a remembered anticipation \citep{Bracis2017}.

Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}. Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change. Because the scenarios underlying migration are multifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.

Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}. These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component. While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics.

Perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy. The interacting role of memory and sociality, in contrast, has been comparatively little studied.

Here, we develop a diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}. Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, including both variability or trends in resource distribution or phenology reflecting climate change.

We anticipate that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments. Over-reliance on either strategy should be maladaptive. We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic (see also \citet{Lin2021}).

\section{Methods}

\subsection{Memory movement model}

In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or \emph{reference}) memory of large-scale movement behavior, (4) a short-term (or \emph{working}) memory that updates movement behavior based on recent experience, and (5) some social aspect to the learned behavior.

A diffusion-advection equation provided a computationally efficient and versatile framework for examining just such a system. We consider a population moving in one dimension in a constrained domain $D$ and distributing itself according to the following equation:

\begin{equation}\label{eq_mainmodel}
{\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
\alpha \frac{\partial}{\partial x}\left(u \frac{\partial h}{\partial x}\right) + 
\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + 
v_m(t)
\end{equation}

\noindent where $u$ represents the population distributed in time and space. The first term is the diffusion term, capturing the fast time-scale exploration and ``random'' movements of individuals, with $\varepsilon$ is the diffusion rate.

The second term represents the attraction to a dynamic resource $h$, with the proportionality of the advection to the gradient of the resource given by the parameter $\alpha$ (note, the population and resource distributions are functions of both space and time $u(x,t)$ and $h(x,t)$ - we omit the dependent variables in the notation for brevity). This is the well-studied standard chemotaxic resource-following behavior. We borrow the general notation from our earlier related work \citep{Fagan2017, Fagan2019}.

The third term captures the collective or social advection term of the population via a non-local, density dependent function $v_s(u,x)$. If this function takes the form of a convolution around a non-local kernel $k$, i.e.\textasciitilde{} $v_s(u) = k(x) * u(x)$ and that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}. We use the kernel analyzed by \citet{Mogilner1999}: $k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$ The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \widehat{u}$, and in a negative direction when $x > \widehat{u}$. The parameter $\lambda$ is a length scale of sociality - roughly one-half the size of the ``swarm'', where $\beta$ is a parameter that quantifies the overall strength of sociality.

Finally, the last term captures the direct advection that emerges from a memory-driven migratory behavior. This term evolves with a set of parameters $\theta_y$ that slowly change each year $y \in \{0,1,2,...\}$, i.e.~the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.

The migration speed is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (nominally, winter and summer) $t_1$, $dt_1$, $t_2$, $dt_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$. Thus the remembered migratory speed term is a simple step function given by:

\begin{equation}
v_m(t, \theta_y) = \begin{cases}
0; & t > t_1  \,\, \text{and} \,\, t \leq t_1 + dt_1 \\ 
s_{12}; & t > t_1 + dt_1 \,\, \text{and} \,\,  t \leq t_2 \\ 
0; & t> t_2 \,\, \text{and} \,\, t \leq t_2 + dt_2 \\ 
s_{21}; &  t > t_2+dt_2 \,\, \text{or} \,\, t \leq t_1\\ 
\end{cases}
\end{equation}

\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + dt_1$, arrive at $x_2$ at $t = t_2$, i.e.~$s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}$ and $s_{21} = \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}$. This step-like migration function is a one-dimensional version of the migration parameters estimated in empirical studies for individuals \citep{Gurarie2017} and, more relevantly, for populations \citep{Gurarie2019} in empirical studies.

We consider these six parameters to be the known or remembered determinants of the migratory behavior, with an initial set $\theta_0$ determining the reference migration behavior. This reference migration is updated each year by the experience of the population. To perform this updating, we estimate a new set of parameters $\widehat{\theta_y}$ after each year, and combine these new parameters with the reference parameters according to the following weighted mean:

$$\theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y}$$

\noindent where each of the six parameters is updated identically. The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$). The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory. When $\kappa = 0$, all of the actionable memory is from the preceding year. When $\kappa = 1$, the actionable memory is entirely the reference memory.

The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.~${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$. As there are no birth or death processes, the total population remains fixed and constant, for convenience integrating to 1. Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process. Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'\,' an intra- and interannually dynamic, seasonal environment.

\subsection{Seasonal resource}

We solved the model numerically on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.\textasciitilde years of 100 days). We were interested in an approximately periodic resource dynamic, i.e.~one in which $h(x,t \approx h(x, t-\tau))$. We generated two types of resource distributions. A ``non-surfable'' resource (\emph{island resource}), and weakly surfable resource (\emph{drifting resource}). Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 30 and -30 at times 25 and 75, respectively). These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively. The island resource is simply two uncorrelated bivariate normal
distributions 
$$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$ 
\noindent where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e.~$\frac{1}{\tau} \int_T\int_X h(x,t) dx\,dt = 1$.

The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1 ($\int_X h(x,t) dx = 1$). This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details). Both types of resources are illustrated in Figure \ref{fig_ResourceExamples}

Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$. However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse. Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses. These trends mirror the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate. The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.

\begin{figure}[t!]
\includegraphics[width = \textwidth]{figures/ResourceExamples.png}

\caption{\label{fig_ResourceExamples}  Examples of various seasonal resource distribution functions, contrasting short duration, but wide pulses ($\sigma_t = 3, \sigma_x = 12$; left panels), long duration but spatially concentrated pulses ($\sigma_t = 12, \sigma_x = 3$; right panels), and isolated resource pulses (upper panels) from the weakly drifting resource (lower panels). The total amount of resource is identical across all scenarios. In the weakly drifting resources, the total amount is constant at all times, and uniform in the middle of the phase (time = 0, 50, 100).}
\end{figure}

\subsection{Metrics}

The main metrics we are interested in are \emph{migration mismatch}, \emph{foraging efficiency} and \emph{adaptation to trends}.

Migration mismatch captures the similarity between the migration phenology and the resource phenology. To do this, we compute the total mismatch, i.e.~is a sum of the difference between the migration coefficients and the resource peaks. Thus, the spatial mismatch is given by the absolute difference between the migration targets and the resource peaks, i.e.~$MM_x = |x_1 - m_x| + |x_2 + m_x|$. The temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.~$MM_t = max\{t_1 - m_1, m_1 - (t_1 + dt_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + dt_2), 0\}$. The total mismatch is the sum of $MM_x$ and $MM_t$. A mismatch of less than 1 is essentially perfect, we consider a mismatch of 1-5 to be ``good'', and beyond 30 the system can be said to have failed to keep up with climate change.

To quantify the foraging efficiency, i.e., the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya Coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions. We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is:
$$FE = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
\noindent where the spatial integral is taken over the domain. This metric is constrained to be between 0 and 1.

For simulations with a constant resource, we ran the model until a quasi-equilibrium state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of 1-1e-5. Once that state was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.

For numerical runs with climate change, we first run a simulation for until equilibrium is attained, and then begin shifting the location of the resource with a trend (slow 0.25, medium 0.5, or rapid 1 units per year), or by adding stochasticity (s.d. 3, 6, 12). For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this as the \emph{spatial adaptation} (SA) index. An SA equal to 1 suggests that the process is keeping up with climate change, an SA of 0 indicated that the process is not responding at all to climate change.  Values greater than 1 (super-adaptation) are possible, as are values less than 1, which correspond to a loss of the migration behavior.

\subsection{Simulation studies}

We explored this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation \ref{eq_mainmodel} with the Runge-Kutte algorithm using the \texttt{deSolve} \citep{Soetaert2010} and \texttt{ReacTran} \citep{Soetaert2012} packages in R. We additionally used the \texttt{nlsLM} function in package \texttt{minpack.LM} \citep{minpack.lm} for robust and fast annual estimation of the migration parameters. The complete code is available as an R package (\texttt{memorymigration}) available on GitHub at \url{https://github.com/EliGurarie/memorymigration} and as an interactive Shiny application at \url{https://url.to.be.determined}.

We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions: (1) Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation? (2) Can this model acquire a migratory behavior from a non-migratory initial condition? (3) What is the role of a reference memory for dealing with stochastic resource dynamics? (4) Can this model adapt when the resource peaks shifts in space?

\begin{table}[b!]
	\caption{\label{tab_defs} Table of parameters, variables and metrics.}

	\centering
	\begin{tabular}{ll|l}
		\hline\hline
		\multicolumn{3}{l}{\textbf{Memory migration model}}\\ 
		& $\epsilon$ & Diffusion \\ 
		& $\alpha$ & Forage following \\ 
		& $\beta$ & Strength of sociality \\ 
		& $\lambda$ & Spatial scale of sociality \\ 
		& $\kappa$ & Proportion of reference versus working memory \\ 
		& $x_1$, $x_2$ & location of population centroids in summer and winter \\ 
		& $t_1$, $dt_1$ & start and duration of summer season \\ 
		& $t_2$, $dt_2$ & start and duration of winter season \\ 
		& $\kappa$ & proportion memory allocated to reference \\
		&& (long-term) memory vs. working (short-term) memory \\ 
		\multicolumn{3}{l}{\textbf{Resource dynamics}} \\ 
		& $\tau$ & duration of period (year) \\ 
		& $m_x$, $-m_x$ & spatial coordinate of resource peak for summer and winter \\ 
		& $m_t$, $\tau - m_t$ & timing of resource peak for the summer and winter \\ 
		& $\sigma_x$, $\sigma_t$ & time duration and and spatial scale of resource pulse \\ 
		& $dx$, $dt$ & rate of change of peak location and timing of resource \\ 
		& $\psi_x$, $\psi_t$ & standard deviation of peak location and timing \\ 
		\multicolumn{3}{l}{\textbf{Metrics}}\\ 
		& $MM_x$ & spatial migration mismatch \\ 
		& $MM_t$ & temporal migration mismatch \\ 
		& $TE$ & total mismatch \\ 
		& $FE$ & foraging efficiency \\ 
		& $SA$ & spatial adaptation index \\ 
		\hline\hline
	\end{tabular}
\end{table}

\section{Results}

\begin{figure}
\includegraphics[width = 0.9\textwidth]{figures/example1_adaptation.png}

\caption{\label{fig_adaptation} Example of adaptation to a shift in resource peak. The initial (year 0) behavior migrates to locations 50 and -50 at days 15 and 60, whereas the resource peak is at 30 and -30, peaking at times 25 and 75. The panels show (a) the first 14 years of the simulation; (b) the centroid of the annual movement of the population is shown in panel b, with dark blue to red colors indicating year 0 to year 40; (c) annual foraging efficiency across years; (d) migration timing parameters for each year, with orange segments indicating arrival and departure from the summering (northern) grounds, and the blue segments indicating timing of arrival and departure at the wintering grounds; (e) migration arrival and departure location across years, with blue and orange indicating winter (southern) and summer (northern) locations.}
\end{figure}

\subsection{Adaptation to resource phenology}

The ability of this system to attain a stable, migratory state that matches the dynamics of the resource is illustrated in figure \ref{fig_adaptation}. In the illustrated scenario, it takes nearly 40 years to attain an equilibrium, and the eventual steady state is one where the centroid of the migration lines up exactly with the centroid of the resource, and the arrival timing coincides with the \emph{peak} of resource availability. Notably, the path to this equilibrium is somewhat indirect, with the later winter range taking more time to stabilize than the earlier summer range. The eventual steady state is one where the foraging efficiency is relatively high.

We ran this process for 8100 parameter combinations (figure \ref{fig_phenologymatching}). The fundamental dynamic of attaining and maintaining a well-matched migration phenology, can occur under many combinations of parameter values, but all parameters play a role. Among the more intuitive results are that greater values of $\alpha$ (resource following) lead to an improved ability to match the migration. Resource peaks with larger spatial extent (higher $\sigma_x$) are generally better for migration matching. A larger set of parameters matched migration with low diffusion than high diffusion.

Less intuitive was the high importance of the sociality parameters, in particular the spatial scale of the swarming. Higher levels of social attraction ($\beta$) led to improved migration matching except in those cases where the sociality scale $\lambda$ was high. Thus, for example, at $\lambda = 20$, no simulations at $\beta >= 200$ managed to acquire or maintain a matched migration. However, at $\lambda = 50$ or $100$, the migration was slightly better matched at high values of $\beta$. (Fig. \textbf{X}) The spatial extent of the swarm was a remarkably significant variable. Smaller swarms were able to match migration only at low values of social attraction ($\beta = 200$), and relatively high values of resource attraction ($\alpha \geq 600$).

Random forest analyses, whether on the log of total mismatch or on the classification of a perfect match, uniformly show that the most important variables \citep{Breiman2001} were $\alpha$ and $\lambda$ (4.14 and 4.02 proportional increase in MSE), and the least important was $\sigma_t$, with a 0.5 proportional increase (table \textbf{X}).

Overall, foraging efficiency was strongly correlated with migration matching, as expected. At high mismatch (\textgreater{} 50), foraging efficiency was low (mean 0.29, s.d. 0.16) compared to the near-perfect matching migrations (mean 0.58, s.d. 0.14). However, somewhat higher mismatch (1 to 5) showed an even higher overall foraging efficiency (mean 0.62, s.d. 0.18 - see also figure \ref{fig_boxplotFE}).

\begin{figure}
\includegraphics[width = \textwidth]{figures/StabilityResults.png} 
\caption{\label{fig_phenologymatching} Migration phenology matching across six model parameters. Low and high diffusion ($\epsilon = 1$ and $8$ in upper and lower panel blocks), tight, medium and loose swarms ($\lambda = 20, 50, 100$) left to right panels. Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top. Within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right. The color scheme reflects the total mismatch, i.e.~the sum of the absolute differences between the migration timing and locations from the resource peak. White squares represent runs that numerically failed to estimate migration parameters.}
\end{figure}

\begin{figure}
\includegraphics[width = \textwidth]{figures/ForagingEfficiency.png}
\caption{\label{fig_boxplotFE} Box-plots of foraging efficiency against mismatch across several values of foraging patch duration.}
\end{figure}

\subsection{Learning to migrate}

\begin{figure}
\includegraphics[width = 0.9\textwidth]{figures/example2_learningtomigrate.png}

\caption{\label{fig_learningtomigrate} Example of model learning to migrate. The resource is a
``weakly drifting'' resource and the initial (year 0) condition is non-migratory. The simulation was run for 100 years, and a sampling of those years (labeled) are presented in panel a: all years from 0 to 10, followed by 20, 40 and 100. Otherwise, panels are as in figure \ref{fig_adaptation}. Parameter values were $\epsilon = 5$, $\alpha = 500$ and $\beta = 50$.}
\end{figure}

Figure \ref{fig_learningtomigrate} illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50. In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$). The qualitative behavior of this process was to start drifting towards the summer resource, while slowly developing a weak pulse towards the winter resource as well. After first locking in on the summer resource, the winter migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource can be consistently reached.

The model had, in general, a difficult time learning migration from a non-migratory initial condition. Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5. Conditions that were more conducive to learning migration were pulses of \emph{longer} duration (high $\sigma_t$), but \emph{smaller} in scope (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact in space but long enough in duration to lock in to the memory.

\subsection{Adapting to climate change}

\begin{figure}
\includegraphics[width = \textwidth]{figures/SpatialClimateChange.png}

\caption{\label{fig_adaptationclimate} Adaptation to a steadily drifting resource. In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively). The y-axis is the index of spatial adaptation (SA), i.e.~the trend of the memory-driven migration divided by the resource drift trend. Values near 1 indicate a behavior that keeps up with climate change, values near indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend. We compare across spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels). The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations. The bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.}
\end{figure}

To assess the ability of the system to adapt to a trending climate, we generated scenarios with slow (0.25 units / year), medium (0.5 units / year), and fast (1 unit / year) drift outward of the two resource pulses. We then assessed 40 parameter combinations for each of those scenarios, a high and low value of resource following ($\alpha = 400$ and $100$) crossed with a high and low value of sociality ($\beta = 400$ and $100$) crossed with 10 values of the spatial scale of sociality ($\lambda = 20$ to $200$). The spatial and temporal scale of the resource pulses were fixed to a relatively ``easy'' to adapt to $\sigma_x = 12$ and $\sigma_t = 6$. We computed the adaptation index and foraging efficiency for each of the 120 runs (figure \ref{fig_adaptationclimate}). We were interested in the dynamics against $\lambda$ due to the uniformly high importance of this parameter for determining the success of this process to match migration in steady states.

As figure \ref{fig_adaptationclimate} shows, higher values of resource following ($\alpha = 400$; orange circles) are nearly universally better for keeping up with climate change (SA values near 1). Furthermore, when combined with high sociality ($\beta = 400$; right panels), nearly all parameter combinations do a good job keeping up with climate change (SA values ranging between 0.53 and 0.85 for a swarm size greater than 50). However, that maximum value is still less than 1, suggesting that truly matching a steadily drifting trend is very difficult. Small swarms ($\lambda < 50$) have a very hard time adapting when the social attraction is high, but do fairly well when social attraction is lower. But larger sized swarms do progressively worse across more parameterizations, e.g.~in the most rapid climate change scenario, the SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to 200 (essentially, the entire spatial domain).

A rather more dramatic pattern is visible for the lower foraging attraction scenario ($\alpha = 100$; blue circles). Notably, no parameter combination at this value comes close to keeping up with the rapid climate change (SA range -0.64 to 0.13). For slower climate change, however, there is a window of values for the swarm size between 40 and 80, where the SA \emph{exceeds} 1, and then crashing quite rapidly to negative values of SA as that swarm size increases. These ``super-adaptive'' processes indicate a unique sweet spot where a swarm is large enough to capture and adapt to the drifting resource, but not so large that the information gathered in a given year is too weak to adjust the migratory behavior in a following year.

As anticipated, better adaptation to the drifting resource correlated strongly with higher foraging efficiency (inset boxplots).

\subsection{Reference memory and stochasticity}

While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, we hypothesized that more conservative approach, relying on a ``reference memory'' may be more beneficial when conditions change stochastically. We tested this hypothesis by solving a set of models across a range of $\kappa$ values fom 0 (all recent memory) to 1 (all reference memory). In these scenarios, we ran the system for as many years as needed with no stochasticity to acquire a stationary state (i.e.~similarity index greater than 1-1e-6). We then used the stationary state as the reference memory, and then ran the process for an additional 50 years with a stochasticity (i.e.~standard deviation in peak location of the resource) ranging from 0 to 12.

Figure \ref{fig_stochasticity} illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency, with some variation across the spatial scale of sociality. When that scale of sociality is high enough (e..g $\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run. Overall, as anticipated, the greater the stochasticity, the lower the foraging efficiency; and the greater the stochasticity, the greater the benefit of a conservative reference memory driven strategy.

\begin{figure}
\includegraphics[width=\textwidth]{figures/Stochasticity.png}

\caption{\label{fig_stochasticity} Foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$. For the stochastic processes ($\psi_x > 0$), the process was run 90 times, and the mean and standard error of the foraging efficiency over a 50 year post-stabilization period is presented (points and bars). In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\epsilon = 4$.}

\end{figure}

\subsection{Stochasticity and trends}

We added trends to the stochastic process described above, adding 30 years of spatial outward trend in the resource. Our goal was to assess the ability of the process to keep up with climate change as a function of the reference memory parameter (figure \ref{fig_trendstochasticity}). Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change, leading to an adaptation index of 0. However, in many cases a balancing of recent and reference memory ($\kappa$ value between 0.6 and 0.8) in many cases was slightly but significantly better than relying entirely on recent memory. The smaller spatial scale (in the selected parameter space) does a generally better job than the larger spatial scale at lower stochasticity. At higher level of stochasticity, however, the larger spatial scale outperforms the smaller spatial scale, which completely loses track of the climate change.

\begin{figure}

\includegraphics[width=\textwidth]{figures/TrendStochasticity.png}

\caption{ \label{fig_trendstochastity} Role of reference memory in adapting to climate change for increasingly stochastic resource dynamics. We ran the model with a moderiate rate of climate change (mean shift: 0.5 / year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels). For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory. Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and diffusion $\epsilon = 4$.}

\end{figure}

\section{Discussion}

Animals navigate complex, dynamic, patchy environments. When there is a strongly localized and seasonal component to the resource dynamics, typical strategies of spatial distribution by straightforward resource-following taxis necessarily fails to maximally exploit the dynamics. It is in these cases - quite common in the natural world -- that seasonal migration becomes a viable, even necessary, strategy. However, when those resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity. It is our conjecture that this plasticity is greatly facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years. To our knowledge, this study is the first attempt to focus in and propose a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics.

We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity. We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior. The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.

Several rather general conclusions can be made from our analyses. One is that migration in general can be acquired, as has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'', but that this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years. On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, or a shift in some of the system parameters, migration can collapse down into a localized non-migratory, residential behavior, at times with only a marginal cost to the foraging efficiency index. This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}.

We also found that the ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics. In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space. Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic. This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.

The sociality parameters - in particular, the spatial scale $\lambda$ - were somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior. In particular, a very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak. In order to adapt to any shift in resources, \emph{some} individuals must experience that resource, and the two sociality-related mechanisms to have that scope are either a relatively large spatial scale or a relatively weak social cohesion. On the other hand, overly large spatial scales compromized the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches.

All of the parameters in our model have well-defined biological interpretations. The diffusion ($\epsilon$) captures short time-scaled randomness of movement, capturing the exploratory and short-term dispersive behavior. The foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale. These two parameters, the basic ingredients in diffusion-advection models of animal movement, have direct parallels to empirically estimated properties of animal behavior: diffusion is closely related to families of random walk models \citep{Gurarie2011} while the advective taxis is related to the step- and resource selection functions that are routinely estimated from movement data \citep{Potts2020}. The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, i.e.~a population-level range. The sociality parameter ($\beta$) quantifies the strength an individual's desire to approach the center of the social group. This is not a parameter that is typically measured, though it would - in principle - be possible to estimate in a manner analogous to step-selection function using only other conspecifics as opposed to environmental variables as a covariate. Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion. It does appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology.

Finally, the migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}. Thus, for example, \citet{Gurarie2019} explicitly esimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration. Finally, we note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that is well-documented, in particular, in birds that exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.

Indvidual-based models have shown that Collective knowledge has been shown to be important, if not essential, to the evolution and process of migration \citep{Shaw2013, Guttal2010, Berdahl2018}. In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior. Synchrony of migration timing and high site fidelity are well documented for migratory species \citep{Gurarie2019, Joly2021}. We note, however, that diffusion-advection models can also be interpreted as a probabilistic description of a single individual's movement. In this case, $\lambda$ would correspond to an individual home-range and $\beta$ would be an individual's tendency to be drawn to the center of that home range, akin to an individual migratory Ornstein-Uhlenbeck process \citep{Gurarie2017}.

The reference memory parameter $\kappa$ is, of course, impossible to observe directly. Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity. In general, our model allows us to consider broadly the risks.

Importantly, our model did not include any selection, inheritance or birth or death processes, i.e.~there was no evolutionary maximization of a fitness, as in many models that explicitly explore the evolution of migration \citep[e.g.][]{Guttal2010, Shaw2013, Anderson2013}. Nonetheless, it is instructive to compare the fundamental assumptions behind such evolutionary individual-based models and a social memory-based model like ours. For example, \citet{Anderson2013} similarly explored the resilience of a population under selective pressure under persistent trends and increased stochasticity of a drifting optimal resource window, showing that a certain amount of heritable phenotypic plasticity is necessary to adapt successfully to climate change even at the cost of efficiency. Our model underscores the fact that some level of resilience and adapatability can be attained with a purely cognitive process that balances sociality, long, and short term collective memory. Importantly, this knowledge must be transmitted through social and cultural, rather than genetic, pathways. The high level of sociality among migratory animals, as well as multi-annual parent offspring bonds, are an evident pathway for that kind of transmission. As with those evolutionary models, however, it is clear that when changes are too rapid, no amount of cognition can help
entirely mitigate against adverse outcomes. Furthermore, if behaviors are not sufficiently plastic (i.e.~if $\kappa$ is too close to 1), then adaptation is that much more difficult.

Rapid environmental change, both global warming and increased anthropogenic development, is causing severe and dramatic impacts to the widespread and generally successful strategy of seasonal migration for many taxa, and the fate of many animal migrations is a topic of increasing concern \citep{Kauffman2021, Wilcove2008}. The ability of animals to respond to these changes is deeply dependent on their behavioral plasticity and cognitive abilities. The importance of those abilities are in direct proportion to the difficulty in studying them directly. By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adapatability and resilience.

\section{Acknowledgments}

EG, WFF, GCC and RSC, were supported in part by NSF award DMS 1853478. EG and WFF were further supported by NSF grant XXXXXXXXXX. The authors are grateful to Quentin Read for computational support and advice for cluster-based multi-node multi-core model runs.

\nolinenumbers

\bibliography{../library/memorymigration.bib}

\linenumbers

\clearpage
\begin{center}
	\Large Supplementary Materials
	\end{center}

\normalsize
\appendix
\section{Drifting resource}
The drifting resource function has the following properties

\begin{enumerate}
\item The total amount of resource across space is constant throughout the year.
\item At the beginning, middle, and end of the year the resource is uniformly distributed. 
\item At some peak time $\mu_t < \tau/2$, the resource concentrates at a location $\mu_x < \chi$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
\item The resource peaks exactly symmetrically at time $\tau - \mu_t$ and location $-\mu_x$ with the same variances.
\end{enumerate}

To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.  Thus:

$$h(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$

where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:

$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$

where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.  These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.

The means and variances themselves are Gaussian pulses, with the mean peaking at $\mu_x$ at time $\mu_t$ with standard deviation $\sigma_t$ and at $-\mu_x$ at time $\tau - \mu_t$ and the standard deviation pulsing from $2 \chi/\sqrt{12}$ (corresponding to a uniform distribution over the domain $-\chi$ to $\chi$) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.

\end{document}

### 2021-07-13 text/ms_backup.tex
98% there

\documentclass[12pt]{article}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Memories of migrations past: Sociality and cognition in dynamic, seasonal environments},
  pdfauthor={E. Gurarie, S. Potluri, C. Cosner, R. Cantrell, W.F. Fagan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Memories of migrations past: Sociality and cognition in dynamic,
seasonal environments}
\author{Eliezer Gurarie, Sriya Potluri, Cris Cosner, Robert Cantrell, William F. Fagan}
\date{2021-07-12}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

\emph{to do.}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Seasonal migrations are extremely widespread among terrestrial, aquatic,
avian and invertebrate species (Dingle 2014). For many species,
migration is an extremely successful strategy, allowing a far greater
number of individuals to inhabit landscape which might not otherwise be
able to support large numbers year round (Fryxell et al. 1988). The
ultimate evolutionary cause of seasonal migration, essentially, is that
the energetic and survival related costs of migration are outweighed by
the fitness benefits of accessing suitable seasonal resources, whether
those are for energetic gain, predator avoidance, a suitable physical,
biotic or social environment for reproduction (Avgar et al. 2014).

Much as the ultimate causes of migration vary across species, the
proximate causes, drivers and mechanism can vary considerably across and
even within species (Berthold 1999, Shaw 2016). Some migrants follow a
``green wave'' of spring vegetation as it flowers across altitudinal or
latitudinal gradients (Bischof et al. 2012, Kölzsch et al. 2015). These
migrations can be considered ``tactical,'' as they can occur - as an
extreme simplification - purely as response to local conditions. Other
migrants perform long-distance migrations in anticipation that critical
resources will be available at the time of arrival at the end point of
migration (Abrahms et al. 2019). This second behavior involves the
greatest trade-off between the costs of migration against the benefits
of accessing resources, whether food, suitable habitat for breeding, or
predator refuge, that are highly seasonal and localized. This approach
can be considered ``strategic'' in the sense that it is driven not by
immediate cues but by a remembered anticipation (Bracis and Mueller
2017).

Migratory species are often considered to be more vulnerable to
environmental change, as a disruption in either of the seasonal ranges
or along a migratory corridor can have significant negative impacts
(Wilcove and Wikelski 2008, Kauffman et al. 2021). On the other hand, it
has been argued that migratory species might be more resilient to
disruptions due to their wide-ranging mobility (Robinson et al. 2009).
Clearly, the ability to perform a migration without local cues is only
possible if the behavior is hard-programmed or remembered. On the other
hand, a strictly programmed behavior can be maladaptive if conditions
change. Because the scenarios underlying migration are multifold and
complex, mathematical modeling may provide some insights and help
clarify where, when, and under what conditions we might expect different
kinds of migrations to operate.

Diffusion-advection models have a long pedigree in animal movement
modeling (Skellam 1951, Turchin 1998, Okubo and Levin 2001). These
models are grounded in the general idea that animal movements - somewhat
like movements of physical particles - combine a random (diffusive)
component with a directed (advective) component. While direct
relationships between diffusion models and movement data are somewhat
tenuous (Gurarie and Ovaskainen 2011), as a theoretical tool for
exploring processes they are invaluable for their versatility and the
relative ease of numeric computation of the partial differential
equations (PDEs) that are used to describe diffusion-advection models
mathematically. Thus, much theoretical and some applied work has been
done on refining the basic assumptions of diffusion models, e.g.~by
including heterogeneity in populations (Skalski and Gilliam 2003,
Gurarie et al. 2009), fat-tailed dispersal kernels (Kot et al. 1996),
non-linear or otherwise complex responses to resources and
consepecifics.

Perhaps the greatest difference between animals and randomly moving
passive particles described by diffusion models is cognition, including
social behavior and memory. Refinements to diffusion-advection equations
have revealed conditions under which non-local information gathering
(Fagan et al. 2017) and behavioral switching may confer foraging
advantages (Fagan et al. 2019), in particular when resources are dynamic
and patchy. The interacting role of memory and sociality, in contrast,
has been comparatively little studied.

Here, we develop a diffusion-advection model with memory to explore the
resilience of a migratory population under various dynamic, seasonal
resource distributions. In formulating the model, our goal is to
identify the minimum set of movement and memory parameters required to
generate an adaptive, migratory behavior. This includes the ability to
learn to migrate from non-migratory initial conditions, simulating the
release of naive animals in a seasonal environment (Jesmer et al. 2018),
or to lose the propensity to migrate if the resource distribution does
not require it, also a commonly observed phenomenon (Wilcove and
Wikelski 2008). Our ultimate goal is to assess the role of long and
short-term memory in the resilience (or fragility) of a migratory
population against changing resource distribution dynamics, including
both variability or trends in resource distribution or phenology
reflecting climate change.

We anticipate that under many conditions a blending of \emph{tactical}
(i.e.~direct response to resource availability or perception) and
\emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will
help foragers navigate dynamic, seasonal environments. Over-reliance on
either strategy should be maladaptive. We further anticipate that a
shorter-term memory updating is needed to navigate trends in resource
spatial distribution and temporal distribution (phenology), but that a
longer-term reference memory is needed to navigate resource
distributions that are stochastic (see also Lin et al. (2021)).

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{memory-movement-model}{%
\subsection{Memory movement model}\label{memory-movement-model}}

In designing our study, our goal was to develop a minimal heuristic in
which the following processes were explicitly modeled: (1) Random or
exploratory movement, (2) attraction to resources, (3) a long-term (or
\emph{reference}) memory of large-scale movement behavior, (4) a
short-term (or \emph{working}) memory that updates movement behavior
based on recent experience, and (5) some social aspect to the learned
behavior.\\
A diffusion-advection equation provided a computationally efficient and
versatile framework for examining just such a system. We consider a
population moving in one dimension in a constrained domain \(D\) and
distributing itself according to the following equation:

$ {\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
\alpha \frac{\partial}{\partial x}\left(u \frac{\partial h}{\partial x}\right) + 
\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + 
v_m(t)
\]

where \(u\) represents the population distributed in time and space. The
first term is the diffusion term, capturing the fast time-scale
exploration and ``random'' movements of individuals, with
\(\varepsilon\) is the diffusion rate.

The second term represents the attraction to a dynamic resource \(h\),
with the proportionality of the advection to the gradient of the
resource given by the parameter \(\alpha\) (note, the population and
resource distributions are functions of both space and time \(u(x,t)\)
and \(h(x,t)\) - we omit the dependent variables in the notation for
brevity). This is the well-studied standard chemotaxic
resource-following behavior. We borrow the general notation from our
earlier related work (Fagan et al. 2017, 2019).

The third term captures the collective or social advection term of the
population via a non-local, density dependent function \(v_s(u,x)\). If
this function takes the form of a convolution around a non-local kernel
\(k\), i.e. $ v_s(u) = k(x) * u(x)$ and that kernel is
odd, an attractive or ``swarming'' behavior can be generated (Mogilner
and Edelstein-Keshet 1999). We use the kernel analyzed by Mogilner and
Edelstein-Keshet (1999):
$ k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$ The convolution
of \(u\) with this kernel has the property of pushing the population in
a positive direction when \(x < \widehat{u}\), and in a negative
direction when \(x > \widehat{u}\). The parameter \(\lambda\) is a
length scale of sociality - roughly one-half the size of the ``swarm,''
where \(\beta\) is a parameter that quantifies the overall strength of
sociality.

Finally, the last term captures the direct advection that emerges from a
memory-driven migratory behavior. This term evolves with a set of
parameters \(\theta_y\) that slowly change each year
\(y \in \{0,1,2,...\}\), i.e.~the count of periods \(\tau\):
\(y = \lfloor t/\tau \rfloor\).

The migration speed is specified by six parameters \(\theta\): the
timing of the start and duration of two anticipated seasons (nominally,
winter and summer) \(t_1\), \(dt_1\), \(t_2\), \(dt_2\), and the spatial
coordinates of the population centroid for each season \(x_1\) and
\(x_2\). Thus the remembered migratory speed term is a simple step
function given by:

$ 
v_m(t, \theta_y) = \begin{cases}
0 &;  & t > t_1 &\text{and} & t \leq t_1 + dt_1 \\ 
s_{12} &;  &  t > t_1 + dt_1 &\text{and} & t \leq t_2 \\ 
0 &; & t> t_2 &\text{and} &t \leq t_2 + dt_2 \\ 
s_{21} &; &  t > t_2+dt_2 & \text{or} & t \leq t_1\\ 
\end{cases}
\]

where the migration speeds \(s_{12}\) and \(s_{21}\) from the respective
ranges are set such that they arrive at \(x_1\) at \(t_1\), leave at
\(t = t_1 + dt_1\), arrive at \(x_2\) at \(t = t_2\),
i.e.~\(s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}\) and
\(s_{21} = \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}\). This step-like
migration function is a one-dimensional version of the migration
parameters estimated in empirical studies for individuals (Gurarie et
al. 2017) and, more relevantly, for populations (Gurarie et al. 2019) in
empirical studies.

We consider these six parameters to be the known or remembered
determinants of the migratory behavior, with an initial set \(\theta_0\)
determining the reference migration behavior. This reference migration
is updated each year by the experience of the population. To perform
this updating, we estimate a new set of parameters
\(\widehat{\theta_y}\) after each year, and combine these new parameters
with the reference parameters according to the following weighted mean:

$ \theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y}\]
where each of the six parameters is updated identically. The estimates
\(\widehat{\theta_y}\) are obtained via a least-squares minimization of
the migration track (\(m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'\))
against the spatial mean of the population process in year \(y\)
(i.e.~\(\widehat{u}(t) = \int_X u_y(t, x) dx\)). The parameter
\(\kappa \in (0,1)\) captures the reliance on that long-term memory.
When \(\kappa = 0\), all of the actionable memory is from the preceding
year. When \(\kappa = 1\), the actionable memory is entirely the
reference memory.

The model is confined to a one-dimensional bounded domain
\([-\chi,\chi]\), with no flux outside of the boundaries
i.e.~\({\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0\).
As there are no birth or death processes, the total population remains
fixed and constant, for convenience integrating to 1. Furthermore, the
parameters remain constant throughout time, with no adaptation or
mutation-selection process. Our interest is entirely in the ability of a
fixed set of memory and movement parameters to ``navigate'\,' an intra-
and interannually dynamic, seasonal environment.

\hypertarget{seasonal-resource}{%
\subsection{Seasonal resource}\label{seasonal-resource}}

We solved the model numerically on a spatial domain
\(x \in [-100,100]\), and a periodicity \(\tau = 100\)
(i.e.\textasciitilde years of 100 days). We were interested in an
approximately periodic resource dynamic, i.e.~one in which
\(h(x,t \approx h(x, t-\tau))\). We generated two types of resource
distributions. A ``non-surfable'' resource (\emph{island resource}), and
weakly surfable resource (\emph{drifting resource}). Both are
characterized by a peak in time and space centered at \(m_x\) at
\(m_t\), and \(-m_x\) at \(\tau - m_t\) (for example, locations 30 and
-30 at times 25 and 75, respectively). These pulses have a shared time
scale of duration \(s_t\) and a spatial scale of extent \(s_x\), the
standard deviation in the time and space dimension respectively. The
island resource is simply two uncorrelated bivariate normal
distributions
$ h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))\]
where \(\Phi\) is the bivariate Gaussian distribution function, and the
normalizing constant \(K\) is selected such that the average total
amount of resource throughout the year is 1,
i.e.~\(\frac{1}{\tau} \int_T\int_X h(x,t) dx\,dt = 1\).

The drifting resource differs from the island resource in that the total
amount of resource at any given time point is 1
(\(\int_X h(x,t) dx = 1\)). This property is attained by distributing
the resource as a re-scaled beta distribution, where the shape and scale
parameters vary sinusoidally in such a way as to make the standard
deviations and means match the desired values of \(m_x, m_t, s_x, s_t\)
(see appendix for details). Both types of resources are illustrated in
Figure 1.

Within a given year, the resource is entirely symmetric:
\(h_y(x,t) = h_y(-x, \tau-t)\). However, in scenarios exploring climate
change we allow the peaks to vary with drift and stochasticity according
to: \(m_x(y) \sim {N}(\mu_x + \beta_x\,y, \sigma_x)\) and
\(m_t(y) \sim {N}(\mu_t + \beta_t\,y, \sigma_t)\), where the \(\mu\),
\(\beta\) and \(\sigma\) terms are the mean, slope and variance,
respectively, for the location and time duration of the pulse. Thus, if
\(\beta=0\) and \(\sigma=0\), the conditions are constant across years,
if \(\beta_x > 0\) there is a drift of the resource towards extremes, if
\(\beta_t < 0\) there is a shift towards earlier resource pulses. These
trends mirror the pole-ward shift of peak resources and the earlier
spring phenology occurring with a warming global climate. The spatial
scales (\(s_x\) and \(s_t\)) remain constant in all of our simulations.

\includegraphics{figures/ResourceExamples.png}

\textbf{Figure 1. Examples of various seasonal resource distribution
functions, contrasting short duration, but wide pulses
(\(\sigma_t = 3, \sigma_x = 12\); left panels), long duration but
spatially concentrated pulses (\(\sigma_t = 12, \sigma_x = 3\); right
panels), and isolated resource pulses (upper panels) from the weakly
drifting resource (lower panels). The total amount of resource is
identical across all scenarios. In the weakly drifting resources, the
total amount is constant at all times, and uniform in the middle of the
phase (time = 0, 50, 100).}

\hypertarget{metrics}{%
\subsection{Metrics}\label{metrics}}

The main metrics we are interested in are \emph{migration mismatch},
\emph{foraging efficiency} and \emph{adaptation to trends}.

Migration mismatch captures the similarity between the migration
phenology and the resource phenology. To do this, we compute the total
mismatch, i.e.~is a sum of the difference between the migration
coefficients and the resource peaks. Thus, the spatial mismatch is given
by the absolute difference between the migration targets and the
resource peaks, i.e.~\(MM_x = |x_1 - m_x| + |x_2 + m_x|\). The temporal
mismatch is the difference between the arrival time and the peak of the
resource if arrival is post-peak, the difference between the departure
time and the peak of the resource if departure is pre-peak, and 0 if the
seasonal duration spans the peak,
i.e.~\(MM_t = max\{t_1 - m_1, m_1 - (t_1 + dt_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + dt_2), 0\}\).
The total mismatch is the sum of \(MM_x\) and \(MM_t\). A mismatch of
less than 1 is essentially perfect, we consider a mismatch of 1-5 to be
``good,'' and beyond 30 the system can be said to have failed to keep up
with climate change.

To quantify the foraging efficiency, i.e., the organisms' ability to
track the distribution of the resources over space and time, we use a
continuous form of the Bhattacharyya Coefficient (Bhattacharyya 1943)
which quantifies the similarity between two distributions. We compute
this coefficient at every time point in a given year, and take the mean
across the equilibrium year to determine foraging efficiency (FE). Thus,
the foraging efficiency index is:
$ FE = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt\]
where the spatial integral is taken over the domain. This metric is
constrained to be between 0 and 1.

For simulations with a constant resource, we ran the model until a
quasi-equilibrium state was achieved, i.e.~where the Bhattacharya index
of the population distribution across subsequent years reached a value
of 0.9999. Once that state was attained, we computed the migration
mismatch and foraging efficiency metrics, as well as the number of years
required to reach stationarity.

For numerical runs with climate change, we first run a simulation for 20
years to attain an equilibrium, and then begin shifting the location of
the resource with a trend (slow 0.25, medium 0.5, or rapid 1 units per
year), or by adding stochasticity (s.d. 3, 6, 12). For the trend
analyses, we then compute the slope of the migration location parameter
against the resource drift and refer to this as the \emph{spatial
adaptation} (SA) index.

\hypertarget{simulation-studies}{%
\subsection{Simulation studies}\label{simulation-studies}}

We explored this model using numerical differencing of a system of
ordinary differential equations (ODE's) approximating the PDE in
equation (2) with the Runge-Kutte algorithm using the \texttt{deSolve}
(Soetaert et al. 2010) and \texttt{ReacTran} (Soetaert and Meysman 2012)
packages in R. We additionally used the \texttt{nlsLM} function in
package \texttt{minpack.LM} (Elzhov et al. 2016) for robust and fast
annual estimation of the migration parameters. The complete code is
available as an R package (\texttt{memorymigration}) available on GitHub
at \url{https://github.com/EliGurarie/memorymigration} and as an
interactive Shiny application at \url{https://url.to.be.determined}.

We assessed a wide range of parameter values and resource geometries and
dynamics with the goal of answering the four main questions: 1. Can this
model adapt to a discrete shift in peak resource location and timing?
What is the relative role of memory and sociality for adaptation? 2. Can
this model acquire a migratory behavior from a non-migratory initial
condition? 3. What is the role of a reference memory for dealing with
stochastic resource dynamics? 4. Can this model adapt when the resource
peaks shifts in space?

\textbf{Table 1} \emph{Parameters, variables and metrics.}

\begin{longtable}[]{@{}lll@{}}
\toprule
\begin{minipage}[b]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Parameter\strut
\end{minipage} &
\begin{minipage}[b]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Description\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\emph{Memory migration model}\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\epsilon\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Diffusion\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\alpha\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Forage following\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\beta\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Strength of sociality\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\lambda\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Spatial scale of sociality\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\kappa\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
Proportion of reference versus working memory\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(x_1\), \(x_2\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
location of population centroids in summer and winter\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(t_1\), \(dt_1\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
start and duration of summer season\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(t_2\), \(dt_2\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
start and duration of winter season\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\kappa\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
proportion memory allocated to reference (long-term) memory
vs.~short-term memory\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\emph{Resource dynamics}\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\tau\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
duration of period (year)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(m_x\), \(-m_x\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
spatial coordinate of resource peak for summer and winter\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(m_t\), \(\tau - m_t\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
timing of resource peak for the summer and winter\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\sigma_x\), \(\sigma_t\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
time duration and and spatial scale of resource pulse\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(dx\), \(dt\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
rate of change of peak locaton and timing of resource\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(\psi_x\), \(\psi_t\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
standard deviation of peak location and timing\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\emph{Metrics}\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(MM_x\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
spatial migration mismatch\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(MM_t\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
temporal migration mismatch\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(TE\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
total mismatch\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(FE\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
foraging efficiency\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.19}}\raggedright
\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
\(SA\)\strut
\end{minipage} &
\begin{minipage}[t]{(\columnwidth - 2\tabcolsep) * \real{0.41}}\raggedright
spatial adaptation index\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{results}{%
\section{Results}\label{results}}

\includegraphics{figures/example1_adaptation.png}

\textbf{Figure 2.} \emph{Example of adaptation to a shift in resource
peak. The initial (year 0) behavior migrates to locations 50 and -50 at
days 15 and 60, whereas the resource peak is at 30 and -30, peaking at
times 25 and 75. The panels show (a) the first 14 years of the
simulation; (b) the centroid of the annual movement of the population is
shown in panel b, with dark blue to red colors indicating year 0 to year
40; (c) annual foraging efficiency across years; (d) migration timing
parameters for each year, with orange segments indicating arrival and
departure from the summering (northern) grounds, and the blue segments
indicating timing of arrival and departure at the wintering grounds; (e)
migration arrival and departure location across years, with blue and
orange indicating winter (southern) and summer (northern) locations.}

\hypertarget{adaptation-to-resource-phenology}{%
\subsection{Adaptation to resource
phenology}\label{adaptation-to-resource-phenology}}

The ability of this system to attain stable, migratory state that
matches the dynamics of the resource is illustrated in Figure 2. In the
illustrated scenario, it takes nearly 40 years to attain an equilibrium,
and the eventual steady state is one where the centroid of the migration
lines up exactly with the centroid of the resource, and the arrival
timing coincides with the \emph{peak} of resource availability. Notably,
the path to this equilibrium is somewhat indirect, with the later winter
range taking more time to stabilize than the earlier summer range. The
eventual steady state is one where the foraging efficiency is relatively
high.

We ran this process for 8100 parameter combinations (Figure 3). The
fundamental dynamic of attaining and maintaining a well-matched
migration phenology, can occur under many combinations of parameter
values, but all parameters play a role. Among the more intuitive results
are that greater values of \(\alpha\) (resource following) lead to an
improved ability to match the migration. Resource peaks with larger
spatial extent (higher \(\sigma_x\)) are generally better for migration
matching. A larger set of parameters matched migration with low
diffusion than high diffusion.

Less intuitive was the high importance of the sociality parameters, in
particular the spatial scale of the swarming. Higher levels of social
attraction (\(\beta\)) led to improved migration matching except in
those cases where the sociality scale \(\lambda\) was high. Thus, for
example, at \(\lambda = 20\), no simulations at \(\beta >= 200\) managed
to acquire or maintain a matched migration. However, at \(\lambda = 50\)
or \(100\), the migration was slightly better matched at high values of
\(\beta\). (Fig. \textbf{X}) The spatial extent of the swarm was a
remarkably significant variable. Smaller swarms were able to match
migration only at low values of social attraction (\(\beta = 200\)), and
relatively high values of resource attraction (\(\alpha \geq 600\)).

Random forest analyses, whether on the log of total mismatch or on the
classification of a perfect match, uniformly show that the most
important variables (Breiman 2001) were \(\alpha\) and \(\lambda\) (4.14
and 4.02 proportional increase in MSE), and the least important was
\(\sigma_t\), with a 0.5 proportional increase (table \textbf{X}).

Overall, foraging efficiency was strongly correlated with migration
matching, as expected. At high mismatch (\textgreater{} 50), foraging
efficiency was low (mean 0.29, s.d. 0.16) compared to the near-perfect
matching migrations (mean 0.58, s.d. 0.14). However, somewhat higher
mismatch (1 to 5) showed an even higher overall foraging efficiency
(mean 0.62, s.d. 0.18 - see also Figure \textbf{X}).

\includegraphics{figures/StabilityResults.png} \textbf{Figure 3.}
\emph{Migration phenology matching across six model parameters. Low and
high diffusion (\(\epsilon = 1\) and \(8\) in upper and lower panel
blocks), tight, medium and loose swarms (\(\lambda = 20, 50, 100\)) left
to right panels. Within each of these blocks, high values of the
resource following parmaeters \(\alpha\) from 0 to 1000 are left to
right, and higher values of the sociality parameter \(\beta\) are bottom
to top. Within each of the combinations of
\(\epsilon, \lambda, \alpha, \beta\), we show results ranging across 6
values of resource duration (\(\sigma_t\), 3-15 left to right), and 6
values of resource extent (\(\sigma_x\), 3-15 bottom to top), as in the
zoomed in panel at bottom right. The color scheme reflects the total
mismatch, i.e.~the sum of the absolute differences between the migration
timing and locations from the resource peak. White squares represent
runs that numerically failed to estimate migration parameters.}

\includegraphics{figures/ForagingEfficiency.png}

\textbf{Figure 4.} \emph{Box-plots of foraging efficiency against
mismatch across several values of foraging patch duration.}

\hypertarget{learning-to-migrate}{%
\subsection{Learning to migrate}\label{learning-to-migrate}}

\includegraphics{figures/example2_learningtomigrate.png} \textbf{Figure
5.} \emph{Example of model learning to migrate. The resource is a
``weakly drifting'' resource and the initial (year 0) condition is non
migratory. The simulation was run for 100 years, and a sampling of those
years (labeled) are presented in panel a: all years from 0 to 10,
followerd by 20, 40 and 100. Otherwise, panels are as in Figure 1.
Parameter values were \(\epsilon = 5\), \(\alpha = 500\) and
\(\beta = 50\).}

Figure \textbf{5} illustrates the ability of the model animals to learn
to migrate in a weakly drifting resource environment with a narrow pulse
of resource peaking at 30 and -30 (at days 25 and 75), but a uniform
distribution of resource at times 0 and 50. In order to learn to
migrate, the system needed to have a higher exploratory impulse (higher
diffusion constant \(\epsilon\)), a stronger resource advection (higher
\(\alpha\)) and somewhat weaker sociality (lower \(\beta\)). The
qualitative behavior of this process was to start drifting towards the
summer resource, while slowly developing a weak pulse towards the winter
resource as well. After first locking in on the summer resource, the
winter migration, driven both by high diffusion and high resource
following, slowly extended itself until both narrow peaks of resource
can be consistently reached.

The model had, in general, a difficult time learning migration from a
non-migratory initial condition. Thus, out of 4047 successful runs, only
4 attained mismatch below 1, and 130 below 5. Conditions that were more
conducive to learning migration were pulses of \emph{longer} duration
(high \(\sigma_t\)), but \emph{smaller} in scope (low \(\sigma_x\)),
suggesting that the feedback that encourages migration needs to be
compact in space but long enough in duration to lock in to the memory.

\hypertarget{adapting-to-climate-change}{%
\subsection{Adapting to climate
change}\label{adapting-to-climate-change}}

\includegraphics{figures/SpatialClimateChange.png}

\textbf{Figure 6.} \emph{Adaptation to a steadily drifting resource. In
three scenarios, the spatial coordinates of the resource drift by 0.25,
0.5, and 1 unit per year (top to bottom, respectively). The y-axis is
the index of spatial adaptation (SA), i.e.~the trend of the
memory-driven migration divided by the resource drift trend. Values near
1 indicate a behavior that keeps up with climate change, values near
indicate no change in migration behavior, and negative values indicate a
trend that is opposite to the climate trend. We compare across spatial
scales of sociality (\(\lambda\) - x-axis), for low and high values
resource following (\(\alpha = 400\) and \(100\) - orange and blue dots)
and low and high values of sociality (\(\beta = 100\) and 400, left and
right panels). The size of the circles is proportional to the foraging
efficiency of the resulting parameter combinations. The bottom-right
boxplots indicate the final year foraging efficiency against the SA
index; purple and blue boxes indicate the highest values, orange and
gray lower values.}

To assess the ability of the system to adapt to a trending climate, we
generated scenarios with slow (0.25 units / year), medium (0.5 units /
year), and fast (1 unit / year) drift outward of the two resource
pulses. We then assessed 40 parameter combinations for each of those
scenarios, a high and low value of resource following (\(\alpha = 400\)
and \(100\)) crossed with a high and low value of sociality
(\(\beta = 400\) and \(100\)) crossed with 10 values of the spatial
scale of sociality (\(\lambda = 20\) to \(200\)). The spatial and
temporal scale of the resource pulses were fixed to a relatively
``easy'' to adapt to \(\sigma_x = 12\) and \(\sigma_t = 6\). We computed
the adaptation index and foraging efficiency for each of the 120 runs
(figure \textbf{Z}). We were interested in the dynamics against
\(\lambda\) due to the uniformly high importance of this parameter for
determining the success of this process to match migration in steady
states.

As figure 6 shows, higher values of resource following
(\(\alpha = 400\); orange circles) are nearly universally better for
keeping up with climate change (SA values near 1). Furthermore, when
combined with high sociality (\(\beta = 400\); right panels), nearly all
parameter combinations do a good job keeping up with climate change (SA
values ranging between 0.53 and 0.85 for a swarm size greater than 50).
However, that maximum value is still less than 1, suggesting that truly
matching a steadily drifting trend is very difficult. Small swarms
(\(\lambda < 50\)) have a very hard time adapting when the social
attraction is high, but do fairly well when social attraction is lower.
But larger sized swarms do progressively worse across more
parameterizations, e.g.~in the most rapid climate change scenario, the
SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to
200 (essentially, the entire spatial domain).

A rather more dramatic pattern is visible for the lower foraging
attraction scenario (\(\alpha = 100\); blue circles). Notably, no
parameter combination at this value comes close to keeping up with the
rapid climate change (SA range -0.64 to 0.13). For slower climate
change, however, there is a window of values for the swarm size -
relatively small, between 40 and 80, where the SA \emph{exceeds} 1, and
then crashing quite rapidly to negative values of SA as that swarm size
increases. These ``super-adapting'' processes indicate a unique sweet
spot where a swarm is large enough to capture and adapt to the drifting
resource, but not so large that the information gathered in a given year
is too weak to adjust the migratory behavior in a following year.

As anticipated, better adaptation to the drifting resource correlated
strongly with higher foraging efficiency (inset boxplots).

\hypertarget{reference-memory-and-stochasticity}{%
\subsection{Reference memory and
stochasticity}\label{reference-memory-and-stochasticity}}

While recent memory can be helpful for adapting to a single novelty or a
smoothly changing conditions, we hypothesized that more conservative
approach, relying on a ``reference memory'' may be more beneficial when
conditions change stochastically. We tested this hypothesis by solving a
set of models across a range of \(\kappa\) values fom 0 (all recent
memory) to 1 (all reference memory). In these scenarios, we ran the
system for as many years as needed with no stochasticity to acquire a
stationary state (i.e.~similarity index greater than 1-1e-6). We then
used the stationary state as the reference memory, and then ran the
process for an additional 50 years with a stochasticity (i.e.~standard
deviation in peak location of the resource) ranging from 0 to 12.

Figure 7 illustrates that, as predicted, the highest level of \(\kappa\)
can significantly help foraging efficiency; however this is highly
sensitive to the spatial scale of sociality. When that scale of
sociality is high enough (e..g \(\lambda = 120\), blue colors) there is
greater probability of overlap with a stochastic resource, and a
conservative, stable migratory regime is much more beneficial in the
long run. Overall, as anticipated, the greater the stochasticity, the
lower the foraging efficiency; and the greater the stochasticity, the
greater the benefit of a conservative reference memory driven strategy.

\includegraphics{figures/stochasticity.png}

\textbf{Figure 7} \emph{Foraging efficiency across various values of
reference memory \(\kappa\) (x-axis) for increasing amounts of
interannual stochasticity (\(\psi_x\), panels left to right), and two
values of spatial scale \(\lambda = 80\) and \(120\). For the stochastic
processes (\(\psi_x > 0\)), the process was run 30 times, and the mean
and standard error of the foraging efficiency over a 50 year
post-stabilization period is presented (points and bars). In these
scenarios, the resource following parameter \(\alpha = 100\), the social
attraction \(\beta = 400\) and diffusion \(\epsilon = 4\).} (NB: we are
running more replicated to get stronger results / narrower error bars
here)

\hypertarget{balancing-stochasticity-with-trends}{%
\subsection{Balancing stochasticity with
trends}\label{balancing-stochasticity-with-trends}}

We added trends to the stochastic process described above, adding 30
years of spatial outward trend in the resource. Our goal was to assess
the ability of the process to keep up with climate change as a function
of the reference memory parameter (Figure 8). Over-reliance on reference
memory (\(\kappa = 1\)) by definition does not allow the system to keep
up with climate change, leading to an adaptation index of 0. However, in
many cases a balancing of recent and reference memory (\(\kappa\) value
between 0.6 and 0.8) in many cases was slightly but significantly better
than relying entirely on recent memory. The smaller spatial scale (in
the selected parameter space) does a generally better job than the
larger spatial scale at lower stochasticity. At higher level of
stochasticity, however, the larger spatial scale outperforms the smaller
spatial scale, which completely loses track of the climate change.

\includegraphics{figures/TrendStochasticity.png}

\textbf{Figure 8}. \emph{Role of reference memory in adapting to climate
change for increasingly stochastic resource dynamics. We ran the model
with a moderiate rate of climate change (mean shift: 0.5 / year) at five
increasing levels of stochasticity (inter-annual standard deviation of
resource peak 0, 3 ,6 and 12, left to right panels). For non-zero
stochasticity, we ran the process 30 times and present the mean and
standard error of the spatial adaptation index across various values of
the reference memory parameter \(\kappa\): where \(\kappa = 0\), the
system modifies its migration based entirely on recent experiences; at
\(\kappa = 1\), the memory never changes from the reference memory.
Other parameter values are resource following \(\alpha = 100\), social
attraction \(\beta = 400\), and social spatial scale \(\lambda = 80\).}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Animals navigate complex, dynamic, patchy environments. When there is a
strongly localized and seasonal component to the resource dynamics,
typical strategies of spatial distribution by straightforward
resource-following taxis necessarily fails to maximally exploit the
dynamics. It is in these cases - quite common in the natural world -
that seasonal migration becomes a viable, even necessary, strategy.
However, when those resources start shifting in space and time - as is
occurring at an accelerated pace with recent global climate change - the
migration phenology itself must exhibit some plasticity. It is our
conjecture that this plasticity is greatly facilitated by a
memory-driven process, in which recent experiences inform strategic
behaviors in subsequent years. To our knowledge, this study is the first
attempt to focus in and propose a relatively simple, memory-driven
mechanism by which a population can respond to long-term changes in
resource dynamics.

We found that the straightforward inclusion of a migration-specific
memory in a social diffusion-advection framework was sufficient to
capture many fundamental phenomena related to migration. In essence, our
model allowed a population with a hybrid migratory and resource
following behavior to adjust the migratory behavior based on recent
experiences with the resource location. The model, while simple, was
able to emulate the successful navigation of an environment with
temporally and spatially isolated seasonal resource patches, the
spontaneous emergence of a migratory behavior, and intrinsic robustness
to changes in those environmental resources, whether steadily shifting
trends or inter-annual stochasticity. We postulated a process defined by
relatively few parameters, and explored the ability of the process to
navigate a fixed set of environments with the goal of understanding the
combinations of factors that lead to a resilient migration behavior. The
foraging efficiency was a convenient metric of the utility of the
migratory behavior, but was not a measure explicitly maximized by the
model.\\
A few rather general conclusions can be made from our model. One is that
migration in general can be acquired, as has been demonstrated for
translocated ungulates (Jesmer et al. 2018), even in environments where
resources are difficult to ``surf,'' but that this requires a very
strong attraction to resources, somewhat higher levels of exploratory
behavior, and many years. On the other hand, migrations can be
considered somewhat fragile, in the sense that under certain conditions,
e.g.~increasing stochasticity, rapidly shifting resources, or a shift in
some of the system parameters, migration can collapse down into a
localized non-migratory, residential behavior, at times with only a
marginal cost to the foraging efficiency index. This sensitivity may
explain why partially migratory populations composed of a mixture of
resident and migratory individuals are so common and, apparently,
evolutionary stable (Berthold 1999, Chapman et al. 2011).

We also found that the ability to maintain, learn and adapt migration
also depends strongly on properties of the resource dynamics. In
particular, the reinforcement of memory and foraging is strongest when
patches are concentrated in time, but relatively large in space.
Interestingly, in most stable patterns, the eventual targeted migration
arrival time coincided with the \emph{peak}, rather than the beginning,
of the resource dynamic. This indicates that the long-distance social
migration behavior may be particularly reinforced when the targeted
resource is very sudden (explosive), which is the case for the rapid
green-up that occurs in high latitudes as snow recedes in tandem with
extended day lengths, leading to a rapid and intense local green-up.

The sociality parameters - in particular, the spatial scale \(\lambda\)
- were somewhat unexpectedly, the most important for determining the
resiliency of the process for stabilizing a migratory behavior. In
particular, a very small spatial scale tended to have a much more
difficult time locking in to an adaptive migratory pattern, and only in
conditions where the social attraction was relatively weak. In order to
adapt to any shift in resources, \emph{some} individuals must experience
that resource, and the two sociality-related mechanisms to have that
scope is either a relatively large spatial scale or a relatively weak
social cohesion. On the other hand, overly large spatial scales
compromized the ability of the process to track climate change, due to a
diluting of the population's ability to concentrate over available
resource patches when made available.

All of the parameters in our model have well-defined biological
interpretations. The diffusion (\(\epsilon\)) captures short time-scaled
randomness of movement, capturing the exploratory and short-term
dispersive behavior. The foraging advection strength (\(\alpha\))
captures the attraction of the population to better quality resources at
a relatively large scale. These two parameters, the basic ingredients in
diffusion-advection models of animal movement, have direct parallels to
empirically estimated properties of animal behavior: diffusion is
closely related to families of random walk models (Gurarie and
Ovaskainen 2011) while the advective taxis is related to the step- and
resource selection functions that are routinely estimated from movement
datas (Potts and Schlägel 2020). The spatial scale of the social group
(\(\lambda\)) captures the spatial extent of the population, i.e.~a
population-level range. The sociality parameter (\(\beta\)) quantifies
the strength an individual's desire to approach the center of the social
group. This is not a parameter that is typically measured, though it
would - in principle - be possible to estimate in a manner analogous to
step-selection function using only other conspecifics as opposed to
environmental variables as a covariate. Because both resources and
populations were normalized to 1, the ratio between \(\beta\) and
\(\alpha\) can be interpreted as the relative importance of foraging to
social cohesion. It does appears a certain balance of both is needed to
successfully maintain, acquire, or adapt a migratory phenology. Finally,
the migration timing and location parameter can be similarly
straightforwardly estimated from movement data (Cagnacci et al. 2015,
Gurarie et al. 2019). Thus, for example, Gurarie et al. (2019)
explicitly esimated the ranging area, timing, and seasonal range
locations for migratory caribou (\emph{Rangifer tarandus}), identifying
the kind of inter-annual variation that is reflected in the stochastic
scenarios explored here as well as some trends in the timing of
migration. Finally, we note that our underlying assumption of the
migration ``urge'' is consistent with the strong endogenous programs to
migrate taht is well-documented, in particular, in birds that exhibit a
seasonal restlessness known as \emph{Zugunruhe} (Berthold 1999, Helm
2006).

Collective knowledge has been shown to be important, if not essential,
to the evolution and process of migration (Guttal and Couzin 2010, Shaw
and Couzin 2013, Berdahl et al. 2018). In our model, we assume that the
migration process itself is driven by a collective trigger for the
timing and locations of seasonal ranges and migration behavior.
Synchrony of migration timing and high site fidelity are well documented
for migratory species (Gurarie et al. 2019, Joly et al. 2021). We note,
however, that diffusion-advection models can also be interpreted as a
probabilistic description of a single individual's movement. In this
case, \(\lambda\) would correspond to an individual home-range and
\(\beta\) would be an individual's tendency to be drawn to the center of
that home range, akin to an individual migratory Ornstein-Uhlenbeck
process (Gurarie et al. 2017).

The reference memory parameter \(\kappa\) is, of course, impossible to
observe directly. Our model does, however, allow us to explore in an
heuristic way the conditions under which a strong cultural tendency to
migrate with certain fixed patterns can help a population hedge against
stochasticity. In general, our model allows us to consider broadly the
risks.

Importantly, our model did not include any selection, inheritance or
birth or death processes, i.e.~there was no evolutionary maximization of
a fitness, as in many models that explicitly explore the evolution of
migration (e.g. Guttal and Couzin 2010, Anderson et al. 2013, Shaw and
Couzin 2013). Nonetheless, it is instructive to compare the fundamental
assumptions behind such evolutionary individual-based models and a
social memory-based model like ours. For example, Anderson et al. (2013)
similarly explored the resilience of a population under selective
pressure under persistent trends and increased stochasticity of a
drifting optimal resource window, showing that a certain amount of
heritable phenotypic plasticity is necessary to adapt successfully to
climate change even at the cost of efficiency. Our model underscores the
fact that some level of resilience and adapatability can be attained
with a purely cognitive process that balances sociality, long, and short
term collective memory. Importantly, this knowledge must be transmitted
through social and cultural, rather than genetic, pathways. The high
level of sociality among migratory animals, as well as multi-annual
parent offspring bonds, are an evident pathway for that kind of
transmission. As with those evolutionary models, however, it is clear
that when changes are too rapid, no amount of cognition can help
entirely mitigate against adverse outcomes. Furthermore, if behaviors
are not sufficiently plastic (i.e.~if \(\kappa\) is too close to 1),
then adaptation is that much more difficult.

Rapid environmental change, both global warming and increased
anthropogenic development, is causing severe and dramatic impacts to the
widespread and generally successful strategy of seasonal migration for
many taxa, and the fate of many animal migrations is a topic of
increasing concern (Wilcove and Wikelski 2008, Kauffman et al. 2021).
The ability of animals to respond to these changes is deeply dependent
on their behavioral plasticity and cognitive abilities. The importance
of those abilities are in direct proportion to the difficulty in
studying them directly. By quantitatively exploring the properties of a
heuristic model that distill many of the main properties of wild
populations in dynamic and seasonal environments, we hope to have
identified some broad patterns that might guide further empirical
exploration of the cognitive underpinnings of adapatability and
resilience.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

NSF grant. Quentin at SESYNC.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Abrahms2019}{}%
Abrahms, B., E. L. Hazen, E. O. Aikens, M. S. Savoca, J. A. Goldbogen,
S. J. Bograd, M. G. Jacox, L. M. Irvine, D. M. Palacios, and B. R. Mate.
2019. Memory and resource tracking drive blue whale migrations.
Proceedings of the National Academy of Sciences 116:5582--5587.

\leavevmode\hypertarget{ref-Anderson2013}{}%
Anderson, J. J., E. Gurarie, C. Bracis, B. J. Burke, and K. L. Laidre.
2013. Modeling climate change impacts on phenology and population
dynamics of migratory marine species. Ecological Modelling 264:83--97.

\leavevmode\hypertarget{ref-Avgar2014}{}%
Avgar, T., G. Street, and J. M. Fryxell. 2014. On the adaptive benefits
of mammal migration. Canadian Journal of Zoology 92:481--490.

\leavevmode\hypertarget{ref-Berdahl2018}{}%
Berdahl, A. M., A. B. Kao, A. Flack, P. A. Westley, E. A. Codling, I. D.
Couzin, A. I. Dell, and D. Biro. 2018. Collective animal navigation and
migratory culture: From theoretical models to empirical evidence.
Philosophical Transactions of the Royal Society B: Biological Sciences
373:20170009.

\leavevmode\hypertarget{ref-Berthold1999}{}%
Berthold, P. 1999. A comprehensive theory for the evolution, control and
adaptability of avian migration. Ostrich 70:1--11.

\leavevmode\hypertarget{ref-Bhattacharyya1943}{}%
Bhattacharyya, A. 1943. On a measure of divergence between two
statistical populations defined by their probability distributions.
Bull. Calcutta Math. Soc. 35:99--109.

\leavevmode\hypertarget{ref-Bischof2012}{}%
Bischof, R., L. E. Loe, E. L. Meisingset, B. Zimmermann, B. Van Moorter,
and A. Mysterud. 2012. A migratory northern ungulate in the pursuit of
spring: Jumping or surfing the green wave? The American Naturalist
180:407--424.

\leavevmode\hypertarget{ref-Bracis2017}{}%
Bracis, C., and T. Mueller. 2017. Memory, not just perception, plays an
important role in terrestrial mammalian migration. Proceedings of the
Royal Society B: Biological Sciences 284:20170449.

\leavevmode\hypertarget{ref-Breiman2001}{}%
Breiman, L. 2001. Random forests. Machine learning 45:5--32.

\leavevmode\hypertarget{ref-Cagnacci2015}{}%
Cagnacci, F., S. Focardi, A. Ghisla, B. van Moorter, E. H. Merrill, E.
Gurarie, M. Heurich, A. Mysterud, J. Linnell, M. Panzacchi, R. May, T.
Nygård, C. Rolandsen, and M. Hebblewhite. 2015. How many routes lead to
migration? Comparison of methods to assess and characterize migratory
movements. Journal of Animal Ecology 85:54--68.

\leavevmode\hypertarget{ref-Chapman2011}{}%
Chapman, B. B., C. Brönmark, J.-Å. Nilsson, and L.-A. Hansson. 2011. The
ecology and evolution of partial migration. Oikos 120:1764--1775.

\leavevmode\hypertarget{ref-Dingle2014}{}%
Dingle, H. 2014. Migration: The biology of life on the move. Oxford
University Press, USA.

\leavevmode\hypertarget{ref-minpack.lm}{}%
Elzhov, T. V., K. M. Mullen, A.-N. Spiess, and B. Bolker. 2016.
Minpack.lm: R interface to the levenberg-marquardt nonlinear
least-squares algorithm found in MINPACK, plus support for bounds.

\leavevmode\hypertarget{ref-Fagan2017}{}%
Fagan, W., E Gurarie, S. Bewick, A. Howard, R. Cantrell, and C. Cosner.
2017. Perceptual ranges, information gathering, and foraging success in
dynamic landscapes. The American Naturalist 189:474--489.

\leavevmode\hypertarget{ref-Fagan2019}{}%
Fagan, W., T. Hoffman, D. Dahiya, E Gurarie, R. Cantrell, and C. Cosner.
2019. Improved foraging by switching between diffusion and advection:
Benefits from movement that depends on spatial context. Theoretical
Ecology 13:127--136.

\leavevmode\hypertarget{ref-Fryxell1988}{}%
Fryxell, J. M., J. Greever, and A. Sinclair. 1988. Why are migratory
ungulates so abundant? The American Naturalist 131:781--798.

\leavevmode\hypertarget{ref-Gurarie2009}{}%
Gurarie, E., J. J. Anderson, and R. W. Zabel. 2009. Continuous models of
population-level heterogeneity inform analysis of animal dispersal and
migration. Ecology 90:2233--2242.

\leavevmode\hypertarget{ref-Gurarie2017}{}%
Gurarie, E., F. Cagnacci, W. Peters, C. H. Fleming, J. M. Calabrese, T.
Mueller, and W. F. Fagan. 2017. A framework for modelling range shifts
and migrations: Asking when, whither, whether and will it return.
Journal of Animal Ecology 86:943--959.

\leavevmode\hypertarget{ref-Gurarie2017}{}%
Gurarie, E., F. Cagnacci, W. Peters, C. H. Fleming, J. M. Calabrese, T.
Mueller, and W. F. Fagan. 2017. A framework for modelling range shifts
and migrations: Asking when, whither, whether and will it return.
Journal of Animal Ecology 86:943--959.

\leavevmode\hypertarget{ref-Gurarie2019}{}%
Gurarie, E., M. Hebblewhite, K. Joly, A. P. Kelly, J. Adamczewski, S. C.
Davidson, T. Davison, A. Gunn, M. J. Suitor, W. F. Fagan, and others.
2019. Tactical departures and strategic arrivals: Divergent effects of
climate and weather on caribou spring migrations. Ecosphere 10:e02971.

\leavevmode\hypertarget{ref-Gurarie2011}{}%
Gurarie, E., and O. Ovaskainen. 2011. Characteristic spatial and
temporal scales unify models of animal movement. The American Naturalist
178:113--123.

\leavevmode\hypertarget{ref-Guttal2010}{}%
Guttal, V., and I. D. Couzin. 2010. Social interactions, information
use, and the evolution of collective migration. Proceedings of the
National Academy of Sciences 107:16172--16177.

\leavevmode\hypertarget{ref-Helm2006}{}%
Helm, B. 2006. Zugunruhe of migratory and non-migratory birds in a
circannual context. Journal of Avian Biology 37:533--540.

\leavevmode\hypertarget{ref-Jesmer2018}{}%
Jesmer, B. R., J. A. Merkle, J. R. Goheen, E. O. Aikens, J. L. Beck, A.
B. Courtemanch, M. A. Hurley, D. E. McWhirter, H. M. Miyasaki, K. L.
Monteith, and others. 2018. {Is ungulate migration culturally
transmitted? Evidence of social learning from translocated animals}.
Science 361:1023--1025.

\leavevmode\hypertarget{ref-Joly2021}{}%
Joly, K., E. Gurarie, D. A. Hansen, and M. D. Cameron. 2021. Seasonal
patterns of spatial fidelity and temporal consistency in the
distribution and movements of a migratory ungulate. Ecology and
Evolution 11:8183--8200.

\leavevmode\hypertarget{ref-Kauffman2021}{}%
Kauffman, M. J., F. Cagnacci, S. Chamaillé-Jammes, M. Hebblewhite, J. G.
C. Hopcraft, J. A. Merkle, T. Mueller, A. Mysterud, W. Peters, C.
Roettger, A. Steingisser, J. E. Meacham, K. Abera, J. Adamczewski, E. O.
Aikens, H. Bartlam-Brooks, E. Bennitt, J. Berger, C. Boyd, S. D. Côté,
L. Debeffe, A. S. Dekrout, N. Dejid, E. Donadio, L. Dziba, W. F. Fagan,
C. Fischer, S. Focardi, J. M. Fryxell, R. W. S. Fynn, C. Geremia, B. A.
González, A. Gunn, E. Gurarie, M. Heurich, J. Hilty, M. Hurley, A.
Johnson, K. Joly, P. Kaczensky, C. J. Kendall, P. Kochkarev, L.
Kolpaschikov, R. Kowalczyk, F. van Langevelde, B. V. Li, A. L. Lobora,
A. Loison, T. H. Madiri, D. Mallon, P. Marchand, R. A. Medellin, E.
Meisingset, E. Merrill, A. D. Middleton, K. L. Monteith, M. Morjan, T.
A. Morrison, S. Mumme, R. Naidoo, A. Novaro, J. O. Ogutu, K. A. Olson,
A. Oteng-Yeboah, R. J. A. Ovejero, N. Owen-Smith, A. Paasivaara, C.
Packer, D. Panchenko, L. Pedrotti, A. J. Plumptre, C. M. Rolandsen, S.
Said, A. Salemgareyev, A. Savchenko, P. Savchenko, H. Sawyer, M.
Selebatso, M. Skroch, E. Solberg, J. A. Stabach, O. Strand, M. J.
Suitor, Y. Tachiki, A. Trainor, A. Tshipa, M. Z. Virani, C. Vynne, S.
Ward, G. Wittemyer, W. Xu, and S. Zuther. 2021. Mapping out a future for
ungulate migrations. Science 372:566--569.

\leavevmode\hypertarget{ref-Kot1996}{}%
Kot, M., M. A. Lewis, and P. van der Driessche. 1996. Dispersal data and
the spread of invading organisms. Ecology 77:2027--2042.

\leavevmode\hypertarget{ref-Kolzsch2015}{}%
Kölzsch, A., S. Bauer, R. De Boer, L. Griffin, D. Cabot, K.-M. Exo, H.
P. Van der Jeugd, and B. A. Nolet. 2015. {Forecasting spring from afar?
Timing of migration and predictability of phenology along different
migration routes of an avian herbivore}. Journal of Animal Ecology
84:272--283.

\leavevmode\hypertarget{ref-Lin2021}{}%
Lin, H.-Y., W. F. Fagan, and P.-E. Jabin. 2021. Memory-driven movement
model for periodic migrations. Journal of Theoretical Biology
508:110486.

\leavevmode\hypertarget{ref-Mogilner1999}{}%
Mogilner, A., and L. Edelstein-Keshet. 1999. A non-local model for a
swarm. Journal of Mathematical Biology 38:534--570.

\leavevmode\hypertarget{ref-Okubo2001}{}%
Okubo, A., and S. Levin. 2001. {Diffusion and Ecological Problems:
Modern Perspectives}. Springer Verlag, New York.

\leavevmode\hypertarget{ref-Potts2020}{}%
Potts, J. R., and U. E. Schlägel. 2020. Parametrizing diffusion-taxis
equations from animal movement trajectories using step selection
analysis. Methods in Ecology and Evolution 11:1092--1105.

\leavevmode\hypertarget{ref-Robinson2009}{}%
Robinson, R., H. Crick, J. Learmonth, I. Maclean, C. Thomas, F.
Bairlein, M. Forchhammer, C. Francis, J. Gill, B. Godley, J. Harwood, G.
Hays, B. Huntley, A. Hutson, G. Pierce, M. Rehfisch, D. Sims, B. Santos,
T. Sparks, D. Stroud, and M. Visser. 2009. Travelling through a warming
world: Climate change and migratory species. Endangered Species Research
7:87--99.

\leavevmode\hypertarget{ref-Shaw2016}{}%
Shaw, A. K. 2016. Drivers of animal migration and implications in
changing environments. Evolutionary Ecology 30:991--1007.

\leavevmode\hypertarget{ref-Shaw2013}{}%
Shaw, A. K., and I. D. Couzin. 2013. Migration or residency? The
evolution of movement behavior and information usage in seasonal
environments. The American Naturalist 181:114--124.

\leavevmode\hypertarget{ref-Skalski2003}{}%
Skalski, G. T., and J. F. Gilliam. 2003. A diffusion-based theory of
organism dispersal in heterogeneous populations. The American Naturalist
161:441--458.

\leavevmode\hypertarget{ref-Skellam1951}{}%
Skellam, J. G. 1951. Random dispersal in theoretical populations.
Biometrika 38:196--218.

\leavevmode\hypertarget{ref-Soetaert2012}{}%
Soetaert, K., and F. Meysman. 2012. {Reactive transport in aquatic
ecosystems: Rapid model prototyping in the open source software R}.
Environmental Modelling \& Software 32:49--60.

\leavevmode\hypertarget{ref-Soetaert2010}{}%
Soetaert, K., T. Petzoldt, and R. W. Setzer. 2010. Solving differential
equations in {R}: Package de{S}olve. Journal of Statistical Software
33:1--25.

\leavevmode\hypertarget{ref-Turchin1998}{}%
Turchin, P. 1998. Quantitative analysis of movement: Measuring and
modeling population redistribution in animals and plants. Sinauer
Associates.

\leavevmode\hypertarget{ref-Wilcove2008}{}%
Wilcove, D. S., and M. Wikelski. 2008. Going, going, gone: Is animal
migration disappearing. {PLoS} Biology 6:e188.

\leavevmode\hypertarget{ref-Wilcove2008}{}%
Wilcove, D. S., and M. Wikelski. 2008. Going, going, gone: Is animal
migration disappearing. {PLoS} Biology 6:e188.

\end{CSLReferences}

\end{document}

### 2021-07-13 text/table.Rmd
98% there

--- 2021-07-09 text/table.Rmd
@@
 
-**Memory migration model**
+*Memory advection-diffusion PDE*
 
-| Parameter   | | Description  
+| Parameter   | Description |
 | ----------- | ----------- |
+| $\epsilon$  | Diffusion | 
 | $\alpha$  | Forage following | 
@@
 | $\lambda$  | Spatial scale of sociality |
+| $\kappa$  | Proportion of reference versus working memory |
+| $x_1$, $x_2$  | location of population centroids in summer and winter| 
+| $t_1$, $dt_1$  | start and duration of summer season |
+| $t_2$, $dt_2$  | start and duration of winter season | 
+| $\kappa$ | proportion memory allocated to reference (long-term) memory vs. short-term memory|
 
-**Resource**
+*Resource dynamics*
+
+| Parameter   | Description |
+| ----------- | ----------- |
+| $\tau$  | duration of period (year) | 
+| $m_x$, $-m_x$  | spatial coordinate of resource peak for summer and winter | 
+| $m_t$, $\tau - m_t$  | timing of resource peak for the summer and winter | 
+| $\sigma_x$, $\sigma_t$   | time duration and and spatial scale of resource pulse | 
+| $dx$, $dt$  | rate of change of peak locaton and timing of resource | 
+| $\psi_x$, $\psi_t$  | standard deviation of peak location and timing |
 
 **Metrics**
+
+| Parameter   | Description |
+| ----------- | ----------- |
+| $MM_x$  | spatial migration mismatch |
+| $MM_t$  | temporal migration mismatch |
+| $TE$  | total mismatch |
+| $FE$  | foraging efficiency |
+| $SA$  | spatial adaptation index |

### 2021-07-13 code/analysis/04_ClimateChangeSpaceAnalysis.R

@@ -109,18 +109,20 @@ if(eval){
   require(gtools)
   
   files <- list.files("results/trendstochasticity/raw/")
-  ccmusigma_compiled <- data.frame()
+  ccmusigma <- data.frame()
   for(i in 1:length(files)){ 
     print(files[i])
-    load(paste0("results/trendstochasticity/raw/", files[i]))
-    ccmusigma_compiled <- ccmusigma_compiled %>% 
-      smartbind(ccmusigma %>% mutate(run = i))
+    a <- load(paste0("results/trendstochasticity/raw/", files[i]))
+    print(a)
+    ccmusigma <- ccmusigma %>% 
+      smartbind(get(a) %>% mutate(run = i))
   }
   
-  ccmusigma <- mutate(ccmusigma_compiled, 
+  
+  ccmusigma <- mutate(ccmusigma, 
                       SAI_total = as.numeric(SAI_total),
                       SAI_recent = as.numeric(SAI_recent),
-                      annualFE = as.numeric(annualFE)) %>% unique
+                      annualFE = as.numeric(avgFE)) %>% unique
   
   save(ccmusigma, file = "results/trendstochasticity/trendstochasticity.rda")
 }
@@ -232,3 +234,35 @@ df %>% ggplot(aes(avgFE, SAI_recent)) +
   geom_hline(yintercept = 0, col = "grey") +
   facet_grid(beta_x~psi_x) + geom_path(col = "grey") + theme_few() +
   geom_point(aes( col = factor(kappa)))
+
+
+
+#  Beta_x 0
+
+
+require(memorymigration)
+eval <- FALSE
+if(eval){
+  require(gtools)
+  
+  files <- list.files("results/trendstochasticity/betax0/")
+  ccsigma <- data.frame()
+  for(i in 1:length(files)){ 
+    print(files[i])
+    a <- load(paste0("results/trendstochasticity/betax0/", files[i]))
+    
+    df <- get(a)
+    df$avgFE <- apply(df$annualFE, 1, mean, na.rm = TRUE)
+    ccsigma <- ccsigma %>% 
+      smartbind(df %>% mutate(run = i))
+  }
+  
+  
+  ccsigma <- mutate(ccsigma, 
+                      SAI_total = as.numeric(SAI_total),
+                      SAI_recent = as.numeric(SAI_recent),
+                      annualFE = as.numeric(avgFE)) %>% unique
+  
+  save(ccsigma, file = "results/trendstochasticity/betax0.rda")
+}
+

### 2021-07-13 text/figures/code/Stochasticity.R

@@ -4,36 +4,85 @@ require(ggplot2)
 require(ggthemes)
 
 rm(list=ls())
-b <- load("results/trendstochasticity/trendstochasticity.rda")
-
-cc <- smartbind(ccmusigma, 
-                subset(ccmusigma, psi_x == 0)) %>% 
-  subset(alpha == 100 & beta == 400 & beta_x == 0)
-
-cc.summary <- cc %>% 
-  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
-        summarize, 
-        n = sum(!is.na(avgFE)), 
-        FE_mean = mean(avgFE, na.rm = TRUE), 
-        FE_sd = sd(avgFE, na.rm = TRUE)) %>% 
-  mutate(FE_se = FE_sd/sqrt(n),
-         FE_high = FE_mean + 2*FE_se,
-         FE_low = FE_mean - 2*FE_se)
 
 
-cc.summary %>% subset(lambda > 40) %>% 
-    ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
-    facet_grid(.~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
-    geom_path()
+  
+  
+  b <- load("results/trendstochasticity/betax0.rda")
+  
+  cc.summary <- ccsigma %>% 
+    ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
+          summarize, 
+          n = sum(!is.na(avgFE)), 
+          FE_mean = mean(avgFE, na.rm = TRUE), 
+          FE_sd = sd(avgFE, na.rm = TRUE)) %>% 
+    mutate(FE_se = FE_sd/sqrt(n),
+           FE_high = FE_mean + 2*FE_se,
+           FE_low = FE_mean - 2*FE_se)
+  cc.summary %>% subset(lambda > 40) %>% 
+      ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
+      facet_grid(beta_x~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
+      geom_path()
+  
+  head(cc.summary)
+  
+  png("text/figures/stochasticity.png", width = 2800*2, height = 1000*2, res = 600)
+  ccsigma %>% subset(lambda > 40) %>% 
+    ggplot(aes(kappa, avgFE, col = factor(lambda))) +
[…]

### 2021-07-13 text/SocialMemoryMigration.tex
ms modifications and figure improvements

--- 2021-07-13 text/ms.tex
@@
 
-\linenumbers
+%\linenumbers
 
 \abstract
-
-\emph{to do.}
 
@@
 Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$.
-However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \beta_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \beta_t\,y, \sigma_t)$, where the $\mu$, $\beta$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
-Thus, if $\beta=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
-These trends mirror the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
+However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {\cal N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
+Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
+These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
 The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.
@@
 		& $\sigma_x$, $\sigma_t$ & time duration and and spatial scale of resource pulse \\ 
-		& $dx$, $dt$ & rate of change of peak location and timing of resource \\ 
+		& $\gamma_x$, $\gamma_t$ & rate of change of peak location and timing of resource \\ 
 		& $\psi_x$, $\psi_t$ & standard deviation of peak location and timing \\ 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/Stochasticity.png}
+\includegraphics[width=\textwidth]{figures/stochasticity.png}
 
-\caption{\label{fig_stochasticity} Foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$.
-For the stochastic processes ($\psi_x > 0$), the process was run 90 times, and the mean and standard error of the foraging efficiency over a 50 year post-stabilization period is presented (points and bars).
+\caption{\label{fig_stochasticity} Foraging efficiency (FE) across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of sociality spatial scale $\lambda = 80$ and $120$.
+For the stochastic processes ($\psi_x > 0$), the process was run 90 times for values of $\kappa$.
+Points represent the means of the FE, and shaded areas represent a loess smooth of FE against $\kappa$.
 In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\epsilon = 4$.}
@@
 EG, WFF, GCC and RSC, were supported in part by NSF award DMS 1853478.
-EG and WFF were further supported by NSF grant XXXXXXXXXX.
+EG and WFF were further partially supported by NSF grant IIBR 1915347.
 The authors are grateful to Quentin Read for computational support and advice for cluster-based multi-node multi-core model runs.

### 2021-07-13 text/figures/code/ClimateChangeSpace.R

@@ -56,9 +56,9 @@ png("text/figures/SpatialClimateChange.png", width = 3000, height = 2400, res = 
       mtext(side = 3, bquote(beta == .(b)), cex = 1.2)
     }
     if(b == 400){
-      if(dx == -.25) rightlabel("slow (dx = 0.25)")
-      if(dx == -.5) rightlabel("medium (dx = 0.5)")
-      if(dx == -1) rightlabel("rapid (dx = 1)")
+      if(dx == -.25) rightlabel(expression("slow"~ (gamma[x]== 0.25)))
+      if(dx == -.5) rightlabel(expression("moderate"~ (gamma[x]== 0.5)))
+      if(dx == -1) rightlabel(expression("rapid"~ (gamma[x]==1)))
     }
     if(b == 100){
       axis(2)

### 2021-07-13 text/figures/code/Stochasticity.R

@@ -50,20 +50,71 @@ rm(list=ls())
     box(bty = "l")
   }
   
-  png("text/figures/Stochasticity.png", width = 2400*2, height = 1000*2, res = 600)
+  df <- subset(ccsigma, psi_x == 12)
+  ylim <- c(0.6, 0.72)
+  cols <- c("darkblue", "darkred")
+    
+  
+  names(df)
+  
+  plotFE <- function(df, ...){
+    kappa <-  seq(0,1,.01)
+    
+    loess80 <- loess(avgFE~kappa, data = subset(df,lambda == 80))
+    p80 <- predict(loess80, se = TRUE, newdata = data.frame(kappa = kappa))
+    loess120 <- loess(avgFE~kappa, data = df %>% subset(lambda == 120))
+    p120 <- predict(loess120, se = TRUE, newdata = data.frame(kappa = kappa))
+    
+    predict.df <- data.frame(
+      with(p80, data.frame(kappa, fit80 = fit, low80 = fit - 2*se.fit, high80 = fit + 2*se.fit)),
+      with(p120, data.frame(fit120 = fit, low120 = fit - 2*se.fit, high120 = fit + 2*se.fit)))
+    
+    
+    n <- nrow(predict.df)
+    plot(c(0,1), ylim, type = "n", xlab = "", ylab = "")
+    with(predict.df,{
+        lines(kappa, fit80, col = cols[1], lwd = 2)
+        polygon(c(kappa, kappa[n:1]), c(low80, high80[n:1]), col = alpha(cols[1], .2), 
+                border = NA)
+        
+        lines(kappa, fit120, ylim = ylim, type = "l", col = cols[2], lwd = 2)
+        polygon(c(kappa, kappa[n:1]), c(low120, high120[n:1]), col = alpha(cols[2], .2), 
+                border = NA)
+        box(bty = "l")
+    })
+    with(ddply(df, c("kappa", "lambda"), summarize, FE = mean(avgFE)),
+         points(kappa, FE, col = cols[as.integer(factor(lambda))], cex= 1.25, lwd = 1.25,
+                pch = 21, bg = 0))
+  }
+  
+  # plotFE <- function(df, ...){
+  #   plot(c(0,1), c(.6,.70), type = "n", xlab = "", ylab = "")
[…]

### 2021-07-15 text/SocialMemoryMigration.tex
msmsmsmsmsmsms

@@
 
-\title{Memories of migrations past: Sociality and cognition in dynamic,
-seasonal environments}
+\title{Memories of migrations past: Sociality and cognition in dynamic, seasonal environments}
 
@@
 \affil[a]{\,\, Department of Biology, University of Maryland, College Park, MD 20742, USA.}
-\affil[b]{\,\,Socio-Economic Synthesis Center, SESYNC, Annapolis, MD, USA.}
-\affil[c]{\,\,Department of Mathematics, The University of Miami, Coral Gables, FL 33146}
+\affil[b]{\,\,Socio-Economic Synthesis Center, SESYNC, Annapolis, MD 21401, USA.}
+\affil[c]{\,\,Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.}
 \date{\today}
@@
 Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}.
-This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized.
-This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by a remembered anticipation \citep{Bracis2017}.
+This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized.
+This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.
 
@@
 On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
-Because the scenarios underlying migration are multifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.
+Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.
 
 Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}.
-These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component.
-While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.
-Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics.
+These models are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components.
+While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically.
+Much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics.
+But perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory.
+Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
+While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality is relatively unexplored.
 
-Perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory.
-Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
-The interacting role of memory and sociality, in contrast, has been comparatively little studied.
-
-Here, we develop a diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
+Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
 In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}.
-Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, including both variability or trends in resource distribution or phenology reflecting climate change.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience (or fragility) of a migratory population with memory against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions.
 
@@
 Over-reliance on either strategy should be maladaptive.
-We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic (see also \citet{Lin2021}).
+We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic \citep{Lin2021}.
 
@@
 The third term captures the collective or social advection term of the population via a non-local, density dependent function $v_s(u,x)$.
-If this function takes the form of a convolution around a non-local kernel $k$, i.e.\textasciitilde{} $v_s(u) = k(x) * u(x)$ and that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}.
+If this function takes the form of a convolution around a non-local kernel $k$, i.e.\textasciitilde{} $v_s(u) = k(x) * u(x)$, and if that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}.
 We use the kernel analyzed by \citet{Mogilner1999}: $k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$ The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \widehat{u}$, and in a negative direction when $x > \widehat{u}$.
@@
 
-The migration speed is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (nominally, winter and summer) $t_1$, $dt_1$, $t_2$, $dt_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$.
+The migration speed is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (nominally, winter and summer) $t_1$, $\Delta t_1$, $t_2$, $\Delta t_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$.
 Thus the remembered migratory speed term is a simple step function given by:
@@
 v_m(t, \theta_y) = \begin{cases}
-0; & t > t_1  \,\, \text{and} \,\, t \leq t_1 + dt_1 \\ 
-s_{12}; & t > t_1 + dt_1 \,\, \text{and} \,\,  t \leq t_2 \\ 
-0; & t> t_2 \,\, \text{and} \,\, t \leq t_2 + dt_2 \\ 
-s_{21}; &  t > t_2+dt_2 \,\, \text{or} \,\, t \leq t_1\\ 
+0; & t > t_1  \,\, \text{and} \,\, t \leq t_1 + \Delta t_1 \\ 
+s_{12}; & t > t_1 + \Delta t_1 \,\, \text{and} \,\,  t \leq t_2 \\ 
+0; & t> t_2 \,\, \text{and} \,\, t \leq t_2 + \Delta t_2 \\ 
+s_{21}; &  t > t_2+\Delta t_2 \,\, \text{or} \,\, t \leq t_1\\ 
 \end{cases}
@@
 
-\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + dt_1$, arrive at $x_2$ at $t = t_2$, i.e.~$s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + dt_1)}$ and $s_{21} = \frac{x_1-x_2}{t_1 - (t_2 - \tau + dt_2)}$.
+\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + \Delta t_1$, arrive at $x_2$ at $t = t_2$, i.e.~$s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + \Delta t_1)}$ and $s_{21} = \frac{x_1-x_2}{t_1 - (t_2 - \tau + \Delta t_2)}$.
 This step-like migration function is a one-dimensional version of the migration parameters estimated in empirical studies for individuals \citep{Gurarie2017} and, more relevantly, for populations \citep{Gurarie2019} in empirical studies.
@@
 \noindent where each of the six parameters is updated identically.
-The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$).
+The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, \Delta t'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$).
 The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory.
@@
 
-The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.~${\frac{\partial u(-\chi,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$.
+The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.~${\partial u(-\chi,t)}/{\partial t} = {\partial u(\chi,t)}/{\partial t} =0$.
 As there are no birth or death processes, the total population remains fixed and constant, for convenience integrating to 1.
 Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process.
-Our interest is entirely in the ability of a fixed set of memory and movement parameters to ``navigate'\,' an intra- and interannually dynamic, seasonal environment.
+Our interest is entirely in the ability of a fixed set of memory and movement parameters to navigate an intra- and interannually dynamic, seasonal environment.
 
@@
 This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).
-Both types of resources are illustrated in Figure \ref{fig_ResourceExamples}
+Both types of resources are illustrated in figure \ref{fig_ResourceExamples}.
 
@@
 However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {\cal N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
-Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years, if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
+Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
 These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
@@
 To do this, we compute the total mismatch, i.e.~is a sum of the difference between the migration coefficients and the resource peaks.
-Thus, the spatial mismatch is given by the absolute difference between the migration targets and the resource peaks, i.e.~$MM_x = |x_1 - m_x| + |x_2 + m_x|$.
-The temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.~$MM_t = max\{t_1 - m_1, m_1 - (t_1 + dt_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + dt_2), 0\}$.
-The total mismatch is the sum of $MM_x$ and $MM_t$.
-A mismatch of less than 1 is essentially perfect, we consider a mismatch of 1-5 to be ``good'', and beyond 30 the system can be said to have failed to keep up with climate change.
+Spatial mismatch $\text{MM}_x$ is given by the absolute difference between the migration targets and the resource peaks: $\text{MM}_x = |x_1 - m_x| + |x_2 + m_x|$.
+Temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.~$\text{MM}_t = max\{t_1 - m_1, m_1 - (t_1 + \Delta t_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + \Delta t_2), 0\}$.
+Finally, the total mismatch is the sum of these: $\text{TM} = \text{MM}_x + \text{MM}_t$.
+A mismatch of less than 1 is essentially perfect, we consider a mismatch of 1-5 to be ``good'', and beyond 50 the system can be said to have failed to keep track of the resource dynamics.
 
-To quantify the foraging efficiency, i.e., the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya Coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions.
+To quantify the foraging efficiency, i.e.
+the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions.
 We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE).
 Thus, the foraging efficiency index is:
-$$FE = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
+$$\text{FE} = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
 \noindent where the spatial integral is taken over the domain. This metric is constrained to be between 0 and 1.
 
-For simulations with a constant resource, we ran the model until a quasi-equilibrium state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of 1-1e-5.
-Once that state was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.
+For simulations with a constant resource, we ran the model until a quasi-equilibrium (stationary) state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of $0.9999$.
+Once stationarity was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.
 
-For numerical runs with climate change, we first run a simulation for until equilibrium is attained, and then begin shifting the location of the resource with a trend (slow 0.25, medium 0.5, or rapid 1 units per year), or by adding stochasticity (s.d.
-3, 6, 12).
-For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this as the \emph{spatial adaptation} (SA) index.
-An SA equal to 1 suggests that the process is keeping up with climate change, an SA of 0 indicated that the process is not responding at all to climate change.
-Values greater than 1 (super-adaptation) are possible, as are values less than 1, which correspond to a loss of the migration behavior.
+For numerical runs with climate change, we first run a simulation with a given parameter set until stationarity, as above, and then begin shifting the location of the resource poleward with a slow, moderate or rapid trend ($\gamma_x = 0.25, 0.5$ and 1, respectively), and / or by adding stochasticity (spatial standard deviation 3, 6, 9 or 12).
+For stochasticity analyses, we compare foraging efficiency across a range of the reference memory parameter $\kappa$.
+For analyses that include trend (with or without stochasticity), we quantified the ability of the system to keep track of climate change with a \emph{spatial adaptation} (SA) index.
+This index is the ratio of the slope of the memory-based migration location over time (i.e.~the fitted slope coefficient of $m_{x,i} = \widehat{\gamma_x} i + m_{x,0}$, where $i$ is the year) against the resource drift parameter $\gamma_x$, i.e.~$\text{SA} = \widehat{\gamma}/\gamma_x$.
+An SA equal to 1 suggests that the process is keeping up with climate change, an SA of 0 indicates that the process is not responding at all to climate change.
+Values greater than 1 (super-adaptation) are possible, as are values less than 1, which correspond to a loss of migration behavior.
+
+All movement model parameters, resource parameters, and metrics are summarized in table \label{table_definitions}.
 
@@
 We additionally used the \texttt{nlsLM} function in package \texttt{minpack.LM} \citep{minpack.lm} for robust and fast annual estimation of the migration parameters.
-The complete code is available as an R package (\texttt{memorymigration}) available on GitHub at \url{https://github.com/EliGurarie/memorymigration} and as an interactive Shiny application at \url{https://url.to.be.determined}.
+The complete code is available as an R package (\texttt{memorymigration}) available on GitHub at \url{https://github.com/EliGurarie/memorymigration} and as an interactive Shiny application at \url{https://spot3512.shinyapps.io/memorymigrationshinyapp/}.
 
@@
 \begin{table}[b!]
-	\caption{\label{tab_defs} Table of parameters, variables and metrics.}
+	\caption{\label{table_definitions} Table of parameters, variables and metrics.}
 
@@
 		& $x_1$, $x_2$ & location of population centroids in summer and winter \\ 
-		& $t_1$, $dt_1$ & start and duration of summer season \\ 
-		& $t_2$, $dt_2$ & start and duration of winter season \\ 
+		& $t_1$, $\Delta t_1$ & start and duration of summer season \\ 
+		& $t_2$, $\Delta t_2$ & start and duration of winter season \\ 
 		& $\kappa$ & proportion memory allocated to reference \\
@@
 		\multicolumn{3}{l}{\textbf{Metrics}}\\ 
-		& $MM_x$ & spatial migration mismatch \\ 
-		& $MM_t$ & temporal migration mismatch \\ 
-		& $TE$ & total mismatch \\ 
-		& $FE$ & foraging efficiency \\ 
-		& $SA$ & spatial adaptation index \\ 
+		& $\text{MM}_x$ & spatial migration mismatch \\ 
+		& $\text{MM}_t$ & temporal migration mismatch \\ 
+		& $\text{TM}$ & total mismatch \\ 
+		& $\text{FE}$ & foraging efficiency \\ 
+		& $\text{SA}$ & spatial adaptation index \\ 
 		\hline\hline
@@
 Higher levels of social attraction ($\beta$) led to improved migration matching except in those cases where the sociality scale $\lambda$ was high.
-Thus, for example, at $\lambda = 20$, no simulations at $\beta >= 200$ managed to acquire or maintain a matched migration.
-However, at $\lambda = 50$ or $100$, the migration was slightly better matched at high values of $\beta$.
-(Fig.
-\textbf{X}) The spatial extent of the swarm was a remarkably significant variable.
+Thus, for example, at $\lambda = 20$, no simulations at $\beta \geq 200$ managed to acquire or maintain a matched migration.
+However, at $\lambda = 50$ or $100$, the migration was slightly better matched at high values of $\beta$ (figure \ref{fig_phenologymatching}).
+The spatial extent of the swarm was a remarkably significant variable.
 Smaller swarms were able to match migration only at low values of social attraction ($\beta = 200$), and relatively high values of resource attraction ($\alpha \geq 600$).
@@
 Overall, foraging efficiency was strongly correlated with migration matching, as expected.
-At high mismatch (\textgreater{} 50), foraging efficiency was low (mean 0.29, s.d.
-0.16) compared to the near-perfect matching migrations (mean 0.58, s.d.
-0.14).
-However, somewhat higher mismatch (1 to 5) showed an even higher overall foraging efficiency (mean 0.62, s.d.
-0.18 - see also figure \ref{fig_boxplotFE}).
+At high mismatch ($>50$), foraging efficiency was low (mean 0.29, s.d.~0.16) compared to the near-perfect matching migrations (mean 0.58, s.d.~0.14).
+However, somewhat higher mismatch (1 to 5) showed an even higher overall foraging efficiency (mean 0.62, s.d.~0.18 - see also figure \ref{fig_boxplotFE}).
 
@@
 Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.
-Within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right.
+Within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3 to 15 left to right), and 6 values of resource extent ($\sigma_x$, 3 to 15 bottom to top), as in the zoomed-in panel at (bottom right).
 The color scheme reflects the total mismatch, i.e.~the sum of the absolute differences between the migration timing and locations from the resource peak.
@@
 
-\caption{\label{fig_learningtomigrate} Example of model learning to migrate. The resource is a
-``weakly drifting'' resource and the initial (year 0) condition is non-migratory.
+\caption{\label{fig_learningtomigrate} Example of model learning to migrate.
+The resource is a ``weakly drifting'' resource and the initial (year 0) condition is non-migratory.
 The simulation was run for 100 years, and a sampling of those years (labeled) are presented in panel a: all years from 0 to 10, followed by 20, 40 and 100.
 Otherwise, panels are as in figure \ref{fig_adaptation}.
-Parameter values were $\epsilon = 5$, $\alpha = 500$ and $\beta = 50$.}
+Additional parameter values were $\epsilon = 5$, $\alpha = 500$, $\beta = 50$ and $\lambda = 40$.}
 \end{figure}
@@
 In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).
-The y-axis is the index of spatial adaptation (SA), i.e.~the trend of the memory-driven migration divided by the resource drift trend.
-Values near 1 indicate a behavior that keeps up with climate change, values near indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend.
+The y-axis is the spatial adaptation index (SA), i.e.~the trend of the memory-driven migration divided by the resource drift trend.
+Values near 1 indicate a behavior that keeps up with climate change, values near 0 indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend.
 We compare across spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels).
 The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations.
-The bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.}
+The bottom-right boxplots indicate the final year foraging efficiency against SA; purple and blue boxes indicate the highest values, orange and gray lower values.}
 \end{figure}
@@
 \caption{\label{fig_stochasticity} Foraging efficiency (FE) across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of sociality spatial scale $\lambda = 80$ and $120$.
-For the stochastic processes ($\psi_x > 0$), the process was run 90 times for values of $\kappa$.
-Points represent the means of the FE, and shaded areas represent a loess smooth of FE against $\kappa$.
+For the processes with non-zero stochasticity ($\psi_x > 0$), the process was run 90 times for values of $\kappa$.
+Points represent the average of the FE's across all 50 years and 90 replicates, and shaded areas represent a locall regression (loess) smooth of FE against $\kappa$.
 In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\epsilon = 4$.}
@@
 
-\caption{ \label{fig_trendstochastity} Role of reference memory in adapting to climate change for increasingly stochastic resource dynamics.
-We ran the model with a moderiate rate of climate change (mean shift: 0.5 / year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels).
+\caption{ \label{fig_trendstochasticity} Role of reference memory in adapting to climate change for increasingly stochastic resource dynamics.
+We ran the model with a moderiate rate of climate change (mean shift: 0.5/year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3, 6 and 12, left to right panels).
 For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.
@@
 Animals navigate complex, dynamic, patchy environments.
-When there is a strongly localized and seasonal component to the resource dynamics, typical strategies of spatial distribution by straightforward resource-following taxis necessarily fails to maximally exploit the dynamics.
+When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit the dynamics.
 It is in these cases - quite common in the natural world -- that seasonal migration becomes a viable, even necessary, strategy.
-However, when those resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity.
+However, when resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity.
 It is our conjecture that this plasticity is greatly facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years.
-To our knowledge, this study is the first attempt to focus in and propose a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics.
+Here, we proposed a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics, and a framework within which interacting memory, movement, social and resource dynamics can be explored.
 
 We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
-In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location.
-The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity.
-We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior.
+In essence, our model allowed a population with a hybrid migratory and resource-following behavior to adjust the migratory behavior based on recent experiences with the resource location.
+The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
+We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments.
+Our goal was to understand the combinations of factors that lead to a resilient migration behavior.
 The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.
+Other metrics, such as foraging efficiency in a given season, may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts (e.g., scenarios where there are differing constraints between 'low' and 'high' seasons or where good resource matching is particularly vital at certain stages in an animal's life history like gestation).
+However, we opted for an overall annually averaged metric in these initial explorations because that provides the broadest linkage between resource dynamics and the animals' locations and was consistent with the smooth and continuous nature of the resource distribution.
 
-Several rather general conclusions can be made from our analyses.
-One is that migration in general can be acquired, as has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'', but that this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years.
+Our analyses afford several rather general conclusions.
+One is that migration in general can be acquired, as has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'' (figure \ref{fig_learningtomigrate}).
+However, this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years, findings that echo empirical observations \citep{Jesmer2018}.
 On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, or a shift in some of the system parameters, migration can collapse down into a localized non-migratory, residential behavior, at times with only a marginal cost to the foraging efficiency index.
@@
 
-We also found that the ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics.
+We also found that the ability to maintain, learn and adapt a migration depends strongly on properties of the resource dynamics.
 In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
 Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic.
-This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.
+This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive).
+This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to a rapid and intense local green-up, or, in a more extreme case, when resources are linked to the short-duration early blooming phenology of very particular plants (\textbf{refs}).
 
-The sociality parameters - in particular, the spatial scale $\lambda$ - were somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior.
-In particular, a very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
+The sociality parameters -- in particular, the spatial scale $\lambda$ -- were, somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior.
+In particular, a population very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
 In order to adapt to any shift in resources, \emph{some} individuals must experience that resource, and the two sociality-related mechanisms to have that scope are either a relatively large spatial scale or a relatively weak social cohesion.
-On the other hand, overly large spatial scales compromized the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches.
+On the other hand, overly large spatial scales compromised the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches.
+\emph{Multiple avenues by which 'sociality' (defined as swarm x attraction interplay) can structure migration depending on the relative balance of diffusion and resource following (figure \ref{fig_}).
+For example this could tie into the relative balance between individuals exploring the landscape to acquire new information about resources and their opportunities to interact with other individuals and communicate (or attract).}
 
@@
 This is not a parameter that is typically measured, though it would - in principle - be possible to estimate in a manner analogous to step-selection function using only other conspecifics as opposed to environmental variables as a covariate.
-Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion.
-It does appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology.
+Because both resources and populations were normalized to 1, the ratio between $\alpha$ and $\beta$ can be interpreted as the relative importance of foraging to social cohesion.
+It does appears a certain balance between these two parameters is needed to successfully maintain, acquire, or adapt a migratory phenology.
 
 Finally, the migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}.
-Thus, for example, \citet{Gurarie2019} explicitly esimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration.
-Finally, we note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that is well-documented, in particular, in birds that exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.
+Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration.
+Finally, we note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that are well-documented.
+Birds in particular exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.
 
-Indvidual-based models have shown that Collective knowledge has been shown to be important, if not essential, to the evolution and process of migration \citep{Shaw2013, Guttal2010, Berdahl2018}.
+Individual based models have shown that collective knowledge is important, if not essential, to the evolution and process of migration \citep{Shaw2013, Guttal2010, Berdahl2018}.
 In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior.
@@
 The reference memory parameter $\kappa$ is, of course, impossible to observe directly.
-Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity.
-In general, our model allows us to consider broadly the risks.
+Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity \citep{Abrahms2019,Fagan2019b}.
 
 Importantly, our model did not include any selection, inheritance or birth or death processes, i.e.~there was no evolutionary maximization of a fitness, as in many models that explicitly explore the evolution of migration \citep[e.g.][]{Guttal2010, Shaw2013, Anderson2013}.
-Nonetheless, it is instructive to compare the fundamental assumptions behind such evolutionary individual-based models and a social memory-based model like ours.
-For example, \citet{Anderson2013} similarly explored the resilience of a population under selective pressure under persistent trends and increased stochasticity of a drifting optimal resource window, showing that a certain amount of heritable phenotypic plasticity is necessary to adapt successfully to climate change even at the cost of efficiency.
-Our model underscores the fact that some level of resilience and adapatability can be attained with a purely cognitive process that balances sociality, long, and short term collective memory.
+Nonetheless, it is instructive to compare the fundamental assumptions behind such evolutionary individual based models and a social memory based model like ours.
+For example, \citet{Anderson2013} explored the resilience of a population under selective pressure under persistent trends and increased stochasticity of a drifting optimal resource window, showing that a certain amount of heritable phenotypic plasticity is necessary to adapt successfully to climate change even at the cost of efficiency.
+Our model underscores the fact that some level of resilience and adapatability can be attained with a purely cognitive process that balances sociality with long and short term collective memory.
 Importantly, this knowledge must be transmitted through social and cultural, rather than genetic, pathways.
 The high level of sociality among migratory animals, as well as multi-annual parent offspring bonds, are an evident pathway for that kind of transmission.
-As with those evolutionary models, however, it is clear that when changes are too rapid, no amount of cognition can help
-entirely mitigate against adverse outcomes. Furthermore, if behaviors are not sufficiently plastic (i.e.~if $\kappa$ is too close to 1), then adaptation is that much more difficult.
+As with those evolutionary models, however, it is clear that when changes are too rapid, no amount of cognition can help entirely mitigate against adverse outcomes.
+Furthermore, if behaviors are not sufficiently plastic (i.e.,~if $\kappa$ is too close to 1), then adaptation is that much more difficult.
 
 Rapid environmental change, both global warming and increased anthropogenic development, is causing severe and dramatic impacts to the widespread and generally successful strategy of seasonal migration for many taxa, and the fate of many animal migrations is a topic of increasing concern \citep{Kauffman2021, Wilcove2008}.
-The ability of animals to respond to these changes is deeply dependent on their behavioral plasticity and cognitive abilities.
-The importance of those abilities are in direct proportion to the difficulty in studying them directly.
-By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adapatability and resilience.
+The ability of animals to respond to these changes depends deeply on their behavioral plasticity and cognitive abilities.
+The importance of those abilities is in direct proportion to the difficulty in studying them directly.
+By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adaptability and resilience.
 

### 2021-07-15 text/SocialMemoryMigration_FEE.tex
msmsmsmsmsmsms

--- 2021-07-15 text/SocialMemoryMigration.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[12pt]{article}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
+% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.   
+%                                              %
+%                                                                                                                                                         %
+% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% Version 3.4 Generated 2018/06/15 %%%
+%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
+%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
+%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%
+
+\documentclass[utf8]{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
+%\documentclass[utf8]{frontiersHLTH} % for Health articles
+%\documentclass[utf8]{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles
+%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
+
+\usepackage{url,lineno,microtype,subcaption}
+
+\usepackage[onehalfspacing]{setspace}
+
+\usepackage[colorlinks]{hyperref}
+\hypersetup{citecolor={blue}}
+\hypersetup{urlcolor={blue}}
+\hypersetup{linkcolor={blue}}
+
 \usepackage{lmodern}
@@
 
-\usepackage[left=1in,top=1.25in,right=1in,bottom=1.25in]{geometry}
-\usepackage[running]{lineno}
-
-\usepackage{authblk}
+\linenumbers
 \usepackage{xcolor}
@@
 \usepackage{setspace}
-\usepackage[colorlinks]{hyperref}
-\hypersetup{citecolor={blue}}
-\hypersetup{urlcolor={blue}}
-\hypersetup{linkcolor={blue}}
 
-\usepackage{lineno}
-\usepackage{graphicx}
-\makeatletter
+% Leave a blank line between paragraphs instead of using \\
 
-\usepackage[]{natbib}
-\bibliographystyle{ecology}
+\def\keyFont{\fontsize{8}{11}\helveticabold }
+\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
+\def\Authors{Eliezer Gurarie\,$^{1,2,*}$, 
+	Sriya Potluri\,$^1$, 
+	G.~Christopher Cosner\,$^3$,
+	R.~Stephen Cantrell\,$^3$,
+	William F.~Fagan\,$^1$
+}
+% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
+% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
+\def\Address{$^{1}$Department of Biology, University of Maryland, College Park, MD 20742, USA.\\
+$^{2}$Department of Environmental and Forest Biology, Syracyse, NY   \\
+$^{3}$Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.
+}
+% The Corresponding Author should be marked with an asterisk
+% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
+\def\corrAuthor{Eliezer Gurarie}
 
-\title{Memories of migrations past: Sociality and cognition in dynamic, seasonal environments}
-
-\author[a,b]{Eliezer Gurarie\footnote{Corresponding author: \texttt{egurarie@umd.edu}}}
-\author[a]{Sriya Potluri}
-\author[c]{G.~Christopher Cosner}
-\author[c]{R.~Stephen Cantrell}
-\author[a,b]{William F.~Fagan}
-\renewcommand\Authands{ and }
-
-\affil[a]{\,\, Department of Biology, University of Maryland, College Park, MD 20742, USA.}
-\affil[b]{\,\,Socio-Economic Synthesis Center, SESYNC, Annapolis, MD 21401, USA.}
-\affil[c]{\,\,Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.}
-\date{\today}
+\def\corrEmail{egurarie@umd.edu}
 
 \begin{document}
-\onehalfspacing
+\onecolumn
+\firstpage{1}
+
+\title[Sociality, memory and migration]{Memories of migrations past: Sociality and cognition in dynamic, seasonal environments}
+
+\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
+\address{} %This field will be automatically populated
+\correspondance{} %This field will be automatically populated
+
+\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
+%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}
 
@@
 
-%\linenumbers
+\begin{abstract}
 
-\abstract
+%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
+\section{}
+For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.
+
+As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership.
+References should not be cited in the abstract.
+Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.
+
+\tiny
+ \keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+\end{abstract}
 
@@
 
-Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}.
+Seasonal migrations are widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}.
 For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}.
-The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction \citep{Avgar2014}.
+Essentially, the evolutionary stability of a migratory strategy relies on the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction outweighting the energetic and survival related costs of migration \citep{Avgar2014}.
 
-Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}.
+Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}.
 Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}.
@@
 On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
-Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.
+Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.
 
@@
 Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
-While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality is relatively unexplored.
+While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
 
 Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
-In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience (or fragility) of a migratory population with memory against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions.
+In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change.
 
-We anticipate that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.
+We anticipated that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.
 Over-reliance on either strategy should be maladaptive.
 We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic \citep{Lin2021}.
+Similarly, we anticipated that a balance between very low sociality and extreme sociality would lead to the most resilient migratory process.
 
@@
 Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$.
-However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {\cal N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
-Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
+However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
+Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
 These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
@@
 
-All movement model parameters, resource parameters, and metrics are summarized in table \label{table_definitions}.
+All movement model parameters, resource parameters, and metrics are summarized in table \ref{table_definitions}.
 
@@
 
-\section{Acknowledgments}
+\section{Nomenclature}
 
-EG, WFF, GCC and RSC, were supported in part by NSF award DMS 1853478.
-EG and WFF were further partially supported by NSF grant IIBR 1915347.
-The authors are grateful to Quentin Read for computational support and advice for cluster-based multi-node multi-core model runs.
+\section*{Conflict of Interest Statement}
 
-\nolinenumbers
+The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
 
+\section*{Author Contributions}
+
+EG and WFF provided the original idea. EG and SP developed and ran models and analysis. All authors contributed to the writing.
+
+\section*{Funding}
+
+EG, WFF, GCC and RSC, were supported in part by NSF award DMS 1853478. EG and WFF were further partially supported by NSF grant IIBR 1915347.
+
+\section*{Acknowledgments}
+The authors are grateful to Quentin Read at SESYNC for computational support and advice for multi-node multi-core model runs.
+
+\section*{Data Availability Statement}
+
+The only data analyzed in this manuscript was generated by simulation scripts made available in the GitHub repository at \url{https://github.com/EliGurarie/memorymigration}.
+
+\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
+%\bibliographystyle{frontiersinHLTH&FPHY} % for Health, Physics and Mathematics articles
 \bibliography{../library/memorymigration.bib}
 
-\linenumbers
-
-\clearpage
-\begin{center}
-	\Large Supplementary Materials
-	\end{center}
-
-\normalsize
-\appendix
-\section{Drifting resource}
-The drifting resource function has the following properties
-
-\begin{enumerate}
-\item The total amount of resource across space is constant throughout the year.
-\item At the beginning, middle, and end of the year the resource is uniformly distributed. 
-\item At some peak time $\mu_t < \tau/2$, the resource concentrates at a location $\mu_x < \chi$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
-\item The resource peaks exactly symmetrically at time $\tau - \mu_t$ and location $-\mu_x$ with the same variances.
-\end{enumerate}
-
-To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.
-Thus:
-
-$$h(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$
-
-where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:
-
-$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
-$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$
-
-where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.
-These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.
-
-The means and variances themselves are Gaussian pulses, with the mean peaking at $\mu_x$ at time $\mu_t$ with standard deviation $\sigma_t$ and at $-\mu_x$ at time $\tau - \mu_t$ and the standard deviation pulsing from $2 \chi/\sqrt{12}$ (corresponding to a uniform distribution over the domain $-\chi$ to $\chi$) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.
-
 \end{document}

### 2021-07-15 text/frontiers.tex
msmsmsmsmsmsms

--- 2021-07-15 text/SocialMemoryMigration_FEE.tex
@@
 
-\usepackage{url,lineno,microtype,subcaption}
-
-\usepackage[onehalfspacing]{setspace}
-
-\usepackage[colorlinks]{hyperref}
-\hypersetup{citecolor={blue}}
-\hypersetup{urlcolor={blue}}
-\hypersetup{linkcolor={blue}}
+\usepackage{url,hyperref,lineno,microtype,subcaption}
+%\usepackage[onehalfspacing]{setspace}
 
@@
 \def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{Eliezer Gurarie\,$^{1,2,*}$, 
+\def\Authors{Eliezer Gurarie\,$^{1,2}$, 
 	Sriya Potluri\,$^1$, 
@@
 % Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
-\def\corrAuthor{Eliezer Gurarie}
+\def\corrAuthor{Corresponding Author}
 
-\def\corrEmail{egurarie@umd.edu}
+\def\corrEmail{email@uni.edu}
 
@@
 
-Seasonal migrations are widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}.
+Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}.
 For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}.
-Essentially, the evolutionary stability of a migratory strategy relies on the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction outweighting the energetic and survival related costs of migration \citep{Avgar2014}.
+The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction \citep{Avgar2014}.
 
-Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}.
+Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}.
 Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}.
@@
 On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
-Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.
+Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.
 
@@
 Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
-While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
+While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality is relatively unexplored.
 
 Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
-In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change.
+In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience (or fragility) of a migratory population with memory against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions.
 
-We anticipated that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.
+We anticipate that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.
 Over-reliance on either strategy should be maladaptive.
 We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic \citep{Lin2021}.
-Similarly, we anticipated that a balance between very low sociality and extreme sociality would lead to the most resilient migratory process.
 
@@
 However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
-Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
+Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
 These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
@@
 
-All movement model parameters, resource parameters, and metrics are summarized in table \ref{table_definitions}.
+All movement model parameters, resource parameters, and metrics are summarized in table \label{table_definitions}.
 
@@
 
+\subsection{Figures}
+Frontiers requires figures to be submitted individually, in the same order as they are referred to in the manuscript.
+Figures will then be automatically embedded at the bottom of the submitted manuscript.
+Kindly ensure that each table and figure is mentioned in the text and in numerical order.
+Figures must be of sufficient resolution for publication \href{http://home.frontiersin.org/about/author-guidelines#ResolutionRequirements}{see here for examples and minimum requirements}.
+Figures which are not according to the guidelines will cause substantial delay during the production process.
+Please see \href{http://home.frontiersin.org/about/author-guidelines#GeneralStyleGuidelinesforFigures}{here} for full figure guidelines.
+Cite figures with subfigures as figure \ref{fig:2}B.
+
+\subsubsection{Permission to Reuse and Copyright}
+Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet).
+It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.
+%%Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet).
+It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.
+
+\subsection{Tables}
+Tables should be inserted at the end of the manuscript.
+Please build your table directly in LaTeX.Tables provided as jpeg/tiff files will not be accepted.
+Please note that very large tables (covering several pages) cannot be included in the final PDF for reasons of space.
+These tables will be published as \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} on the online article page at the time of acceptance.
+The author will be notified during the typesetting of the final article if this is the case.
+
 \section{Nomenclature}
 
+\subsection{Resource Identification Initiative}
+To take part in the Resource Identification Initiative, please use the corresponding catalog number and RRID in your current manuscript.
+For more information about the project and for steps on how to search for an RRID, please click \href{http://www.frontiersin.org/files/pdf/letter_to_author.pdf}{here}.
+
+\subsection{Life Science Identifiers}
+Life Science Identifiers (LSIDs) for ZOOBANK registered names or nomenclatural acts should be listed in the manuscript before the keywords.
+For more information on LSIDs please see \href{http://www.frontiersin.org/about/AuthorGuidelines#InclusionofZoologicalNomenclature}{Inclusion of Zoological Nomenclature} section of the guidelines.
+
+\section{Additional Requirements}
+
+For additional requirements for specific article types and further information please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines#AdditionalRequirements}{Author Guidelines}.
+
 \section*{Conflict of Interest Statement}
+%All financial, commercial or other relationships that might be perceived by the academic community as representing a potential conflict of interest must be disclosed.
+If no such relationship exists, authors will be asked to confirm the following statement:
 
@@
 
-EG and WFF provided the original idea. EG and SP developed and ran models and analysis. All authors contributed to the writing.
+The Author Contributions section is mandatory for all articles, including articles by sole authors.
+If an appropriate statement is not provided on submission, a standard one will be inserted during the production process.
+The Author Contributions statement must describe the contributions of individual authors referred to by their initials and, in doing so, all authors agree to be accountable for the content of the work.
+Please see  \href{http://home.frontiersin.org/about/author-guidelines#AuthorandContributors}{here} for full authorship criteria.
 
 \section*{Funding}
-
-EG, WFF, GCC and RSC, were supported in part by NSF award DMS 1853478. EG and WFF were further partially supported by NSF grant IIBR 1915347.
+Details of all funding sources should be provided, including grant numbers if applicable. Please ensure to add all necessary funding information, as after publication this is no longer possible.
 
 \section*{Acknowledgments}
-The authors are grateful to Quentin Read at SESYNC for computational support and advice for multi-node multi-core model runs.
+This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.
+
+\section*{Supplemental Data}
+ \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} should be uploaded separately on submission, if there are Supplementary Figures, please include the caption in the same file as the figure.
+LaTeX Supplementary Material templates can be found in the Frontiers LaTeX folder.
 
 \section*{Data Availability Statement}
-
-The only data analyzed in this manuscript was generated by simulation scripts made available in the GitHub repository at \url{https://github.com/EliGurarie/memorymigration}.
+The datasets [GENERATED/ANALYZED] for this study can be found in the [NAME OF REPOSITORY] [LINK].
+% Please see the availability of data guidelines for more information, at https://www.frontiersin.org/about/author-guidelines#AvailabilityofData
 
@@
 
+%%% Make sure to upload the bib file along with the tex file and PDF
+%%% Please see the test.bib file for some examples of references
+
+\section*{Figure captions}
+
+%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
+%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
+%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
+%%% If using *.tif files convert them to .jpg or .png
+%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=10cm]{logo1}% This is a *.eps file
+\end{center}
+\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=15cm]{logos}
+\end{center}
+\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
+\end{figure}
+
+%%% If you are submitting a figure with subfigures please combine these into one image file with part labels integrated.
+%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
+%%% Frontiers will add the figures at the end of the provisional pdf automatically
+%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.
+
 \end{document}

### 2021-07-15 text/SocialMemoryMigration_FEE.tex
bibibbibib

@@
 \section{}
-For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.
-
-As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership.
-References should not be cited in the abstract.
-Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.
+Seasonal migrations are a widespread and broadly successful strategy for animals to exploit
 
@@
 Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}.
-Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}.
+Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}.
 These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions.
@@
 Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
-On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}.
+On the other hand, migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}.
 Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered.
@@
 In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
-This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change.
+This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change \citep{Park2020}.
 
@@
 
-The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1 ($\int_X h(x,t) dx = 1$).
-This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details).
+The drifting resource differs from the island resource in that the total amount of resource at any given time $\int_X h(x,t) dx = 1$.
+This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see Supplementary Materials for details).
 Both types of resources are illustrated in figure \ref{fig_ResourceExamples}.
@@
 Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic.
-This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive).
-This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to a rapid and intense local green-up, or, in a more extreme case, when resources are linked to the short-duration early blooming phenology of very particular plants (\textbf{refs}).
+This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden.
+This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to an intense green-up period \citep{Park2020} or, for example, when resources are linked to the short-duration early blooming phenology of very particular plants \citep{Post2007}.
 
@@
 
-Finally, the migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}.
+Migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}.
 Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration.
-Finally, we note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that are well-documented.
+We note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that are well-documented.
 Birds in particular exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.
@@
 Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity \citep{Abrahms2019,Fagan2019b}.
+An extremely conservative behavior may be the best way to hedge against stochasticity (high $\kappa$ values in figure \ref{fig_stochasticity}) as it serves no benefit to change behavior based on recent experiences if they do not inform what might happen in the future.
+However, this extreme conservatism is, by definition, incapable of adapting when there is a consistent shift in resource disrtibution (figure \ref{fig_trendstochasticity}).
+In cases where both processes are occurring, we did, as predicted, see a slight improvement in adaptability when some long-term reference memory was used to stability the inclination to adapt, visible as peaks in figure \ref{fig_trendstochasticity} at $\kappa$ values around 0.8.
 
@@
 
+\clearpage
+\appendix
+
+\section{Supplementary material}
+\normalsize
+
+\subsection{Drifting resource}
+The drifting resource function has the following properties
+
+\begin{enumerate}
+	\item The total amount of resource across space is constant throughout the year.
+	\item At the beginning, middle, and end of the year the resource is uniformly distributed. 
+	\item At some peak time $\mu_t < \tau/2$, the resource concentrates at a location $\mu_x < \chi$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
+	\item The resource peaks exactly symmetrically at time $\tau - \mu_t$ and location $-\mu_x$ with the same variances.
+\end{enumerate}
+
+To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.
+Thus:
+
+$$h(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$
+
+where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:
+
+$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
+$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$
+
+where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.
+These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.
+
+The means and variances themselves are Gaussian pulses, with the mean peaking at $\mu_x$ at time $\mu_t$ with standard deviation $\sigma_t$ and at $-\mu_x$ at time $\tau - \mu_t$ and the standard deviation pulsing from $2 \chi/\sqrt{12}$ (corresponding to a uniform distribution over the domain $-\chi$ to $\chi$) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.
+
 \end{document}

### 2021-07-15 text/SocialMemoryMigration_FEE.tex
...

@@
 
-Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
-On the other hand, migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}.
+Whether long distance migration is a Migratory species are often considered to be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
+On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}.
 Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered.

### 2021-07-16 text/SocialMemoryMigration_FEE.tex
...

@@
 This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized.
-This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.
+This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017, Merkle2019, Bauer2020}.
 
-Whether long distance migration is a Migratory species are often considered to be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
+	Migratory species may be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
 On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}.
-Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered.
-On the other hand, a strictly programmed behavior can be maladaptive if conditions change.
-Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.
+A long-distance migration can not rely entirely on local cues, and is therefore only possible if the behavior is hard-programmed or remembered.
+On the other hand, a strictly programmed behavior can be maladaptive if conditions change.While migration can clearly be a highly successful strategy, the resilience of migratory populations and migratory behavior is an open question and depends on the plasticity and adaptability of a population \citep{Xu2021}.
+That plasticity can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citet{Gurarie2017,Xu2021}.
 
-Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}.
-These models are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components.
-While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically.
-Much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics.
-But perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory.
-Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
-While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
+	Cognitive processes, in particular spatial memory, have been shown and argued to be important mechanisms for the reinforcement and maintenance of migration \citep{Merkle2019, Bauer2020}.
+Similarly, sociality and social learning are likely essential to maintaining migration \citep{Guttal2020,Fagan2011,Jesmer2018,Berdahl2018}.
+However, the interacting role of sociality and spatial memory for the plasticity of migration and the resilience of the behavior when faced with a changing environment is largely unstudied.
+Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.
 
-Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
+	Here, we develop a diffusion-advection model with sociality and memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.
 In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior.
@@
 
-Animals navigate complex, dynamic, patchy environments.
+	Animals navigate complex, dynamic and patchy environments.
 When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit the dynamics.
-It is in these cases - quite common in the natural world -- that seasonal migration becomes a viable, even necessary, strategy.
+It is in these cases, quite common in the natural world, that seasonal migration becomes a viable, even necessary, strategy.
 However, when resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity.
-It is our conjecture that this plasticity is greatly facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years.
-Here, we proposed a relatively simple, memory-driven mechanism by which a population can respond to long-term changes in resource dynamics, and a framework within which interacting memory, movement, social and resource dynamics can be explored.
+It is our conjecture that this plasticity is facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years, combined with a social dynamic.
+The relatively simple, social and memory-driven mechanism we  propose here was, in fact, able to adapt to long-term changes in resource dynamics, even with inter-annual stochasticity.
+The model thereby provides a framework with which the interaction  of memory, movement, social and resource dynamics can be explored.
 
@@
 In essence, our model allowed a population with a hybrid migratory and resource-following behavior to adjust the migratory behavior based on recent experiences with the resource location.
-The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
-We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments.
+The model was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
+In a generalized way, it captured all three fundamental types of migration plasticity: the \textit{where}, the \textit{when} and the \textit{whether} \citep{Xu2021}.
+
 	Our goal was to understand the combinations of factors that lead to a resilient migration behavior.
-The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.
-Other metrics, such as foraging efficiency in a given season, may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts (e.g., scenarios where there are differing constraints between 'low' and 'high' seasons or where good resource matching is particularly vital at certain stages in an animal's life history like gestation).
-However, we opted for an overall annually averaged metric in these initial explorations because that provides the broadest linkage between resource dynamics and the animals' locations and was consistent with the smooth and continuous nature of the resource distribution.
+We used foraging efficiency as a convenient metric of the utility of migration, but this was not a measure explicitly maximized by the model.
+Other metrics, such as foraging efficiency in a given season, or probability of survival or reproduction relative to resource availability \citep{Bauer2020} may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts.
+However, the overall annually averaged metric provided the broadest linkage between resource dynamics and animals' locations and was consistent with the minimal biological assumptions and generality of our framework.
 
-Our analyses afford several rather general conclusions.
-One is that migration in general can be acquired, as has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'' (figure \ref{fig_learningtomigrate}).
-However, this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years, findings that echo empirical observations \citep{Jesmer2018}.
-On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, or a shift in some of the system parameters, migration can collapse down into a localized non-migratory, residential behavior, at times with only a marginal cost to the foraging efficiency index.
-This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}.
+	Among the general conclusions from our model, one is that migration can be acquired in a seasonal environment, even with no strong intrinsic propopensity to migrate and a weak phenological resource pulse to follow.
+This has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'' (figure \ref{fig_learningtomigrate}).
+Learning migration, howewer, requires a very strong resource attraction, higher levels of exploratory behavior (e.g.~diffusion, and larger spatial scale of sociliaty), and -- often --- many more years.
+These findings echo empirical observations \citep{Jesmer2018}.
+On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, a shift in some of the system parameters, or even a shift in the spatial and temporal extent of resources, migration can collapse and turn into a non-migratory, residential behavior (figure \ref{fig_phenologymatching}).
+This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}, as well as the wide range of migration plasticity shown in wild populations, even within a species \citep{Xu2021}.
 
-We also found that the ability to maintain, learn and adapt a migration depends strongly on properties of the resource dynamics.
+	The ability to maintain, learn and adapt a migration also depends strongly on properties of the resource dynamics.
 In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
@@
 This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden.
-This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to an intense green-up period \citep{Park2020} or, for example, when resources are linked to the short-duration early blooming phenology of very particular plants \citep{Post2007}.
+This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to an intense green-up period \citep{Park2020} or, for example, when resources are linked to the short-duration early blooming phenology of very particular plants \citep{Post2007, Renner2018}.
 
-The sociality parameters -- in particular, the spatial scale $\lambda$ -- were, somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior.
-In particular, a population very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
+	\textit{The sociality parameters -- in particular, the spatial scale $\lambda$ -- were, unexpectedly, the most important parameters for determining the resiliency of the process.
+In particular, a population with a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
 In order to adapt to any shift in resources, \emph{some} individuals must experience that resource, and the two sociality-related mechanisms to have that scope are either a relatively large spatial scale or a relatively weak social cohesion.
 On the other hand, overly large spatial scales compromised the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches.
-\emph{Multiple avenues by which 'sociality' (defined as swarm x attraction interplay) can structure migration depending on the relative balance of diffusion and resource following (figure \ref{fig_}).
-For example this could tie into the relative balance between individuals exploring the landscape to acquire new information about resources and their opportunities to interact with other individuals and communicate (or attract).}
+Thus, there appear to be multiple avenues by which 'sociality' can structure migration.
+For example this could tie into the relative balance between individuals exploring the landscape to acquire new information about resources and their opportunities to interact with other individuals and communicate (or attract).
+	}
 
@@
 By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adaptability and resilience.
+
+	Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}.
+These models are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components.
+While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically.
+Much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics.
+But perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory.
+Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
+While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
 

### 2021-07-16 text/SocialMemoryMigration_FEE.tex
...

@@
 
-	Migratory species may be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
-On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}.
+	Migratory species may be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Seebacher2015, Kauffman2021}.
+On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}, and the cognitive and social factors determining the resilience of migrations to climate change are generally unknown.
 A long-distance migration can not rely entirely on local cues, and is therefore only possible if the behavior is hard-programmed or remembered.
 On the other hand, a strictly programmed behavior can be maladaptive if conditions change.While migration can clearly be a highly successful strategy, the resilience of migratory populations and migratory behavior is an open question and depends on the plasticity and adaptability of a population \citep{Xu2021}.
-That plasticity can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citet{Gurarie2017,Xu2021}.
+That plasticity can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citet{Gurarie2017, Xu2021}.
 

### 2021-07-20 text/SocialMemoryMigration_FEE.tex
pushy pushy

--- 2021-07-16 text/SocialMemoryMigration_FEE.tex
@@
 \usepackage[onehalfspacing]{setspace}
+
+\usepackage{braket}
 
@@
 \def\Address{$^{1}$Department of Biology, University of Maryland, College Park, MD 20742, USA.\\
-	$^{2}$Department of Environmental and Forest Biology, Syracyse, NY   \\
+	$^{2}$Department of Environmental and Forest Biology, SUNY College of Environmental Science and Forestry, Syracyse, NY 13210, USA.\\
 	$^{3}$Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.
@@
 \def\corrAuthor{Eliezer Gurarie}
-
 \def\corrEmail{egurarie@umd.edu}
@@
 		%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
-		\section{}
-		Seasonal migrations are a widespread and broadly successful strategy for animals to exploit
+	Seasonal migrations are a widespread and broadly successful strategy for animals to exploit periodic and localized resources over large spatial scales.
+It remains an open and largely case-specific question whether long-distance migrations are resilient to environmental disruptions.
+High levels of mobility suggest an ability to shift ranges that can confer resilience.
+On the the other hand, a conservative, hard-wired commitment to a risky behavior can be costly if conditions change.
+
+	Mechanisms that contribute to migration include identification and responsiveness to resources, sociality, and cognitive processes spatial memory and learning.
+Our goal was to explore the extent to which these factors interact not only to maintain a migratory behavior but also to provide resilience against environmental changes.
+We develop a relatively simple diffusion-advection model of animal movement in which an endogenous migratory behavior is modified by recent experiences via a memory process, and animals have a social swarming-like behavior over a range of spatial scales.
+
+	We found that this relatively simple framework was able to adapt to a stable, seasonal resource dynamic under a fairly broad range of parameter values.
+Furthermore, the model was able to acquire an adaptive migration behavior with time.
+However, the resilience of the process depended on all the parameters under consideration, with many complex trade-offs.
+For example, the spatial scale of sociality needed to be large enough to capture changes in the resource, but not so large that the acquired collective information was overly diluted.
+A long-term reference memory was important for hedging against a highly stochastic process, but a higher weighting of more recent memory was needed for adapting to directional changes in resource phenology.
+
+	Our model provides a general and versatile framework for exploring the interaction of memory, movement, social and resource dynamics, even as environmental conditions globally are undergoing rapid change.
 
 		\tiny
-		\keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+		\keyFont{ \section{Keywords:} Sociality, seasonal migration, collective knowledge, diffusion-advection models, spatial memory} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
 	\end{abstract}
@@
 For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}.
-Essentially, the evolutionary stability of a migratory strategy relies on the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction outweighting the energetic and survival related costs of migration \citep{Avgar2014}.
+The evolutionary stability of a migratory strategy essentially relies on the fitness benefits of accessing seasonal resources, whether for energetic gain, predator avoidance, or a suitable  environment for reproduction, outweighing the energetic and survival related costs of migration \citep{Avgar2014}.
 
-	Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}.
+	Proximate causes, drivers and mechanisms for migration vary widely across and even within species \citep{Berthold1999, Shaw2016}.
 Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}.
@@
 Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}.
-This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized.
+This second behavior involves the greatest trade-off between the costs and benefits of accessing those highly seasonal and localized resources.
 This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017, Merkle2019, Bauer2020}.
 
-	Migratory species may be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}.
-On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}.
-A long-distance migration can not rely entirely on local cues, and is therefore only possible if the behavior is hard-programmed or remembered.
-On the other hand, a strictly programmed behavior can be maladaptive if conditions change.While migration can clearly be a highly successful strategy, the resilience of migratory populations and migratory behavior is an open question and depends on the plasticity and adaptability of a population \citep{Xu2021}.
-That plasticity can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citet{Gurarie2017,Xu2021}.
+	While migration can be a very successful strategy, with migratory ecotypes of the same species often outnumbering non-migratory conspecifics.
+Migratory caribou and reindeer \emph{Rangifer tarandus}, for example, are several orders of magnitude more abundant than non-migratory woodland, mountain and forest ecotypes \citep{FestaBianchet2011,Uboni2016}.
+However, the question of whether migratory animals are more or less resilient to environmental disruptions in the environment remains open and largely case-specific \citep{Moore2008, HardestyMoore2017, Xu2021}.
+On the one hand, migratory species may be more vulnerable as disruptions in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Seebacher2015, Kauffman2021}.
+On the other hand, migratory species might be more resilient due to their general wide-ranging mobility \citep{Robinson2009}.
+The resilience of a migratory populations depends on the plasticity and adaptability of a population, which can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citep{Gurarie2017, Xu2021}.
 
-	Cognitive processes, in particular spatial memory, have been shown and argued to be important mechanisms for the reinforcement and maintenance of migration \citep{Merkle2019, Bauer2020}.
-Similarly, sociality and social learning are likely essential to maintaining migration \citep{Guttal2020,Fagan2011,Jesmer2018,Berdahl2018}.
-However, the interacting role of sociality and spatial memory for the plasticity of migration and the resilience of the behavior when faced with a changing environment is largely unstudied.
+	Cognitive processes, in particular spatial memory, have been shown to be important mechanisms for the reinforcement and maintenance of migration \citep{Merkle2019, Bauer2020}.
+Similarly, sociality and social learning are likely essential to maintaining migration \citep{Guttal2010, Fagan2011, Jesmer2018, Berdahl2018}.
+However, the interacting role of sociality and spatial memory for the plasticity of migration and the resilience of the behavior when faced with a changing environment are generally unknown, though it has been hypothesized that the importance of these cognitive processes depend on the predictability of these resources \citep{RiotteLambert2020}.
 Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.
@@
 \section{Methods}
-
+~
 \subsection{Memory movement model}
 
-	In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or \emph{reference}) memory of large-scale movement behavior, (4) a short-term (or \emph{working}) memory that updates movement behavior based on recent experience, and (5) some social aspect to the learned behavior.
+	In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) sociality in the movements, (4) a long-term (or \emph{reference}) memory of large-scale movement behavior, and (5) a short-term (or \emph{working}) memory that updates movement behavior based on recent experience.
 
@@
 	\begin{equation}\label{eq_mainmodel}
-		{\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
+		-{\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
 		\alpha \frac{\partial}{\partial x}\left(u \frac{\partial h}{\partial x}\right) + 
-		\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + 
-		v_m(t)
+		\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + \frac{\partial}{\partial x}\left(u\,v_m(t)\right)
 		\end{equation}
@@
 This is the well-studied standard chemotaxic resource-following behavior.
-We borrow the general notation from our earlier related work \citep{Fagan2017, Fagan2019}.
+We borrow the general notation from earlier related work expanding diffusion-advection models to incorporate non-local information \citep{Fagan2017} and behavioral switching \citep{Fagan2019}.
 
 	The third term captures the collective or social advection term of the population via a non-local, density dependent function $v_s(u,x)$.
-If this function takes the form of a convolution around a non-local kernel $k$, i.e.\textasciitilde{} $v_s(u) = k(x) * u(x)$, and if that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}.
-We use the kernel analyzed by \citet{Mogilner1999}: $k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$ The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \widehat{u}$, and in a negative direction when $x > \widehat{u}$.
-The parameter $\lambda$ is a length scale of sociality - roughly one-half the size of the ``swarm'', where $\beta$ is a parameter that quantifies the overall strength of sociality.
+If this function takes the form of a convolution around a non-local kernel $k$, i.e.~$v_s(u) = k(x) * u(x)$, and if that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}.
+We use the kernel analyzed by \citet{Mogilner1999}:
+
+	$$k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$$
+
+	\noindent The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \braket{u}$, and in a negative direction when $x > \braket{u}$, where $\braket{u}$ is the mean location of the population.
+The parameter $\lambda$ is a length scale of sociality, roughly one-half the size of the swarm, and $\beta$ is a parameter that quantifies the overall strength of sociality.
 
@@
 This term evolves with a set of parameters $\theta_y$ that slowly change each year $y \in \{0,1,2,...\}$, i.e.~the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.
-
-	The migration speed is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (nominally, winter and summer) $t_1$, $\Delta t_1$, $t_2$, $\Delta t_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$.
-Thus the remembered migratory speed term is a simple step function given by:
+The migration is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (e.g.~summer and winter) $t_1$, $\Delta t_1$, $t_2$, $\Delta t_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$.
+The remembered migratory speed term is a simple step function given by:
 
@@
 
-	\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + \Delta t_1$, arrive at $x_2$ at $t = t_2$, i.e.~$s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + \Delta t_1)}$ and $s_{21} = \frac{x_1-x_2}{t_1 - (t_2 - \tau + \Delta t_2)}$.
-This step-like migration function is a one-dimensional version of the migration parameters estimated in empirical studies for individuals \citep{Gurarie2017} and, more relevantly, for populations \citep{Gurarie2019} in empirical studies.
+	\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, depart at $t = t_1 + \Delta t_1$, arrive at $x_2$ at $t = t_2$, and depart at $t_2 + \Delta t_2$.
+Thus, $s_{12} = (x_2-x_1)/ (t_2 - (t_1 + \Delta t_1))$ and $s_{21} = (x_1-x_2)/(t_1 - (t_2 - \tau + \Delta t_2))$.
+This step-like migration function is a one-dimensional version of the migration parameters estimated for individuals \citep{Gurarie2017} and populations \citep{Gurarie2019} in empirical studies.
 
@@
 
-	$$\theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y}$$
+	\begin{equation}\label{eq_memory}
+	\theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y} 
+\end{equation}
 
-	\noindent where each of the six parameters is updated identically.
+	\noindent where each of the six parameters is updated according to equaton \ref{eq_memory} identically.
 The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, \Delta t'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$).
@@
 
-	The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.~${\partial u(-\chi,t)}/{\partial t} = {\partial u(\chi,t)}/{\partial t} =0$.
+	The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries. Formally, this no-flux condition means the following conditions must be met
+
+	$$
+	\begin{cases}
+		\varepsilon \frac{\partial u}{\partial x} - 		
+	\alpha \left(u \frac{\partial h}{\partial x}\right) - \beta(v_x(u)) - (uv_m(t)) = 0 & \text{where}\,\, x = \chi\\
+	-\varepsilon \frac{\partial u}{\partial x} - 
+	\alpha \left(u \frac{\partial h}{\partial x}\right) - \beta(v_x(u)) - (uv_m(t)) = 0 & \text{where}\,\, x = -\chi
+	\end{cases}
+$$
+
+	\noindent In practice, the design of our resource space (see below) and other parameterization lead 0 or near 0 values of both $h(x)$ and $u(x)$, and the simpler ${\partial u(-\chi,t)}/{\partial t} = {\partial u(\chi,t)}/{\partial t} =0$ boundary condition provides a good approximation.
+
 	As there are no birth or death processes, the total population remains fixed and constant, for convenience integrating to 1.
 Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process.
-Our interest is entirely in the ability of a fixed set of memory and movement parameters to navigate an intra- and interannually dynamic, seasonal environment.
+Our interest is in the ability of a fixed set of movement and memory parameters to navigate an intra- and interannually dynamic, seasonal environment.
 
@@
 
-	We solved the model numerically on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.\textasciitilde years of 100 days).
+	We ran this model on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day year~s).
 We were interested in an approximately periodic resource dynamic, i.e.~one in which $h(x,t \approx h(x, t-\tau))$.
@@
 	Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$.
-However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
-Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses.
-These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate.
-The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.
+However, in scenarios exploring climate change we allow the peaks to vary with directional trend and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse.
+Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a shift of the resource towards the extremes of the domain.
+While we did note explore phenological shifts in timing, those can readily be modeled as well.
+These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate \citep{Renner2018}.
+The spatial and temporal scales of the resource peak ($s_x$ and $s_t$) remain constant in all of our simulations.
 
@@
 
-	The main metrics we are interested in are \emph{migration mismatch}, \emph{foraging efficiency} and \emph{adaptation to trends}.
+	The main metrics we were interested in are \emph{migration mismatch}, \emph{foraging efficiency} and \emph{adaptation to directional trends}.
 
-	Migration mismatch captures the similarity between the migration phenology and the resource phenology.
-To do this, we compute the total mismatch, i.e.~is a sum of the difference between the migration coefficients and the resource peaks.
-Spatial mismatch $\text{MM}_x$ is given by the absolute difference between the migration targets and the resource peaks: $\text{MM}_x = |x_1 - m_x| + |x_2 + m_x|$.
+	Migration mismatch captures the combined difference between the migration phenology and the resource phenology in time and space.
+Spatial mismatch $\text{MM}_x$ is the absolute difference between the migration targets and the resource peaks: $\text{MM}_x = |x_1 - m_x| + |x_2 + m_x|$.
 Temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.~$\text{MM}_t = max\{t_1 - m_1, m_1 - (t_1 + \Delta t_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + \Delta t_2), 0\}$.
-Finally, the total mismatch is the sum of these: $\text{TM} = \text{MM}_x + \text{MM}_t$.
-A mismatch of less than 1 is essentially perfect, we consider a mismatch of 1-5 to be ``good'', and beyond 50 the system can be said to have failed to keep track of the resource dynamics.
+Thus, the total mismatch is the sum of these: $\text{TM} = \text{MM}_x + \text{MM}_t$.
+A mismatch of less than 1 is essentially perfect, a mismatch of 1-5 we consider excellent, and beyond 50 the system can be said to have failed to keep track of the resource dynamics.
 
-	To quantify the foraging efficiency, i.e.
-the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions.
+	To quantify the foraging efficiency, i.e.~the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions.
 We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE).
 Thus, the foraging efficiency index is:
+
 	$$\text{FE} = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
+
 	\noindent where the spatial integral is taken over the domain. This metric is constrained to be between 0 and 1.
 
-	For simulations with a constant resource, we ran the model until a quasi-equilibrium (stationary) state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of $0.9999$.
+	For simulations with a constant resource, we ran the model until a quasi-equilibrium (stationary) state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of $0.99999$.
 Once stationarity was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.
@@
 For stochasticity analyses, we compare foraging efficiency across a range of the reference memory parameter $\kappa$.
-For analyses that include trend (with or without stochasticity), we quantified the ability of the system to keep track of climate change with a \emph{spatial adaptation} (SA) index.
-This index is the ratio of the slope of the memory-based migration location over time (i.e.~the fitted slope coefficient of $m_{x,i} = \widehat{\gamma_x} i + m_{x,0}$, where $i$ is the year) against the resource drift parameter $\gamma_x$, i.e.~$\text{SA} = \widehat{\gamma}/\gamma_x$.
+For analyses that included directional trends, with or without stochasticity, we quantified the ability of the system to keep track of climate change with a \emph{spatial adaptation} (SA) index.
+This index is the ratio of the slope of the memory-based migration location over time, i.e.~$\text{SA} = \widehat{\gamma}/\gamma_x$  where the adaptation slope estimate is the regression coefficient of the spatial coordinate of the migration against time (i.e.~$m_{x,i} = \widehat{\gamma_x} i + m_{x,0}$, where $i$ is the year), and $\gamma_x$ is the rate of drift of the resource peak (table \ref{table_definitions}).
 An SA equal to 1 suggests that the process is keeping up with climate change, an SA of 0 indicates that the process is not responding at all to climate change.
@@
 (4) Can this model adapt when the resource peaks shifts in space?
+Details of parameter combinations and reported metrics are provided in respective results sections.
 
@@
 			\multicolumn{3}{l}{\textbf{Memory migration model}}\\ 
-			& $\epsilon$ & Diffusion \\ 
-			& $\alpha$ & Forage following \\ 
+			& $\varepsilon$ & Diffusion \\ 
+			& $\alpha$ & Strength of resource following \\ 
 			& $\beta$ & Strength of sociality \\ 
@@
 
-	\subsection{Adaptation to resource phenology}
+\subsection{Adaptation to resource phenology}\label{results_adaptation}
 
@@
 Notably, the path to this equilibrium is somewhat indirect, with the later winter range taking more time to stabilize than the earlier summer range.
-The eventual steady state is one where the foraging efficiency is relatively high.
+The eventual steady state is one where the foraging efficiency is relatively high, near 0.6 compared to an initial value of 0.3.
+However, the increase in the foraging efficiency was not entirely monotonic, as the system moved through some slightly sub-optimal stages in adjusting its migration behavior.
 
-	We ran this process for 8100 parameter combinations (figure \ref{fig_phenologymatching}).
-The fundamental dynamic of attaining and maintaining a well-matched migration phenology, can occur under many combinations of parameter values, but all parameters play a role.
+	We ran this process for 8100 parameter combinations crossing different values of the movement process ($\alpha$, $\beta$ and $\lambda$) and resource characteristics ($\sigma_x$ and $\sigma_t$), and present the total mismatch (TM) against all those combinations in figure \ref{fig_phenologymatching}.
+In all of these simulations, memory was entirely recent ($\kappa = 0$), since there can be no benefit to relying on a sub-optimal reference memory.
+We compared a set of diffusion rates $\varepsilon$ between 1 and 8, but only illustrate results for $\varepsilon = 4$.
+
+	A well-matched migration phenology (TM < 5) occurred under very many combinations of parameter values, but all parameters play interacting roles.
 Among the more intuitive results are that greater values of $\alpha$ (resource following) lead to an improved ability to match the migration.
 Resource peaks with larger spatial extent (higher $\sigma_x$) are generally better for migration matching.
-A larger set of parameters matched migration with low diffusion than high diffusion.
 
@@
 
-	Random forest analyses, whether on the log of total mismatch or on the classification of a perfect match, uniformly show that the most important variables \citep{Breiman2001} were $\alpha$ and $\lambda$ (4.14 and 4.02 proportional increase in MSE), and the least important was $\sigma_t$, with a 0.5 proportional increase (table \textbf{X}).
+	Random forest analyses, whether on the log of total mismatch or on the classification of a perfect match, uniformly show that the most important variables \citep{Breiman2001} were $\alpha$ and $\lambda$ (4.14 and 4.02 proportional increase in MSE), and the least important was $\sigma_t$, with a 0.5 proportional increase.
 
@@
 		\caption{\label{fig_phenologymatching} Migration phenology matching across six model parameters.
-Low and high diffusion ($\epsilon = 1$ and $8$ in upper and lower panel blocks), tight, medium and loose swarms ($\lambda = 20, 50, 100$) left to right panels.
+Low and high diffusion ($\varepsilon = 1$ and $8$ in upper and lower panel blocks), tight, medium and loose swarms ($\lambda = 20, 50, 100$) left to right panels.
 Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.
-Within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3 to 15 left to right), and 6 values of resource extent ($\sigma_x$, 3 to 15 bottom to top), as in the zoomed-in panel at (bottom right).
-The color scheme reflects the total mismatch, i.e.~the sum of the absolute differences between the migration timing and locations from the resource peak.
-White squares represent runs that numerically failed to estimate migration parameters.}
+Within each of the combinations of $\varepsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3 to 15 left to right), and 6 values of resource extent ($\sigma_x$, 3 to 15 bottom to top), as in the zoomed-in panel (bottom right; drawn from $\alpha = 200$, $\beta = 200$, $\lambda = 100$).
+The color scheme reflects the total mismatch, i.e.~the sum of the absolute differences between the migration timing and locations from the resource peak.}
 	\end{figure}
@@
 Otherwise, panels are as in figure \ref{fig_adaptation}.
-Additional parameter values were $\epsilon = 5$, $\alpha = 500$, $\beta = 50$ and $\lambda = 40$.}
+Additional parameter values were $\varepsilon = 5$, $\alpha = 500$, $\beta = 50$ and $\lambda = 40$.}
 	\end{figure}
@@
 	Figure \ref{fig_learningtomigrate} illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.
-In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
+In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\varepsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).
 The qualitative behavior of this process was to start drifting towards the summer resource, while slowly developing a weak pulse towards the winter resource as well.
-After first locking in on the summer resource, the winter migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource can be consistently reached.
+After first locking in on the summer resource, the winter migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource were consistently reached.
 
 	The model had, in general, a difficult time learning migration from a non-migratory initial condition.
-Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5.
+Out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5.
 Conditions that were more conducive to learning migration were pulses of \emph{longer} duration (high $\sigma_t$), but \emph{smaller} in scope (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact in space but long enough in duration to lock in to the memory.
 
-	\subsection{Adapting to climate change}
+\subsection{Directional climate change}
 
@@
 
-	To assess the ability of the system to adapt to a trending climate, we generated scenarios with slow (0.25 units / year), medium (0.5 units / year), and fast (1 unit / year) drift outward of the two resource pulses.
-We then assessed 40 parameter combinations for each of those scenarios, a high and low value of resource following ($\alpha = 400$ and $100$) crossed with a high and low value of sociality ($\beta = 400$ and $100$) crossed with 10 values of the spatial scale of sociality ($\lambda = 20$ to $200$).
-The spatial and temporal scale of the resource pulses were fixed to a relatively ``easy'' to adapt to $\sigma_x = 12$ and $\sigma_t = 6$.
+	To assess the ability of the system to adapt to a trending climate, we generated scenarios with slow, moderate and fast outward directional shifts in the resource peak (0.25, 0.5 and 1 units / year, respectively).
+We then assessed 40 parameter combinations for each of those scenarios, high and low values of resource following ($\alpha = 400$ and $100$), high and low values of sociality ($\beta = 400$ and $100$) and 10 values of the spatial scale of sociality ($\lambda = 20$ to $200$).
+The spatial and temporal scale of the resource pulses were fixed to $\sigma_x = 12$ and $\sigma_t = 6$, a combination which analyses in section \ref{results_adaptation} showed were generally ``easy'' to adapt to.
 We computed the adaptation index and foraging efficiency for each of the 120 runs (figure \ref{fig_adaptationclimate}).
-We were interested in the dynamics against $\lambda$ due to the uniformly high importance of this parameter for determining the success of this process to match migration in steady states.
+We were interested in the dynamics against $\lambda$ due to the consistently high importance of this parameter for matching migration in steady states.
+Our main index of interest was the spatial adaptation (SA) to trends.
 
@@
 However, that maximum value is still less than 1, suggesting that truly matching a steadily drifting trend is very difficult.
-Small swarms ($\lambda < 50$) have a very hard time adapting when the social attraction is high, but do fairly well when social attraction is lower.
-But larger sized swarms do progressively worse across more parameterizations, e.g.~in the most rapid climate change scenario, the SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to 200 (essentially, the entire spatial domain).
+Smaller social spatial scales ($\lambda < 50$) have a very hard time adapting when the social attraction is high, but do fairly well when social attraction is low.
+Larger sized swarms do progressively worse across more parameterizations, e.g.~in the most rapid climate change scenario, the SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to 200 (encompassing, essentially, the entire spatial domain).
 
@@
 Notably, no parameter combination at this value comes close to keeping up with the rapid climate change (SA range -0.64 to 0.13).
-For slower climate change, however, there is a window of values for the swarm size between 40 and 80, where the SA \emph{exceeds} 1, and then crashing quite rapidly to negative values of SA as that swarm size increases.
+For slower climate change, however, there is a window of values for the swarm size between 40 and 80, where the SA \emph{exceeds} 1, but then crashes quite rapidly to negative values of SA as that swarm size increases.
 These ``super-adaptive'' processes indicate a unique sweet spot where a swarm is large enough to capture and adapt to the drifting resource, but not so large that the information gathered in a given year is too weak to adjust the migratory behavior in a following year.
@@
 
-	While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, we hypothesized that more conservative approach, relying on a ``reference memory'' may be more beneficial when conditions change stochastically.
+	While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, we hypothesized that more conservative approach that relies on a reference memory may be beneficial when conditions change stochastically.
 We tested this hypothesis by solving a set of models across a range of $\kappa$ values fom 0 (all recent memory) to 1 (all reference memory).
 In these scenarios, we ran the system for as many years as needed with no stochasticity to acquire a stationary state (i.e.~similarity index greater than 1-1e-6).
-We then used the stationary state as the reference memory, and then ran the process for an additional 50 years with a stochasticity (i.e.~standard deviation in peak location of the resource) ranging from 0 to 12.
+We then used the stationary state as the reference memory, and then ran the process for an additional 50 years with a stochasticity (i.e.~standard deviation in peak location of the resource) ranging from 0 to 12, and present the resulting average foraging efficiency (figure \ref{fig_stochasticity}).
 
-	Figure \ref{fig_stochasticity} illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency, with some variation across the spatial scale of sociality.
-When that scale of sociality is high enough (e..g $\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.
-Overall, as anticipated, the greater the stochasticity, the lower the foraging efficiency; and the greater the stochasticity, the greater the benefit of a conservative reference memory driven strategy.
+	Overall, as expected, the greater the stochasticity, the lower the foraging efficiency.
+Further, as we predicted, highest level of $\kappa$ can significantly help foraging efficiency, with some variation across the spatial scale of sociality, especially in more highly stochastic scenarios.
+When that scale of sociality is high enough ($\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.
 
@@
 Points represent the average of the FE's across all 50 years and 90 replicates, and shaded areas represent a locall regression (loess) smooth of FE against $\kappa$.
-In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\epsilon = 4$.}
+In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\varepsilon = 4$.}
 
@@
 
-	We added trends to the stochastic process described above, adding 30 years of spatial outward trend in the resource.
-Our goal was to assess the ability of the process to keep up with climate change as a function of the reference memory parameter (figure \ref{fig_trendstochasticity}).
+	We added 30 years of directional trends to the variously stochastic process described above, and assessed the adaptation index against the reference memory parameter $\kappa$ (figure \ref{fig_trendstochasticity}).
 Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change, leading to an adaptation index of 0.
@@
 For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.
-Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and diffusion $\epsilon = 4$.}
+Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and diffusion $\varepsilon = 4$.}
 
@@
 	Animals navigate complex, dynamic and patchy environments.
-When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit the dynamics.
+When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit available resources.
 It is in these cases, quite common in the natural world, that seasonal migration becomes a viable, even necessary, strategy.
 However, when resources start shifting in space and time -- as is occurring at an accelerated pace with recent global climate change -- the migration phenology itself must exhibit some plasticity.
-It is our conjecture that this plasticity is facilitated by a memory-driven process, in which recent experiences inform strategic behaviors in subsequent years, combined with a social dynamic.
+It is our conjecture that this plasticity is facilitated by a memory-driven process in which recent experiences inform strategic behaviors in subsequent years.
 The relatively simple, social and memory-driven mechanism we propose here was, in fact, able to adapt to long-term changes in resource dynamics, even with inter-annual stochasticity.
-The model thereby provides a framework with which the interaction  of memory, movement, social and resource dynamics can be explored.
+The model thereby provides a framework with which the interaction  of memory, movement, social and resource dynamics can be further explored.
 
-	We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration.
-In essence, our model allowed a population with a hybrid migratory and resource-following behavior to adjust the migratory behavior based on recent experiences with the resource location.
+\subsection{Adaptation and resiliency}
+
+	By allowing a population to adjust its migratory behavior based on recent experiences with the resource location, our model captured many fundamental phenomena related to migration.
 The model was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
@@
 	Our goal was to understand the combinations of factors that lead to a resilient migration behavior.
-We used foraging efficiency as a convenient metric of the utility of migration, but this was not a measure explicitly maximized by the model.
+We used foraging efficiency as a convenient metric of the utility of migration, though this was not a measure explicitly maximized by the model.
 Other metrics, such as foraging efficiency in a given season, or probability of survival or reproduction relative to resource availability \citep{Bauer2020} may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts.
@@
 
-	Among the general conclusions from our model, one is that migration can be acquired in a seasonal environment, even with no strong intrinsic propopensity to migrate and a weak phenological resource pulse to follow.
-This has been demonstrated for translocated ungulates \citep{Jesmer2018}, even in environments where resources are difficult to ``surf'' (figure \ref{fig_learningtomigrate}).
-Learning migration, howewer, requires a very strong resource attraction, higher levels of exploratory behavior (e.g.~diffusion, and larger spatial scale of sociliaty), and -- often --- many more years.
-These findings echo empirical observations \citep{Jesmer2018}.
-On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g.~increasing stochasticity, rapidly shifting resources, a shift in some of the system parameters, or even a shift in the spatial and temporal extent of resources, migration can collapse and turn into a non-migratory, residential behavior (figure \ref{fig_phenologymatching}).
-This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}, as well as the wide range of migration plasticity shown in wild populations, even within a species \citep{Xu2021}.
+	Sociality parameters -- in particular, the spatial scale $\lambda$ -- were, unexpectedly among the most important parameters for determining the resiliency of the process.
+Populations with small spatial scales tended to have a more difficult time locking in to an adaptive migratory pattern, and only when social attraction was relatively weak.
+On the other hand, overly large spatial scales compromised the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches and remember the corresponding benefits.
 
-	The ability to maintain, learn and adapt a migration also depends strongly on properties of the resource dynamics.
+	The ability to adapt a migration also depended strongly on properties of the resource dynamics.
 In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.
@@
 
-	\textit{The sociality parameters -- in particular, the spatial scale $\lambda$ -- were, unexpectedly, the most important parameters for determining the resiliency of the process.
-In particular, a population with a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak.
-In order to adapt to any shift in resources, \emph{some} individuals must experience that resource, and the two sociality-related mechanisms to have that scope are either a relatively large spatial scale or a relatively weak social cohesion.
-On the other hand, overly large spatial scales compromised the ability of the process to track climate change, due to a dilution of the population's ability to concentrate over available resource patches.
-Thus, there appear to be multiple avenues by which 'sociality' can structure migration.
-For example this could tie into the relative balance between individuals exploring the landscape to acquire new information about resources and their opportunities to interact with other individuals and communicate (or attract).
-	}
+	Even with no strong intrinsic propensity to migrate and a weak phenological resource pulse to follow, our model captured the abililty to acquire a strong and adaptive migration behavior (figure \ref{fig_learningtomigrate}).
+Learning migration, howewer, requires a very strong resource attraction, higher levels of exploratory behavior (e.g.~diffusion, and larger spatial scale of sociliaty), and -- often --- many more years, findings that echo empirical observations \citep{Jesmer2018}.
+
+	Despite the ability of the process to adapt under many stable conditions, our migration model (and, perhaps, migration behaviors in general) can also be considered somewhat fragile.
+Under many shifting conditions, e.g.~increasing stochasticity, rapidly shifting resources, a shift in some of the system parameters, or even a shift in the spatial and temporal extent of resources, migration can collapse and turn into a non-migratory, residential behavior (figure \ref{fig_phenologymatching}).
+This sensitivity may explain why partially migratory populations are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}, as well as the wide range of migration plasticity shown in wild populations, even within a species \citep{Xu2021}.
+
+\subsection{Biological interpretation of parameters}
+
+%	Diffusion-advection models of animal movement and redistribution are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components \citep{Skellam1951, Turchin1998, Okubo2001}.
+While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically.
+Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
+While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986, Gruenbaum1994, Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
 
 	All of the parameters in our model have well-defined biological interpretations.
-The diffusion ($\epsilon$) captures short time-scaled randomness of movement, capturing the exploratory and short-term dispersive behavior.
+The diffusion ($\varepsilon$) captures short time-scaled randomness of movement, reflecting exploratory and short-term dispersive behavior.
 The foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.
 These two parameters, the basic ingredients in diffusion-advection models of animal movement, have direct parallels to empirically estimated properties of animal behavior: diffusion is closely related to families of random walk models \citep{Gurarie2011} while the advective taxis is related to the step- and resource selection functions that are routinely estimated from movement data \citep{Potts2020}.
-The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, i.e.~a population-level range.
-The sociality parameter ($\beta$) quantifies the strength an individual's desire to approach the center of the social group.
-This is not a parameter that is typically measured, though it would - in principle - be possible to estimate in a manner analogous to step-selection function using only other conspecifics as opposed to environmental variables as a covariate.
-Because both resources and populations were normalized to 1, the ratio between $\alpha$ and $\beta$ can be interpreted as the relative importance of foraging to social cohesion.
-It does appears a certain balance between these two parameters is needed to successfully maintain, acquire, or adapt a migratory phenology.
-
-	Migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}.
-Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration.
-We note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that are well-documented.
-Birds in particular exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.
-
-	Individual based models have shown that collective knowledge is important, if not essential, to the evolution and process of migration \citep{Shaw2013, Guttal2010, Berdahl2018}.
-In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior.
-Synchrony of migration timing and high site fidelity are well documented for migratory species \citep{Gurarie2019, Joly2021}.
+The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, i.e.~a population-level home range \citep{Noonan2019}.
 	We note, however, that diffusion-advection models can also be interpreted as a probabilistic description of a single individual's movement.
@@
 
+	The sociality parameter ($\beta$) quantifies the strength an individual's desire to approach the center of the social group.
+While this parameter is not typically measured, it may in principle be possible to estimate in a manner analogous to step-selection function by replacing environmental variables with presence of conspecifics as a covariate.
+The ratio between $\alpha$ and $\beta$ can be interpreted as the relative importance of foraging to social cohesion, which appears to be important in predicting the resilience of migration.
+
+	Migration timing and location parameter can be straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019} and synchrony of migration timing and site fidelity are well-documented for many migratory species \citep{Joly2021}.
+Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou, identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as trends in timing.
+
 	The reference memory parameter $\kappa$ is, of course, impossible to observe directly.
 Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity \citep{Abrahms2019,Fagan2019b}.
-An extremely conservative behavior may be the best way to hedge against stochasticity (high $\kappa$ values in figure \ref{fig_stochasticity}) as it serves no benefit to change behavior based on recent experiences if they do not inform what might happen in the future.
-However, this extreme conservatism is, by definition, incapable of adapting when there is a consistent shift in resource disrtibution (figure \ref{fig_trendstochasticity}).
-In cases where both processes are occurring, we did, as predicted, see a slight improvement in adaptability when some long-term reference memory was used to stability the inclination to adapt, visible as peaks in figure \ref{fig_trendstochasticity} at $\kappa$ values around 0.8.
+An extremely conservative behavior is the best way to hedge against stochasticity with no directional changes (high $\kappa$ values in figure \ref{fig_stochasticity}), as there is no benefit to change behavior based on recent experiences if they provide no information about future conditions.
+However, this extreme conservatism is, by definition, incapable of adapting when there is a consistent shift in resource distribution (figure \ref{fig_trendstochasticity}).
+In cases where both processes are occurring, we did see a slight improvement in adaptability when long-term reference memory was balanced against a strong response to recent experience (see peaks in \ref{fig_trendstochasticity}).
 
-	Importantly, our model did not include any selection, inheritance or birth or death processes, i.e.~there was no evolutionary maximization of a fitness, as in many models that explicitly explore the evolution of migration \citep[e.g.][]{Guttal2010, Shaw2013, Anderson2013}.
-Nonetheless, it is instructive to compare the fundamental assumptions behind such evolutionary individual based models and a social memory based model like ours.
+	We'll note that our exploration of the model was not exhaustive.
+We did not explore, for example, the resilience of the migration process to changes in phenology timing.
+We hope that making the model available, including via an interactive interface, will facilitate those explorations.
+
+\subsection{Social learning and collective knowledge}
+
+	Models have shown that collective knowledge is important, if not essential, to the evolution and process of migration \citep{Shaw2013, Guttal2010, Berdahl2018}.
+Many migratory organisms are social, and social learning is an acknowledged, non-genetic method for transmitting information \citep{Kashetsky}.
+Because our model is not individual-based, we can not identify any specific mechanism (e.g.~leader-follower) of social information transfer.
+But, in a generic way, our model assumes that migration is driven by a collective trigger for the timing and locations of seasonal ranges, which is consistent with the known social and exogenous (e.g.~daylength related) triggers for migration.
+Further, the underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate, e.g.~the seasonal restlessness known as \emph{Zugunruhe}  exhibited by many birds \citep{Berthold1999, Helm2006}.
+
+	In contrast to the many individual-based models of the evolution of migration \citep[e.g.][]{Guttal2010, Shaw2013, Anderson2013}, our model did not include any selection, inheritance or birth or death processes.
 For example, \citet{Anderson2013} explored the resilience of a population under selective pressure under persistent trends and increased stochasticity of a drifting optimal resource window, showing that a certain amount of heritable phenotypic plasticity is necessary to adapt successfully to climate change even at the cost of efficiency.
-Our model underscores the fact that some level of resilience and adapatability can be attained with a purely cognitive process that balances sociality with long and short term collective memory.
-Importantly, this knowledge must be transmitted through social and cultural, rather than genetic, pathways.
+Our model underscores the fact that some level of resilience and adaptability can be attained with a purely cognitive process that balances sociality with long and short term collective memory.
+Importantly, this knowledge can be transmitted through social and cultural, rather than genetic, pathways.
 The high level of sociality among migratory animals, as well as multi-annual parent offspring bonds, are an evident pathway for that kind of transmission.
 As with those evolutionary models, however, it is clear that when changes are too rapid, no amount of cognition can help entirely mitigate against adverse outcomes.
-Furthermore, if behaviors are not sufficiently plastic (i.e.,~if $\kappa$ is too close to 1), then adaptation is that much more difficult.
+Furthermore, if behaviors are not sufficiently plastic (i.e.,~if $\kappa$ is too close to 1), then adaptation is very difficult.
 
-	Rapid environmental change, both global warming and increased anthropogenic development, is causing severe and dramatic impacts to the widespread and generally successful strategy of seasonal migration for many taxa, and the fate of many animal migrations is a topic of increasing concern \citep{Kauffman2021, Wilcove2008}.
+	Given the slow scale of fitness selection and the constant change in environmental conditions, it is possible that certain inherent properties of populations, for example the ``conservatism'' captured by the $\kappa$ parameter, are themselves selected for to maximize resilience over a long time scale in stochastic environments.
+
+\subsection{Summary}
+
+	Rapid environmental change, both global warming and increased anthropogenic development, is causing severe and dramatic impacts to the widespread and generally successful strategy of seasonal migration for many taxa, and the fate of many animal migrations is a topic of increasing concern \citep{Wilcove2008, Kauffman2021}.
 The ability of animals to respond to these changes depends deeply on their behavioral plasticity and cognitive abilities.
-The importance of those abilities is in direct proportion to the difficulty in studying them directly.
-By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adaptability and resilience.
-
-	Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}.
-These models are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components.
-While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically.
-Much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics.
-But perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory.
-Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
-While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
-
-	\section{Nomenclature}
+The importance of those abilities is in direct proportion to the difficulty in studying them directly.	By quantitatively exploring the properties of a heuristic model that distill many of the main properties of wild populations in dynamic and seasonal environments, we hope to have identified some broad patterns that might guide further empirical exploration of the cognitive underpinnings of adaptability and resilience.
 
@@
 	%\bibliographystyle{frontiersinHLTH&FPHY} % for Health, Physics and Mathematics articles
-	\bibliography{../library/memorymigration.bib}
+	\bibliography{memorymigration.bib}
 

### 2021-08-16 text/SocialMemoryMigration_FEE.tex
addressing comments

@@
 Migratory caribou and reindeer \emph{Rangifer tarandus}, for example, are several orders of magnitude more abundant than non-migratory woodland, mountain and forest ecotypes \citep{FestaBianchet2011,Uboni2016}.
-However, the question of whether migratory animals are more or less resilient to environmental disruptions in the environment remains open and largely case-specific \citep{Moore2008, HardestyMoore2017, Xu2021}.
+However, the question of whether migratory animals are more or less resilient to environmental disruptions in the environment remains open and largely case-specific \citep{Moore2008, HardestyMoore2018, Xu2021}.
 On the one hand, migratory species may be more vulnerable as disruptions in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Seebacher2015, Kauffman2021}.
@@
 	\noindent where each of the six parameters is updated according to equaton \ref{eq_memory} identically.
-The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, \Delta t'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$).
+The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, dt'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$).
 The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory.
@@
 
-	We ran this model on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day year~s).
-We were interested in an approximately periodic resource dynamic, i.e.~one in which $h(x,t \approx h(x, t-\tau))$.
+	We ran this model on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day years).
+We were interested in an approximately periodic resource dynamic, i.e.~one in which $h(x,t) \approx h(x, t-\tau))$.
 We generated two types of resource distributions.
 A ``non-surfable'' resource (\emph{island resource}), and weakly surfable resource (\emph{drifting resource}).
-Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 30 and -30 at times 25 and 75, respectively).
+Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 30 and $-30$ at times 25 and 75, respectively).
 These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively.
@@
 	distributions
+
 	$$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$
+
 	\noindent where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e.~$\frac{1}{\tau} \int_T\int_X h(x,t) dx\,dt = 1$.
@@
 Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a shift of the resource towards the extremes of the domain.
-While we did note explore phenological shifts in timing, those can readily be modeled as well.
+While we did not explore phenological shifts in timing, those can readily be modeled as well.
 These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate \citep{Renner2018}.
@@
 
-	As anticipated, better adaptation to the drifting resource correlated strongly with higher foraging efficiency (inset boxplots).
+	As anticipated, better adaptation to the drifting resource correlated strongly with higher foraging efficiency (inset boxplots).\\
 
@@
 		\caption{ \label{fig_trendstochasticity} Role of reference memory in adapting to climate change for increasingly stochastic resource dynamics.
-We ran the model with a moderiate rate of climate change (mean shift: 0.5/year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3, 6 and 12, left to right panels).
+We ran the model with a moderate rate of climate change (mean shift: 0.5/year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3, 6 and 12, left to right panels).
 For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.
@@
 	By allowing a population to adjust its migratory behavior based on recent experiences with the resource location, our model captured many fundamental phenomena related to migration.
-The model was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
+The model was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the emergence of a migratory behavior from an essentially resident or naive initial condition, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
 In a generalized way, it captured all three fundamental types of migration plasticity: the \textit{where}, the \textit{when} and the \textit{whether} \citep{Xu2021}.
@@
 
-	\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
-	%\bibliographystyle{frontiersinHLTH&FPHY} % for Health, Physics and Mathematics articles
+	\bibliographystyle{frontiersinSCNS_ENG_HUMS} 
 	\bibliography{memorymigration.bib}

### 2021-08-24 text/SocialMemoryMigration_FEE.tex
more ms finalization

@@
 
-		%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
 	Seasonal migrations are a widespread and broadly successful strategy for animals to exploit periodic and localized resources over large spatial scales.
@@
 		\tiny
-		\keyFont{ \section{Keywords:} Sociality, seasonal migration, collective knowledge, diffusion-advection models, spatial memory} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+		\keyFont{ \section{Keywords:} Sociality, seasonal migration, collective knowledge, diffusion-advection models, spatial memory} 
 	\end{abstract}
@@
 Values greater than 1 (super-adaptation) are possible, as are values less than 1, which correspond to a loss of migration behavior.
-
 All movement model parameters, resource parameters, and metrics are summarized in table \ref{table_definitions}.
+\\
 
@@
 			& $\lambda$ & Spatial scale of sociality \\ 
-			& $\kappa$ & Proportion of reference versus working memory \\ 
+			& $\kappa$ & Initial weighting of reference versus working memory \\ 
 			& $x_1$, $x_2$ & location of population centroids in summer and winter \\ 
@@
 			& $t_2$, $\Delta t_2$ & start and duration of winter season \\ 
-			& $\kappa$ & proportion memory allocated to reference \\
 			&& (long-term) memory vs. working (short-term) memory \\ 
@@
 			& $m_t$, $\tau - m_t$ & timing of resource peak for the summer and winter \\ 
-			& $\sigma_x$, $\sigma_t$ & time duration and and spatial scale of resource pulse \\ 
+			& $\sigma_x$, $\sigma_t$ & time duration and spatial scale of resource pulse \\ 
 			& $\gamma_x$, $\gamma_t$ & rate of change of peak location and timing of resource \\ 
@@
 	\begin{figure}
-		\includegraphics[width = 0.9\textwidth]{figures/example1_adaptation.png}
+		\includegraphics[width = 0.9\textwidth]{figures/example1_adaptation_lettered.png}
 
@@
 Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.
-Within each of the combinations of $\varepsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3 to 15 left to right), and 6 values of resource extent ($\sigma_x$, 3 to 15 bottom to top), as in the zoomed-in panel (bottom right; drawn from $\alpha = 200$, $\beta = 200$, $\lambda = 100$).
-The color scheme reflects the total mismatch, i.e.~the sum of the absolute differences between the migration timing and locations from the resource peak.}
+Within each of the combinations of $\varepsilon, \lambda, \alpha, \beta$, we show results ranging across 5 values of resource duration ($\sigma_t$, 3 to 15 left to right), and 6 values of resource extent ($\sigma_x$, 3 to 15 bottom to top), as in the zoomed-in panel (bottom right; drawn from $\alpha = 200$, $\beta = 200$, $\lambda = 100$).
+The color scheme reflects the total mismatch, i.e.~the sum of the absolute differences between the migration timing and locations from the resource peak.
+The white squares represents parameter combinations where the PDE could not be solved for artifactual numerical reasons, that all correspond to a failed adaptation (high mismatch).
+}
 	\end{figure}
@@
 	\begin{figure}
-		\includegraphics[width = 0.9\textwidth]{figures/example2_learningtomigrate.png}
+		\includegraphics[width = 0.9\textwidth]{figures/example2_learningtomigrate_lettered.png}
 
@@
 When that scale of sociality is high enough ($\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.
+\\
 
@@
 For the processes with non-zero stochasticity ($\psi_x > 0$), the process was run 90 times for values of $\kappa$.
-Points represent the average of the FE's across all 50 years and 90 replicates, and shaded areas represent a locall regression (loess) smooth of FE against $\kappa$.
-In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\varepsilon = 4$.}
+Points represent the average of the FE's across all 50 years and 90 replicates.
+In these scenarios, the resource following parameter $\alpha = 100$, the social attraction $\beta = 400$ and diffusion $\varepsilon = 4$.
+Note that the y-axis scale is constrained over a relatively narrow range (0.6-0.7).
+}
 
@@
 	Animals navigate complex, dynamic and patchy environments.
-When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to maximally exploit available resources.
+When there is a strongly localized and seasonal component to the resource dynamics, movement strategies limited to straightforward resource-following taxis necessarily fail to efficiently exploit available resources.
 It is in these cases, quite common in the natural world, that seasonal migration becomes a viable, even necessary, strategy.
@@
 It is our conjecture that this plasticity is facilitated by a memory-driven process in which recent experiences inform strategic behaviors in subsequent years.
-The relatively simple, social and memory-driven mechanism we propose here was, in fact, able to adapt to long-term changes in resource dynamics, even with inter-annual stochasticity.
-The model thereby provides a framework with which the interaction  of memory, movement, social and resource dynamics can be further explored.
+
+	By allowing a population to adjust its migratory behavior based on recent experiences with the resource location, the model we presented here emulated (a) the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, (b) the emergence of a migratory behavior from an essentially resident or naive initial condition, and (c) some intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity.
+The relatively simple, social and memory-driven mechanism was able to adapt to long-term changes in resource dynamics, even with inter-annual stochasticity, and may thereby provide a framework with which the interaction of memory, movement, social and resource dynamics can be further explored.
+% In a generalized way, it captured all three fundamental types of migration plasticity: the \textit{where}, the \textit{when} and the \textit{whether} \citep{Xu2021}.
+
+	Importantly, our model was in no ways evolutionary, as it contained no birth-death processes or selection pressures.
+Thus, we used foraging efficiency as a convenient metric of the utility of migration, though this was not a measure explicitly maximized by the model.
+Other metrics, such as foraging efficiency in a given season, or probability of survival or reproduction relative to resource availability \citep{Bauer2020} may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts.
+However, the overall annually averaged foraging efficiency metric provided the broadest linkage between resource dynamics and animals' locations and was consistent with the minimal biological assumptions and generality of our framework.
+\\
 
@@
 
-	By allowing a population to adjust its migratory behavior based on recent experiences with the resource location, our model captured many fundamental phenomena related to migration.
-The model was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the emergence of a migratory behavior from an essentially resident or naive initial condition, and intrinsic robustness to changes in those environmental resources, whether those changes were steadily shifting trends or inter-annual stochasticity.
-In a generalized way, it captured all three fundamental types of migration plasticity: the \textit{where}, the \textit{when} and the \textit{whether} \citep{Xu2021}.
+	Our goal was to understand the combinations of factors that lead to a resilient migration behavior.
+The model we describe was a final iteration of a sequence of models which failed to develop or maintain social migration behavior.
+For example, in earlier versions memory was modeled as an attractive advection mathematically identical to the resource attraction, but with the attractor being the location of the population in previous years.
+These models proved to be inefficient at generating a consistent social migratory behavior, i.e.
+only under very specific parameter combinations and ``easy'' conditions was a migratory equilibrium, and that equilibrium was highly unstable to perturbations.
+Only a clear, directed advective process with an explicit seasonal signal (i.e.~the remembered migration timing, rates, and targets which were remembered in our model) could generate the patterns we aimed to capture.
+This suggests, somewhat indirectly, that migration behavior is unique as a fundamental, long-term, and risky strategy, profoundly different from the kind of tactical resource response which governs shorter-scaled animal redistributions.
 
-	Our goal was to understand the combinations of factors that lead to a resilient migration behavior.
-We used foraging efficiency as a convenient metric of the utility of migration, though this was not a measure explicitly maximized by the model.
-Other metrics, such as foraging efficiency in a given season, or probability of survival or reproduction relative to resource availability \citep{Bauer2020} may respond differently across model parameters and could be useful in understanding the relative success of alternative migratory strategies in different contexts.
-However, the overall annually averaged metric provided the broadest linkage between resource dynamics and animals' locations and was consistent with the minimal biological assumptions and generality of our framework.
-
-	Sociality parameters -- in particular, the spatial scale $\lambda$ -- were, unexpectedly among the most important parameters for determining the resiliency of the process.
+	Similarly, iterations of the model that did not have some amount of social cohesion tended to diffuse away without establishing a consistent, migratory stationary state.
+In fact, sociality parameters -- in particular, the spatial scale $\lambda$ -- were, unexpectedly among the most important parameters for determining the resiliency of the process.
 Populations with small spatial scales tended to have a more difficult time locking in to an adaptive migratory pattern, and only when social attraction was relatively weak.
@@
 
-	Even with no strong intrinsic propensity to migrate and a weak phenological resource pulse to follow, our model captured the abililty to acquire a strong and adaptive migration behavior (figure \ref{fig_learningtomigrate}).
+	Even with no strong intrinsic propensity to migrate and a weak phenological resource pulse to follow, our model captured the ability to acquire a strong and adaptive migration behavior (figure \ref{fig_learningtomigrate}).
 Learning migration, howewer, requires a very strong resource attraction, higher levels of exploratory behavior (e.g.~diffusion, and larger spatial scale of sociliaty), and -- often --- many more years, findings that echo empirical observations \citep{Jesmer2018}.
@@
 Under many shifting conditions, e.g.~increasing stochasticity, rapidly shifting resources, a shift in some of the system parameters, or even a shift in the spatial and temporal extent of resources, migration can collapse and turn into a non-migratory, residential behavior (figure \ref{fig_phenologymatching}).
-This sensitivity may explain why partially migratory populations are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}, as well as the wide range of migration plasticity shown in wild populations, even within a species \citep{Xu2021}.
+This sensitivity may explain why partially migratory populations are so common and, apparently, evolutionary stable \citep{Berthold1999, Chapman2011}, as well as the wide range of migration plasticity shown in wild populations, even within a species \citep{Xu2021}.\\
 
@@
 
-%	Diffusion-advection models of animal movement and redistribution are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components \citep{Skellam1951, Turchin1998, Okubo2001}.
+	Diffusion-advection models of animal movement and redistribution are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components \citep{Skellam1951, Turchin1998, Okubo2001}.
 While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically.
-Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
+% Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy.
 While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986, Gruenbaum1994, Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.
 
-	All of the parameters in our model have well-defined biological interpretations.
+	Despite its evident abstraction, our goal was to develop a model where all parameters have well-defined biological interpretations.
 The diffusion ($\varepsilon$) captures short time-scaled randomness of movement, reflecting exploratory and short-term dispersive behavior.
 The foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.
-These two parameters, the basic ingredients in diffusion-advection models of animal movement, have direct parallels to empirically estimated properties of animal behavior: diffusion is closely related to families of random walk models \citep{Gurarie2011} while the advective taxis is related to the step- and resource selection functions that are routinely estimated from movement data \citep{Potts2020}.
+These two parameters, the basic ingredients in diffusion-advection models of animal movement, have direct parallels to empirically estimated properties of animal behavior: diffusion is closely related to families of random walk models \citep{Gurarie2011} while the advective taxis is related to the step and resource selection functions that are routinely estimated from movement data \citep{Potts2020}.
 The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, i.e.~a population-level home range \citep{Noonan2019}.
@@
 
-	The sociality parameter ($\beta$) quantifies the strength an individual's desire to approach the center of the social group.
+	The sociality parameter ($\beta$) quantifies the strength of an individual's desire to approach the center of the social group.
 While this parameter is not typically measured, it may in principle be possible to estimate in a manner analogous to step-selection function by replacing environmental variables with presence of conspecifics as a covariate.
@@
 
-	Migration timing and location parameter can be straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019} and synchrony of migration timing and site fidelity are well-documented for many migratory species \citep{Joly2021}.
-Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou, identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as trends in timing.
+	Migration timing, rate, and seasonal range location parameters can be straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019} and synchrony of migration timing and site fidelity are well-documented for many migratory species \citep{Joly2021}.
+Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou, identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here, as well as trends in timing.
 
@@
 
-	We'll note that our exploration of the model was not exhaustive.
-We did not explore, for example, the resilience of the migration process to changes in phenology timing.
-We hope that making the model available, including via an interactive interface, will facilitate those explorations.
+	Clearly, our exploration of the model was not exhaustive.
+We did not explore, for example, the resilience of the migration process to changes in resource timing, which would correspond to the widely observed earlier onset of spring as measured by green-up and flowering phenology \citep{Cleland2007}.
+We hope that making the model available, including via the interactive interface, will facilitate further independent exploration of these processes.
+\\
 
@@
 Many migratory organisms are social, and social learning is an acknowledged, non-genetic method for transmitting information \citep{Kashetsky}.
+Furthermore, the general role of social learning for improving a population's ability to track resources has been studied not just in animal systems, but in synthetic systems inspired by social behavior of animals such as optimization heuristics algorithms and the study of swarm robotics \citep{Sahin2005, Brambilla2013}.
 Because our model is not individual-based, we can not identify any specific mechanism (e.g.~leader-follower) of social information transfer.
-But, in a generic way, our model assumes that migration is driven by a collective trigger for the timing and locations of seasonal ranges, which is consistent with the known social and exogenous (e.g.~daylength related) triggers for migration.
+But, in a generic way, our model assumes that migration is driven by a collective decision for the timing and locations of seasonal ranges, consistent with the known social and exogenous (e.g.~daylength related) triggers for migration.
 Further, the underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate, e.g.~the seasonal restlessness known as \emph{Zugunruhe} exhibited by many birds \citep{Berthold1999, Helm2006}.
+However, in its generic diffusion-based approach to randomness, our model indirectly captures individual-level variation in migration parameters, an inevitable property of any population-level process \citep{Gurarie2019}.
 
@@
 	Given the slow scale of fitness selection and the constant change in environmental conditions, it is possible that certain inherent properties of populations, for example the ``conservatism'' captured by the $\kappa$ parameter, are themselves selected for to maximize resilience over a long time scale in stochastic environments.
+The structure of the reference memory in our model was a rather simplistic approach to introduce conservatism or lag to the shifting migration parameters.
+In our model that reference memory is eventually entirely forgotten, whereas a more sophisticated approach would separate a slowly varying cultural memory, perhaps that is transmitted genetically or culturally, i.e.~on the scale of generations, against shorter-scaled responses.
+In an evolutionary model, we might hypothesize that the overall rates of long- and short-term memory shifts would be related both to the scales of short and long-term fluctuation of the resource, i.e.~the auto-correlation scale, strength of trends, and stochasticity of the resource dynamics.
+\\
 
@@
 	\subsection{Drifting resource}
-	The drifting resource function has the following properties
+	The drifting resource function has the following properties:
 
